---
title: "The two languages"
---

{{< include ../_setup.qmd >}}

A pipeline is the same sentence in both languages.
[Your first pipeline](../chapters/first-pipeline.qmd) showed the two
places they differ. This appendix is the rest of it: why the pipe is what it is,
the three rules Python adds, how R's own habits are read, and what god does to
the names already in your session.

Everything else in the book is written once, with a tab for each language,
because everywhere else the two really are the same sentence.

## The two differences

`>>` is the pipe in Python, and it is an operator rather than a method call for
the reason the preface gives: an arrow points at the next step and a dot ends a
sentence. Python cannot have `|>` itself. Its operator set is fixed by the
language, and `|>` does not even tokenize.

Of the operators a frame leaves free, `>>` is the only one available everywhere,
which makes it the choice rather than a preference. `|` was considered and would
have been a disaster, because pandas defines it, so `frame | verb` never reaches
the verb at all. It does an elementwise or and hands back a frame of `True`,
which is a wrong answer with no error attached.

`col.name` is how Python names a column. R can look a name up in your data before
it looks in your session, so a bare `revenue` is unambiguous there. Python has no
such hook, so a column says that it is one.

A name that is not a Python identifier is still a column. Write it in
brackets, `col["order date"]`, and it is the same expression the dot would
have made:

```{python}
orders = pd.DataFrame({"order date": ["2026-01-02", "2026-01-05"],
                       "total": [40, 90]})
orders >> keep(col["order date"] > "2026-01-03")
```

R's spelling for the same name is its own, backticks, and R removes them
before god ever sees the name:

```{r}
orders <- data.frame(`order date` = c("2026-01-02", "2026-01-05"),
                     total = c(40, 90), check.names = FALSE)
orders |> keep(`order date` > "2026-01-03")
```

## Three rules Python adds

None of them comes from the grammar, and this book states them here rather
than letting you find them in an error message.

Comparisons joined by `&` need parentheses, because `&` binds more tightly than
`==` in Python. Negation is `~`, because `not` is a keyword whose result Python
turns into a bool, so an object cannot see it.

```{python}
sales >> keep((col.revenue > 150) & (col.cost < 100))
```

```{python}
sales >> keep(~(col.product == "Widget"))
```

Asking whether a value is one of several is a method, since Python's `in` cannot
be reached either.

```{python}
sales >> keep(col.region.is_in(["West", "East"]))
```

Write the values as a list when the order matters to you. A set works too, and
gets sorted before it is written, because Python's hashing changes between runs
and the same pipeline has to read the same way every time.

The third rule is a refusal rather than a spelling. Python's own `and`, `or`,
`in` and `if` all ask a value to become one plain yes or no, and a column
expression refuses, because its honest answer is one per row:

```{python}
#| classes: refusal
try:
    if col.revenue > 150:
        print("large")
except GodError as refusal:
    print(refusal)
```

The message names `&`, `|` and `~` at the moment you reach the wrong
spelling, which is earlier than a wrong answer would have told you.

One word used to be on this list and no longer is. The marker that inverts a
`pick` was `except`, which Python had to spell `except_`, because `except` is a
keyword here. It is `all_but` in both languages now, and in the text form. No
word in the vocabulary is spelled differently in the two languages.

Two host-side calls do differ, and neither is a word of the vocabulary.
Asking a pipeline for its text form is `format(p)` in R and `p.written()` in
Python: `format` is the word R already owns for turning a thing into text,
and Python's `format` means something else, so its method says what it
gives. And `show_as` prints in R while it returns in Python, for the
notebook reason [What it wrote](../chapters/what-it-wrote.qmd) explains.

## R's habits, in the grammar's words

The grammar has one spelling for each idea, and some of R's spellings are not it.
You still write R, and the verbs translate as they build the sentence. `format`
shows what they wrote.

```{r}
cat(format(sales |> keep(region != "West" & !is.na(cost))))
```

`!=` became `is not` and `is.na` became `is missing`. The full set of
replacements is small.

```{r}
#| echo: false
translations <- data.frame(
  `You write` = c("==", "!=", "&", "|", "!", "%in%", "is.na(x)", "TRUE", "NA"),
  `god writes` = c("is", "is not", "and", "or", "not", "in { }",
                   "x is missing", "yes", "missing"),
  check.names = FALSE
)
knitr::kable(translations)
```

A set of values is written out, because the grammar has no variables yet and
will not guess at one.

```{r}
sales |>
  keep(region %in% c("West", "East")) |>
  summarize(n = row_count(), by = region)
```

## The names god takes over

Attaching god in R replaces `sort` for the rest of your session. That is
deliberate: a prefix on every sentence is a worse trade than a good message on
the few calls that go wrong. The message names what was shadowed and how to reach
it.

```{r}
#| error: true
sort(c(3, 1, 2))
```

What it does not take over is every other package you have. R looks a name up
inside a package through that package's own namespace, never through the search
path you attached to, so a function that calls `sort` in its own body is
untouched. `median` and `quantile` both do.

```{r}
c(median(c(3, 1, 2)), quantile(1:9, 0.5))
```

So the cost is your own calls to `sort`, at the console or in a script, and
nothing further. `base::sort` is always there, and the message above says so at
the moment you need it.

Python has namespaces where R does not, so `import god` collides with nothing at
all and only `from god import *` can.

## The names other packages take back

Four of god's names are dplyr's as well: `collect`, `pick`, `rename` and
`summarize`. `keep` is purrr's. Attach one of those packages after god and its
version of the name is what your next line reaches.

**Three of the four cost you nothing.** `collect`, `rename` and `summarize` are
generics, and god gives each of them a method for a pipeline, so the sentence
runs whichever package you attached last.

```{r}
sales |>
  keep(region == "West") |>
  dplyr::summarize(takings = total(revenue), by = product)
```

That is dplyr's `summarize`, handed a god pipeline and answering as god's. Which
order you wrote your `library()` calls in is not visible from here.

`pick` is the exception, and the reason belongs to dplyr rather than to god: its
`pick` is not a generic, so there is no method to give it. Write `god::pick` when
both packages are attached.

## If you also use gog

`col` is a good enough name that other people picked it too. polars uses it, and
so does gog, this project's sibling for drawing plots, where it does the same job
for the same reason. `from god import *` and `from gog import *` in one file
means the second one wins and the first is gone.

Import them by name when you want both, which is what Python's namespaces are
for:

```python
import god
import gog

sales_by_product = god.collect(
    sales >> god.summarize(revenue = god.total(god.col.revenue), by = god.col.product)
)
```

Or bring in one of them and reach for the other by name. The failure is loud if
you forget: god's verbs have no idea what a gog column is, so a mixed-up pipeline
stops rather than guessing.

## Asking which god you have

Each language answers in its own idiom, and a release is refused when the two
would disagree.

```{r}
packageVersion("god")
```

```{python}
import god
god.__version__
```

Same number, two idioms, which is this appendix in one line: the differences
between the languages are spellings, and never the sentence.
