---
title: "Every word in the grammar"
---

{{< include ../_setup.qmd >}}

The whole vocabulary, on one page. If you are deciding whether to learn this,
read the list and see how long it takes: that is the argument, and it is shorter
than the argument for it.

Nothing here is typed by hand. The engine is asked what words it has when this
page is built, so the list cannot describe a word that was removed or miss one
that was added.

```{r}
#| echo: false
vocabulary <- read.delim(
  text = system2(god:::god_binary(), "--vocabulary", stdout = TRUE),
  header = FALSE, col.names = c("kind", "word"), stringsAsFactors = FALSE
)
counts <- table(vocabulary$kind)
```

```{r}
#| echo: false
knitr::kable(
  data.frame(
    Kind = c("Verbs", "Aggregations", "Windows", "Scalar functions", "Grammar words"),
    Count = as.integer(counts[c("verb", "aggregate", "window", "scalar", "word")]),
    `What they do` = c(
      "Steps in a pipeline. Every one is an imperative English verb",
      "Collapse a group to one value",
      "Answer for every row by looking at the other rows",
      "Answer for one row from that row",
      "The small words that join the rest together"
    ),
    check.names = FALSE
  )
)
```

## The verbs

A pipeline is a table, then `then`, then one of these, repeated.

```{r}
#| echo: false
paste0("`", vocabulary$word[vocabulary$kind == "verb"], "`", collapse = " · ") |> cat()
```

`keep` and `pick` are the two that choose: rows and columns. `add` and
`summarize` are the two that make a column: one per row, and one per group.
`sort` and `take` order and cut. Everything after those six is either a second
table, a missing value, or a change of shape.

## The aggregations

These go inside `summarize`, and inside `add` when it carries a `by`.

```{r}
#| echo: false
paste0("`", vocabulary$word[vocabulary$kind == "aggregate"], "`", collapse = " · ") |> cat()
```

## The windows

These answer once per row by looking along the other rows. Several of them
require a `sort` in front, because they mean nothing until something has said in
what order.

```{r}
#| echo: false
paste0("`", vocabulary$word[vocabulary$kind == "window"], "`", collapse = " · ") |> cat()
```

## The scalar functions

One row in, one value out. These go anywhere a value goes.

```{r}
#| echo: false
paste0("`", vocabulary$word[vocabulary$kind == "scalar"], "`", collapse = " · ") |> cat()
```

## The grammar words

Not verbs and not functions. These are the joints: they say how the words on
either side of them are related.

```{r}
#| echo: false
paste0("`", vocabulary$word[vocabulary$kind == "word"], "`", collapse = " · ") |> cat()
```

## That is all of it

There is no second list, no set of helpers, and no dialect that adds a few more
on a cluster. A word that is not above is not in the grammar, and asking for one
gets a refusal that names the nearest word that is.
