---
title: "Every word in the grammar"
---

{{< include ../_setup.qmd >}}

The whole vocabulary, on one page. If you are deciding whether to learn this,
read the list and see how long it takes: that is the argument, and it is shorter
than the argument for it.

Nothing here is typed by hand. The engine is asked what words it has when this
page is built, so the list cannot describe a word that was removed or miss one
that was added.

```{r}
#| echo: false
vocabulary <- read.delim(
  text = system2(god:::god_binary(), "--vocabulary", stdout = TRUE),
  header = FALSE, col.names = c("kind", "word"), stringsAsFactors = FALSE
)
counts <- table(vocabulary$kind)
```

```{r}
#| echo: false
knitr::kable(
  data.frame(
    Kind = c("Verbs", "Aggregations", "Windows", "Scalar functions", "Grammar words"),
    Count = as.integer(counts[c("verb", "aggregate", "window", "scalar", "word")]),
    `What they do` = c(
      "Steps in a pipeline. Every one is an imperative English verb",
      "Collapse a group to one value",
      "Answer for every row by looking at the other rows",
      "Answer for one row from that row",
      "The small words that join the rest together"
    ),
    check.names = FALSE
  )
)
```

## The verbs

A pipeline is a table, then `then`, then one of these, repeated.

```{r}
#| echo: false
paste0("`", vocabulary$word[vocabulary$kind == "verb"], "`", collapse = " · ") |>
  cat()
```

`keep` and `pick` are the two that choose: rows and columns. `add` and
`summarize` are the two that make a column: one per row, and one per group.
`sort` and `take` order and cut. Everything after those six is a second
table, a missing value, a change of shape, or the housekeeping pair that
renames a column and drops repeated rows.

## The aggregations

These go inside `summarize`, and inside `add` when it carries a `by`.

```{r}
#| echo: false
paste0("`", vocabulary$word[vocabulary$kind == "aggregate"], "`", collapse = " · ") |>
  cat()
```

## The windows

These answer once per row by looking along the other rows. Several of them
require a `sort` in front, because they mean nothing until something has said in
what order.

```{r}
#| echo: false
paste0("`", vocabulary$word[vocabulary$kind == "window"], "`", collapse = " · ") |>
  cat()
```

## The scalar functions

One row in, one value out. These go anywhere a value goes.

```{r}
#| echo: false
paste0("`", vocabulary$word[vocabulary$kind == "scalar"], "`", collapse = " · ") |>
  cat()
```

## The grammar words

Not verbs and not functions. These are the joints: they say how the words on
either side of them are related.

```{r}
#| echo: false
paste0("`", vocabulary$word[vocabulary$kind == "word"], "`", collapse = " · ") |>
  cat()
```

## Where each word is first used

Each verb and function below links to the page whose running example uses it
first, in the book's own order. The table is worked out from the chapters
when the page is built, the same way the lists above come from the engine. A
moved example moves its link, and a word no chapter runs cannot ship at all,
because a build check refuses the book first.

```{r}
#| echo: false
# The chapter order is _quarto.yml's, teaching pages only; the first
# executable chunk to use a word, in that order, is where a reader meets it.
listed <- readLines("../_quarto.yml", warn = FALSE)
entries <- grep("^\\s*-\\s*(part:\\s*)?[A-Za-z0-9_/-]+\\.qmd\\s*(#.*)?$",
                listed, value = TRUE)
entries <- sub("\\s*#.*$", "", sub("^\\s*-\\s*(part:\\s*)?", "", trimws(entries)))
teaching <- entries[grepl("^(chapters|cookbook)/", entries)]

read_page <- function(f) readLines(file.path("..", f), warn = FALSE)
chunk_text <- vapply(teaching, function(f) {
  ln <- read_page(f)
  keep <- logical(length(ln)); inside <- FALSE
  for (i in seq_along(ln)) {
    if (grepl("^\\s*```\\{", ln[i])) { inside <- TRUE; next }
    if (grepl("^\\s*```\\s*$", ln[i])) { inside <- FALSE; next }
    keep[i] <- inside
  }
  paste(ln[keep], collapse = "\n")
}, character(1))
title_of <- vapply(teaching, function(f) {
  found <- grep("^title:", read_page(f), value = TRUE)[1]
  gsub('^title:\\s*"?|"?\\s*$', "", found)
}, character(1))

written <- vocabulary[vocabulary$kind != "word", ]
home <- vapply(written$word, function(w) {
  hit <- which(grepl(paste0("\\b", w, "\\b"), chunk_text, perl = TRUE))[1]
  if (is.na(hit)) "the appendices"
  else sprintf("[%s](../%s)", title_of[[hit]], teaching[[hit]])
}, character(1))

knitr::kable(data.frame(
  Word = paste0("`", written$word, "`"),
  `First running example` = home,
  check.names = FALSE
))
```

The grammar words are not in the table on purpose: `in`, `is`, `not` and
their siblings are also words the host languages use, so a scan would
confidently attribute the wrong page. They are everywhere above instead.

## Every word answers for itself

You do not have to come back here. Each word above has a page where you are
already sitting, and every one of them carries an example you can paste.

```r
?keep
?add_combinations
```

```python
help(god.keep)
help(god.add_combinations)
```

The two show the same example in the two spellings, down to the same table and
the same numbers, which is this book's claim stated one word at a time. They are
run rather than written: the R suite executes the example on every help page and
the Python suite executes every one in every docstring, so a page that falls
behind the engine fails a build instead of misleading the reader who went
looking.

One difference between those pages and these can surprise you. This book tells
Python to print a whole number without its decimal point, so that the two tabs
of an example agree. Nothing has told your own session that, and a total of
whole numbers arrives there as `250.0` rather than `250`. The help pages show
the second, because that is what you will see.

## That is all of it

There is no second list, no set of helpers, and no dialect that adds a few more
on a cluster. A word that is not above is not in the grammar, and asking for one
gets a refusal that names the nearest word that is.
