---
title: "The same sentence in six dialects"
---

{{< include ../_setup.qmd >}}

If you already write dplyr, pandas, polars, PySpark, SQL or Spark SQL, this
appendix is the translation. Each row is one god sentence and what it becomes
in each of them.

Nothing here is written by hand. Every cell is produced by asking god to print
the pipeline in that language when this page is built, which is the same
machinery [What it wrote](../chapters/what-it-wrote.qmd) describes. A translation table
maintained by a person goes out of date; this one cannot, because it is
generated from the pipeline it describes.

```{r}
#| echo: false
# Ask the engine for one sentence in one language. `show_as` prints as well as
# returning, which is right at a prompt and not here, so the printing is caught
# and thrown away.
spoken <- function(pipeline, language) {
  text <- NULL
  invisible(capture.output(text <- show_as(pipeline, language, sales = sales)))
  gsub("\n", " ", trimws(paste(text, collapse = " ")))
}

# A cell written as HTML rather than as markdown, because of what a markdown
# table has to do with a pipe. `kable` escapes every `|` in a cell to `&#124;`
# so it cannot be read as a column separator, and pandoc leaves an entity alone
# inside a code span, so every dplyr row came out reading
# `sales &#124;> filter(...)`. Writing the cell as HTML keeps the pipe a pipe.
code <- function(x) {
  x <- gsub("&", "&amp;", x, fixed = TRUE)
  x <- gsub("<", "&lt;", x, fixed = TRUE)
  x <- gsub(">", "&gt;", x, fixed = TRUE)
  paste0("<code>", x, "</code>")
}

sentences <- c(
  "sales then keep where [region] is \"West\"",
  "sales then pick [product, revenue]",
  "sales then add [margin] as [revenue] - [cost]",
  "sales then summarize [total] as total([revenue]) by [product]",
  "sales then sort [revenue] descending",
  "sales then take 3",
  "sales then drop_duplicates",
  "sales then keep where [revenue] > 150 and [cost] < 100"
)
```

## Row by row

```{r}
#| echo: false
short <- function(x, n = 60) ifelse(nchar(x) > n, paste0(substr(x, 1, n - 1), "…"), x)
knitr::kable(
  data.frame(
    god    = code(sub("^sales then ", "", sentences)),
    dplyr  = code(short(vapply(sentences, spoken, "", "dplyr"))),
    pandas = code(short(vapply(sentences, spoken, "", "pandas"))),
    check.names = FALSE, row.names = NULL
  ),
  format = "html", escape = FALSE
)
```

```{r}
#| echo: false
knitr::kable(
  data.frame(
    god     = code(sub("^sales then ", "", sentences)),
    polars  = code(short(vapply(sentences, spoken, "", "polars"))),
    pyspark = code(short(vapply(sentences, spoken, "", "pyspark"))),
    check.names = FALSE, row.names = NULL
  ),
  format = "html", escape = FALSE
)
```

```{r}
#| echo: false
knitr::kable(
  data.frame(
    god   = code(sub("^sales then ", "", sentences)),
    sql   = code(short(vapply(sentences, spoken, "", "sql"))),
    spark = code(short(vapply(sentences, spoken, "", "spark"))),
    check.names = FALSE, row.names = NULL
  ),
  format = "html", escape = FALSE
)
```

## One sentence in full

The tables above shorten the long lines, and take the six languages two at a
time, because six columns do not fit across a page. Here is one sentence
written out completely in each, so you can see the shape rather than the
abbreviation.

```{r}
#| echo: false
#| output: asis
# `asis` is what makes the headings below headings. Without it the whole loop
# lands on the page as one block of literal text, hashes and fences included.
one <- "sales then keep where [region] is \"West\" then summarize [total] as total([revenue]) by [product]"
for (language in c("sql", "spark", "dplyr", "pandas", "polars", "pyspark")) {
  cat("### `", language, "`\n\n```\n", sep = "")
  invisible(capture.output(text <- show_as(one, language, sales = sales)))
  cat(paste(text, collapse = "\n"), "\n```\n\n", sep = "")
}
```

## A whole pipeline, all seven

The sentence above is two steps, chosen to stay readable. The translation
does not thin as a pipeline grows, and a technical reader deciding whether
to trust that should see it on a hard case. Six steps this time: a join, a
computed column, a grouped total, a sort, a rank, and a cut. The seventh
answer is the grammar's own text form, and it is printed first. It is what
`show_as` gives when you ask for `"god"`, and it is the one spelling that runs
anywhere.

```{r}
#| echo: false
#| output: asis
long <- paste(
  "sales",
  "then join products by [product]",
  "then add [margin] as [revenue] - [cost]",
  "then summarize [margin] as total([margin]) by [maker]",
  "then sort [margin] descending",
  "then add [place] as rank([margin] descending)",
  "then take 3"
)
for (language in c("god", "sql", "spark", "dplyr", "pandas", "polars", "pyspark")) {
  cat("### `", language, "`\n\n```\n", sep = "")
  invisible(capture.output(
    text <- show_as(long, language, sales = sales, products = products)
  ))
  cat(paste(text, collapse = "\n"), "\n```\n\n", sep = "")
}
```

## What the table is showing

Read down a column and you are reading one library's idea of how a pipeline is
written. Read one god sentence across all three tables and you are reading one idea,
six ways.

The lengths differ, and the differences are the honest result rather than a
selection. polars and PySpark have a method per verb, so a pipeline maps onto a
chain and stays short. pandas has no single way to chain, so the idiom that does
is `.assign` and `.loc[lambda d: ...]`, and every column inside one of those
names the frame again. SQL puts the steps in an order that is not the order they
happen in.

This is also the way out. A small vocabulary covers most of what people do and
never all of it. When you reach the end of what god can say, ask it what your
pipeline would have been in the tool you already use, and carry on there.
