---
title: "Where each word can stand"
---

{{< include ../_setup.qmd >}}

[Every word in the grammar](b-every-word.qmd) says what the words are. This
page says where they may stand.

Every seat below is a place in a step that takes a computed value. The grid
is asked from the engine while this page is built, from the same table its
checker answers to. So what you read here cannot drift from what the engine
does. The engine's own tests also run every cell through the parser and the
checker. A cell that moved without its rule moving would fail those tests
before it could mislead this page.

```{r}
#| echo: false
cells <- read.delim(
  text = system2(god:::god_binary(), "--seats", stdout = TRUE),
  header = FALSE,
  col.names = c("seat", "shown", "kind", "answer", "note"),
  stringsAsFactors = FALSE
)
cells$note[is.na(cells$note)] <- ""
seats <- unique(cells$shown)
grid <- data.frame(
  `The seat` = paste0("`", seats, "`"),
  check.names = FALSE
)
for (k in c("scalar", "aggregate", "window")) {
  grid[[k]] <- vapply(seats, function(s) {
    cells$answer[cells$shown == s & cells$kind == k]
  }, character(1))
}
knitr::kable(grid, row.names = FALSE)
```

Read a row and you are reading what one seat is for.

A **condition** asks one row a question. A scalar can decide it and a group's
answer cannot, because the group's answer does not exist until the rows are
chosen.

A **value** in `add` is the one seat that accepts every kind of answer. A group's
total is repeated on every row of its group there, and a running total can
walk the rows once a `sort` has said in what order.

A **group's answer** in `summarize` must span its group. Its head is
therefore an aggregate, and a scalar stands inside an aggregate's argument.

An **ordering position** names a column and never an expression, in `sort`
and inside `rank` alike. The computed key is one `add` away, and the grammar
refuses to offer a second spelling of that `add`.

A **filler** in `fill_missing` holds one value, and only a scalar may stand
there. A **cell** in `widen` takes a scalar or an aggregate, and the
aggregate is the answer when two rows want the same cell. Neither seat takes
a look along the rows, and for anything else that is refused, make it a
column first.

Some cells carry one more sentence, and the engine says those too:

```{r}
#| echo: false
noted <- cells[cells$note != "", c("shown", "kind", "note")]
knitr::kable(
  noted,
  row.names = FALSE,
  col.names = c("The seat", "The kind", "What the cell says")
)
```

What the grid never says is *useful*. A sentence these seats accept is
answered however strange it is. Well-formed and worthwhile are different
judgments, and they belong to different people: the engine settles the
first, always, and the second is yours. That split is a law of its own, and
[the laws](../chapters/design-laws.qmd) show it working beside the rest.
