---
title: "Where each word can stand"
---

{{< include ../_setup.qmd >}}

The vocabulary says what the words are; this page says where they may stand.
Every seat below is a place in a step that takes a computed value, and the
grid is asked from the engine while this page is built, from the same table
its checker answers to, so what you are reading cannot drift from what the
engine does. The engine's own tests run every cell of it through the parser
and checker besides, so a cell that moved without its rule moving would fail
the build before it could mislead this page.

```{r}
#| echo: false
cells <- read.delim(
  text = system2(god:::god_binary(), "--seats", stdout = TRUE),
  header = FALSE,
  col.names = c("seat", "shown", "kind", "answer", "note"),
  stringsAsFactors = FALSE
)
cells$note[is.na(cells$note)] <- ""
seats <- unique(cells$shown)
grid <- data.frame(
  `The seat` = paste0("`", seats, "`"),
  check.names = FALSE
)
for (k in c("scalar", "aggregate", "window")) {
  grid[[k]] <- vapply(seats, function(s) {
    cells$answer[cells$shown == s & cells$kind == k]
  }, character(1))
}
knitr::kable(grid, row.names = FALSE)
```

Read a row and you are reading what one seat is for. A **condition** asks one
row a question, so a scalar can decide it and a group's answer cannot: the
group's answer does not exist until the rows are chosen. A **value** in `add`
is the one seat with a row for every kind of answer, which is why a group's
total can land on every row of its group there, and a running total can walk
the rows once a `sort` has said in what order. A **group's answer** in
`summarize` must span its group, so its head is an aggregate and scalars ride
inside it. An **ordering position** names a column and never an expression,
in `sort` and inside `rank` alike, because the computed key is one `add`
away and a second spelling of that `add` is what the refusal declines to be.
A **filler** and a **cell** each hold one value, and where a group's answer
or a look along the rows would help, the road is the same as everywhere
else: make it a column first.

Some cells carry one more sentence, and the engine says those too:

```{r}
#| echo: false
noted <- cells[cells$note != "", c("shown", "kind", "note")]
knitr::kable(
  noted,
  row.names = FALSE,
  col.names = c("The seat", "The kind", "What the cell says")
)
```

What the grid never says is *useful*. A sentence these seats accept is
answered however strange it is, because well-formed and worthwhile are
different judgments belonging to different people: the engine settles the
first, always, and the second is yours. That split is a law of its own, and
[the laws](../chapters/design-laws.qmd) show it working beside the rest.
