---
title: "Where the speed lives"
---

{{< include ../_setup.qmd >}}

A benchmark of god against another tool is really a benchmark of god's
engine against that tool, because god's own work ends when the query is
written. So this page refuses to blur the two. The half that belongs to god
is measured live, while this page is built. The half that belongs to the
engines is a dated record, run once on real data with the date printed
beside it. A speed table without its date is a rumor.

## The grammar's toll, clocked on this render

What god costs is compiling a pipeline: reading the sentence, checking it
against the columns, writing the query. That cost is fixed per pipeline
rather than per row, and rather than assert it, this page measures it while
rendering, on a five-step sentence:

```{r}
sentence <- paste(
  'sales then keep where [revenue] > 100',
  'then add [margin] as ([revenue] - [cost])',
  'then summarize [margin] as total([margin]) by [region]',
  'then sort [margin] descending then take 5')
columns <- "region:text,product:text,quantity:number,revenue:number,cost:number"

timings <- vapply(1:50, function(i) {
  started <- Sys.time()
  god_sql(sentence, columns)
  as.numeric(Sys.time() - started)
}, 0)

cat(sprintf("%.1f milliseconds per pipeline, on the machine that built this page",
            median(timings) * 1000))
```

A hundred rows and a hundred million compile in the same time, because the
compiler reads the sentence and the column names, never the data.
Everything after that belongs to the engine.

## The engines' race, on record

The record below was measured on 2026-08-10: six operations on twenty million
real rows of New York taxi trips. They run from a simple filter-and-group to a
seven-step capstone that filters, derives, joins, groups, ranks within each
borough and takes the top two. Seven implementations were run, and
every answer was proven identical across all seven before any time was
read. A benchmark whose answers are not first proven equal measures
nothing. The numbers are read from the measurement files, not typed, so
this table cannot drift from what was measured.

```{r}
#| echo: false
bench <- normalizePath(file.path(dirname(find_up("book/_quarto.yml")), "bench/2026-08-10"))
meta <- jsonlite::fromJSON(file.path(bench, "meta.json"))
tools <- names(meta$tools)
runs <- lapply(tools, function(t) jsonlite::fromJSON(file.path(bench, paste0(t, ".json"))))
names(runs) <- tools
ops <- names(meta$ops)
grid <- data.frame(`The operation` = unlist(meta$ops[ops]), check.names = FALSE)
for (t in tools) {
  grid[[meta$tools[[t]]]] <-
    sprintf("%.2f", vapply(ops, function(o) runs[[t]]$ops[[o]]$median_s, 0))
}
totals <- c("all six together",
            vapply(tools, function(t)
              sprintf("%.2f", sum(vapply(ops, function(o) runs[[t]]$ops[[o]]$median_s, 0))),
              ""))
grid <- rbind(grid, setNames(as.list(totals), names(grid)))
knitr::kable(grid, row.names = FALSE,
             caption = sprintf("Median seconds over %s rows, %s. %s.",
                               format(meta$rows, big.mark = ","),
                               meta$date, meta$machine))
```

```{r}
#| echo: false
versions <- paste(sprintf("%s %s", names(meta$versions), unlist(meta$versions)),
                  collapse = ", ")
threads <- paste(sprintf("%s on %s", names(meta$threads), unlist(meta$threads)),
                 collapse = ", ")
knitr::kable(data.frame(
  What = c("Versions", "Threads, as shipped", "Method"),
  Detail = c(versions, threads, meta$reps),
  check.names = FALSE
))
```

## What the table says, read honestly

The engine is the story, not the language. The fastest columns above are god
through DuckDB and polars run lazily, one reached from R and text, one from
Python. The widest gap in the whole table is not between the two languages
people argue about. pandas and dplyr finish within a second of each other,
and R's own data.table beats both. Whoever hands their rows to a columnar
engine wins, from either language. The folk ranking of languages is a decade
out of date; the ranking that holds is a ranking of engines.

Both spellings of polars are shown on purpose. The eager style people
type interactively and the lazy style that lets the planner fuse steps
differ by more than four times here. A table that showed only one of them
would be marketing.

god's two columns are one engine. The same sentences, the same DuckDB,
identical answers; the R side hands a frame across the boundary more
cheaply than the Python side, and that boundary is the whole difference.
What the grammar itself adds is the toll clocked live above, in either
language.

And the number that matters most is not a time. Seven implementations,
six operations, twenty million rows, identical answers, checked before the
clock was trusted. Speed varies by machine and by release; that agreement
is the part this book stakes its name on.

The record ages, and it says so: the date sits in the caption, the
libraries move, and a number this page shows is a photograph rather than a
law. When you want today's number for your own engine, ask `show_as` for
the exact query this grammar writes, and measure it where your data lives.
