---
title: "The same task, five ways"
---

{{< include ../_setup.qmd >}}

```{r}
#| include: false
# The four neighbours, attached quietly. `data.table` is the one that needs
# saying: it is not a `show_as` target and is not going to be one, so its column
# here is written by hand where the other three could have been generated. That
# is why every entry on this page is *executed and compared* rather than read:
# a hand-written cell can be wrong in a way no reader would see.
#
# **The neighbours are attached first and god is attached after them, and the
# order is load-bearing.** god and dplyr export four names in common
# (`collect`, `pick`, `rename` and `summarize`), so whichever is attached last
# wins. With dplyr last, a god pipeline handed to `collect()` reaches dplyr's
# generic, which has no method for it and says so in R's words rather than the
# book's. Attaching god last is what lets every column below be written the
# bare way its own readers write it. Where a dplyr example needs one of the
# four, it says `dplyr::` outright.
suppressMessages({
  library(dplyr, warn.conflicts = FALSE)
  library(tidyr, warn.conflicts = FALSE)
  library(data.table)
  pkgload::load_all(find_up("r-pkg/god"), export_all = FALSE, quiet = TRUE)
})

# Two tables agree when they hold the same values, whatever order they arrived
# in and whatever each library calls its integers.
#
# **Row order is normalized away here, and the entries that are about order
# check it separately.** For an unordered question, five libraries disagreeing
# about which row comes first is a fact about them rather than a defect in any
# of them, so this comparison asks only whether the same rows and values came
# back.
#
# That is not enough for a task whose answer *is* an order, and this comment
# claimed it was handled when nothing handled it: the top-five entry passed
# against its own answer reversed. `.in_order` below is the missing half, and
# it compares the sort key's sequence rather than whole rows, because ties are
# ordered differently by different libraries and that is not a disagreement
# about the answer.
.tidy <- function(d) {
  d <- as.data.frame(d, stringsAsFactors = FALSE)
  rownames(d) <- NULL
  d <- d[, order(names(d)), drop = FALSE]
  for (j in seq_along(d)) {
    if (is.factor(d[[j]])) d[[j]] <- as.character(d[[j]])
    if (is.integer(d[[j]])) d[[j]] <- as.double(d[[j]])
  }
  if (nrow(d) > 1L) d <- d[do.call(order, lapply(d, as.character)), , drop = FALSE]
  rownames(d) <- NULL
  d
}

# Every entry ends here. A disagreement stops the render, which is the whole
# point: this page's claim is that the answers are the same, and a claim the
# build does not test is a claim nobody is keeping.
.agree <- function(task, god, ...) {
  others <- list(...)
  g <- .tidy(god)
  for (nm in names(others)) {
    o <- .tidy(others[[nm]])
    if (!isTRUE(all.equal(g, o, tolerance = 1e-8, check.attributes = FALSE))) {
      stop(sprintf(
        "%s: god and %s returned different tables.\n  god has %d row(s), %d column(s); %s has %d, %d.",
        task, nm, nrow(g), ncol(g), nm, nrow(o), ncol(o)), call. = FALSE)
    }
  }
  .checked <<- .checked + length(others)
  invisible(TRUE)
}
.checked <- 0L

# **god's three spellings are checked against each other before anything else.**
# The book's central claim is that the R sentence, the Python sentence and the
# text form are one sentence. This page shows all three on every task, and a
# tab that is shown without being run is the defect the tabset guard exists to
# stop, so all three execute and all three are compared here. `answer` is the R
# one and stays the page's answer; the other two have to match it.
.agree_all <- function(task, god, ..., ordered = NULL) {
  others <- c(list(`god: run` = answer_run,
                   `god: Python` = py_eval("for_r(collect(answer_py))")),
              list(...))
  do.call(.agree, c(list(task, god), others))
  if (!is.null(ordered)) do.call(.in_order, c(list(task, ordered, god), others))
  invisible(TRUE)
}

# The sequence one column comes back in, which is the whole answer for a task
# about sorting. Compared as text so a column of any kind can be checked.
.in_order <- function(task, column, god, ...) {
  seq_of <- function(d) as.character(as.data.frame(d)[[column]])
  g <- seq_of(god)
  for (nm in names(list(...))) {
    o <- seq_of(list(...)[[nm]])
    if (!identical(g, o)) {
      stop(sprintf(
        "%s: god and %s returned `%s` in different orders.\n  god: %s\n  %s: %s",
        task, nm, column, paste(g, collapse = ", "), nm, paste(o, collapse = ", ")),
        call. = FALSE)
    }
  }
  .checked <<- .checked + length(list(...))
  invisible(TRUE)
}
```

```{python}
#| include: false
import polars as pl

# The two Python neighbours run on the same rows as the R ones. pandas already
# arrived with the setup; polars is built from it so that no fixture is defined
# twice and drifts.
pl_sales = pl.from_pandas(sales)
pl_products = pl.from_pandas(products)
pl_marks = pl.from_pandas(marks)
pl_messy = pl.from_pandas(messy)


def for_r(df):
    """A pandas frame in a shape R can actually read.

    **pandas 3.0 stores text as `StringDtype`, and reticulate does not convert
    it.** A string column crosses into R as the Python extension array itself
    rather than as a character vector, so a comparison on the R side sees an
    object where it expected two values and stops on a length mismatch that
    says nothing about the tables. Casting back to plain object dtype is what
    makes the column ordinary again. Numbers and booleans already cross
    cleanly and are left alone.
    """
    out = df.copy()
    for column in out.columns:
        if not (pd.api.types.is_numeric_dtype(out[column])
                or pd.api.types.is_bool_dtype(out[column])):
            out[column] = out[column].astype(object)
    return out
```

If you write dplyr, tidyr, pandas, polars or data.table today, this appendix is
the one that answers the question you actually have: *can it do what I already
do?* Each entry states a task in English, gives god's sentence in all three of
its spellings, gives what you would write in each of the five libraries, and
then shows the answer once, because all seven produce the same table and the
page refuses to build if any of them does not.

god's three come first, in the book's usual order: the text form that `run`
takes, then R, then Python. They are one sentence, and this page is where that
is checked rather than claimed: all three execute and all three are compared
against each other before any neighbour is looked at.

That last part is the difference between this appendix and a comparison table.
Nothing here is asserted. Every cell runs when the book is built, and the
tables are compared value by value; a disagreement stops the build rather than
reaching you. The code in each neighbour's column is written the way somebody
fluent in it would write it, not generated, so what you are reading is a fair
comparison rather than god's own output wearing five costumes.

**What this page is not** is a claim that god does everything these libraries
do. It does not, on purpose, and
[what this grammar does not do](e-coverage.qmd) is the list: regular
expressions, arbitrary functions in a cell, and eight more, each with the exit
named. What this page claims is narrower and more useful: the
ordinary work, the operations that make up most days, are all sayable in one
small vocabulary that reads the same in two languages.

The tables are the book's usual cast, described in
[the preface](../index.qmd). `sales` is fifteen rows of orders.

## Keep the rows above a value

Which orders brought in more than 150?

::: {.panel-tabset}

### god: run

```{r}
answer_run <- run('
  sales
    then keep where ([revenue] > 150)
')
```

### god: R

```{r}
answer <- sales |> keep(revenue > 150) |> collect()
```

### god: Python

```{python}
answer_py = sales >> keep(col.revenue > 150)
```

### dplyr

```{r}
by_dplyr <- sales |> filter(revenue > 150)
```

### pandas

```{python}
by_pandas = sales[sales.revenue > 150]
```

### polars

```{python}
by_polars = pl_sales.filter(pl.col("revenue") > 150)
```

### data.table

```{r}
by_dt <- as.data.table(sales)[revenue > 150]
```

:::

```{r}
#| echo: false
.agree_all("keep the rows above a value", answer,
       dplyr = by_dplyr, pandas = py_eval("for_r(by_pandas)"),
       polars = py_eval("for_r(by_polars.to_pandas())"), data.table = by_dt)
answer
```

Seven spellings of one idea, and three of them are god's. Its four neighbours
each read as an operation on a frame; god's reads as a sentence, and reads the
same whichever of the three you write it in.

## Total a column for each group

What did each product earn in total, and across how many orders?

::: {.panel-tabset}

### god: run

```{r}
answer_run <- run('
  sales
    then summarize [earned] as total([revenue]), [orders] as row_count() by [product]
')
```

### god: R

```{r}
answer <- sales |>
  summarize(earned = total(revenue), orders = row_count(), by = product) |>
  collect()
```

### god: Python

```{python}
answer_py = sales >> summarize(earned = total(col.revenue), orders = row_count(),
                               by = col.product)
```

### dplyr

```{r}
by_dplyr <- sales |>
  group_by(product) |>
  summarise(earned = sum(revenue), orders = n(), .groups = "drop")
```

### pandas

```{python}
by_pandas = (sales
  .groupby("product")
  .agg(earned=("revenue", "sum"), orders=("revenue", "size"))
  .reset_index())
```

### polars

```{python}
by_polars = pl_sales.group_by("product").agg(
    pl.col("revenue").sum().alias("earned"),
    pl.len().alias("orders"))
```

### data.table

```{r}
by_dt <- as.data.table(sales)[, .(earned = sum(revenue), orders = .N), by = product]
```

:::

```{r}
#| echo: false
.agree_all("total a column for each group", answer,
       dplyr = by_dplyr, pandas = py_eval("for_r(by_pandas)"),
       polars = py_eval("for_r(by_polars.to_pandas())"), data.table = by_dt)
answer
```

Note how each library spells *count the rows*: `n()`, `size`, `pl.len()`, `.N`.
god says `row_count()`, which is the same word in R and in Python.

## Turn rows into columns

Each student answered two questions. Put one row per student, one column per
question.

::: {.panel-tabset}

### god: run

```{r}
answer_run <- run('
  marks
    then widen name [question], value [mark] by [student]
')
```

### god: R

```{r}
answer <- marks |> widen(name = question, value = mark, by = student) |> collect()
```

### god: Python

```{python}
answer_py = marks >> widen(name = col.question, value = col.mark,
                           by = col.student)
```

### tidyr

```{r}
by_dplyr <- marks |> pivot_wider(names_from = question, values_from = mark)
```

### pandas

```{python}
by_pandas = (marks
  .pivot(index="student", columns="question", values="mark")
  .reset_index()
  .rename_axis(None, axis=1))
```

### polars

```{python}
by_polars = pl_marks.pivot(on="question", index="student", values="mark")
```

### data.table

```{r}
by_dt <- dcast(as.data.table(marks), student ~ question, value.var = "mark")
```

:::

```{r}
#| echo: false
.agree_all("turn rows into columns", answer,
       dplyr = by_dplyr, pandas = py_eval("for_r(by_pandas)"),
       polars = py_eval("for_r(by_polars.to_pandas())"), data.table = by_dt)
answer
```

Five names for one idea: `pivot_wider`, `pivot`, `pivot`, `dcast`, and `widen`.
Only one of them says in the verb what happens to the table.

The `by` is written out here, and it did not have to be. Left off, god works
out that the rows going together are the ones agreeing on every other column,
and then *says so*: the answer arrives with the assumption printed above it,
naming `student` and offering the phrasing that would have removed the doubt.
That is the behaviour, not a warning about it. It is written out above so that
all three of god's spellings on this page are the same sentence.

## Keep the rows matching a word

Which orders came from the West?

::: {.panel-tabset}

### god: run

```{r}
answer_run <- run('
  sales
    then keep where ([region] is "West")
')
```

### god: R

```{r}
answer <- sales |> keep(region == "West") |> collect()
```

### god: Python

```{python}
answer_py = sales >> keep(col.region == "West")
```

### dplyr

```{r}
by_dplyr <- sales |> filter(region == "West")
```

### pandas

```{python}
by_pandas = sales[sales.region == "West"]
```

### polars

```{python}
by_polars = pl_sales.filter(pl.col("region") == "West")
```

### data.table

```{r}
by_dt <- as.data.table(sales)[region == "West"]
```

:::

```{r}
#| echo: false
.agree_all("keep the rows matching a word", answer,
       dplyr = by_dplyr, pandas = py_eval("for_r(by_pandas)"),
       polars = py_eval("for_r(by_polars.to_pandas())"), data.table = by_dt)
answer
```

god writes `is` in the text form and `==` in both host languages, because
that is what each host already means by it.

## Choose some columns

Just the product and what it earned.

::: {.panel-tabset}

### god: run

```{r}
answer_run <- run('
  sales
    then pick [product, revenue]
')
```

### god: R

```{r}
answer <- sales |> pick(product, revenue) |> collect()
```

### god: Python

```{python}
answer_py = sales >> pick(col.product, col.revenue)
```

### dplyr

```{r}
by_dplyr <- sales |> select(product, revenue)
```

### pandas

```{python}
by_pandas = sales[["product", "revenue"]]
```

### polars

```{python}
by_polars = pl_sales.select("product", "revenue")
```

### data.table

```{r}
by_dt <- as.data.table(sales)[, .(product, revenue)]
```

:::

```{r}
#| echo: false
.agree_all("choose some columns", answer,
       dplyr = by_dplyr, pandas = py_eval("for_r(by_pandas)"),
       polars = py_eval("for_r(by_polars.to_pandas())"), data.table = by_dt)
answer
```

Four of the five name the columns they want. god does too, and says `pick`,
which is a verb where `select` and `[[ ]]` are not.

## Drop a column

Everything except what it cost.

::: {.panel-tabset}

### god: run

```{r}
answer_run <- run('
  sales
    then pick all_but [cost]
')
```

### god: R

```{r}
answer <- sales |> pick(all_but(cost)) |> collect()
```

### god: Python

```{python}
answer_py = sales >> pick(all_but(col.cost))
```

### dplyr

```{r}
by_dplyr <- sales |> select(-cost)
```

### pandas

```{python}
by_pandas = sales.drop(columns=["cost"])
```

### polars

```{python}
by_polars = pl_sales.drop("cost")
```

### data.table

```{r}
by_dt <- as.data.table(sales)[, !"cost"]
```

:::

```{r}
#| echo: false
.agree_all("drop a column", answer,
       dplyr = by_dplyr, pandas = py_eval("for_r(by_pandas)"),
       polars = py_eval("for_r(by_polars.to_pandas())"), data.table = by_dt)
answer
```

`all_but` is the same word wherever it stands, where the neighbours reach for
a minus sign, a `drop`, and a `!`.

## Add a computed column

What did each order make?

::: {.panel-tabset}

### god: run

```{r}
answer_run <- run('
  sales
    then add [margin] as ([revenue] - [cost])
    then pick [product, margin]
')
```

### god: R

```{r}
answer <- sales |> add(margin = revenue - cost) |> pick(product, margin) |> collect()
```

### god: Python

```{python}
answer_py = (sales >> add(margin = col.revenue - col.cost)
                   >> pick(col.product, col.margin))
```

### dplyr

```{r}
by_dplyr <- sales |> mutate(margin = revenue - cost) |> select(product, margin)
```

### pandas

```{python}
by_pandas = sales.assign(margin=sales.revenue - sales.cost)[["product", "margin"]]
```

### polars

```{python}
by_polars = (pl_sales
  .with_columns((pl.col("revenue") - pl.col("cost")).alias("margin"))
  .select("product", "margin"))
```

### data.table

```{r}
by_dt <- as.data.table(sales)[, .(product, margin = revenue - cost)]
```

:::

```{r}
#| echo: false
.agree_all("add a computed column", answer,
       dplyr = by_dplyr, pandas = py_eval("for_r(by_pandas)"),
       polars = py_eval("for_r(by_polars.to_pandas())"), data.table = by_dt)
answer
```

`mutate`, `assign`, `with_columns`, `:=`, and `add`. Only the last one is a
word a reader who has never seen it would guess correctly.

## Average a column for each group

What is the typical order in each region?

::: {.panel-tabset}

### god: run

```{r}
answer_run <- run('
  sales
    then summarize [typical] as average([revenue]) by [region]
')
```

### god: R

```{r}
answer <- sales |> summarize(typical = average(revenue), by = region) |> collect()
```

### god: Python

```{python}
answer_py = sales >> summarize(typical = average(col.revenue), by = col.region)
```

### dplyr

```{r}
by_dplyr <- sales |>
  group_by(region) |>
  summarise(typical = mean(revenue), .groups = "drop")
```

### pandas

```{python}
by_pandas = (sales
  .groupby("region")
  .agg(typical=("revenue", "mean"))
  .reset_index())
```

### polars

```{python}
by_polars = pl_sales.group_by("region").agg(
    pl.col("revenue").mean().alias("typical"))
```

### data.table

```{r}
by_dt <- as.data.table(sales)[, .(typical = mean(revenue)), by = region]
```

:::

```{r}
#| echo: false
.agree_all("average a column for each group", answer,
       dplyr = by_dplyr, pandas = py_eval("for_r(by_pandas)"),
       polars = py_eval("for_r(by_polars.to_pandas())"), data.table = by_dt)
answer
```

`by` is an argument in god rather than a step, so grouping cannot outlive the
sentence that asked for it.

## Sort and take the top few

The five biggest orders.

::: {.panel-tabset}

### god: run

```{r}
answer_run <- run('
  sales
    then sort [revenue] descending, [product]
    then take 5
    then pick [product, revenue]
')
```

### god: R

```{r}
answer <- sales |>
  sort(descending(revenue), product) |>
  take(5) |>
  pick(product, revenue) |>
  collect()
```

### god: Python

```{python}
answer_py = (sales >> sort(descending(col.revenue), col.product) >> take(5)
                   >> pick(col.product, col.revenue))
```

### dplyr

```{r}
by_dplyr <- sales |> arrange(desc(revenue), product) |> head(5) |> select(product, revenue)
```

### pandas

```{python}
by_pandas = (sales
  .sort_values(["revenue", "product"], ascending=[False, True])
  .head(5)[["product", "revenue"]])
```

### polars

```{python}
by_polars = (pl_sales
  .sort(["revenue", "product"], descending=[True, False])
  .head(5)
  .select("product", "revenue"))
```

### data.table

```{r}
by_dt <- head(as.data.table(sales)[order(-revenue, product)], 5)[, .(product, revenue)]
```

:::

```{r}
#| echo: false
.agree_all("sort and take the top few", answer,
       dplyr = by_dplyr, pandas = py_eval("for_r(by_pandas)"),
       polars = py_eval("for_r(by_polars.to_pandas())"), data.table = by_dt,
       ordered = "revenue")
answer
```

Every library here spells *reverse the order* differently: `desc`,
`ascending=False`, `descending=True`, a minus sign. god says `descending`.

The second sort key is not decoration, and this page needed it to build.
Two orders earned 150, so the fifth row of a top five is a tie, and which of
them wins is whatever each library's sort happened to do. pandas sorts with an
unstable algorithm by default, so it answered one way on one machine and the
other way on another, and the comparison here refused to build until the
question was made a fair one. A top five over a column with repeats is not a
settled answer, which is the law this grammar is built on, and the repair is the
same in all six spellings: name a second key.

## Take the rows at the far end

The three smallest orders, smallest last.

::: {.panel-tabset}

### god: run

```{r}
answer_run <- run('
  sales
    then sort [revenue] descending
    then take_last 3
    then pick [product, revenue]
')
```

### god: R

```{r}
answer <- sales |> sort(descending(revenue)) |> take_last(3) |> pick(product, revenue) |> collect()
```

### god: Python

```{python}
answer_py = (sales >> sort(descending(col.revenue)) >> take_last(3)
                   >> pick(col.product, col.revenue))
```

### dplyr

```{r}
by_dplyr <- sales |> arrange(desc(revenue)) |> tail(3) |> select(product, revenue)
```

### pandas

```{python}
by_pandas = (sales
  .sort_values("revenue", ascending=False)
  .tail(3)[["product", "revenue"]])
```

### polars

```{python}
by_polars = (pl_sales
  .sort("revenue", descending=True)
  .tail(3)
  .select("product", "revenue"))
```

### data.table

```{r}
by_dt <- tail(as.data.table(sales)[order(-revenue)], 3)[, .(product, revenue)]
```

:::

```{r}
#| echo: false
.agree_all("take the rows at the far end", answer,
       dplyr = by_dplyr, pandas = py_eval("for_r(by_pandas)"),
       polars = py_eval("for_r(by_polars.to_pandas())"), data.table = by_dt,
       ordered = "revenue")
answer
```

This is the entry that produced a word. `take_last` came out of the polars and
pandas sweeps, both of which called it `tail`, because sorting the other way
and taking the first three returns the same rows backwards.

## Bring in a column from another table

Who makes each product?

::: {.panel-tabset}

### god: run

```{r}
answer_run <- run('
  sales
    then join products by [product]
    then pick [product, maker, revenue]
')
```

### god: R

```{r}
answer <- sales |>
  join(products, by = product) |>
  pick(product, maker, revenue) |>
  collect()
```

### god: Python

```{python}
answer_py = (sales >> join(products, by = col.product)
                   >> pick(col.product, col.maker, col.revenue))
```

### dplyr

```{r}
by_dplyr <- sales |>
  left_join(products, by = "product") |>
  select(product, maker, revenue)
```

### pandas

```{python}
by_pandas = sales.merge(products, on="product", how="left")[["product", "maker", "revenue"]]
```

### polars

```{python}
by_polars = (pl_sales
  .join(pl_products, on="product", how="left")
  .select("product", "maker", "revenue"))
```

### data.table

```{r}
by_dt <- merge(as.data.table(sales), as.data.table(products),
               by = "product", all.x = TRUE)[, .(product, maker, revenue)]
```

:::

```{r}
#| echo: false
.agree_all("bring in a column from another table", answer,
       dplyr = by_dplyr, pandas = py_eval("for_r(by_pandas)"),
       polars = py_eval("for_r(by_polars.to_pandas())"), data.table = by_dt)
answer
```

god has one `join` and an `unmatched` word for whose rows survive, where the
neighbours have four verbs or a `how=` argument. Note which way the default
falls: god keeps the rows it started with, so `Sprocket` survives with no
maker rather than disappearing. Writing this entry with `inner_join` is what
the page's own comparison caught, on the row count.

## Ask whether a row has a partner

Which orders are for a product the catalogue lists?

::: {.panel-tabset}

### god: run

```{r}
answer_run <- run('
  sales
    then keep where matching(products, by [product])
    then pick [product, revenue]
')
```

### god: R

```{r}
answer <- sales |>
  keep(matching(products, by = product)) |>
  pick(product, revenue) |>
  collect()
```

### god: Python

```{python}
answer_py = (sales >> keep(matching(products, by = col.product))
                   >> pick(col.product, col.revenue))
```

### dplyr

```{r}
by_dplyr <- sales |> semi_join(products, by = "product") |> select(product, revenue)
```

### pandas

```{python}
by_pandas = sales[sales["product"].isin(products["product"])][["product", "revenue"]]
```

### polars

```{python}
by_polars = (pl_sales
  .join(pl_products.select("product"), on="product", how="semi")
  .select("product", "revenue"))
```

### data.table

```{r}
by_dt <- as.data.table(sales)[product %chin% products$product, .(product, revenue)]
```

:::

```{r}
#| echo: false
.agree_all("ask whether a row has a partner", answer,
       dplyr = by_dplyr, pandas = py_eval("for_r(by_pandas)"),
       polars = py_eval("for_r(by_polars.to_pandas())"), data.table = by_dt)
answer
```

`matching` is a question rather than a join, so it cannot multiply rows. That
is a promise `merge` and `isin` do not make in the same breath.

## Turn columns into rows

Two measures per order, one row each.

::: {.panel-tabset}

### god: run

```{r}
answer_run <- run('
  sales
    then pick [product, revenue, cost]
    then lengthen [revenue, cost] as name [measure], value [amount]
')
```

### god: R

```{r}
answer <- sales |>
  pick(product, revenue, cost) |>
  lengthen(revenue, cost, name = measure, value = amount) |>
  collect()
```

### god: Python

```{python}
answer_py = (sales >> pick(col.product, col.revenue, col.cost)
                   >> lengthen(col.revenue, col.cost, name = col.measure,
                               value = col.amount))
```

### tidyr

```{r}
by_dplyr <- sales |>
  select(product, revenue, cost) |>
  pivot_longer(c(revenue, cost), names_to = "measure", values_to = "amount")
```

### pandas

```{python}
by_pandas = (sales[["product", "revenue", "cost"]]
  .melt(id_vars="product", var_name="measure", value_name="amount"))
```

### polars

```{python}
by_polars = (pl_sales
  .select("product", "revenue", "cost")
  .unpivot(index="product", variable_name="measure", value_name="amount"))
```

### data.table

```{r}
by_dt <- melt(as.data.table(sales)[, .(product, revenue, cost)],
              id.vars = "product", variable.name = "measure",
              value.name = "amount", variable.factor = FALSE)
```

:::

```{r}
#| echo: false
.agree_all("turn columns into rows", answer,
       dplyr = by_dplyr, pandas = py_eval("for_r(by_pandas)"),
       polars = py_eval("for_r(by_polars.to_pandas())"), data.table = by_dt)
answer
```

`lengthen` and `widen` are one idea read in both directions, and they are the
only pair here that says so in the words themselves.

## Choose between values

Label each order by size.

::: {.panel-tabset}

### god: run

```{r}
answer_run <- run('
  sales
    then add [size] as when(([revenue] >= 200), "big", ([revenue] >= 100), "medium", otherwise "small")
    then pick [product, revenue, size]
')
```

### god: R

```{r}
answer <- sales |>
  add(size = when(revenue >= 200, "big", revenue >= 100, "medium", otherwise = "small")) |>
  pick(product, revenue, size) |>
  collect()
```

### god: Python

```{python}
answer_py = (sales >> add(size = when(col.revenue >= 200, "big",
                                      col.revenue >= 100, "medium",
                                      otherwise = "small"))
                   >> pick(col.product, col.revenue, col.size))
```

### dplyr

```{r}
by_dplyr <- sales |>
  mutate(size = case_when(revenue >= 200 ~ "big",
                          revenue >= 100 ~ "medium",
                          .default = "small")) |>
  select(product, revenue, size)
```

### pandas

```{python}
import numpy as np
by_pandas = sales.assign(size=np.select(
    [sales.revenue >= 200, sales.revenue >= 100],
    ["big", "medium"], default="small"))[["product", "revenue", "size"]]
```

### polars

```{python}
by_polars = (pl_sales
  .with_columns(pl.when(pl.col("revenue") >= 200).then(pl.lit("big"))
                  .when(pl.col("revenue") >= 100).then(pl.lit("medium"))
                  .otherwise(pl.lit("small")).alias("size"))
  .select("product", "revenue", "size"))
```

### data.table

```{r}
by_dt <- as.data.table(sales)[, .(product, revenue,
  size = fifelse(revenue >= 200, "big", fifelse(revenue >= 100, "medium", "small")))]
```

:::

```{r}
#| echo: false
.agree_all("choose between values", answer,
       dplyr = by_dplyr, pandas = py_eval("for_r(by_pandas)"),
       polars = py_eval("for_r(by_polars.to_pandas())"), data.table = by_dt)
answer
```

polars needs `pl.lit` around every one of those words, because a bare string
there is read as a column name. That is the defect class this book's printed-code
harness was built to catch, and it is why this page runs everything it shows.

## Number the rows in order

Where does each order place, biggest first?

::: {.panel-tabset}

### god: run

```{r}
answer_run <- run('
  sales
    then add [place] as rank([revenue] descending)
    then pick [product, revenue, place]
')
```

### god: R

```{r}
answer <- sales |>
  add(place = rank(descending(revenue))) |>
  pick(product, revenue, place) |>
  collect()
```

### god: Python

```{python}
answer_py = (sales >> add(place = rank(descending(col.revenue)))
                   >> pick(col.product, col.revenue, col.place))
```

### dplyr

```{r}
by_dplyr <- sales |>
  mutate(place = min_rank(desc(revenue))) |>
  select(product, revenue, place)
```

### pandas

```{python}
by_pandas = sales.assign(
    place=sales.revenue.rank(ascending=False, method="min").astype("int64")
)[["product", "revenue", "place"]]
```

### polars

```{python}
by_polars = (pl_sales
  .with_columns(pl.col("revenue").rank(method="min", descending=True)
                  .cast(pl.Int64).alias("place"))
  .select("product", "revenue", "place"))
```

### data.table

```{r}
by_dt <- as.data.table(sales)[, .(product, revenue,
                                  place = frank(-revenue, ties.method = "min"))]
```

:::

```{r}
#| echo: false
.agree_all("number the rows in order", answer,
       dplyr = by_dplyr, pandas = py_eval("for_r(by_pandas)"),
       polars = py_eval("for_r(by_polars.to_pandas())"), data.table = by_dt)
answer
```

`rank` is the one window word that needs no `sort` in front of it, because its
argument already says what the order is.

## Total as you go down the rows

How does revenue accumulate, smallest order first?

::: {.panel-tabset}

### god: run

```{r}
answer_run <- run('
  sales
    then sort [revenue]
    then add [so_far] as running_total([revenue])
    then pick [revenue, so_far]
')
```

### god: R

```{r}
answer <- sales |>
  sort(revenue) |>
  add(so_far = running_total(revenue)) |>
  pick(revenue, so_far) |>
  collect()
```

### god: Python

```{python}
answer_py = (sales >> sort(col.revenue)
                   >> add(so_far = running_total(col.revenue))
                   >> pick(col.revenue, col.so_far))
```

### dplyr

```{r}
by_dplyr <- sales |>
  arrange(revenue) |>
  mutate(so_far = cumsum(revenue)) |>
  select(revenue, so_far)
```

### pandas

```{python}
_sorted = sales.sort_values("revenue", kind="stable")
by_pandas = _sorted.assign(so_far=_sorted.revenue.cumsum())[["revenue", "so_far"]]
```

### polars

```{python}
by_polars = (pl_sales
  .sort("revenue")
  .with_columns(pl.col("revenue").cum_sum().alias("so_far"))
  .select("revenue", "so_far"))
```

### data.table

```{r}
by_dt <- as.data.table(sales)[order(revenue), .(revenue, so_far = cumsum(revenue))]
```

:::

```{r}
#| echo: false
.agree_all("total as you go down the rows", answer,
       dplyr = by_dplyr, pandas = py_eval("for_r(by_pandas)"),
       polars = py_eval("for_r(by_polars.to_pandas())"), data.table = by_dt,
       ordered = "so_far")
answer
```

god refuses `running_total` without a `sort` in front of it. The other four
compute happily on whatever order the rows happen to be in, which is an answer
that can change between runs.

## Trim the spaces off a text column

The names arrived padded.

::: {.panel-tabset}

### god: run

```{r}
answer_run <- run('
  messy
    then add [name] as trim([raw])
    then pick [name, n]
')
```

### god: R

```{r}
answer <- messy |> add(name = trim(raw)) |> pick(name, n) |> collect()
```

### god: Python

```{python}
answer_py = messy >> add(name = trim(col.raw)) >> pick(col.name, col.n)
```

### dplyr

```{r}
by_dplyr <- messy |> mutate(name = trimws(raw)) |> select(name, n)
```

### pandas

```{python}
by_pandas = messy.assign(name=messy.raw.str.strip())[["name", "n"]]
```

### polars

```{python}
by_polars = (pl_messy
  .with_columns(pl.col("raw").str.strip_chars().alias("name"))
  .select("name", "n"))
```

### data.table

```{r}
by_dt <- as.data.table(messy)[, .(name = trimws(raw), n)]
```

:::

```{r}
#| echo: false
.agree_all("trim the spaces off a text column", answer,
       dplyr = by_dplyr, pandas = py_eval("for_r(by_pandas)"),
       polars = py_eval("for_r(by_polars.to_pandas())"), data.table = by_dt)
answer
```

`trimws`, `str.strip`, `str.strip_chars`, `trimws`, `trim`. The same operation,
five spellings, and no two libraries agree.

## Change a column's case

Shout the region.

::: {.panel-tabset}

### god: run

```{r}
answer_run <- run('
  sales
    then add [loud] as upper([region])
    then pick [region, loud]
')
```

### god: R

```{r}
answer <- sales |> add(loud = upper(region)) |> pick(region, loud) |> collect()
```

### god: Python

```{python}
answer_py = (sales >> add(loud = upper(col.region))
                   >> pick(col.region, col.loud))
```

### dplyr

```{r}
by_dplyr <- sales |> mutate(loud = toupper(region)) |> select(region, loud)
```

### pandas

```{python}
by_pandas = sales.assign(loud=sales.region.str.upper())[["region", "loud"]]
```

### polars

```{python}
by_polars = (pl_sales
  .with_columns(pl.col("region").str.to_uppercase().alias("loud"))
  .select("region", "loud"))
```

### data.table

```{r}
by_dt <- as.data.table(sales)[, .(region, loud = toupper(region))]
```

:::

```{r}
#| echo: false
.agree_all("change a column's case", answer,
       dplyr = by_dplyr, pandas = py_eval("for_r(by_pandas)"),
       polars = py_eval("for_r(by_polars.to_pandas())"), data.table = by_dt)
answer
```

`upper` and `lower` are the two everyone has. god spells them the shortest
way that still says which direction.

## Take a piece of a text column

The part of each product name before the letter `d`.

::: {.panel-tabset}

### god: run

```{r}
answer_run <- run('
  sales
    then add [head] as split_text([product], "d", 1)
    then pick [product, head]
')
```

### god: R

```{r}
answer <- sales |> add(head = split_text(product, "d", 1)) |> pick(product, head) |> collect()
```

### god: Python

```{python}
answer_py = (sales >> add(head = split_text(col.product, "d", 1))
                   >> pick(col.product, col.head))
```

### dplyr

```{r}
by_dplyr <- sales |>
  mutate(head = vapply(strsplit(product, "d", fixed = TRUE), `[`, "", 1L)) |>
  select(product, head)
```

### pandas

```{python}
by_pandas = sales.assign(
    head=sales["product"].str.split("d").str[0])[["product", "head"]]
```

### polars

```{python}
by_polars = (pl_sales
  .with_columns(pl.col("product").str.split("d").list.get(0).alias("head"))
  .select("product", "head"))
```

### data.table

```{r}
by_dt <- as.data.table(sales)[, .(product,
  head = vapply(strsplit(product, "d", fixed = TRUE), `[`, "", 1L))]
```

:::

```{r}
#| echo: false
.agree_all("take a piece of a text column", answer,
       dplyr = by_dplyr, pandas = py_eval("for_r(by_pandas)"),
       polars = py_eval("for_r(by_polars.to_pandas())"), data.table = by_dt)
answer
```

god counts pieces from 1, and says which piece it wants in the call. The two
R spellings need a `vapply` around a list, which is the shape this grammar is
trying to spare you.

## Put text back together

Label each order with its region and product.

::: {.panel-tabset}

### god: run

```{r}
answer_run <- run('
  sales
    then add [label] as join_text([region], " ", [product])
    then pick [label, revenue]
')
```

### god: R

```{r}
answer <- sales |>
  add(label = join_text(region, " ", product)) |>
  pick(label, revenue) |>
  collect()
```

### god: Python

```{python}
answer_py = (sales >> add(label = join_text(col.region, " ", col.product))
                   >> pick(col.label, col.revenue))
```

### dplyr

```{r}
by_dplyr <- sales |>
  mutate(label = paste0(region, " ", product)) |>
  select(label, revenue)
```

### pandas

```{python}
by_pandas = sales.assign(
    label=sales.region + " " + sales["product"])[["label", "revenue"]]
```

### polars

```{python}
by_polars = (pl_sales
  .with_columns(pl.concat_str([pl.col("region"), pl.lit(" "), pl.col("product")])
                  .alias("label"))
  .select("label", "revenue"))
```

### data.table

```{r}
by_dt <- as.data.table(sales)[, .(label = paste0(region, " ", product), revenue)]
```

:::

```{r}
#| echo: false
.agree_all("put text back together", answer,
       dplyr = by_dplyr, pandas = py_eval("for_r(by_pandas)"),
       polars = py_eval("for_r(by_polars.to_pandas())"), data.table = by_dt)
answer
```

`join_text` is the sweep's second finding: god could take text apart with
`split_text` and could not put it back, which is an asymmetry no law asked for.

## Pull the year out of a date

Which year did each order land in?

::: {.panel-tabset}

### god: run

```{r}
answer_run <- run('
  sales
    then add [d] as to_date([date])
    then add [y] as year([d]), [m] as month([d])
    then pick [date, y, m]
')
```

### god: R

```{r}
answer <- sales |>
  add(d = to_date(date)) |>
  add(y = year(d), m = month(d)) |>
  pick(date, y, m) |>
  collect()
```

### god: Python

```{python}
answer_py = (sales >> add(d = to_date(col.date))
                   >> add(y = year(col.d), m = month(col.d))
                   >> pick(col.date, col.y, col.m))
```

### dplyr

```{r}
by_dplyr <- sales |>
  mutate(d = as.Date(date),
         y = as.numeric(format(d, "%Y")),
         m = as.numeric(format(d, "%m"))) |>
  select(date, y, m)
```

### pandas

```{python}
_d = pd.to_datetime(sales.date)
by_pandas = sales.assign(y=_d.dt.year, m=_d.dt.month)[["date", "y", "m"]]
```

### polars

```{python}
by_polars = (pl_sales
  .with_columns(pl.col("date").str.to_date().alias("d"))
  .with_columns(pl.col("d").dt.year().alias("y"),
                pl.col("d").dt.month().alias("m"))
  .select("date", "y", "m"))
```

### data.table

```{r}
by_dt <- as.data.table(sales)[, {
  d <- as.Date(date)
  .(date, y = as.numeric(format(d, "%Y")), m = as.numeric(format(d, "%m")))
}]
```

:::

```{r}
#| echo: false
.agree_all("pull the year out of a date", answer,
       dplyr = by_dplyr, pandas = py_eval("for_r(by_pandas)"),
       polars = py_eval("for_r(by_polars.to_pandas())"), data.table = by_dt)
answer
```

Every library here needs the text turned into a date first, and god is the
only one that makes you say so: `to_date` is never applied on your behalf.

## Fill the holes in a column

The catalogue does not list every product. Say so rather than leaving a blank.

::: {.panel-tabset}

### god: run

```{r}
answer_run <- run('
  sales
    then join products by [product]
    then fill_missing [maker] as "unlisted"
    then pick [product, maker]
')
```

### god: R

```{r}
answer <- sales |>
  join(products, by = product) |>
  fill_missing(maker = "unlisted") |>
  pick(product, maker) |>
  collect()
```

### god: Python

```{python}
answer_py = (sales >> join(products, by = col.product)
                   >> fill_missing(maker = "unlisted")
                   >> pick(col.product, col.maker))
```

### dplyr

```{r}
by_dplyr <- sales |>
  left_join(products, by = "product") |>
  mutate(maker = coalesce(maker, "unlisted")) |>
  select(product, maker)
```

### pandas

```{python}
by_pandas = (sales
  .merge(products, on="product", how="left")
  .fillna({"maker": "unlisted"})[["product", "maker"]])
```

### polars

```{python}
by_polars = (pl_sales
  .join(pl_products, on="product", how="left")
  .with_columns(pl.col("maker").fill_null("unlisted"))
  .select("product", "maker"))
```

### data.table

```{r}
by_dt <- merge(as.data.table(sales), as.data.table(products),
               by = "product", all.x = TRUE)[
  , .(product, maker = fifelse(is.na(maker), "unlisted", maker))]
```

:::

```{r}
#| echo: false
.agree_all("fill the holes in a column", answer,
       dplyr = by_dplyr, pandas = py_eval("for_r(by_pandas)"),
       polars = py_eval("for_r(by_polars.to_pandas())"), data.table = by_dt)
answer
```

`fill_missing` refuses a filler of the wrong kind: a number into a text
column is a refusal rather than a silent coercion, which is the one place
this list's neighbours differ from each other most.

## Drop the rows with a hole

Only the orders whose product the catalogue lists.

::: {.panel-tabset}

### god: run

```{r}
answer_run <- run('
  sales
    then join products by [product]
    then drop_missing [maker]
    then pick [product, maker, revenue]
')
```

### god: R

```{r}
answer <- sales |>
  join(products, by = product) |>
  drop_missing(maker) |>
  pick(product, maker, revenue) |>
  collect()
```

### god: Python

```{python}
answer_py = (sales >> join(products, by = col.product)
                   >> drop_missing(col.maker)
                   >> pick(col.product, col.maker, col.revenue))
```

### dplyr

```{r}
by_dplyr <- sales |>
  left_join(products, by = "product") |>
  filter(!is.na(maker)) |>
  select(product, maker, revenue)
```

### pandas

```{python}
by_pandas = (sales
  .merge(products, on="product", how="left")
  .dropna(subset=["maker"])[["product", "maker", "revenue"]])
```

### polars

```{python}
by_polars = (pl_sales
  .join(pl_products, on="product", how="left")
  .drop_nulls("maker")
  .select("product", "maker", "revenue"))
```

### data.table

```{r}
by_dt <- merge(as.data.table(sales), as.data.table(products),
               by = "product", all.x = TRUE)[!is.na(maker),
                                             .(product, maker, revenue)]
```

:::

```{r}
#| echo: false
.agree_all("drop the rows with a hole", answer,
       dplyr = by_dplyr, pandas = py_eval("for_r(by_pandas)"),
       polars = py_eval("for_r(by_polars.to_pandas())"), data.table = by_dt)
answer
```

`drop_missing` names the column it is judging. A bare `dropna()` judges every
column at once, which is a different question wearing the same name.

## Remove exact repeats

Which regions appear at all?

::: {.panel-tabset}

### god: run

```{r}
answer_run <- run('
  sales
    then pick [region]
    then drop_duplicates
')
```

### god: R

```{r}
answer <- sales |> pick(region) |> drop_duplicates() |> collect()
```

### god: Python

```{python}
answer_py = sales >> pick(col.region) >> drop_duplicates()
```

### dplyr

```{r}
by_dplyr <- sales |> select(region) |> distinct()
```

### pandas

```{python}
by_pandas = sales[["region"]].drop_duplicates()
```

### polars

```{python}
by_polars = pl_sales.select("region").unique()
```

### data.table

```{r}
by_dt <- unique(as.data.table(sales)[, .(region)])
```

:::

```{r}
#| echo: false
.agree_all("remove exact repeats", answer,
       dplyr = by_dplyr, pandas = py_eval("for_r(by_pandas)"),
       polars = py_eval("for_r(by_polars.to_pandas())"), data.table = by_dt)
answer
```

`distinct`, `drop_duplicates`, `unique`, `unique`, `drop_duplicates`. god took
the longest of the five names because it is the one that says what happens to
the table rather than what the table becomes.

## Make the missing combinations appear

Every region crossed with every product, including the pairs nobody sold.

::: {.panel-tabset}

### god: run

```{r}
answer_run <- run('
  sales
    then summarize [earned] as total([revenue]) by [region, product]
    then add_combinations [region, product]
    then fill_missing [earned] as 0
')
```

### god: R

```{r}
answer <- sales |>
  summarize(earned = total(revenue), by = c(region, product)) |>
  add_combinations(region, product) |>
  fill_missing(earned = 0) |>
  collect()
```

### god: Python

```{python}
answer_py = (sales >> summarize(earned = total(col.revenue),
                                by = [col.region, col.product])
                   >> add_combinations(col.region, col.product)
                   >> fill_missing(earned = 0))
```

### tidyr

```{r}
by_dplyr <- sales |>
  group_by(region, product) |>
  summarise(earned = sum(revenue), .groups = "drop") |>
  complete(region, product, fill = list(earned = 0))
```

### pandas

```{python}
_g = sales.groupby(["region", "product"], as_index=False).agg(
    earned=("revenue", "sum"))
_full = pd.MultiIndex.from_product(
    [sorted(sales.region.unique()), sorted(sales["product"].unique())],
    names=["region", "product"])
by_pandas = (_g.set_index(["region", "product"])
  .reindex(_full, fill_value=0)
  .reset_index())
```

### polars

```{python}
_g = pl_sales.group_by("region", "product").agg(
    pl.col("revenue").sum().alias("earned"))
by_polars = (pl_sales.select("region").unique()
  .join(pl_sales.select("product").unique(), how="cross")
  .join(_g, on=["region", "product"], how="left")
  .with_columns(pl.col("earned").fill_null(0)))
```

### data.table

```{r}
by_dt <- as.data.table(sales)[, .(earned = sum(revenue)), by = .(region, product)][
  CJ(region = unique(sales$region), product = unique(sales$product)),
  on = .(region, product)][is.na(earned), earned := 0][]
```

:::

```{r}
#| echo: false
.agree_all("make the missing combinations appear", answer,
       dplyr = by_dplyr, pandas = py_eval("for_r(by_pandas)"),
       polars = py_eval("for_r(by_polars.to_pandas())"), data.table = by_dt)
answer
```

This is the entry where the five diverge most, and the reason is that the
absent rows have to be conjured rather than found. god splits it in two on
purpose: `add_combinations` makes the rows and `fill_missing` says what a new
one holds, because whether an absent pair means zero is a question about your
data rather than about the verb.

```{r}
#| echo: false
#| output: asis
cat(sprintf(
  "*This page compared %d answers against god's while it was built, and every one of them agreed.*\n",
  .checked))
```

Whichever of the five you arrived with, the sentence you would write in god is
the same sentence, in either language, and it was checked against your own
tool's answer before this page reached you.
