---
title: "The same task, five ways"
---

{{< include ../_setup.qmd >}}

```{r}
#| include: false
# The four neighbours, attached quietly. `data.table` is the one that needs
# saying: it is not a `show_as` target and is not going to be one, so its column
# here is written by hand where the other three could have been generated. That
# is why every entry on this page is *executed and compared* rather than read:
# a hand-written cell can be wrong in a way no reader would see.
#
# **The neighbours are attached first and god is attached after them, and the
# order is load-bearing.** god and dplyr export four names in common
# (`collect`, `pick`, `rename` and `summarize`), so whichever is attached last
# wins. With dplyr last, a god pipeline handed to `collect()` reaches dplyr's
# generic, which has no method for it and says so in R's words rather than the
# book's. Attaching god last is what lets every column below be written the
# bare way its own readers write it. Where a dplyr example needs one of the
# four, it says `dplyr::` outright.
suppressMessages({
  library(dplyr, warn.conflicts = FALSE)
  library(tidyr, warn.conflicts = FALSE)
  library(data.table)
  pkgload::load_all(find_up("r-pkg/god"), export_all = FALSE, quiet = TRUE)
})

# Two tables agree when they hold the same values, whatever order they arrived
# in and whatever each library calls its integers.
#
# **Row order is normalized away on purpose, and only here.** Order is a
# meaningful part of an answer, and the sorting entries below check it by
# comparing the sorted column itself. What this page is asking of every other
# entry is whether the same rows and values came back, and five libraries
# disagreeing about row order for an unordered question is a fact about them
# rather than a defect in any of them.
.tidy <- function(d) {
  d <- as.data.frame(d, stringsAsFactors = FALSE)
  rownames(d) <- NULL
  d <- d[, order(names(d)), drop = FALSE]
  for (j in seq_along(d)) {
    if (is.factor(d[[j]])) d[[j]] <- as.character(d[[j]])
    if (is.integer(d[[j]])) d[[j]] <- as.double(d[[j]])
  }
  if (nrow(d) > 1L) d <- d[do.call(order, lapply(d, as.character)), , drop = FALSE]
  rownames(d) <- NULL
  d
}

# Every entry ends here. A disagreement stops the render, which is the whole
# point: this page's claim is that the answers are the same, and a claim the
# build does not test is a claim nobody is keeping.
.agree <- function(task, god, ...) {
  others <- list(...)
  g <- .tidy(god)
  for (nm in names(others)) {
    o <- .tidy(others[[nm]])
    if (!isTRUE(all.equal(g, o, tolerance = 1e-8, check.attributes = FALSE))) {
      stop(sprintf(
        "%s: god and %s returned different tables.\n  god has %d row(s), %d column(s); %s has %d, %d.",
        task, nm, nrow(g), ncol(g), nm, nrow(o), ncol(o)), call. = FALSE)
    }
  }
  .checked <<- .checked + length(others)
  invisible(TRUE)
}
.checked <- 0L
```

```{python}
#| include: false
import polars as pl

# The two Python neighbours run on the same rows as the R ones. pandas already
# arrived with the setup; polars is built from it so that no fixture is defined
# twice and drifts.
pl_sales = pl.from_pandas(sales)
pl_products = pl.from_pandas(products)
pl_marks = pl.from_pandas(marks)
pl_messy = pl.from_pandas(messy)


def for_r(df):
    """A pandas frame in a shape R can actually read.

    **pandas 3.0 stores text as `StringDtype`, and reticulate does not convert
    it.** A string column crosses into R as the Python extension array itself
    rather than as a character vector, so a comparison on the R side sees an
    object where it expected two values and stops on a length mismatch that
    says nothing about the tables. Casting back to plain object dtype is what
    makes the column ordinary again. Numbers and booleans already cross
    cleanly and are left alone.
    """
    out = df.copy()
    for column in out.columns:
        if not (pd.api.types.is_numeric_dtype(out[column])
                or pd.api.types.is_bool_dtype(out[column])):
            out[column] = out[column].astype(object)
    return out
```

If you write dplyr, tidyr, pandas, polars or data.table today, this appendix is
the one that answers the question you actually have: *can it do what I already
do?* Each entry states a task in English, shows what you would write in each of
the five, shows god's sentence, and then shows the answer once, because all six
produce the same table, and the page refuses to build if any of them does not.

That last part is the difference between this appendix and a comparison table.
Nothing here is asserted. Every cell runs when the book is built, and the
tables are compared value by value; a disagreement stops the build rather than
reaching you. The code in each neighbour's column is written the way somebody
fluent in it would write it, not generated, so what you are reading is a fair
comparison rather than god's own output wearing five costumes.

**What this page is not** is a claim that god does everything these libraries
do. It does not, on purpose, and
[what this grammar does not do](e-coverage.qmd) is the list: regular
expressions, rolling windows, arbitrary functions in a cell, and six more, each
with the exit named. What this page claims is narrower and more useful: the
ordinary work, the operations that make up most days, are all sayable in one
small vocabulary that reads the same in two languages.

The tables are the book's usual cast, described in
[the preface](../index.qmd). `sales` is fifteen rows of orders.

## Keep the rows above a value

Which orders brought in more than 150?

::: {.panel-tabset}

### god

```{r}
answer <- sales |> keep(revenue > 150) |> collect()
```

### dplyr

```{r}
by_dplyr <- sales |> filter(revenue > 150)
```

### pandas

```{python}
by_pandas = sales[sales.revenue > 150]
```

### polars

```{python}
by_polars = pl_sales.filter(pl.col("revenue") > 150)
```

### data.table

```{r}
by_dt <- as.data.table(sales)[revenue > 150]
```

:::

```{r}
#| echo: false
.agree("keep the rows above a value", answer,
       dplyr = by_dplyr, pandas = py_eval("for_r(by_pandas)"),
       polars = py_eval("for_r(by_polars.to_pandas())"), data.table = by_dt)
answer
```

Five spellings of one idea. god's is the only one that reads as a sentence,
and the only one that is the same in both languages.

## Total a column for each group

What did each product earn in total, and across how many orders?

::: {.panel-tabset}

### god

```{r}
answer <- sales |>
  summarize(earned = total(revenue), orders = row_count(), by = product) |>
  collect()
```

### dplyr

```{r}
by_dplyr <- sales |>
  group_by(product) |>
  summarise(earned = sum(revenue), orders = n(), .groups = "drop")
```

### pandas

```{python}
by_pandas = (sales
  .groupby("product")
  .agg(earned=("revenue", "sum"), orders=("revenue", "size"))
  .reset_index())
```

### polars

```{python}
by_polars = pl_sales.group_by("product").agg(
    pl.col("revenue").sum().alias("earned"),
    pl.len().alias("orders"))
```

### data.table

```{r}
by_dt <- as.data.table(sales)[, .(earned = sum(revenue), orders = .N), by = product]
```

:::

```{r}
#| echo: false
.agree("total a column for each group", answer,
       dplyr = by_dplyr, pandas = py_eval("for_r(by_pandas)"),
       polars = py_eval("for_r(by_polars.to_pandas())"), data.table = by_dt)
answer
```

Note how each library spells *count the rows*: `n()`, `size`, `pl.len()`, `.N`.
god says `row_count()`, which is the same word in R and in Python.

## Turn rows into columns

Each student answered two questions. Put one row per student, one column per
question.

::: {.panel-tabset}

### god

```{r}
answer <- marks |> widen(name = question, value = mark) |> collect()
```

### tidyr

```{r}
by_dplyr <- marks |> pivot_wider(names_from = question, values_from = mark)
```

### pandas

```{python}
by_pandas = (marks
  .pivot(index="student", columns="question", values="mark")
  .reset_index()
  .rename_axis(None, axis=1))
```

### polars

```{python}
by_polars = pl_marks.pivot(on="question", index="student", values="mark")
```

### data.table

```{r}
by_dt <- dcast(as.data.table(marks), student ~ question, value.var = "mark")
```

:::

```{r}
#| echo: false
.agree("turn rows into columns", answer,
       dplyr = by_dplyr, pandas = py_eval("for_r(by_pandas)"),
       polars = py_eval("for_r(by_polars.to_pandas())"), data.table = by_dt)
answer
```

Five names for one idea: `pivot_wider`, `pivot`, `pivot`, `dcast`, and `widen`.
Only one of them says in the verb what happens to the table.

## Keep the rows matching a word

Which orders came from the West?

::: {.panel-tabset}

### god

```{r}
answer <- sales |> keep(region == "West") |> collect()
```

### dplyr

```{r}
by_dplyr <- sales |> filter(region == "West")
```

### pandas

```{python}
by_pandas = sales[sales.region == "West"]
```

### polars

```{python}
by_polars = pl_sales.filter(pl.col("region") == "West")
```

### data.table

```{r}
by_dt <- as.data.table(sales)[region == "West"]
```

:::

```{r}
#| echo: false
.agree("keep the rows matching a word", answer,
       dplyr = by_dplyr, pandas = py_eval("for_r(by_pandas)"),
       polars = py_eval("for_r(by_polars.to_pandas())"), data.table = by_dt)
answer
```

god writes `is` in the text form and `==` in both host languages, because
that is what each host already means by it.

## Choose some columns

Just the product and what it earned.

::: {.panel-tabset}

### god

```{r}
answer <- sales |> pick(product, revenue) |> collect()
```

### dplyr

```{r}
by_dplyr <- sales |> select(product, revenue)
```

### pandas

```{python}
by_pandas = sales[["product", "revenue"]]
```

### polars

```{python}
by_polars = pl_sales.select("product", "revenue")
```

### data.table

```{r}
by_dt <- as.data.table(sales)[, .(product, revenue)]
```

:::

```{r}
#| echo: false
.agree("choose some columns", answer,
       dplyr = by_dplyr, pandas = py_eval("for_r(by_pandas)"),
       polars = py_eval("for_r(by_polars.to_pandas())"), data.table = by_dt)
answer
```

Four of the five name the columns they want. god does too, and says `pick`,
which is a verb where `select` and `[[ ]]` are not.

## Drop a column

Everything except what it cost.

::: {.panel-tabset}

### god

```{r}
answer <- sales |> pick(all_but(cost)) |> collect()
```

### dplyr

```{r}
by_dplyr <- sales |> select(-cost)
```

### pandas

```{python}
by_pandas = sales.drop(columns=["cost"])
```

### polars

```{python}
by_polars = pl_sales.drop("cost")
```

### data.table

```{r}
by_dt <- as.data.table(sales)[, !"cost"]
```

:::

```{r}
#| echo: false
.agree("drop a column", answer,
       dplyr = by_dplyr, pandas = py_eval("for_r(by_pandas)"),
       polars = py_eval("for_r(by_polars.to_pandas())"), data.table = by_dt)
answer
```

`all_but` is the same word wherever it stands, where the neighbours reach for
a minus sign, a `drop`, and a `!`.

## Add a computed column

What did each order make?

::: {.panel-tabset}

### god

```{r}
answer <- sales |> add(margin = revenue - cost) |> pick(product, margin) |> collect()
```

### dplyr

```{r}
by_dplyr <- sales |> mutate(margin = revenue - cost) |> select(product, margin)
```

### pandas

```{python}
by_pandas = sales.assign(margin=sales.revenue - sales.cost)[["product", "margin"]]
```

### polars

```{python}
by_polars = (pl_sales
  .with_columns((pl.col("revenue") - pl.col("cost")).alias("margin"))
  .select("product", "margin"))
```

### data.table

```{r}
by_dt <- as.data.table(sales)[, .(product, margin = revenue - cost)]
```

:::

```{r}
#| echo: false
.agree("add a computed column", answer,
       dplyr = by_dplyr, pandas = py_eval("for_r(by_pandas)"),
       polars = py_eval("for_r(by_polars.to_pandas())"), data.table = by_dt)
answer
```

`mutate`, `assign`, `with_columns`, `:=`, and `add`. Only the last one is a
word a reader who has never seen it would guess correctly.

## Average a column for each group

What is the typical order in each region?

::: {.panel-tabset}

### god

```{r}
answer <- sales |> summarize(typical = average(revenue), by = region) |> collect()
```

### dplyr

```{r}
by_dplyr <- sales |>
  group_by(region) |>
  summarise(typical = mean(revenue), .groups = "drop")
```

### pandas

```{python}
by_pandas = (sales
  .groupby("region")
  .agg(typical=("revenue", "mean"))
  .reset_index())
```

### polars

```{python}
by_polars = pl_sales.group_by("region").agg(
    pl.col("revenue").mean().alias("typical"))
```

### data.table

```{r}
by_dt <- as.data.table(sales)[, .(typical = mean(revenue)), by = region]
```

:::

```{r}
#| echo: false
.agree("average a column for each group", answer,
       dplyr = by_dplyr, pandas = py_eval("for_r(by_pandas)"),
       polars = py_eval("for_r(by_polars.to_pandas())"), data.table = by_dt)
answer
```

`by` is an argument in god rather than a step, so grouping cannot outlive the
sentence that asked for it.

## Sort and take the top few

The five biggest orders.

::: {.panel-tabset}

### god

```{r}
answer <- sales |> sort(descending(revenue)) |> take(5) |> pick(product, revenue) |> collect()
```

### dplyr

```{r}
by_dplyr <- sales |> arrange(desc(revenue)) |> head(5) |> select(product, revenue)
```

### pandas

```{python}
by_pandas = (sales
  .sort_values("revenue", ascending=False)
  .head(5)[["product", "revenue"]])
```

### polars

```{python}
by_polars = (pl_sales
  .sort("revenue", descending=True)
  .head(5)
  .select("product", "revenue"))
```

### data.table

```{r}
by_dt <- head(as.data.table(sales)[order(-revenue)], 5)[, .(product, revenue)]
```

:::

```{r}
#| echo: false
.agree("sort and take the top few", answer,
       dplyr = by_dplyr, pandas = py_eval("for_r(by_pandas)"),
       polars = py_eval("for_r(by_polars.to_pandas())"), data.table = by_dt)
answer
```

Every library here spells *reverse the order* differently: `desc`,
`ascending=False`, `descending=True`, a minus sign. god says `descending`.

## Take the rows at the far end

The three smallest orders, smallest last.

::: {.panel-tabset}

### god

```{r}
answer <- sales |> sort(descending(revenue)) |> take_last(3) |> pick(product, revenue) |> collect()
```

### dplyr

```{r}
by_dplyr <- sales |> arrange(desc(revenue)) |> tail(3) |> select(product, revenue)
```

### pandas

```{python}
by_pandas = (sales
  .sort_values("revenue", ascending=False)
  .tail(3)[["product", "revenue"]])
```

### polars

```{python}
by_polars = (pl_sales
  .sort("revenue", descending=True)
  .tail(3)
  .select("product", "revenue"))
```

### data.table

```{r}
by_dt <- tail(as.data.table(sales)[order(-revenue)], 3)[, .(product, revenue)]
```

:::

```{r}
#| echo: false
.agree("take the rows at the far end", answer,
       dplyr = by_dplyr, pandas = py_eval("for_r(by_pandas)"),
       polars = py_eval("for_r(by_polars.to_pandas())"), data.table = by_dt)
answer
```

This is the entry that produced a word. `take_last` came out of the polars and
pandas sweeps, both of which called it `tail`, because sorting the other way
and taking the first three returns the same rows backwards.

## Bring in a column from another table

Who makes each product?

::: {.panel-tabset}

### god

```{r}
answer <- sales |>
  join(products, by = product) |>
  pick(product, maker, revenue) |>
  collect()
```

### dplyr

```{r}
by_dplyr <- sales |>
  left_join(products, by = "product") |>
  select(product, maker, revenue)
```

### pandas

```{python}
by_pandas = sales.merge(products, on="product", how="left")[["product", "maker", "revenue"]]
```

### polars

```{python}
by_polars = (pl_sales
  .join(pl_products, on="product", how="left")
  .select("product", "maker", "revenue"))
```

### data.table

```{r}
by_dt <- merge(as.data.table(sales), as.data.table(products),
               by = "product", all.x = TRUE)[, .(product, maker, revenue)]
```

:::

```{r}
#| echo: false
.agree("bring in a column from another table", answer,
       dplyr = by_dplyr, pandas = py_eval("for_r(by_pandas)"),
       polars = py_eval("for_r(by_polars.to_pandas())"), data.table = by_dt)
answer
```

god has one `join` and an `unmatched` word for whose rows survive, where the
neighbours have four verbs or a `how=` argument. Note which way the default
falls: god keeps the rows it started with, so `Sprocket` survives with no
maker rather than disappearing. Writing this entry with `inner_join` is what
the page's own comparison caught, on the row count.

## Ask whether a row has a partner

Which orders are for a product the catalogue lists?

::: {.panel-tabset}

### god

```{r}
answer <- sales |>
  keep(matching(products, by = product)) |>
  pick(product, revenue) |>
  collect()
```

### dplyr

```{r}
by_dplyr <- sales |> semi_join(products, by = "product") |> select(product, revenue)
```

### pandas

```{python}
by_pandas = sales[sales["product"].isin(products["product"])][["product", "revenue"]]
```

### polars

```{python}
by_polars = (pl_sales
  .join(pl_products.select("product"), on="product", how="semi")
  .select("product", "revenue"))
```

### data.table

```{r}
by_dt <- as.data.table(sales)[product %chin% products$product, .(product, revenue)]
```

:::

```{r}
#| echo: false
.agree("ask whether a row has a partner", answer,
       dplyr = by_dplyr, pandas = py_eval("for_r(by_pandas)"),
       polars = py_eval("for_r(by_polars.to_pandas())"), data.table = by_dt)
answer
```

`matching` is a question rather than a join, so it cannot multiply rows. That
is a promise `merge` and `isin` do not make in the same breath.

## Turn columns into rows

Two measures per order, one row each.

::: {.panel-tabset}

### god

```{r}
answer <- sales |>
  pick(product, revenue, cost) |>
  lengthen(revenue, cost, name = measure, value = amount) |>
  collect()
```

### tidyr

```{r}
by_dplyr <- sales |>
  select(product, revenue, cost) |>
  pivot_longer(c(revenue, cost), names_to = "measure", values_to = "amount")
```

### pandas

```{python}
by_pandas = (sales[["product", "revenue", "cost"]]
  .melt(id_vars="product", var_name="measure", value_name="amount"))
```

### polars

```{python}
by_polars = (pl_sales
  .select("product", "revenue", "cost")
  .unpivot(index="product", variable_name="measure", value_name="amount"))
```

### data.table

```{r}
by_dt <- melt(as.data.table(sales)[, .(product, revenue, cost)],
              id.vars = "product", variable.name = "measure",
              value.name = "amount", variable.factor = FALSE)
```

:::

```{r}
#| echo: false
.agree("turn columns into rows", answer,
       dplyr = by_dplyr, pandas = py_eval("for_r(by_pandas)"),
       polars = py_eval("for_r(by_polars.to_pandas())"), data.table = by_dt)
answer
```

`lengthen` and `widen` are one idea read in both directions, and they are the
only pair here that says so in the words themselves.

## Choose between values

Label each order by size.

::: {.panel-tabset}

### god

```{r}
answer <- sales |>
  add(size = when(revenue >= 200, "big", revenue >= 100, "medium", otherwise = "small")) |>
  pick(product, revenue, size) |>
  collect()
```

### dplyr

```{r}
by_dplyr <- sales |>
  mutate(size = case_when(revenue >= 200 ~ "big",
                          revenue >= 100 ~ "medium",
                          .default = "small")) |>
  select(product, revenue, size)
```

### pandas

```{python}
import numpy as np
by_pandas = sales.assign(size=np.select(
    [sales.revenue >= 200, sales.revenue >= 100],
    ["big", "medium"], default="small"))[["product", "revenue", "size"]]
```

### polars

```{python}
by_polars = (pl_sales
  .with_columns(pl.when(pl.col("revenue") >= 200).then(pl.lit("big"))
                  .when(pl.col("revenue") >= 100).then(pl.lit("medium"))
                  .otherwise(pl.lit("small")).alias("size"))
  .select("product", "revenue", "size"))
```

### data.table

```{r}
by_dt <- as.data.table(sales)[, .(product, revenue,
  size = fifelse(revenue >= 200, "big", fifelse(revenue >= 100, "medium", "small")))]
```

:::

```{r}
#| echo: false
.agree("choose between values", answer,
       dplyr = by_dplyr, pandas = py_eval("for_r(by_pandas)"),
       polars = py_eval("for_r(by_polars.to_pandas())"), data.table = by_dt)
answer
```

polars needs `pl.lit` around every one of those words, because a bare string
there is read as a column name. That is the defect class this book's printed-code
harness was built to catch, and it is why this page runs everything it shows.

## Number the rows in order

Where does each order place, biggest first?

::: {.panel-tabset}

### god

```{r}
answer <- sales |>
  add(place = rank(descending(revenue))) |>
  pick(product, revenue, place) |>
  collect()
```

### dplyr

```{r}
by_dplyr <- sales |>
  mutate(place = min_rank(desc(revenue))) |>
  select(product, revenue, place)
```

### pandas

```{python}
by_pandas = sales.assign(
    place=sales.revenue.rank(ascending=False, method="min").astype("int64")
)[["product", "revenue", "place"]]
```

### polars

```{python}
by_polars = (pl_sales
  .with_columns(pl.col("revenue").rank(method="min", descending=True)
                  .cast(pl.Int64).alias("place"))
  .select("product", "revenue", "place"))
```

### data.table

```{r}
by_dt <- as.data.table(sales)[, .(product, revenue,
                                  place = frank(-revenue, ties.method = "min"))]
```

:::

```{r}
#| echo: false
.agree("number the rows in order", answer,
       dplyr = by_dplyr, pandas = py_eval("for_r(by_pandas)"),
       polars = py_eval("for_r(by_polars.to_pandas())"), data.table = by_dt)
answer
```

`rank` is the one window word that needs no `sort` in front of it, because its
argument already says what the order is.

## Total as you go down the rows

How does revenue accumulate, smallest order first?

::: {.panel-tabset}

### god

```{r}
answer <- sales |>
  sort(revenue) |>
  add(so_far = running_total(revenue)) |>
  pick(revenue, so_far) |>
  collect()
```

### dplyr

```{r}
by_dplyr <- sales |>
  arrange(revenue) |>
  mutate(so_far = cumsum(revenue)) |>
  select(revenue, so_far)
```

### pandas

```{python}
_sorted = sales.sort_values("revenue", kind="stable")
by_pandas = _sorted.assign(so_far=_sorted.revenue.cumsum())[["revenue", "so_far"]]
```

### polars

```{python}
by_polars = (pl_sales
  .sort("revenue")
  .with_columns(pl.col("revenue").cum_sum().alias("so_far"))
  .select("revenue", "so_far"))
```

### data.table

```{r}
by_dt <- as.data.table(sales)[order(revenue), .(revenue, so_far = cumsum(revenue))]
```

:::

```{r}
#| echo: false
.agree("total as you go down the rows", answer,
       dplyr = by_dplyr, pandas = py_eval("for_r(by_pandas)"),
       polars = py_eval("for_r(by_polars.to_pandas())"), data.table = by_dt)
answer
```

god refuses `running_total` without a `sort` in front of it. The other four
compute happily on whatever order the rows happen to be in, which is an answer
that can change between runs.

## Trim the spaces off a text column

The names arrived padded.

::: {.panel-tabset}

### god

```{r}
answer <- messy |> add(name = trim(raw)) |> pick(name, n) |> collect()
```

### dplyr

```{r}
by_dplyr <- messy |> mutate(name = trimws(raw)) |> select(name, n)
```

### pandas

```{python}
by_pandas = messy.assign(name=messy.raw.str.strip())[["name", "n"]]
```

### polars

```{python}
by_polars = (pl_messy
  .with_columns(pl.col("raw").str.strip_chars().alias("name"))
  .select("name", "n"))
```

### data.table

```{r}
by_dt <- as.data.table(messy)[, .(name = trimws(raw), n)]
```

:::

```{r}
#| echo: false
.agree("trim the spaces off a text column", answer,
       dplyr = by_dplyr, pandas = py_eval("for_r(by_pandas)"),
       polars = py_eval("for_r(by_polars.to_pandas())"), data.table = by_dt)
answer
```

`trimws`, `str.strip`, `str.strip_chars`, `trimws`, `trim`. The same operation,
five spellings, and no two libraries agree.

## Change a column's case

Shout the region.

::: {.panel-tabset}

### god

```{r}
answer <- sales |> add(loud = upper(region)) |> pick(region, loud) |> collect()
```

### dplyr

```{r}
by_dplyr <- sales |> mutate(loud = toupper(region)) |> select(region, loud)
```

### pandas

```{python}
by_pandas = sales.assign(loud=sales.region.str.upper())[["region", "loud"]]
```

### polars

```{python}
by_polars = (pl_sales
  .with_columns(pl.col("region").str.to_uppercase().alias("loud"))
  .select("region", "loud"))
```

### data.table

```{r}
by_dt <- as.data.table(sales)[, .(region, loud = toupper(region))]
```

:::

```{r}
#| echo: false
.agree("change a column's case", answer,
       dplyr = by_dplyr, pandas = py_eval("for_r(by_pandas)"),
       polars = py_eval("for_r(by_polars.to_pandas())"), data.table = by_dt)
answer
```

`upper` and `lower` are the two everyone has. god spells them the shortest
way that still says which direction.

## Take a piece of a text column

The part of each product name before the letter `d`.

::: {.panel-tabset}

### god

```{r}
answer <- sales |> add(head = split_text(product, "d", 1)) |> pick(product, head) |> collect()
```

### dplyr

```{r}
by_dplyr <- sales |>
  mutate(head = vapply(strsplit(product, "d", fixed = TRUE), `[`, "", 1L)) |>
  select(product, head)
```

### pandas

```{python}
by_pandas = sales.assign(
    head=sales["product"].str.split("d").str[0])[["product", "head"]]
```

### polars

```{python}
by_polars = (pl_sales
  .with_columns(pl.col("product").str.split("d").list.get(0).alias("head"))
  .select("product", "head"))
```

### data.table

```{r}
by_dt <- as.data.table(sales)[, .(product,
  head = vapply(strsplit(product, "d", fixed = TRUE), `[`, "", 1L))]
```

:::

```{r}
#| echo: false
.agree("take a piece of a text column", answer,
       dplyr = by_dplyr, pandas = py_eval("for_r(by_pandas)"),
       polars = py_eval("for_r(by_polars.to_pandas())"), data.table = by_dt)
answer
```

god counts pieces from 1, and says which piece it wants in the call. The two
R spellings need a `vapply` around a list, which is the shape this grammar is
trying to spare you.

## Put text back together

Label each order with its region and product.

::: {.panel-tabset}

### god

```{r}
answer <- sales |>
  add(label = join_text(region, " ", product)) |>
  pick(label, revenue) |>
  collect()
```

### dplyr

```{r}
by_dplyr <- sales |>
  mutate(label = paste0(region, " ", product)) |>
  select(label, revenue)
```

### pandas

```{python}
by_pandas = sales.assign(
    label=sales.region + " " + sales["product"])[["label", "revenue"]]
```

### polars

```{python}
by_polars = (pl_sales
  .with_columns(pl.concat_str([pl.col("region"), pl.lit(" "), pl.col("product")])
                  .alias("label"))
  .select("label", "revenue"))
```

### data.table

```{r}
by_dt <- as.data.table(sales)[, .(label = paste0(region, " ", product), revenue)]
```

:::

```{r}
#| echo: false
.agree("put text back together", answer,
       dplyr = by_dplyr, pandas = py_eval("for_r(by_pandas)"),
       polars = py_eval("for_r(by_polars.to_pandas())"), data.table = by_dt)
answer
```

`join_text` is the sweep's second finding: god could take text apart with
`split_text` and could not put it back, which is an asymmetry no law asked for.

## Pull the year out of a date

Which year did each order land in?

::: {.panel-tabset}

### god

```{r}
answer <- sales |>
  add(d = to_date(date)) |>
  add(y = year(d), m = month(d)) |>
  pick(date, y, m) |>
  collect()
```

### dplyr

```{r}
by_dplyr <- sales |>
  mutate(d = as.Date(date),
         y = as.numeric(format(d, "%Y")),
         m = as.numeric(format(d, "%m"))) |>
  select(date, y, m)
```

### pandas

```{python}
_d = pd.to_datetime(sales.date)
by_pandas = sales.assign(y=_d.dt.year, m=_d.dt.month)[["date", "y", "m"]]
```

### polars

```{python}
by_polars = (pl_sales
  .with_columns(pl.col("date").str.to_date().alias("d"))
  .with_columns(pl.col("d").dt.year().alias("y"),
                pl.col("d").dt.month().alias("m"))
  .select("date", "y", "m"))
```

### data.table

```{r}
by_dt <- as.data.table(sales)[, {
  d <- as.Date(date)
  .(date, y = as.numeric(format(d, "%Y")), m = as.numeric(format(d, "%m")))
}]
```

:::

```{r}
#| echo: false
.agree("pull the year out of a date", answer,
       dplyr = by_dplyr, pandas = py_eval("for_r(by_pandas)"),
       polars = py_eval("for_r(by_polars.to_pandas())"), data.table = by_dt)
answer
```

Every library here needs the text turned into a date first, and god is the
only one that makes you say so: `to_date` is never applied on your behalf.

## Fill the holes in a column

The catalogue does not list every product. Say so rather than leaving a blank.

::: {.panel-tabset}

### god

```{r}
answer <- sales |>
  join(products, by = product) |>
  fill_missing(maker = "unlisted") |>
  pick(product, maker) |>
  collect()
```

### dplyr

```{r}
by_dplyr <- sales |>
  left_join(products, by = "product") |>
  mutate(maker = coalesce(maker, "unlisted")) |>
  select(product, maker)
```

### pandas

```{python}
by_pandas = (sales
  .merge(products, on="product", how="left")
  .fillna({"maker": "unlisted"})[["product", "maker"]])
```

### polars

```{python}
by_polars = (pl_sales
  .join(pl_products, on="product", how="left")
  .with_columns(pl.col("maker").fill_null("unlisted"))
  .select("product", "maker"))
```

### data.table

```{r}
by_dt <- merge(as.data.table(sales), as.data.table(products),
               by = "product", all.x = TRUE)[
  , .(product, maker = fifelse(is.na(maker), "unlisted", maker))]
```

:::

```{r}
#| echo: false
.agree("fill the holes in a column", answer,
       dplyr = by_dplyr, pandas = py_eval("for_r(by_pandas)"),
       polars = py_eval("for_r(by_polars.to_pandas())"), data.table = by_dt)
answer
```

`fill_missing` refuses a filler of the wrong kind: a number into a text
column is a refusal rather than a silent coercion, which is the one place
this list's neighbours differ from each other most.

## Drop the rows with a hole

Only the orders whose product the catalogue lists.

::: {.panel-tabset}

### god

```{r}
answer <- sales |>
  join(products, by = product) |>
  drop_missing(maker) |>
  pick(product, maker, revenue) |>
  collect()
```

### dplyr

```{r}
by_dplyr <- sales |>
  left_join(products, by = "product") |>
  filter(!is.na(maker)) |>
  select(product, maker, revenue)
```

### pandas

```{python}
by_pandas = (sales
  .merge(products, on="product", how="left")
  .dropna(subset=["maker"])[["product", "maker", "revenue"]])
```

### polars

```{python}
by_polars = (pl_sales
  .join(pl_products, on="product", how="left")
  .drop_nulls("maker")
  .select("product", "maker", "revenue"))
```

### data.table

```{r}
by_dt <- merge(as.data.table(sales), as.data.table(products),
               by = "product", all.x = TRUE)[!is.na(maker),
                                             .(product, maker, revenue)]
```

:::

```{r}
#| echo: false
.agree("drop the rows with a hole", answer,
       dplyr = by_dplyr, pandas = py_eval("for_r(by_pandas)"),
       polars = py_eval("for_r(by_polars.to_pandas())"), data.table = by_dt)
answer
```

`drop_missing` names the column it is judging. A bare `dropna()` judges every
column at once, which is a different question wearing the same name.

## Remove exact repeats

Which regions appear at all?

::: {.panel-tabset}

### god

```{r}
answer <- sales |> pick(region) |> drop_duplicates() |> collect()
```

### dplyr

```{r}
by_dplyr <- sales |> select(region) |> distinct()
```

### pandas

```{python}
by_pandas = sales[["region"]].drop_duplicates()
```

### polars

```{python}
by_polars = pl_sales.select("region").unique()
```

### data.table

```{r}
by_dt <- unique(as.data.table(sales)[, .(region)])
```

:::

```{r}
#| echo: false
.agree("remove exact repeats", answer,
       dplyr = by_dplyr, pandas = py_eval("for_r(by_pandas)"),
       polars = py_eval("for_r(by_polars.to_pandas())"), data.table = by_dt)
answer
```

`distinct`, `drop_duplicates`, `unique`, `unique`, `drop_duplicates`. god took
the longest of the five names because it is the one that says what happens to
the table rather than what the table becomes.

## Make the missing combinations appear

Every region crossed with every product, including the pairs nobody sold.

::: {.panel-tabset}

### god

```{r}
answer <- sales |>
  summarize(earned = total(revenue), by = c(region, product)) |>
  add_combinations(region, product) |>
  fill_missing(earned = 0) |>
  collect()
```

### tidyr

```{r}
by_dplyr <- sales |>
  group_by(region, product) |>
  summarise(earned = sum(revenue), .groups = "drop") |>
  complete(region, product, fill = list(earned = 0))
```

### pandas

```{python}
_g = sales.groupby(["region", "product"], as_index=False).agg(
    earned=("revenue", "sum"))
_full = pd.MultiIndex.from_product(
    [sorted(sales.region.unique()), sorted(sales["product"].unique())],
    names=["region", "product"])
by_pandas = (_g.set_index(["region", "product"])
  .reindex(_full, fill_value=0)
  .reset_index())
```

### polars

```{python}
_g = pl_sales.group_by("region", "product").agg(
    pl.col("revenue").sum().alias("earned"))
by_polars = (pl_sales.select("region").unique()
  .join(pl_sales.select("product").unique(), how="cross")
  .join(_g, on=["region", "product"], how="left")
  .with_columns(pl.col("earned").fill_null(0)))
```

### data.table

```{r}
by_dt <- as.data.table(sales)[, .(earned = sum(revenue)), by = .(region, product)][
  CJ(region = unique(sales$region), product = unique(sales$product)),
  on = .(region, product)][is.na(earned), earned := 0][]
```

:::

```{r}
#| echo: false
.agree("make the missing combinations appear", answer,
       dplyr = by_dplyr, pandas = py_eval("for_r(by_pandas)"),
       polars = py_eval("for_r(by_polars.to_pandas())"), data.table = by_dt)
answer
```

This is the entry where the five diverge most, and the reason is that the
absent rows have to be conjured rather than found. god splits it in two on
purpose: `add_combinations` makes the rows and `fill_missing` says what a new
one holds, because whether an absent pair means zero is a question about your
data rather than about the verb.

```{r}
#| echo: false
#| output: asis
cat(sprintf(
  "*This page compared %d answers against god's while it was built, and every one of them agreed.*\n",
  .checked))
```

Whichever of the five you arrived with, the sentence you would write in god is
the same sentence, in either language, and it was checked against your own
tool's answer before this page reached you.
