---
title: "How much is there?"
---

{{< include ../_setup.qmd >}}

How much of something does the table hold? Every recipe on this page is that
question wearing different clothes, and every answer is `summarize` with a
word or two around it. Every verb here arrived in [the six
verbs](../parts/six-verbs.qmd); the chapter that owns the arguments is
[Summarizing by group](../chapters/summarizing-by-group.qmd).

## How much, in total, per group?

::: {.panel-tabset}

### R

```{r}
sales |> summarize(revenue = total(revenue), by = product)
```

### Python

```{python}
sales >> summarize(revenue = total(col.revenue), by = col.product)
```


### run

```{r}
run('
  sales
    then summarize [revenue] as total([revenue]) by [product]
')
```

:::

*`sales then summarize [revenue] as total([revenue]) by [product]`*

**One verb away.** Group by two things: `by = c(region, product)` in R,
`by = [col.region, col.product]` in Python. Average instead of total: swap
`total` for `average`. Count instead of sum: `row_count()`, which takes no
column at all.

## How much of the whole is each group?

Two steps: collapse to totals, then divide each total by their sum.

::: {.panel-tabset}

### R

```{r}
sales |>
  summarize(revenue = total(revenue), by = region) |>
  add(share = revenue / total(revenue)) |>
  sort(descending(share))
```

### Python

```{python}
(sales
  >> summarize(revenue = total(col.revenue), by = col.region)
  >> add(share = col.revenue / total(col.revenue))
  >> sort(descending(col.share)))
```


### run

```{r}
run('
  sales
    then summarize [revenue] as total([revenue]) by [region]
    then add [share] as ([revenue] / total([revenue]))
    then sort [share] descending
')
```

:::

*`sales then summarize [revenue] as total([revenue]) by [region] then add [share] as [revenue] / total([revenue]) then sort [share] descending`*

The `add` uses an aggregate with no `by`, so the divisor spans the whole
table: each region's total over the grand total. The shares sum to one,
which is the check to run with your eye before you trust any share column.

**One verb away.** Share within a group instead of the whole: keep the rows
and move the `by` onto the `add`, which is the idiom
[the small-words chapter](../chapters/words-that-repeat.qmd) reads aloud.

## How many different values does a column hold?

::: {.panel-tabset}

### R

```{r}
sales |> summarize(products = unique_count(product), orders = row_count())
```

### Python

```{python}
sales >> summarize(products = unique_count(col.product), orders = row_count())
```


### run

```{r}
run('
  sales
    then summarize [products] as unique_count([product]), [orders] as row_count()
')
```

:::

*`sales then summarize [products] as unique_count([product]), [orders] as row_count()`*

**One verb away.** The values themselves rather than their count:
`pick` then `drop_duplicates`, from
[the housekeeping chapter](../chapters/renaming-and-repeats.qmd).

## How much, after a cost comes out?

Make the derived number first, then collapse it like any other column.

::: {.panel-tabset}

### R

```{r}
sales |>
  add(margin = revenue - cost) |>
  summarize(margin = total(margin), by = region)
```

### Python

```{python}
(sales
  >> add(margin = col.revenue - col.cost)
  >> summarize(margin = total(col.margin), by = col.region))
```


### run

```{r}
run('
  sales
    then add [margin] as ([revenue] - [cost])
    then summarize [margin] as total([margin]) by [region]
')
```

:::

*`sales then add [margin] as [revenue] - [cost] then summarize [margin] as total([margin]) by [region]`*

**One verb away.** Only profitable orders in the totals: a
`keep(margin > 0)` between the two steps, and nothing else moves.

## What this section refuses

The recipe most often reached for and least often meant: a total of a column
that is not a number.

::: {.panel-tabset}

### R

```{r}
#| error: true
sales |> summarize(t = total(product), by = region) |> collect()
```

### Python

```{python}
#| classes: refusal
try:
    collect(sales >> summarize(t = total(col.product), by = col.region))
except GodError as refusal:
    print(refusal)
```

:::

Some tools would concatenate the text, some would error from inside the
engine, and one famously returns zero. Here the sentence is refused whole,
with the column's kind named, before anything runs. If the column *should*
be a number, the conversion words in
[saying what something is](../chapters/converting-kinds.qmd) are the repair.
