---
title: "Which rows do you want?"
---

{{< include ../_setup.qmd >}}

Which rows does the question actually need? Half of daily work is narrowing.
The recipes here are the narrowings that recur: the top few, the top one per
group, the ones above a bar, the ones without a twin. The owning chapters
are [Keeping rows](../chapters/keeping-rows.qmd) and
[Sorting and taking](../chapters/sorting-and-taking.qmd).

## The top few

::: {.panel-tabset}

### run

```{r}
run('
  sales
    then sort [revenue] descending
    then take 5
')
```

### R

```{r}
sales |> sort(descending(revenue)) |> take(5)
```

### Python

```{python}
sales >> sort(descending(col.revenue)) >> take(5)
```

:::

*`sales then sort [revenue] descending then take 5`*

**One verb away.** The bottom few: `take_last(5)`, with the sort left as it
is. Dropping `descending` instead reaches the same rows in the opposite
order, which is a different table. The top few of a different measure: change
the sort key, and nothing else. The top five *scores* rather than five rows:
`take(5, ties = TRUE)`, which keeps every row level with the fifth.

## The top one per group

::: {.panel-tabset}

### run

```{r}
run('
  sales
    then sort [revenue] descending
    then take 1 by [region]
')
```

### R

```{r}
sales |> sort(descending(revenue)) |> take(1, by = region)
```

### Python

```{python}
sales >> sort(descending(col.revenue)) >> take(1, by = col.region)
```

:::

*`sales then sort [revenue] descending then take 1 by [region]`*

**One verb away.** The top two: `take(2, by = region)`. The most recent per
group instead of the largest: sort by `date` instead of `revenue`, which has
its own section in sorting and taking.

## The rows above a bar

::: {.panel-tabset}

### run

```{r}
run('
  sales
    then keep where ([quantity] >= 5)
    then sort [quantity] descending
')
```

### R

```{r}
sales |> keep(quantity >= 5) |> sort(descending(quantity))
```

### Python

```{python}
sales >> keep(col.quantity >= 5) >> sort(descending(col.quantity))
```

:::

*`sales then keep where [quantity] >= 5 then sort [quantity] descending`*

**One verb away.** A bar on a derived number: `add` the number first, then
keep on it. A bar per group ("orders above their region's average") is the
same two steps with `by` on the `add`:
`add(avg = average(revenue), by = region)`, then keep the rows above it.
[The words that repeat](../chapters/words-that-repeat.qmd) teaches that use
of `by`.

## The rows without a twin

::: {.panel-tabset}

### run

```{r}
run('
  sales
    then pick [region, product]
    then drop_duplicates
')
```

### R

```{r}
sales |> pick(region, product) |> drop_duplicates()
```

### Python

```{python}
sales >> pick(col.region, col.product) >> drop_duplicates()
```

:::

*`sales then pick [region, product] then drop_duplicates`*

Every region-product pairing that occurred, once each. Narrow first, then
drop: the columns you pick are the definition of what counts as a twin. The
owning chapter is
[Renaming and repeats](../chapters/renaming-and-repeats.qmd). The pairings
that never occurred are the opposite question, which
[the rows that are not there](../chapters/rows-that-are-not-there.qmd)
answers.

**One verb away.** Whole rows, one per group, is a different question with a
forced honesty in it: `sort` then `take 1 by`, saying which row you mean.

## What this section refuses

"Some rows" is not a question, and the grammar declines to pretend it is.
The first of each group, with nothing said about order:

::: {.panel-tabset}

### R

```{r}
#| error: true
sales |> take(1, by = product) |> collect()
```

### Python

```{python}
#| classes: refusal
try:
    collect(sales >> take(1, by = col.product))
except GodError as refusal:
    print(refusal)
```

:::

The repair is one sort, and the refusal names it. Every recipe on this page
that took a first or a top said its order aloud. Saying the order is what
makes a sentence mean the same thing on every run, on every engine, this
year and next. One pick is still the engine's: a cut that falls inside a
tie. `with ties`, from sorting and taking, keeps the whole tie, so nothing
is left to pick.
