---
title: "Renaming, and dropping repeats"
---

{{< include ../_setup.qmd >}}

The new name goes first, the way assignment reads and the way every other verb
that makes a column already works. If you are arriving from pandas, read that
pair twice: `rename(columns={"old": "new"})` is the other way round, and both
spellings are legal here.

::: {.panel-tabset}

### R

```{r}
sales |> rename(earned = revenue) |> pick(product, earned)
```

### Python

```{python}
sales >> rename(earned = col.revenue) >> pick(col.product, col.earned)
```

:::

`drop_duplicates` drops rows that are identical across every column. The answer
comes back in a settled order, because dropping repeats says nothing about which
order the rest should be in, and an answer that reorders itself between runs is
not predictable.

::: {.panel-tabset}

### R

```{r}
data.frame(a = c(1, 1, 2), b = c("x", "x", "y")) |> drop_duplicates()
```

### Python

```{python}
pd.DataFrame({"a": [1, 1, 2], "b": ["x", "x", "y"]}) >> drop_duplicates()
```

:::
