---
title: "The words that repeat"
---

{{< include ../_setup.qmd >}}

You have met six verbs and a handful of small words that go with them: `by`,
`where`, `as`, `all_but`, `descending`. This chapter is about those small words,
and it is the shortest useful thing in the book.

Each of them means one thing. Not one thing per verb, one thing. When you meet a
verb you have never seen, the small words in it already make sense, and that is
most of what makes this grammar quick to get back after a year away.

## `by` says which rows go together

You saw `by` on `summarize`, where it made the groups. It is the same word on
`add`, where it says which rows a value is worked out within, and on `take`,
where it says which rows the first few are taken from.

::: {.panel-tabset}

### R

```{r}
sales |> summarize(total = total(revenue), by = product)
```

### Python

```{python}
sales >> summarize(total = total(col.revenue), by = col.product)
```

:::

::: {.panel-tabset}

### R

```{r}
sales |> add(share = revenue / total(revenue), by = product) |> pick(product, revenue, share)
```

### Python

```{python}
sales >> add(share = col.revenue / total(col.revenue), by = col.product) >> pick(col.product, col.revenue, col.share)
```

:::

::: {.panel-tabset}

### R

```{r}
sales |> sort(descending(revenue)) |> take(1, by = product)
```

### Python

```{python}
sales >> sort(descending(col.revenue)) >> take(1, by = col.product)
```

:::

Three verbs, one word, one meaning. Later you will meet `by` on `join` and on
`widen`, and it will mean the same thing there: the columns that say which rows
correspond.

## `as` gives something a name

`add [name] as [value]`. The name goes first, the way assignment reads, and the
same shape names the result of `rename` and the filler in `fill_missing`.

::: {.panel-tabset}

### R

```{r}
sales |> add(margin = revenue - cost) |> rename(earned = revenue) |> pick(product, earned, margin)
```

### Python

```{python}
sales >> add(margin = col.revenue - col.cost) >> rename(earned = col.revenue) >> pick(col.product, col.earned, col.margin)
```

:::

## `all_but` turns a list inside out

Wrap the names you do *not* want. It is the same word wherever a set of columns
is chosen, which so far is `pick` and, in the next part, `lengthen`.

::: {.panel-tabset}

### R

```{r}
sales |> pick(all_but(cost, region))
```

### Python

```{python}
sales >> pick(all_but(col.cost, col.region))
```

:::

## `descending` reverses an ordering

It is a modifier on a column, in a position where an ordering is being given. So
it reads the same on `sort` and inside `rank`, and it needs no argument of its
own on either.

::: {.panel-tabset}

### R

```{r}
sales |> add(place = rank(descending(revenue))) |> sort(place) |> pick(product, revenue, place)
```

### Python

```{python}
sales >> add(place = rank(descending(col.revenue))) >> sort(col.place) >> pick(col.product, col.revenue, col.place)
```

:::

## `where` asks a question

On `keep` the question is about a row. Later you will see it on `pick`, where the
question is about a column, and the difference is written down rather than
guessed: inside `pick` you say `name` or `kind` to name what you are asking
about.

::: {.panel-tabset}

### R

```{r}
sales |> keep(revenue > 150)
```

### Python

```{python}
sales >> keep(col.revenue > 150)
```

:::

## Why this matters more than it looks

A vocabulary is easy to learn and easy to lose. What survives a year away is not
the word list, it is the rules for putting words together, and those only survive
if there are few of them and they have no exceptions.

So the test this grammar sets itself is not whether it has a word for everything.
It is whether a word you learned in one place still means the same thing in the
next place you meet it. Every part after this one is that test being run.
