---
title: "Nothing runs until you ask"
---

{{< include ../_setup.qmd >}}

A verb does not return a table. It returns a pipeline, which is a plan for one.

::: {.panel-tabset}

### R

```{r}
answer <- sales |> keep(region == "West") |> take(2)
class(answer)
```

### Python

```{python}
answer = sales >> keep(col.region == "West") >> take(2)
type(answer).__name__
```

:::

Printing it runs it, which is why every example above shows a table. When you
want the table itself, ask for it with `collect`.

::: {.panel-tabset}

### R

```{r}
nrow(collect(answer))
```

### Python

```{python}
len(collect(answer))
```

:::

In R this matters for more than speed. `sort` and `keep` are names other packages
use too, and a pipeline handed to the wrong `sort` has no meaning there, so it
stops instead of returning a plausible answer.

You can also see what the verbs wrote, which is the text form, and it is the same
sentence in both languages.

::: {.panel-tabset}

### R

```{r}
cat(format(sales |> keep(region == "West") |> take(10)))
```

### Python

```{python}
print((sales >> keep(col.region == "West") >> take(10)).written())
```

:::
