---
title: "Where this ends, and gog begins"
---

{{< include ../_setup.qmd >}}

```{r}
#| include: false
suppressMessages(library(gog))
```

```{python}
#| include: false
import gog
```

god manipulates tables. It does not draw them, and it is not going to, because
drawing is a different job with a different vocabulary. A grammar that tried to
do both would stop being small enough to hold in your head, which is the one
thing this project will not trade.

That other grammar exists. **gog** is this project's sibling: a grammar of
graphics, by the same author, built on the same discipline and holding the same
promise about day one and day two. The two are meant to sit beside each other,
and most real questions use both, because a question usually starts as a table
that is the wrong shape and ends as a picture.

## One question, end to end

Here is the whole arc. A messy table arrives, god gets it into shape, and gog
draws the answer. Nothing else is loaded.

The table is quarterly revenue, one column per quarter, which is the shape a
spreadsheet gives you and the wrong shape for almost everything else.

::: {.panel-tabset}

### R

```{r}
quarters <- data.frame(
  product = c("Widget", "Gadget", "Gizmo"),
  q1 = c(100, 300, 250),
  q2 = c(140, 280, 300),
  q3 = c(180, 260, 360),
  q4 = c(220, 240, 410)
)

quarters
```

### Python

```{python}
quarters = pd.DataFrame({
    "product": ["Widget", "Gadget", "Gizmo"],
    "q1": [100, 300, 250],
    "q2": [140, 280, 300],
    "q3": [180, 260, 360],
    "q4": [220, 240, 410],
})

quarters.style.hide(axis="index")
```

:::

Four columns that are really one variable measured four times. Turn them into
rows, which is one verb, and the table is ready to be drawn.

::: {.panel-tabset}

### R

```{r}
tidy <- quarters |>
  lengthen(all_but(product), name = quarter, value = revenue) |>
  collect()

tidy |> head(4)
```

### Python

```{python}
tidy = collect(quarters
  >> lengthen(all_but(col.product), name = col.quarter, value = col.revenue))

tidy.head(4).style.hide(axis="index")
```

:::

Now gog. The table goes in, a mark and two channels say what to draw, and the
sentence reads the same way a god pipeline does.

::: {.panel-tabset}

### R

```{r}
data(tidy) + x(quarter) + y(revenue) + color(product) + line
```

### Python

```{python}
gog.data(tidy) + gog.x(gog.col.quarter) + gog.y(gog.col.revenue) + gog.color(gog.col.product) + gog.line
```

:::

That is the whole analysis. One reshape and one plot, in two packages that were
designed to be read the same way.

## What it replaced

Written in the tools this book keeps comparing itself to, the same two steps
need a reshaping function from one package and a plotting system from another,
each with its own vocabulary, its own idea of what a column reference looks like,
and its own rules about which arguments are quoted.

The saving is not the number of lines. It is that `lengthen` and `line` come from
the same design, so the second half of the analysis does not ask you to switch
how you are thinking. `all_but` means in gog what it means here. So does `by`.

## The one thing to know about using both

In R, load them and go. The two packages take care not to collide, and R looks a
column name up in your data before it looks anywhere else.

In Python, one name is claimed by both. `col` is how each of them names a column,
so `from god import *` followed by `from gog import *` would leave you with one
of them and no warning. Import gog by name instead, which is what the examples
above do:

```python
from god import *
import gog
```

Then `col.revenue` is god's and `gog.col.revenue` is gog's, and nothing is
ambiguous. Appendix A says the same thing about every other name god takes over.

## Where to go next

If you want the graphics grammar, its manual is
[the gog book](https://psychometrician.github.io/gog-book/), and it is written
the same way this one is: every plot on the page was drawn by the engine while
the page was being built.

Read it in the order you would use them. You start in god, with a table that is
the wrong shape. You end in gog, looking at the answer.
