---
title: "A whole question, answered"
---

{{< include ../_setup.qmd >}}

The chapters before this one take one verb at a time, which is the right way to
meet them and not how anyone uses them. Here is a question that needs most of
the book.

The table is a set of survey returns. Each respondent answered three questions,
some answers are missing, a backup file filled in some of the gaps, and the
regions have long names that a lookup table holds.

::: {.panel-tabset}

### R

```{r}
returns <- data.frame(
  id     = c(1, 2, 3, 4, 5),
  code   = c("W", "W", "E", "E", "N"),
  q1     = c(5, 3, NA, 5, 2),
  q2     = c(4, NA, 5, 4, 3),
  q3     = c(NA, 2, 5, 5, 1),
  q1_alt = c(NA, NA, 4, NA, NA),
  stringsAsFactors = FALSE
)

places <- data.frame(
  code = c("W", "E"),
  area = c("West", "East"),
  stringsAsFactors = FALSE
)
```

```{r}
#| echo: false
returns
```

### Python

```{python}
returns = pd.DataFrame({
    "id":     [1, 2, 3, 4, 5],
    "code":   ["W", "W", "E", "E", "N"],
    "q1":     [5.0, 3.0, None, 5.0, 2.0],
    "q2":     [4.0, None, 5.0, 4.0, 3.0],
    "q3":     [None, 2.0, 5.0, 5.0, 1.0],
    "q1_alt": [None, None, 4.0, None, None],
})

places = pd.DataFrame({
    "code": ["W", "E"],
    "area": ["West", "East"],
})
```

```{python}
#| echo: false
returns.style.hide(axis="index")
```

:::

The question is this. For each area we have a name for, what did the average
answer to each question look like, and which area scored highest on question
one?

Getting there takes seven steps, and each one is a verb from this chapter.

::: {.panel-tabset}

### R

```{r}
answers <- returns |>
  keep(matching(places, by = code)) |>
  add(q1 = first_present(q1, q1_alt)) |>
  pick(all_but(q1_alt)) |>
  join(places, by = code) |>
  summarize(where(startsWith(name, "q"), average(value)), by = area) |>
  add(place = rank(descending(q1))) |>
  sort(place)

answers
```

### Python

```{python}
answers = (returns
  >> keep(matching(places, by = col.code))
  >> add(q1 = first_present(col.q1, col.q1_alt))
  >> pick(all_but(col.q1_alt))
  >> join(places, by = col.code)
  >> summarize(where(name.starts("q"), average(value)), by = col.area)
  >> add(place = rank(descending(col.q1)))
  >> sort(col.place))

answers
```

### run

```{r}
run('
  returns
    then keep where matching(places, by [code])
    then add [q1] as first_present([q1], [q1_alt])
    then pick all_but [q1_alt]
    then join places by [code]
    then summarize where (name starts "q") as average(value) by [area]
    then add [place] as rank([q1] descending)
    then sort [place]
')
```

:::

Read it aloud and it is the question. Keep the rows matching a place we know.
Take the backup answer for question one where the first is missing. Drop the
backup column. Bring in the area names. For each area, average every column whose
name starts with `q`. Rank the areas by question one, highest first. Sort by that
rank.

Two things that would have been easy to get wrong are done by single clauses.
`matching` drops respondent 5, whose code `N` is in no lookup row, and it cannot
multiply the rows the way a join could. And the `summarize` names no question
column at all, so a fourth question arriving next month is included without this
line changing.

Here is the whole thing as SQL, which is what actually ran.

```{r}
show_as(returns |>
  keep(matching(places, by = code)) |>
  add(q1 = first_present(q1, q1_alt)) |>
  pick(all_but(q1_alt)) |>
  join(places, by = code) |>
  summarize(where(startsWith(name, "q"), average(value)), by = area) |>
  add(place = rank(descending(q1))) |>
  sort(place), "sql")
```

A misspelling anywhere in those seven steps would have stopped the whole
sentence before anything ran, with the caret on the step that carries it.
Here is the grouping mistyped:

::: {.panel-tabset}

### R

```{r}
#| error: true
collect(returns |> join(places, by = code) |> summarize(q1 = average(q1), by = aera))
```

### Python

```{python}
#| classes: refusal
try:
    collect(returns >> join(places, by = col.code)
            >> summarize(q1 = average(col.q1), by = col.aera))
except GodError as refusal:
    print(refusal)
```

:::

Seven verbs became one query. The query is the proof rather than the point:
the sentence above it is the version a person can read aloud, and still
write, a year from now.
