---
title: "Adding a column"
---

{{< include ../_setup.qmd >}}

What did each order actually earn, and what was the price per unit? Neither
number is in the table, but everything needed to work them out is. `add`
makes a column from the ones already there, and one `add` can make several at
once.

::: {.panel-tabset}

### R

```{r}
sales |>
  add(margin = revenue - cost, price = revenue / quantity) |>
  pick(product, quantity, margin, price)
```

### Python

```{python}
(sales
  >> add(margin = col.revenue - col.cost, price = col.revenue / col.quantity)
  >> pick(col.product, col.quantity, col.margin, col.price))
```

:::

*`sales then add [margin] as [revenue] - [cost], [price] as [revenue] / [quantity] then pick [product, quantity, margin, price]`*

Read the price column down and you will notice it repeats by product; each
product sells at one price all year in this table, which is what makes its
totals easy to check by eye. The new columns are ordinary columns from here
on: pickable, keepable, sortable, and summable like anything the table
arrived with.

## The name goes first

`as` is the word that names what was made, and the name comes first: new name,
then the value. That is the order assignment reads in almost every language,
and it is the same order `rename` and `summarize` will use, so there is one
direction to remember across the whole grammar rather than one per verb.

## A step is one step

A column made in a step is not there yet for the rest of that same step.
Every value in a step is worked out from the table as it arrives, so a second
value cannot read the first. When one column needs another, give it a step of
its own, and the pipeline says what it does in the order it does it:

::: {.panel-tabset}

### R

```{r}
sales |>
  add(margin = revenue - cost) |>
  add(doubled = margin * 2) |>
  keep(margin > 50) |>
  sort(descending(margin))
```

### Python

```{python}
(sales
  >> add(margin = col.revenue - col.cost)
  >> add(doubled = col.margin * 2)
  >> keep(col.margin > 50)
  >> sort(descending(col.margin)))
```

:::

This is worth meeting early if you are arriving from dplyr or pandas, because
`mutate` and `assign` both let the second column read the first. This grammar
does not, for the same reason SQL does not: a step is one step, not a
sequence hiding inside one. The moment two values in a step can depend on
each other, the order they are written in starts to matter, and a reader has
to trace a hidden pipeline inside every line. Two short steps cost you
nothing at all when the book reaches laziness in Part VII, and they read as
what they are.

## Replacing a column

Replacing is a different thing from self-reference, and it works exactly as
you would expect, because the old value is on the table when the step begins:

::: {.panel-tabset}

### R

```{r}
sales |> add(revenue = revenue * 2) |> pick(product, revenue)
```

### Python

```{python}
sales >> add(revenue = col.revenue * 2) >> pick(col.product, col.revenue)
```

:::

The doubled `revenue` exists only in the pipeline's answer. The `sales` your
session holds is unchanged, the way it always is.

## What travels with it

- `as` names what a step makes, name first, and it means exactly this again
  in `summarize` and in `rename`.
- The whole expression language rides inside `add`: arithmetic here, and in
  Part III the text repairs, conversions, dates, and conditionals, all
  written in this same seat.
- `by` can follow an `add` whose value summarizes a group, which turns a
  collapse into a broadcast; the chapter on the small words shows it, and
  share-of-group is the idiom it exists for.

## What it refuses

The law above is enforced, not advised. A value that reads a column made in
its own step is refused whole, before anything runs:

::: {.panel-tabset}

### R

```{r}
#| error: true
sales |> add(margin = revenue - cost, doubled = margin * 2) |> collect()
```

### Python

```{python}
try:
    collect(sales >> add(margin = col.revenue - col.cost, doubled = col.margin * 2))
except GodError as refusal:
    print(refusal)
```

:::

The message does not stop at no: it names the two-step spelling that works.
The same law has a second face the text form can reach, making one name twice
in a single step, and it is refused with the same directness, because a
column silently thrown away is work silently thrown away. If you remember
that a step is one step, you will never meet either message again.
