---
title: "Adding rows"
---

{{< include ../_setup.qmd >}}

```{r}
#| include: false
more <- god_table("more")
```

```{python}
#| include: false
more = god_table("more")
```

A late order arrives in a table of its own, so where do its rows go?
`add_rows` stacks another table underneath this one. Nothing is matched and
no column is added; the answer holds every row of both tables.

The late order is `more`, one row with the same columns as `sales`:

```{r}
#| echo: false
more
```

::: {.panel-tabset}

### run

```{r}
run('
  sales
    then add_rows more
    then sort [date] descending
    then take 3
')
```

### R

```{r}
sales |> add_rows(more) |> sort(descending(date)) |> take(3)
```

### Python

```{python}
sales >> add_rows(more) >> sort(descending(col.date)) >> take(3)
```

:::

*`sales then add_rows more then sort [date] descending then take 3`*

The `sort` is there so the page can show the stacking worked: newest first,
and the late order is the first row of the answer.

The other table can even be this one, which is how a table is doubled:

::: {.panel-tabset}

### run

```{r}
run('
  sales
    then add_rows sales
    then summarize [rows] as row_count()
')
```

### R

```{r}
sales |> add_rows(sales) |> summarize(rows = row_count())
```

### Python

```{python}
sales >> add_rows(sales) >> summarize(rows = row_count())
```

:::

Every row appears twice, and both copies stay: `add_rows` never removes
anything. Dropping repeats has a verb of its own, `drop_duplicates`.

Both tables need the same columns, and each column has to hold the same kind
of thing on both sides. A column on one side only is refused rather than
filled in with missing values, because a half-empty column that says nothing
is how a mistake survives to the end of a pipeline.

::: {.panel-tabset}

### R

```{r}
#| error: true
late <- data.frame(region = "North")

collect(sales |> add_rows(late))
```

### Python

```{python}
#| classes: refusal
late = pd.DataFrame({"region": ["North"]})

try:
    collect(sales >> add_rows(late))
except GodError as refusal:
    print(refusal)
```

:::

The message names the columns one side lacks, so the repair starts with the
tables rather than with the sentence: make them agree, and the verb has
nothing left to refuse.
