---
title: "Adding a column"
---

{{< include ../_setup.qmd >}}

What did each order actually earn, and what was the price per unit? Neither
number is in the table, but everything needed to work them out is. `add`
makes a column from the ones already there, and one `add` can make several at
once.

::: {.panel-tabset}

### run

```{r}
run('
  sales
    then add [margin] as ([revenue] - [cost]),
         [price] as ([revenue] / [quantity])
    then pick [product, quantity, margin, price]
')
```

### R

```{r}
sales |>
  add(margin = revenue - cost, price = revenue / quantity) |>
  pick(product, quantity, margin, price)
```

### Python

```{python}
(sales
  >> add(margin = col.revenue - col.cost,
         price = col.revenue / col.quantity)
  >> pick(col.product, col.quantity, col.margin, col.price))
```

:::

*`sales then add [margin] as [revenue] - [cost], [price] as [revenue] / [quantity] then pick [product, quantity, margin, price]`*

Read the price column down and you will notice it repeats by product; each
product sells at one price all year in this table, which is what makes its
totals easy to check by eye. The new columns are ordinary columns from here
on: pickable, keepable, sortable, and summable like anything the table
arrived with.

## The name goes first

`as` is the word that names what was made, and the name comes first: new name,
then the value. In R and Python you type `=`, and it is playing `as`'s part;
the text form writes the word out. That is the order assignment reads in
almost every language, and it is the same order `rename` and `summarize`
will use, so there is one direction to remember across the whole grammar
rather than one per verb.

## A step is one step

A column made in a step is not there yet for the rest of that same step.
Every value in a step is worked out from the table as it arrives, so a second
value cannot read the first. When one column needs another, give it a step of
its own, and the pipeline says what it does in the order it does it:

::: {.panel-tabset}

### run

```{r}
run('
  sales
    then add [margin] as ([revenue] - [cost])
    then add [doubled] as ([margin] * 2)
    then keep where ([margin] > 50)
    then sort [margin] descending
')
```

### R

```{r}
sales |>
  add(margin = revenue - cost) |>
  add(doubled = margin * 2) |>
  keep(margin > 50) |>
  sort(descending(margin))
```

### Python

```{python}
(sales
  >> add(margin = col.revenue - col.cost)
  >> add(doubled = col.margin * 2)
  >> keep(col.margin > 50)
  >> sort(descending(col.margin)))
```

:::

This is worth meeting early if you are arriving from dplyr or pandas, because
`mutate` and `assign` both let the second column read the first. This grammar
does not, for the same reason SQL does not: a step is one step, not a
sequence hiding inside one. The moment two values in a step can depend on
each other, the order they are written in starts to matter, and a reader has
to trace a hidden pipeline inside every line. Two short steps cost you
nothing now: no step runs until the whole pipeline is asked for, which is
[the chapter on laziness](nothing-runs-yet.qmd), so the two become one
query. And they read as what they are.

## Replacing a column

Replacing is a different thing from self-reference, and it works exactly as
you would expect, because the old value is on the table when the step begins:

::: {.panel-tabset}

### run

```{r}
run('
  sales
    then add [revenue] as ([revenue] * 2)
    then pick [product, revenue]
')
```

### R

```{r}
sales |> add(revenue = revenue * 2) |> pick(product, revenue)
```

### Python

```{python}
sales >> add(revenue = col.revenue * 2) >> pick(col.product, col.revenue)
```

:::

The doubled `revenue` exists only in the pipeline's answer. The `sales` your
session holds is unchanged, the way it always is.

## What is left over

`remainder` divides and hands back what did not divide evenly. It is how you
say "every third row", or sort rows into a fixed number of buckets, or ask
whether a number is even.

::: {.panel-tabset}

### run

```{r}
run('
  sales
    then sort [date]
    then add [n] as row_number()
    then keep where remainder([n], 3) is 0
    then pick [n, product, revenue]
')
```

### R

```{r}
sales |>
  sort(date) |>
  add(n = row_number()) |>
  keep(remainder(n, 3) == 0) |>
  pick(n, product, revenue)
```

### Python

```{python}
(sales
  >> sort(col.date)
  >> add(n = row_number())
  >> keep(remainder(col.n, 3) == 0)
  >> pick(col.n, col.product, col.revenue))
```

:::

*`sales then sort [date] then add [n] as row_number() then keep where remainder([n], 3) is 0`*

The answer takes the sign of what you divided by, so `remainder(-7, 2)` is 1
rather than -1. That is R's answer and Python's; SQL engines give the other
one, and the grammar names which it means so the same sentence cannot come
back two ways depending on where it ran.

It is the only arithmetic word the grammar has beyond `+`, `-`, `*` and `/`,
and that is because it is the only one you cannot already write. Whole
division is `([n] - remainder([n], 3)) / 3`, and a square is `[x] * [x]`, so
neither gets a word of its own.

## What travels with it

- `as` names what a step makes, name first, and it means exactly this again
  in `summarize` and in `rename`.
- The whole expression language rides inside `add`: arithmetic here, and in
  [the part on values](../parts/values.qmd) the text repairs, conversions,
  dates, and conditionals, all written in this same seat.
- `by` can follow an `add` whose value summarizes a group, which turns a
  collapse into a broadcast: the group's one answer written onto every row
  of the group. The chapter on the small words shows it, and share-of-group
  is the idiom it exists for.

## What it refuses

The law above is enforced, not advised. A value that reads a column made in
its own step is refused whole, before anything runs:

::: {.panel-tabset}

### R

```{r}
#| error: true
sales |> add(margin = revenue - cost, doubled = margin * 2) |> collect()
```

### Python

```{python}
#| classes: refusal
try:
    collect(sales
      >> add(margin = col.revenue - col.cost, doubled = col.margin * 2))
except GodError as refusal:
    print(refusal)
```

:::

The message does not stop at no: it names the two-step spelling that works.
The same law refuses a second mistake that only the text form can spell,
making one name twice in a single step, and it is refused with the same
directness, because a column silently thrown away is work silently thrown
away. If you remember that a step is one step, you will never meet either
message again.
