---
title: "Adding rows from another table"
---

{{< include ../_setup.qmd >}}

A late order arrives in a table of its own, so where do its rows go?
`add_rows` stacks another table underneath this one. Nothing is matched and
no column is added; the rows just carry on where the first table stopped.

::: {.panel-tabset}

### R

```{r}
more <- data.frame(
  date = "2026-10-30", region = "North", product = "Sprocket",
  quantity = 20L, revenue = 300, cost = 240,
  stringsAsFactors = FALSE
)

sales |> add_rows(more) |> take(6)
```

### Python

```{python}
more = pd.DataFrame({
    "date": ["2026-10-30"], "region": ["North"], "product": ["Sprocket"],
    "quantity": [20], "revenue": [300], "cost": [240],
})

sales >> add_rows(more) >> take(6)
```

:::

*`sales then add_rows more then take 6`*

The other table can even be this one, which is how a table is doubled:

::: {.panel-tabset}

### R

```{r}
sales |> add_rows(sales) |> summarize(rows = row_count())
```

### Python

```{python}
sales >> add_rows(sales) >> summarize(rows = row_count())
```

:::

Both tables need the same columns. A column on one side only is refused rather
than filled in with missing values, because a half-empty column that says nothing
is how a mistake survives to the end of a pipeline.

::: {.panel-tabset}

### R

```{r}
#| error: true
collect(sales |> add_rows(data.frame(region = "North")))
```

### Python

```{python}
#| classes: refusal
try:
    collect(sales >> add_rows(pd.DataFrame({"region": ["North"]})))
except GodError as refusal:
    print(refusal)
```

:::

The message names the columns one side lacks, so the repair starts with the
tables rather than with the sentence: make them agree, and the verb has
nothing left to refuse.
