---
title: "Adding a column"
---

{{< include ../_setup.qmd >}}

`add` makes a column, or replaces one that is already there, and one `add` can
make several at once.

::: {.panel-tabset}

### R

```{r}
sales |> add(margin = revenue - cost, paid = revenue * 2) |> pick(margin, paid)
```

### Python

```{python}
sales >> add(margin = col.revenue - col.cost, paid = col.revenue * 2) >> pick(col.margin, col.paid)
```

:::

A column made in one step is not there yet for the rest of that same step. Every
value in a step is worked out from the table as it arrives, so the second one
cannot see the first.

::: {.panel-tabset}

### R

```{r}
#| error: true
sales |> add(margin = revenue - cost, doubled = margin * 2) |> collect()
```

### Python

```{python}
try:
    collect(sales >> add(margin = col.revenue - col.cost, doubled = col.margin * 2))
except GodError as refusal:
    print(refusal)
```

:::

Give it a step of its own and it works, and from there on the new column is an
ordinary column that any later step can use.

::: {.panel-tabset}

### R

```{r}
sales |>
  add(margin = revenue - cost) |>
  add(doubled = margin * 2) |>
  keep(margin > 50) |>
  sort(descending(margin))
```

### Python

```{python}
(sales
  >> add(margin = col.revenue - col.cost)
  >> add(doubled = col.margin * 2)
  >> keep(col.margin > 50)
  >> sort(descending(col.margin)))
```

:::

This is worth knowing early if you are arriving from dplyr or pandas, because
`mutate` and `assign` both let the second column read the first. This grammar
does not, for the same reason SQL does not: a step is one step, rather than a
sequence hiding inside one. Replacing a column is a different thing and works as
you would expect, because the old value is on the table when the step begins.

::: {.panel-tabset}

### R

```{r}
sales |> add(revenue = revenue * 2) |> pick(product, revenue)
```

### Python

```{python}
sales >> add(revenue = col.revenue * 2) >> pick(col.product, col.revenue)
```

:::
