---
title: "All of it at once"
---

{{< include ../_setup.qmd >}}

How much is there to learn? Here is the whole of it, on one page, before you
meet any of it properly. The pipeline you just read used a handful of these
words; every pipeline you will ever write uses words from this page and no
others.

::: {.panel-tabset}

### run

```{r}
run('
  sales
    then keep where ([revenue] > 150)
    then summarize [sold] as total([quantity]) by [region]
    then sort [sold] descending
')
```

### R

```{r}
sales |>
  keep(revenue > 150) |>
  summarize(sold = total(quantity), by = region) |>
  sort(descending(sold))
```

### Python

```{python}
(sales
  >> keep(col.revenue > 150)
  >> summarize(sold = total(col.quantity), by = col.region)
  >> sort(descending(col.sold)))
```

:::

*`sales then keep where [revenue] > 150 then summarize [sold] as total([quantity]) by [region] then sort [sold] descending`*

One sentence, and it already drew on three of the five shelves below. The lists
are asked from the engine while this page is built. So what you are reading is
the vocabulary, not a summary of one.

```{r}
#| echo: false
vocabulary <- read.delim(
  text = system2(god:::god_binary(), "--vocabulary", stdout = TRUE),
  header = FALSE, col.names = c("kind", "word"), stringsAsFactors = FALSE
)
shelf <- function(kind) {
  paste0("`", vocabulary$word[vocabulary$kind == kind], "`", collapse = " · ")
}
```

**The verbs**, each a step that takes a table and gives a table back. The next
chapters of this part carry the six that do most of the work. The rest arrive
in the parts that own them.

```{r}
#| echo: false
shelf("verb") |> cat()
```

**The aggregations**, which collapse a group of rows to one answer. They
live inside `summarize`, and inside `add` when a group's answer should land
on every row.

```{r}
#| echo: false
shelf("aggregate") |> cat()
```

**The windows**, which answer once per row by looking along the other rows:
a place, a running total, the value one row back.

```{r}
#| echo: false
shelf("window") |> cat()
```

**The scalar functions**, one value in and one value out: text repairs,
conversions, the parts of a date, one conditional companion.

```{r}
#| echo: false
shelf("scalar") |> cat()
```

**The grammar words**, the joints that say how the words on either side
relate. Five of them travel between the verbs so often they get a chapter of
their own at the end of this part.

```{r}
#| echo: false
shelf("word") |> cat()
```

That is all of it. There is no second list waiting after this one, no helper
family, and no extra dialect on a cluster. A word that is not on this page is
not in the grammar, and asking for one gets a refusal that names the nearest
word that is. The count is the argument. A vocabulary this size can be met in
a sitting and carried whole, and the rest of this book is those shelves taken
down one or two words at a time, each with the design decision that put it
there. When you want this page again as a reference,
[every word in the grammar](../appendices/b-every-word.qmd) is the same list
with a map of where each word first runs.
