---
title: "Another table"
---

{{< include ../_setup.qmd >}}

The sale knows its product, so who knows the maker? Another table does.
`join` adds that table's columns to this one, matching rows by the columns
that say which rows correspond. `products` holds a maker for each product.

::: {.panel-tabset}

### R

```{r}
sales |> join(products, by = product) |> pick(product, maker, revenue)
```

### Python

```{python}
(sales
  >> join(products, by = col.product)
  >> pick(col.product, col.maker, col.revenue))
```


### run

```{r}
run('
  sales
    then join products by [product]
    then pick [product, maker, revenue]
')
```

:::

*`sales then join products by [product] then pick [product, maker, revenue]`*

Leave `by` out and god matches on the column names both tables share, then
prints the choice back as a note beside the answer, the way [the
laws](design-laws.qmd) show. It is never silent about a choice you did not
make.

::: {.panel-tabset}

### R

```{r}
sales |> join(products) |> take(2)
```

### Python

```{python}
sales >> join(products) >> take(2)
```


### run

```{r}
run('
  sales
    then join products
    then take 2
')
```

:::

What varies between the four joins you may know by name is only this: what
happens to a row that found no match. So that is what the argument is called.

| `unmatched =` | Which unmatched rows survive | Called elsewhere |
|---|---|---|
| `"this"` (the default) | this table's | left join |
| `"none"` | neither table's | inner join |
| `"both"` | both tables' | full join |

There is no `"other"`, and that is not an omission. A right join is this join
with the tables the other way round, so it adds no meaning and gets no word.

::: {.panel-tabset}

### R

```{r}
sales |>
  join(products, unmatched = "none") |>
  summarize(revenue = total(revenue), by = maker)
```

### Python

```{python}
(sales
  >> join(products, unmatched = "none")
  >> summarize(revenue = total(col.revenue), by = col.maker))
```


### run

```{r}
run('
  sales
    then join products unmatched "none"
    then summarize [revenue] as total([revenue]) by [maker]
')
```

:::

A column on both tables that is not being matched on would arrive twice, so it
is refused rather than quietly renamed to something you did not choose.

::: {.panel-tabset}

### R

```{r}
#| error: true
overlap <- data.frame(product = "Widget", revenue = 1,
                      stringsAsFactors = FALSE)
collect(sales |> join(overlap, by = product))
```

### Python

```{python}
#| classes: refusal
overlap = pd.DataFrame({"product": ["Widget"], "revenue": [1]})

try:
    collect(sales >> join(overlap, by = col.product))
except GodError as refusal:
    print(refusal)
```

:::

Two more refusals guard the key itself, because a join that could never match
would come back empty and look like a fact about your data. The key must hold
the same kind of thing in both tables:

::: {.panel-tabset}

### R

```{r}
#| error: true
codes <- data.frame(product = c(1, 2), code = c("A", "B"))
collect(sales |> join(codes, by = product))
```

### Python

```{python}
#| classes: refusal
codes = pd.DataFrame({"product": [1, 2], "code": ["A", "B"]})

try:
    collect(sales >> join(codes, by = col.product))
except GodError as refusal:
    print(refusal)
```

:::

And the key must exist on both sides, with the usual suggestion when it is
only a spelling away:

::: {.panel-tabset}

### R

```{r}
#| error: true
collect(sales |> join(products, by = maker))
```

### Python

```{python}
#| classes: refusal
try:
    collect(sales >> join(products, by = col.maker))
except GodError as refusal:
    print(refusal)
```

:::

One thing the checker cannot see is a key that repeats. If the other table
holds two rows for one key, your rows multiply, and that is a fact about the
rows rather than about the sentence. The checker reads column names and
never a row, so this one stays yours to know.
