---
title: "Choosing columns"
---

{{< include ../_setup.qmd >}}

The table has six columns and your answer needs two of them. Which product,
and what did it bring in? `pick` names the columns you want, and the rest do
not travel any further down the pipeline.

::: {.panel-tabset}

### R

```{r}
sales |> pick(product, revenue)
```

### Python

```{python}
sales >> pick(col.product, col.revenue)
```

:::

*`sales then pick [product, revenue]`*

All fifteen rows are still here. `keep` and `pick` are the two axes of the
same act of narrowing: `keep` decides which rows survive and never touches
the columns, `pick` decides which columns survive and never touches the rows.
Between them you can cut any table down to exactly the piece your question is
about, and the two read as what they are, one verb per axis.

## Saying which ones to drop

When the columns you want outnumber the ones you do not, name the leavers.
Wrap the names in `all_but`:

::: {.panel-tabset}

### R

```{r}
sales |> pick(all_but(cost, region))
```

### Python

```{python}
sales >> pick(all_but(col.cost, col.region))
```

:::

It is the same verb, deliberately. Choosing columns is choosing columns
whichever way you say which ones, so there is no second verb called `drop`
here for the same reason there was no second verb for dropping rows in the
last chapter. `all_but` is a marker on the list, not a new idea, and it is
the same word in both languages and in the text form.

## Column order

`pick` also settles column order, because the order you name the columns is
the order the table comes back in. There is no separate rearranging verb: if
you want `revenue` first, ask for it first. One small consequence is worth
knowing now: a `pick` is a statement about the table it receives, so a column
a previous step made is as pickable as one the table arrived with, and a
column you picked away is gone for every later step. Each step sees only the
table the last step handed it.

## When names are not enough

Naming columns one at a time holds up until the table is wide, or until next
month's file arrives with two more columns than this month's. There is a
second form, `pick where`, that chooses columns by a question about them
instead of by name, and it reuses the condition language you met in the last
chapter. It has a part of its own near the end of the book, because the
interesting questions (what is the name shaped like, what kind of value does
it hold) deserve more than a footnote. Until then, every table in this part
is narrow enough to name.

## What travels with it

- `all_but` inverts any list of columns, here and anywhere a list of columns
  appears; you will meet it again in `lengthen`.
- `where` waits inside this verb too: `pick where` takes the same conditions
  `keep` takes, aimed at columns instead of rows, in [choosing columns
  without naming them](../parts/choosing-columns.qmd).
- The order of names is meaningful, which is the general rule about lists in
  this grammar: nothing you write is decoration.

## What it refuses

A misspelled column is caught before anything runs, and the refusal does not
stop at no:

::: {.panel-tabset}

### R

```{r}
#| error: true
collect(sales |> pick(product, reveune))
```

### Python

```{python}
#| classes: refusal
try:
    collect(sales >> pick(col.product, col.reveune))
except GodError as refusal:
    print(refusal)
```

:::

The message names the nearest real column and then lists everything the
table has, because the second most common cause of this error is not a typo
at all: it is being wrong about which table you are holding. Both repairs are
on the screen, and neither requires opening another window.
