---
title: "Turning columns into rows"
---

{{< include ../_setup.qmd >}}

A survey arrives with a row per person and a column per question, so how do
you total by question? Most things you want to ask of a wide table need the
opposite shape.

::: {.panel-tabset}

### R

```{r}
answers <- data.frame(
  student = c("ann", "bob"),
  q1 = c(1, 4), q2 = c(2, 5), q3 = c(3, 6),
  stringsAsFactors = FALSE
)

answers |> lengthen(q1, q2, q3)
```

### Python

```{python}
answers = pd.DataFrame({
    "student": ["ann", "bob"],
    "q1": [1, 4], "q2": [2, 5], "q3": [3, 6],
})

answers >> lengthen(col.q1, col.q2, col.q3)
```

:::

*`answers then lengthen [q1, q2, q3]`*

The table grew taller, which is what the verb's name says. Direction is in the
name because the alternative is a pair of words you have to look up every time.

Two new columns arrived, called `name` and `value`. Those are the grammar's own
words for what a column is called and what it holds, and they are the defaults
here, so the shortest sentence is also the one that teaches the vocabulary.

Give them names of your own when the table is about something:

::: {.panel-tabset}

### R

```{r}
answers |> lengthen(q1, q2, q3, name = question, value = mark)
```

### Python

```{python}
answers >> lengthen(col.q1, col.q2, col.q3, name = col.question, value = col.mark)
```


### run

```{r}
run('
  answers
    then lengthen [q1, q2, q3] as name [question], value [mark]
')
```

:::

## Choosing the columns

The columns are chosen the same three ways `pick` chooses them, so there is
nothing new to learn. The commonest one by far is everything except the column
that identifies a row:

::: {.panel-tabset}

### R

```{r}
answers |> lengthen(all_but(student), name = question, value = mark) |> take(3)
```

### Python

```{python}
answers >> lengthen(all_but(col.student), name = col.question, value = col.mark) >> take(3)
```


### run

```{r}
run('
  answers
    then lengthen all_but [student] as name [question], value [mark]
    then take 3
')
```

:::

A question about the name works too, and it is the same `where` that chooses
columns anywhere else:

::: {.panel-tabset}

### R

```{r}
answers |> lengthen(where(startsWith(name, "q")), name = question, value = mark) |> take(2)
```

### Python

```{python}
answers >> lengthen(where(name.starts("q")), name = col.question, value = col.mark) >> take(2)
```


### run

```{r}
run('
  answers
    then lengthen where (name starts "q") as name [question], value [mark]
    then take 2
')
```

:::

The columns being stacked have to hold the same kind of thing, because the one
column they become can only hold one kind. Stacking a text column onto a number
column is refused, and the refusal names both:

::: {.panel-tabset}

### R

```{r}
#| error: true
collect(answers |> lengthen(student, q1))
```

### Python

```{python}
#| classes: refusal
try:
    collect(answers >> lengthen(col.student, col.q1))
except GodError as refusal:
    print(refusal)
```

:::

## When a name holds two things

Column names often encode more than one thing: `q1_2020` is a question and a
year. Say what the names look like, and name each piece:

::: {.panel-tabset}

### R

```{r}
terms <- data.frame(
  id = c(1, 2),
  q1_2020 = c(10, 20), q1_2021 = c(11, 21),
  q2_2020 = c(12, 22), q2_2021 = c(13, 23)
)

terms |> lengthen(all_but(id), name = "{question}_{year}", value = mark) |> take(4)
```

### Python

```{python}
terms = pd.DataFrame({
    "id": [1, 2],
    "q1_2020": [10, 20], "q1_2021": [11, 21],
    "q2_2020": [12, 22], "q2_2021": [13, 23],
})

terms >> lengthen(all_but(col.id), name = "{question}_{year}", value = col.mark) >> take(4)
```

:::

The braces mark a piece, and everything outside them is the text that separates
the pieces. That is the whole rule. It is not a regular expression, and it is not
going to become one: a pattern you cannot read two months later is a pattern that
costs more than it saves.

A piece written `{value}` says something different. It says that piece picks
which value column the row belongs to, rather than becoming a value itself. That
is what to write when the names encode both what was measured and which measure
it is:

::: {.panel-tabset}

### R

```{r}
readings <- data.frame(
  site = c("north", "south"),
  air_mean = c(12.5, 15.0), air_max = c(20.0, 24.0),
  sea_mean = c(8.5, 9.0),   sea_max = c(11.0, 12.5)
)

readings |> lengthen(all_but(site), name = "{medium}_{value}")
```

### Python

```{python}
readings = pd.DataFrame({
    "site": ["north", "south"],
    "air_mean": [12.5, 15.0], "air_max": [20.0, 24.0],
    "sea_mean": [8.5, 9.0],   "sea_max": [11.0, 12.5],
})

readings >> lengthen(all_but(col.site), name = "{medium}_{value}")
```

:::

Four columns became one column saying which medium, and two columns holding the
two measures.
