---
title: "Saying what something is"
---

{{< include ../_setup.qmd >}}

The digits are right there in the column, so why refuse to count them?
Conversions are always written out. Nothing here converts a column on your
behalf, because a column that quietly changes what it holds is the hardest
kind of mistake to find later.

::: {.panel-tabset}

### R

```{r}
messy |> add(word = to_text(n)) |> add(digits = characters(word))
```

### Python

```{python}
messy >> add(word = to_text(col.n)) >> add(digits = characters(col.word))
```

:::

*`messy then add [word] as to_text([n]) then add [digits] as characters([word])`*

**Every conversion begins `to_`**, and nothing else does: `to_number`,
`to_whole`, `to_text`, `to_date`. Ask for the characters in a number without
converting it first and you are told which conversion you wanted:

::: {.panel-tabset}

### R

```{r}
#| error: true
collect(messy |> add(size = characters(n)))
```

### Python

```{python}
try:
    collect(messy >> add(size = characters(col.n)))
except GodError as refusal:
    print(refusal)
```

:::

## The four conversions

`to_text`, `to_number`, `to_whole` and `to_date` are the whole set. Each one says
what it makes rather than what it takes, so there is nothing to remember about
direction.

`to_number` gives a number that may have a decimal point. `to_whole` gives one
that may not, and it truncates rather than rounding, which is worth knowing
before you use it on money.

::: {.panel-tabset}

### R

```{r}
messy |>
  add(as_text = to_text(n), rounded = to_whole(to_number(n) / 2)) |>
  pick(n, as_text, rounded)
```

### Python

```{python}
(messy
  >> add(as_text = to_text(col.n), rounded = to_whole(to_number(col.n) / 2))
  >> pick(col.n, col.as_text, col.rounded))
```

:::

A conversion that cannot be made is not a silent missing value in either
language. The engine underneath decides what to do with `to_number("abc")`, and
this is one of the few places where what you get depends on where the pipeline
ran.

## Between two ends

`between` is not a conversion, and it is here because it is the other word you
reach for when a column has just become a number. It asks whether a value falls
in a range.

::: {.panel-tabset}

### R

```{r}
sales |> keep(between(revenue, 150, 400)) |> sort(revenue)
```

### Python

```{python}
sales >> keep(between(col.revenue, 150, 400)) >> sort(col.revenue)
```

:::

Both ends count, the way SQL and dplyr both have it, so nothing here needs
checking against a manual.
