---
title: "Saying what something is"
---

{{< include ../_setup.qmd >}}

The digits are right there in the column, so why refuse to count them?
Conversions are always written out. Nothing here converts a column on your
behalf, because a column that quietly changes what it holds is the hardest
kind of mistake to find later.

::: {.panel-tabset}

### run

```{r}
run('
  messy
    then add [word] as to_text([n])
    then add [digits] as characters([word])
')
```

### R

```{r}
messy |> add(word = to_text(n)) |> add(digits = characters(word))
```

### Python

```{python}
messy >> add(word = to_text(col.n)) >> add(digits = characters(col.word))
```

:::

*`messy then add [word] as to_text([n]) then add [digits] as characters([word])`*

**Every conversion begins `to_`**, and nothing else does: `to_number`,
`to_whole`, `to_text`, `to_date`. Ask for the characters in a number without
converting it first and you are told which conversion you wanted:

::: {.panel-tabset}

### R

```{r}
#| error: true
collect(messy |> add(size = characters(n)))
```

### Python

```{python}
#| classes: refusal
try:
    collect(messy >> add(size = characters(col.n)))
except GodError as refusal:
    print(refusal)
```

:::

## The four conversions

`to_text`, `to_number`, `to_whole` and `to_date` are the whole set. Each one says
what it makes rather than what it takes, so there is nothing to remember about
direction.

`to_number` gives a number that may have a decimal point. `to_whole` gives one
that may not, and it truncates rather than rounding, which is worth knowing
before you use it on money.

::: {.panel-tabset}

### run

```{r}
run('
  messy
    then add [as_text] as to_text([n]),
         [halved] as to_whole((to_number([n]) / 2))
    then pick [n, as_text, halved]
')
```

### R

```{r}
messy |>
  add(as_text = to_text(n), halved = to_whole(to_number(n) / 2)) |>
  pick(n, as_text, halved)
```

### Python

```{python}
(messy
  >> add(as_text = to_text(col.n), halved = to_whole(to_number(col.n) / 2))
  >> pick(col.n, col.as_text, col.halved))
```

:::

### Which way `to_whole` goes, and why

Half of 7 is 3.5, and `to_whole` makes that 3. It throws the fraction away
rather than rounding to the nearest, and it does that in both directions: -3.5
becomes -3, not -4. The word for that is truncation, and it is toward zero.

This is worth stating rather than leaving to be discovered, because the other
answer is just as reasonable and you cannot tell which you have by reading the
name. Two reasons decided it.

The first is that this is what a conversion does everywhere else. Every
host language god reaches already truncates when it turns a number into a whole
one: R's `as.integer(7.5)` is 7, Python's `int(7.5)` is 7, and pandas, polars
and Spark all agree. If you learned `to_whole` here and then read the dplyr or
pandas version of your own pipeline, nothing changes under you.

The second is that rounding is a different question wearing the same word.
Rounding has to answer what happens exactly halfway, and there are at least
three defensible answers to that: away from zero, toward the even number, or
always up. A word that has to pick one of those silently is a word you have to
look up, and the point of this vocabulary is that you should not have to. If
rounding is ever added it will be its own word, saying which rule it follows.

So `to_whole` is a conversion, not a rounding, and if what you want is the
nearest whole number, add half before converting.

::: {.panel-tabset}

### run

```{r}
run('
  messy
    then add [down] as to_whole((to_number([n]) / 2))
    then add [nearest] as to_whole((to_number([n]) / 2) + 0.5)
    then pick [n, down, nearest]
')
```

### R

```{r}
messy |>
  add(down = to_whole(to_number(n) / 2)) |>
  add(nearest = to_whole(to_number(n) / 2 + 0.5)) |>
  pick(n, down, nearest)
```

### Python

```{python}
(messy
  >> add(down = to_whole(to_number(col.n) / 2))
  >> add(nearest = to_whole(to_number(col.n) / 2 + 0.5))
  >> pick(col.n, col.down, col.nearest))
```

:::

A conversion that cannot be made is the one answer this grammar does not own.
`to_number("abc")` goes to the engine underneath, and engines disagree: one
stops with an error, another hands back a missing value. It is one of the few
places where the answer depends on where the pipeline ran, and the honest
sentence is exactly that one.

## Between two ends

`between` is not a conversion, and it is here because it is the other word you
reach for when a column has just become a number. It asks whether a value falls
in a range.

::: {.panel-tabset}

### run

```{r}
run('
  sales
    then keep where between([revenue], 150, 400)
    then sort [revenue]
')
```

### R

```{r}
sales |> keep(between(revenue, 150, 400)) |> sort(revenue)
```

### Python

```{python}
sales >> keep(between(col.revenue, 150, 400)) >> sort(col.revenue)
```

:::

Both ends count, the way SQL and dplyr both have it, so nothing here needs
checking against a manual.
