---
title: "Saying what something is"
---

{{< include ../_setup.qmd >}}

You can see the digits in the column, so why refuse to count them?
Conversions are always written out. Nothing here converts a column on your
behalf, because a column that quietly changes what it holds is the hardest
kind of mistake to find later.

::: {.panel-tabset}

### run

```{r}
run('
  messy
    then add [word] as to_text([n])
    then add [digits] as characters([word])
')
```

### R

```{r}
messy |> add(word = to_text(n)) |> add(digits = characters(word))
```

### Python

```{python}
messy >> add(word = to_text(col.n)) >> add(digits = characters(col.word))
```

:::

*`messy then add [word] as to_text([n]) then add [digits] as characters([word])`*

**Every conversion begins `to_`**, and nothing else does: `to_number`,
`to_text`, `to_date`. Ask for the characters in a number without converting it
first and you are told which conversion you wanted:

::: {.panel-tabset}

### R

```{r}
#| error: true
collect(messy |> add(size = characters(n)))
```

### Python

```{python}
#| classes: refusal
try:
    collect(messy >> add(size = characters(col.n)))
except GodError as refusal:
    print(refusal)
```

:::

## The three conversions

`to_text`, `to_number` and `to_date` are the whole set. Each one says what it
makes rather than what it takes, so there is nothing to remember about
direction.

There were four of these, and the fourth was a mistake of the kind that hides.
`to_whole` turned a number into a whole one, which sounds like a conversion
and is not: this grammar has one kind of number, so `to_whole` converted
nothing. It was a rounding wearing a conversion's name, and because the name
said nothing about direction, no reader could tell whether -5.5 came back as
-5 or as -6. It is two words now, `round_below` and `round_above`, and they
are in [the chapter on adding a column](adding-a-column.qmd), where the
arithmetic lives.

::: {.panel-tabset}

### run

```{r}
run('
  messy
    then add [as_text] as to_text([n]), [half] as to_number([n]) / 2
    then pick [n, as_text, half]
')
```

### R

```{r}
messy |>
  add(as_text = to_text(n), half = to_number(n) / 2) |>
  pick(n, as_text, half)
```

### Python

```{python}
(messy
  >> add(as_text = to_text(col.n), half = to_number(col.n) / 2)
  >> pick(col.n, col.as_text, col.half))
```

:::

A conversion that cannot be made is the one answer this grammar does not own.
`to_number("abc")` goes to the engine underneath, and engines disagree: one
stops with an error, another hands back a missing value. It is one of the few
places where the answer depends on where the pipeline ran, and the honest
sentence is exactly that one.

## Between two ends

`between` is not a conversion, and it is here because it is the other word you
reach for when a column has just become a number. It asks whether a value falls
in a range.

::: {.panel-tabset}

### run

```{r}
run('
  sales
    then keep where between([revenue], 150, 400)
    then sort [revenue]
')
```

### R

```{r}
sales |> keep(between(revenue, 150, 400)) |> sort(revenue)
```

### Python

```{python}
sales >> keep(between(col.revenue, 150, 400)) >> sort(col.revenue)
```

:::

Both ends are included, the way SQL and dplyr both have it, so nothing here
needs checking against a manual.

`to_date` is the conversion this chapter has not run. It reads a date out of
text, and [the next chapter](dates.qmd) opens with it.
