---
title: "Dates"
---

{{< include ../_setup.qmd >}}

```{r}
#| include: false
stamps <- god_table("stamps")
```

```{python}
#| include: false
stamps = god_table("stamps")
```

What year was that, what month, and did it land on a weekday? A date usually
arrives as text. `to_date` reads one, and once a column holds a date there is
a word for each part you might want out of it.

The table is `diary`, three days held as text:

```{r}
#| echo: false
diary
```

::: {.panel-tabset}

### run

```{r}
run('
  diary
    then add [d] as to_date([on_])
    then add [year] as year([d]), [month] as month([d]), [weekday] as weekday([d])
')
```

### R

```{r}
diary |>
  add(d = to_date(on_)) |>
  add(year = year(d), month = month(d), weekday = weekday(d))
```

### Python

```{python}
(diary
 >> add(d = to_date(col.on_))
 >> add(year = year(col.d), month = month(col.d), weekday = weekday(col.d)))
```

:::

*`diary then add [d] as to_date([on_]) then add [year] as year([d]), [month] as month([d]), [weekday] as weekday([d])`*

**`weekday` counts Monday as 1**, wherever you run it. That is the grammar's
numbering rather than the engine's, and it has to be: asked plainly, one engine
calls a Friday 5 and another calls it 4, and neither says anything is wrong.

## The five parts

`year`, `month`, `day`, `weekday` and `hour`. Each one pulls a number out of a
date, and each is named for what it returns.

::: {.panel-tabset}

### run

```{r}
run('
  diary
    then add [on] as to_date([on_])
    then add [y] as year([on]), [m] as month([on]), [d] as day([on]),
         [wd] as weekday([on])
    then pick [on_, y, m, d, wd]
')
```

### R

```{r}
diary |>
  add(on = to_date(on_)) |>
  add(y = year(on), m = month(on), d = day(on), wd = weekday(on)) |>
  pick(on_, y, m, d, wd)
```

### Python

```{python}
(diary
  >> add(on = to_date(col.on_))
  >> add(y = year(col.on), m = month(col.on), d = day(col.on),
         wd = weekday(col.on))
  >> pick(col.on_, col.y, col.m, col.d, col.wd))
```

:::

`hour` is the fifth, and it wants a value that carries a time. A plain date has
no hour in it, so on the dates above it answers zero rather than refusing.

```{r}
#| echo: false
stamps
```

::: {.panel-tabset}

### run

```{r}
run('
  stamps
    then add [h] as hour(to_date([at]))
    then pick [at, h]
')
```

### R

```{r}
stamps |>
  add(h = hour(to_date(at))) |>
  pick(at, h)
```

### Python

```{python}
(stamps
  >> add(h = hour(to_date(col.at)))
  >> pick(col.at, col.h))
```

:::

*`stamps then add [h] as hour(to_date([at])) then pick [at, h]`*

## Text is not yet a date

The five words read a date, and text is not one until `to_date` has said so.
Skip the conversion and the refusal points straight at it:

::: {.panel-tabset}

### R

```{r}
#| error: true
collect(diary |> add(y = year(on_)))
```

### Python

```{python}
#| classes: refusal
try:
    collect(diary >> add(y = year(col.on_)))
except GodError as refusal:
    print(refusal)
```

:::

That is the whole date vocabulary: `to_date` reads the text, five words take
the part you ask for, and the answers agree wherever the sentence runs,
Monday counting as 1 on every engine.
