---
title: "Dates"
---

{{< include ../_setup.qmd >}}

```{r}
#| include: false
# **`at` is made a real timestamp here, in R and in Python**, because that is
# what `hour` reads and a CSV cannot carry one: every value in a CSV is text.
# The chapter says so in prose below rather than hiding it.
stamps <- god_table("stamps")
stamps$at <- as.POSIXct(stamps$at, tz = "UTC")
```

```{python}
#| include: false
stamps = god_table("stamps")
stamps["at"] = pd.to_datetime(stamps["at"])
```

What year was that, what month, and did it land on a weekday? A date usually
arrives as text. `to_date` reads one, and once a column holds a date there is
a word for each part you might want out of it.

The table is `diary`, three days held as text:

```{r}
#| echo: false
diary
```

::: {.panel-tabset}

### run

```{r}
run('
  diary
    then add [d] as to_date([on_])
    then add [year] as year([d]), [month] as month([d]), [weekday] as weekday([d])
')
```

### R

```{r}
diary |>
  add(d = to_date(on_)) |>
  add(year = year(d), month = month(d), weekday = weekday(d))
```

### Python

```{python}
(diary
 >> add(d = to_date(col.on_))
 >> add(year = year(col.d), month = month(col.d), weekday = weekday(col.d)))
```

:::

*`diary then add [d] as to_date([on_]) then add [year] as year([d]), [month] as month([d]), [weekday] as weekday([d])`*

**`weekday` counts Monday as 1**, wherever you run it. That is the grammar's
numbering rather than the engine's, and it has to be: asked plainly, one engine
calls a Friday 5 and another calls it 4, and neither says anything is wrong.

## The five parts

`year`, `month`, `day`, `weekday` and `hour`. Each one pulls a number out of a
date, and each is named for what it returns.

::: {.panel-tabset}

### run

```{r}
run('
  diary
    then add [on] as to_date([on_])
    then add [y] as year([on]), [m] as month([on]), [d] as day([on]),
         [wd] as weekday([on])
    then pick [on_, y, m, d, wd]
')
```

### R

```{r}
diary |>
  add(on = to_date(on_)) |>
  add(y = year(on), m = month(on), d = day(on), wd = weekday(on)) |>
  pick(on_, y, m, d, wd)
```

### Python

```{python}
(diary
  >> add(on = to_date(col.on_))
  >> add(y = year(col.on), m = month(col.on), d = day(col.on),
         wd = weekday(col.on))
  >> pick(col.on_, col.y, col.m, col.d, col.wd))
```

:::

`hour` is the fifth, and it is the one that needs a column carrying a time. The
four above read parts of the calendar, which every date has. An hour is not part
of the calendar, so a column has to have brought one with it.

`stamps` did. Its `at` column arrived as a timestamp rather than as a date:

```{r}
#| echo: false
stamps
```

::: {.panel-tabset}

### run

```{r}
run('
  stamps
    then add [h] as hour([at])
    then pick [at, h]
')
```

### R

```{r}
stamps |>
  add(h = hour(at)) |>
  pick(at, h)
```

### Python

```{python}
(stamps
  >> add(h = hour(col.at))
  >> pick(col.at, col.h))
```

:::

*`stamps then add [h] as hour([at]) then pick [at, h]`*

## Why `hour` will not read a plain date

A column either carries a time or it does not, and the grammar knows which. Ask
`hour` of one that does not and it refuses, rather than hand back a column of
noughts that quietly assumed midnight:

::: {.panel-tabset}

### R

```{r}
#| error: true
collect(stamps |> add(h = hour(to_date(at))))
```

### Python

```{python}
#| classes: refusal
try:
    collect(stamps >> add(h = hour(to_date(col.at))))
except GodError as refusal:
    print(refusal)
```

:::

That refusal is the shape worth having. The sentence is legal English and reads
as though it should work, the answer would have been a number, and nothing about
a column of zeros would have looked wrong on the page. Handing one back is what
the grammar will not do.

`to_date` is no way around it, which is worth seeing once because it is the
repair everybody reaches for. It makes a **date**, which is the thing that has
no time in it, so it moves the sentence no closer:

::: {.panel-tabset}

### R

```{r}
#| error: true
collect(diary |> add(h = hour(to_date(on_))))
```

### Python

```{python}
#| classes: refusal
try:
    collect(diary >> add(h = hour(to_date(col.on_))))
except GodError as refusal:
    print(refusal)
```

:::

Splitting it over two steps does not help either, and the grammar says so in the
same words: what it reads is the column, not the spelling that made it.

A time has to reach the table as a time, then. Where the values are text, the
conversion belongs to the language you are calling from, above the pipeline: R
has `as.POSIXct` and Python has `pd.to_datetime`, and that is how `at` above
became a timestamp before any of these examples ran.

The four calendar words are untouched by all of this. A year, a month, a day and
a weekday are parts of the calendar, and every date has them whether or not a
time came with it.

## Text is not yet a date

The five words read a date, and text is not one until `to_date` has said so.
Skip the conversion and the refusal points straight at it:

::: {.panel-tabset}

### R

```{r}
#| error: true
collect(diary |> add(y = year(on_)))
```

### Python

```{python}
#| classes: refusal
try:
    collect(diary >> add(y = year(col.on_)))
except GodError as refusal:
    print(refusal)
```

:::

That is the whole date vocabulary: `to_date` reads the text, five words take
the part you ask for, and the answers agree wherever the sentence runs,
Monday counting as 1 on every engine.
