---
title: "Dates"
---

{{< include ../_setup.qmd >}}

What year was that, what month, and did it land on a weekday? A date usually
arrives as text. `to_date` reads one, and once a column holds a date there is
a word for each part you might want out of it.

::: {.panel-tabset}

### R

```{r}
diary <- data.frame(
  g = c("a", "a", "b"),
  on_ = c("2026-01-02", "2026-01-05", "2026-01-06"),
  x = c(10, 20, 30),
  stringsAsFactors = FALSE
)

diary |>
  add(d = to_date(on_)) |>
  add(year = year(d), month = month(d), weekday = weekday(d))
```

### Python

```{python}
diary = pd.DataFrame({
    "g": ["a", "a", "b"],
    "on_": ["2026-01-02", "2026-01-05", "2026-01-06"],
    "x": [10, 20, 30],
})

(diary
 >> add(d = to_date(col.on_))
 >> add(year = year(col.d), month = month(col.d), weekday = weekday(col.d)))
```

:::

*`diary then add [d] as to_date([on_]) then add [year] as year([d]), [month] as month([d]), [weekday] as weekday([d])`*

**`weekday` counts Monday as 1**, wherever you run it. That is the grammar's
numbering rather than the engine's, and it has to be: asked plainly, one engine
calls a Friday 5 and another calls it 4, and neither says anything is wrong.

## The five parts

`year`, `month`, `day`, `weekday` and `hour`. Each one pulls a number out of a
date, and each is named for what it returns.

::: {.panel-tabset}

### R

```{r}
diary |>
  add(on = to_date(on_)) |>
  add(y = year(on), m = month(on), d = day(on), wd = weekday(on)) |>
  pick(on_, y, m, d, wd)
```

### Python

```{python}
(diary
  >> add(on = to_date(col.on_))
  >> add(y = year(col.on), m = month(col.on), d = day(col.on), wd = weekday(col.on))
  >> pick(col.on_, col.y, col.m, col.d, col.wd))
```

:::

`hour` is the fifth, and it wants a value that carries a time. A plain date has
no hour in it, so on the dates above it answers zero rather than refusing.

::: {.panel-tabset}

### R

```{r}
data.frame(at = "2026-01-02 14:30:00") |> add(h = hour(to_date(at))) |> pick(at, h)
```

### Python

```{python}
pd.DataFrame({"at": ["2026-01-02 14:30:00"]}) >> add(h = hour(to_date(col.at))) >> pick(col.at, col.h)
```

:::
