---
title: "Giving each row a place"
---

{{< include ../_setup.qmd >}}

Who came first, and what happens to second place when there is a tie? `rank`
gives every row its place by some column. It is a value like any other, so it
goes in `add`.

The table below has a tie in it on purpose, because the tie is where ranking
functions differ from each other and where most of the confusion lives.

::: {.panel-tabset}

### R

```{r}
races <- data.frame(
  heat  = c("x", "x", "y", "y"),
  name  = c("ana", "bo", "cy", "di"),
  score = c(20, 20, 5, 50),
  stringsAsFactors = FALSE
)

races |> add(place = rank(score)) |> sort(place)
```

### Python

```{python}
races = pd.DataFrame({
    "heat":  ["x", "x", "y", "y"],
    "name":  ["ana", "bo", "cy", "di"],
    "score": [20, 20, 5, 50],
})

races >> add(place = rank(col.score)) >> sort(col.place)
```

:::

*`races then add [place] as rank([score]) then sort [place]`*

Read the places down the column: 1, 2, 2, 4. The lowest score is first, the two
20s **share** second place, and the next value takes fourth. Third place is not
awarded, because two rows already stand ahead of the 50.

The `sort` at the end is doing real work. Adding a column says nothing about what
order the rows come back in, so a sentence that wants a particular order asks for
it.

That is how a race is scored, and it is what people mean when they say rank. It
is worth saying out loud because the alternative is just as common in other
tools: numbering the *distinct values* instead would give 1, 2, 2, 3 and award a
third place to the runner who came fourth.

`descending` reverses it, and it is the same word `sort` takes, because a column
in an ordering position is one idea wherever it appears.

::: {.panel-tabset}

### R

```{r}
races |> add(place = rank(descending(score)))
```

### Python

```{python}
races >> add(place = rank(descending(col.score)))
```

:::

`by` restarts the numbering inside each group, so every heat gets a winner.

::: {.panel-tabset}

### R

```{r}
races |> add(place = rank(descending(score)), by = heat) |> sort(heat, place)
```

### Python

```{python}
(races
  >> add(place = rank(descending(col.score)), by = col.heat)
  >> sort(col.heat, col.place))
```

:::

## Numbering the rows instead

`row_number()` numbers the rows 1, 2, 3, 4 and never ties. That is the whole
difference from `rank`, and it shows on the same table.

::: {.panel-tabset}

### R

```{r}
races |> sort(score) |> add(n = row_number()) |> sort(n)
```

### Python

```{python}
races >> sort(col.score) >> add(n = row_number()) >> sort(col.n)
```

:::

`row_number()` takes no argument, so it can only mean the order the rows are
already in, and a table has no order until a `sort` gives it one. Without one it
is refused rather than answered, because two runs could disagree and both look
right.

::: {.panel-tabset}

### R

```{r}
#| error: true
collect(races |> add(n = row_number()))
```

### Python

```{python}
try:
    collect(races >> add(n = row_number()))
except GodError as refusal:
    print(refusal)
```

:::

That is the practical difference between the two. `rank` says what it goes by, so
it never needs a `sort` and does not disturb the order of your table.
`row_number()` says nothing, so it needs one.

## Where a place can stand

A place is worked out by looking at every row in the group, which means it cannot
be the thing that decides which rows are in the group. Asking for the top three
inside `keep` is refused, and the message names both ways to say it.

::: {.panel-tabset}

### R

```{r}
#| error: true
collect(races |> keep(rank(score) <= 2))
```

### Python

```{python}
try:
    collect(races >> keep(rank(col.score) <= 2))
except GodError as refusal:
    print(refusal)
```

:::

Make the column first, then filter on it like any other column.

::: {.panel-tabset}

### R

```{r}
races |> add(place = rank(descending(score))) |> keep(place <= 2)
```

### Python

```{python}
races >> add(place = rank(descending(col.score))) >> keep(col.place <= 2)
```

:::

If you only want the top rows and not the place itself, `take` says it in one
step and leaves no extra column behind.

::: {.panel-tabset}

### R

```{r}
races |> sort(descending(score)) |> take(1, by = heat)
```

### Python

```{python}
races >> sort(descending(col.score)) >> take(1, by = col.heat)
```

:::

`summarize` also refuses a place, for the opposite reason: it returns one row per
group, and a place is worked out for every row, so there is nowhere to put the
answer. Ask for the value you actually want instead, with `largest`, `first` or
`row_count`.

## Two words where dplyr has six

dplyr offers `row_number`, `min_rank`, `dense_rank`, `percent_rank`, `cume_dist`
and `ntile`. This grammar has two, and the other four are refused until somebody
asks for one.

The reason is that `rank` is already what a person means by the word. dplyr calls
this one `min_rank`, which names how it is computed rather than what it is, and
leaves the plain word `rank` free to mean something else. Here the plain word
means the plain thing.

The other four are real but specialist. `dense_rank` is the 1, 2, 2, 3 numbering
described above; `percent_rank` and `cume_dist` are positions expressed as a
fraction; `ntile` cuts the rows into buckets. Carrying four more words so that
every reader has to learn which is which, in order to serve the cases that
occasionally need one, is a poor trade. If you need one of them, the pipeline
that gets you most of the way is `rank` plus arithmetic, and `show_as` will hand
you the query to take the rest of the way.
