---
title: "Every refusal is the grammar's"
---

{{< include ../_setup.qmd >}}

The verbs build a sentence and check nothing. Whether a column exists, whether
an aggregation belongs where you put it, whether the types line up: all of that
is answered in one place, so both languages get the same answer, in the same
words.

::: {.panel-tabset}

### R

```{r}
#| error: true
collect(sales |> keep(reveune > 100))
```

### Python

```{python}
#| classes: refusal
try:
    collect(sales >> keep(col.reveune > 100))
except GodError as refusal:
    print(refusal)
```

:::

The whole pipeline is read before any of it runs, so a mistake at step two is
reported as a mistake at step two rather than partway through the work.

Because the reading happens before anything runs, a sentence that will not run
can still be drawn, as far as it got.

::: {.panel-tabset}

### R

```{r}
sales |>
  summarize(total = total(revenue), by = region) |>
  keep(product == "Widget") |>
  show_steps()
```

### Python

```{python}
show_steps(sales
           >> summarize(total = total(col.revenue), by = col.region)
           >> keep(col.product == "Widget"))
```

:::

The message says `product` is not a column, and it is right. The drawing says
where it went: the `summarize` above took it away, along with everything else
that was not grouped by. That is the question a message on its own cannot
answer, and it is why the refusal and the drawing belong together.

There is one refusal that cannot work this way. When `widen` finds two rows
wanting the same cell, that is a fact about the data rather than about the
sentence, and the grammar reads column names rather than rows. So that one
arrives while the query runs. The words are still the grammar's, and they
still name what to write instead.