---
title: "Your first pipeline"
---

{{< include ../_setup.qmd >}}

A pipeline is a sentence. It names a table, then says what happens to it, one
step at a time, in the order the steps happen.

Here is one. Read it aloud before you read anything about it.

::: {.panel-tabset}

### R

```{r}
sales |>
  keep(region == "West") |>
  add(margin = revenue - cost) |>
  summarize(margin = total(margin), orders = row_count(), by = product) |>
  sort(descending(margin))
```

### Python

```{python}
(sales
  >> keep(col.region == "West")
  >> add(margin = col.revenue - col.cost)
  >> summarize(margin = total(col.margin), orders = row_count(), by = col.product)
  >> sort(descending(col.margin)))
```

:::

Keep the West rows. Add a margin. Total the margin by product and count the
orders. Sort by margin, descending. That is the whole sentence, and the table
under it is what the engine returned when this page was built.

## The table

Every example in this book runs on a table called `sales`, unless a chapter says
otherwise. It is fifteen rows, small enough to check the arithmetic by hand.

::: {.panel-tabset}

### R

```{r}
#| echo: false
sales
```

### Python

```{python}
#| echo: false
# A frame of its own rather than a pipeline, so it is asked for the table view
# here. Everything else is a pipeline and renders as a table on its own. Hidden
# because it is the book showing you the data, not god's grammar.
sales.style.hide(axis="index")
```

:::

## The same sentence, twice

The two tabs above are not translations of each other. They are the same
sentence, and only two things about them differ.

The first is the pipe. R writes `|>`, which is R's own. Python writes `>>`,
because Python cannot have `|>` at all, and of the operators a data frame leaves
free that is the one available everywhere.

The second is how a column is named. R can look a name up in your data before it
looks in your session, so a bare `revenue` is unambiguous there. Python has no
such hook, so a column says that it is one: `col.revenue`.

Every verb, every keyword and every argument position is the same. Someone who
learned the grammar in one language has learned it in the other. Both tabs on
every page of this book are executed when the book is built, so that claim is
checked on every page rather than asserted once here.

[Appendix A](../appendices/a-two-languages.qmd) has the rest of the differences
between the two languages, and you do not need it yet.

## How to read the rest

Read the tab for the language you write. Glance at the other one when you are
curious. The explanation under a tabset is written once and covers both.

The next six chapters are one verb each. They are short, and by the end of them
you can answer real questions about a table.
