---
title: "Your first pipeline"
---

{{< include ../_setup.qmd >}}

You have a year of sales in front of you. Which product earned the most in
the West, after costs? That is a real question, one of a hundred like it you
will ask this year, and one sentence answers it.

A pipeline is that sentence. It names a table, then says what happens to it,
one step at a time, in the order the steps happen. Here is the whole answer.
Read it aloud before you read anything about it.

::: {.panel-tabset}

### run

```{r}
run('
  sales
    then keep where ([region] is "West")
    then add [margin] as ([revenue] - [cost])
    then summarize [margin] as total([margin]),
         [orders] as row_count() by [product]
    then sort [margin] descending
')
```

### R

```{r}
sales |>
  keep(region == "West") |>
  add(margin = revenue - cost) |>
  summarize(margin = total(margin), orders = row_count(), by = product) |>
  sort(descending(margin))
```

### Python

```{python}
(sales
  >> keep(col.region == "West")
  >> add(margin = col.revenue - col.cost)
  >> summarize(margin = total(col.margin), orders = row_count(),
               by = col.product)
  >> sort(descending(col.margin)))
```

:::

*`sales then keep where [region] is "West" then add [margin] as [revenue] - [cost] then summarize [margin] as total([margin]), [orders] as row_count() by [product] then sort [margin] descending`*

Keep the West rows. Add a margin. Total the margin by product and count the
orders. Sort by margin, largest first. That is the whole sentence. The table
under it is what the engine returned when this page was built, and its top
row is the answer to the question you started with.

Notice what you did not have to know. No method names, no order of clauses
to memorize. You read a sentence in plain English, and if you can read this
one you can already read most of the pipelines in this book, including the
ones no page has shown you yet.

## The gloss, and why it runs

The italic line under the example is the same sentence again, in the
grammar's own words. This book prints one under the first example of every
teaching chapter. Reading it aloud is not a ritual for its own sake: it is
the preface's retention promise being rehearsed. A sentence you can say is a
sentence you can still say after a year away.

The gloss is not decoration, and it is not pseudocode. The text form is the
grammar's main spelling, the one `run` takes, and the one this book points a
new reader at first. It is also the only spelling that still means something
where there is neither R nor Python: a database cell, a configuration file, a
stored pipeline. In it, `then` is the pipe, written as a word. A column goes
in square brackets, `[margin]`, so that any name at all can be a column. And
because it is real, this book's build hands every gloss on every page to the
engine and refuses to publish if one stops parsing. A gloss cannot quietly
go stale; the build will not let it.

## The table

Every example in this book runs on a table called `sales`, unless a chapter
says otherwise. It is fifteen rows: three regions, four products, a year of
dates, and for each order a quantity, the revenue, and the cost. That is
small enough to check any total by hand, and you should. The whole cast of
tables this book uses is declared in the preface, along with how to put the
same rows in your session. None of them changes shape from one chapter to
the next.

::: {.panel-tabset}

### R

```{r}
#| echo: false
sales
```

### Python

```{python}
#| echo: false
# A frame of its own rather than a pipeline, so it is asked for the table view
# here. Everything else is a pipeline and renders as a table on its own. Hidden
# because it is the book showing you the data, not god's grammar.
sales.style.hide(axis="index")
```

:::

## The same sentence, twice

The R and Python tabs of the first example are not translations of each
other. They are the same sentence, and only two things about them differ.

The first is the pipe. R writes `|>`, which is R's own. Python writes `>>`,
because Python cannot have `|>` at all, and of the operators a data frame
leaves free that is the one available everywhere.

The second is how a column is named. R can look a name up in your data before
it looks in your session, so a bare `revenue` is unambiguous there. Python
has no such hook, so a column says that it is one: `col.revenue`.

Every verb, every keyword and every argument position is the same. Someone
who learned the grammar in one language has learned it in the other. Every
tab on every page of this book is executed when the book is built, so that
claim is checked on every page rather than asserted once here.

[Appendix A](../appendices/a-two-languages.qmd) has the rest of the
differences between the two languages, and you do not need it yet.

## What a step is

Each step in the sentence takes a table and gives a table back. That one
fact matters more than it looks like it should. It means you can stop the
pipeline anywhere and look at what you have so far. It means the order you
read is the order things happen, top to bottom. Nothing runs earlier than
it reads, the way a SQL `WHERE` runs before the `SELECT` written above it.
And it means every chapter in this book can teach one verb at a time,
because a verb never needs to know which verbs came before it.

There is one more fact about the sentence you just read, and it can wait for
its own chapter: nothing ran until the page asked to see the answer. A
pipeline is a plan, not a command, and [the chapter on
laziness](nothing-runs-yet.qmd) shows what that buys you.

## The first refusal, met on purpose

You will mistype something within the hour, so see now what that costs. A
`keep` needs a question, and a column on its own is not one:

::: {.panel-tabset}

### R

```{r}
#| error: true
collect(sales |> keep(revenue))
```

### Python

```{python}
#| classes: refusal
try:
    collect(sales >> keep(col.revenue))
except GodError as refusal:
    print(refusal)
```

:::

Nothing ran and nothing broke. The message says what stopped and what to
write instead, and every refusal in this book is held to that standard.

## How to read the rest

Read the tab for the language you write. Glance at the others when you are
curious. The explanation under a tabset is written once and covers every tab.

The next page lays out the whole vocabulary at once, so you can see how
little there is to meet. The teaching chapters after it take the verbs one
or two at a time. They share one shape, declared on the part page you just
passed: a question, the verb that answers it, what the verb can vary, then
what travels with it and what it refuses. They are short. By the end of
them you can answer real questions about a table, and the part's practice
chapters will ask you to prove it.
