---
title: "Has a partner"
---

{{< include ../_setup.qmd >}}

```{r}
#| include: false
listed <- god_table("listed")
```

```{python}
#| include: false
listed = god_table("listed")
```

```{r}
#| include: false
stocked <- god_table("stocked")
```

```{python}
#| include: false
stocked = god_table("stocked")
```

Which of your rows have a partner in the other table? Sometimes that is the
whole question: you do not want the other table's columns, only to keep the
rows that appear in it. That is a question about a row, so it is written as
one, and the verb that takes a question is `keep`.

The other table is `stocked`, a list of one product:

```{r}
#| echo: false
stocked
```

::: {.panel-tabset}

### run

```{r}
run('
  sales
    then keep where matching(stocked, by [product])
')
```

### R

```{r}
sales |> keep(matching(stocked, by = product))
```

### Python

```{python}
sales >> keep(matching(stocked, by = col.product))
```

:::

*`sales then keep where matching(stocked, by [product])`*

Nothing was added. The columns are the ones `sales` started with, and only the
rows changed, which is why this is not spelled `join`. Negate it and you get
exactly the rows the first one dropped.

::: {.panel-tabset}

### run

```{r}
run('
  sales
    then keep where (not matching(stocked, by [product]))
')
```

### R

```{r}
sales |> keep(!matching(stocked, by = product))
```

### Python

```{python}
sales >> keep(~matching(stocked, by = col.product))
```

:::

`by` works here exactly as it does on `join`, and leaving it out means the
same thing: god matches on the column names both tables share. The note
naming its choice arrives beside the answer, as it does for `join`.

That sameness is the whole point, so it extends to keys the two tables name
differently. Write `is` between them, this table's name first, exactly as
[another table](another-table.qmd) does.

`listed` names the same products under a column of its own:

```{r}
#| echo: false
listed
```

::: {.panel-tabset}

### run

```{r}
run('
  sales
    then keep where matching(listed, by [product] is [item])
    then take 3
')
```

### R

```{r}
sales |> keep(matching(listed, by = product == item)) |> take(3)
```

### Python

```{python}
sales >> keep(matching(listed, by = col.product == col.item)) >> take(3)
```

:::

*`sales then keep where matching(listed, by [product] is [item]) then take 3`*

With no `by` at all, god says which column it used:

::: {.panel-tabset}

### run

```{r}
run('
  sales
    then keep where matching(stocked)
')
```

### R

```{r}
sales |> keep(matching(stocked))
```

### Python

```{python}
sales >> keep(matching(stocked))
```

:::

*`sales then keep where matching(stocked)`*

There is one thing `matching` guarantees that `join` cannot. If the other table
has three rows for a product, a join gives you three rows back for each of
yours. A question does not: a row either has a partner or it does not, and how
many it has never reaches the answer.

::: {.panel-tabset}

### R

```{r}
restocks <- data.frame(
  product   = c("Widget", "Widget", "Widget"),
  restocked = c(10, 20, 30),
  stringsAsFactors = FALSE
)

nrow(collect(sales |> keep(matching(restocks, by = product))))
nrow(collect(sales |> join(restocks, by = product, unmatched = "none")))
```

### Python

```{python}
restocks = pd.DataFrame({
    "product":   ["Widget", "Widget", "Widget"],
    "restocked": [10, 20, 30],
})

print(len(collect(sales >> keep(matching(restocks, by = col.product)))))
print(len(collect(
    sales >> join(restocks, by = col.product, unmatched = "none"))))
```

:::

`matching` is the whole question `keep` asks rather than one part of it, so
it does not combine with `and` or `or`. Ask it in its own step.

::: {.panel-tabset}

### R

```{r}
#| error: true
collect(sales |> keep(matching(products, by = product) & revenue > 100))
```

### Python

```{python}
#| classes: refusal
try:
    collect(sales
      >> keep(matching(products, by = col.product) & (col.revenue > 100)))
except GodError as refusal:
    print(refusal)
```

:::

::: {.panel-tabset}

### run

```{r}
run('
  sales
    then keep where matching(products, by [product])
    then keep where ([revenue] > 100)
')
```

### R

```{r}
sales |> keep(matching(products, by = product)) |> keep(revenue > 100)
```

### Python

```{python}
(sales
  >> keep(matching(products, by = col.product))
  >> keep(col.revenue > 100))
```

:::

The two `keep` steps read as one sentence: first the rows with a partner,
then the rows whose revenue is over 100. However long the pipeline grows,
`matching` can only remove rows, which is the property the word exists to
guarantee.

That property is what separates it from a join, because the two sentences look
alike and mean different things.

::: {.panel-tabset}

### R

```{r}
sales |> keep(matching(products, by = product)) |> show_steps()
```

### Python

```{python}
show_steps(sales >> keep(matching(products, by = col.product)))
```

:::

`products` gets a line of its own, the same as it would under a join, and its
`maker` column stays where it is. Nothing crossed over, and the last line says
the rows can only be fewer. A join in the same place says they may multiply.
That difference is the whole reason both words exist.
