---
title: "Asking whether a row has a partner"
---

{{< include ../_setup.qmd >}}

Which of your rows have a partner in the other table? Sometimes that is the
whole question: you do not want the other table's columns, only to keep the
rows that appear in it. That is a question about a row, so it is written as
one, and the verb that takes a question is `keep`.

::: {.panel-tabset}

### R

```{r}
stocked <- data.frame(product = "Widget", stringsAsFactors = FALSE)

sales |> keep(matching(stocked, by = product))
```

### Python

```{python}
stocked = pd.DataFrame({"product": ["Widget"]})

sales >> keep(matching(stocked, by = col.product))
```

:::

*`sales then keep where matching(stocked, by [product])`*

Nothing was added. The columns are the ones `sales` started with, and only the
rows changed, which is why this is not spelled `join`. Negate it and you get
exactly the rows the first one dropped.

::: {.panel-tabset}

### R

```{r}
sales |> keep(!matching(stocked, by = product))
```

### Python

```{python}
sales >> keep(~matching(stocked, by = col.product))
```


### run

```{r}
run('
  sales
    then keep where (not matching(stocked, by [product]))
')
```

:::

`by` works here exactly as it does on `join`, and leaving it out means the same
thing: god matches on the column names both tables share, and the note naming
its choice arrives beside the answer, as it does for `join`.

::: {.panel-tabset}

### R

```{r}
sales |> keep(matching(stocked))
```

### Python

```{python}
sales >> keep(matching(stocked))
```


### run

```{r}
run('
  sales
    then keep where matching(stocked)
')
```

:::

*`sales then keep where matching(stocked)`*

There is one thing `matching` guarantees that `join` cannot. If the other table
has three rows for a product, a join gives you three rows back for each of
yours. A question does not: a row either has a partner or it does not, and how
many it has never reaches the answer.

::: {.panel-tabset}

### R

```{r}
restocks <- data.frame(
  product   = c("Widget", "Widget", "Widget"),
  restocked = c(10, 20, 30),
  stringsAsFactors = FALSE
)

nrow(collect(sales |> keep(matching(restocks, by = product))))
nrow(collect(sales |> join(restocks, by = product, unmatched = "none")))
```

### Python

```{python}
restocks = pd.DataFrame({
    "product":   ["Widget", "Widget", "Widget"],
    "restocked": [10, 20, 30],
})

print(len(collect(sales >> keep(matching(restocks, by = col.product)))))
print(len(collect(
    sales >> join(restocks, by = col.product, unmatched = "none"))))
```

:::

`matching` is the whole question `keep` asks rather than one part of one, so it
does not combine with `and`. Ask it in its own step.

::: {.panel-tabset}

### R

```{r}
#| error: true
collect(sales |> keep(matching(products, by = product) & revenue > 100))
```

### Python

```{python}
#| classes: refusal
try:
    collect(sales
      >> keep(matching(products, by = col.product) & (col.revenue > 100)))
except GodError as refusal:
    print(refusal)
```

:::

::: {.panel-tabset}

### R

```{r}
sales |> keep(matching(products, by = product)) |> keep(revenue > 100)
```

### Python

```{python}
(sales
  >> keep(matching(products, by = col.product))
  >> keep(col.revenue > 100))
```


### run

```{r}
run('
  sales
    then keep where matching(products, by [product])
    then keep where ([revenue] > 100)
')
```

:::

The two `keep` steps read as one sentence: first the rows with a partner,
then the rows that clear the bar. However long the pipeline grows, `matching`
can only remove rows, which is the property the word exists to keep.
