---
title: "Keeping rows"
---

{{< include ../_setup.qmd >}}

Which of the year's orders happened in the West? Most questions about a table
begin this way, by wanting fewer rows than you have. `keep` is the verb: it
takes a condition, asks it of every row, and keeps the rows where the answer
is yes.

::: {.panel-tabset}

### run

```{r}
run('
  sales
    then keep where ([region] is "West")
')
```

### R

```{r}
sales |> keep(region == "West")
```

### Python

```{python}
sales >> keep(col.region == "West")
```

:::

*`sales then keep where [region] is "West"`*

Six of the fifteen rows survive. The other nine were not deleted from
anything; `sales` is untouched, and the pipeline handed back a new, smaller
table. Every verb in this grammar works that way, so no step can damage the
table a colleague is also using.

## Combining conditions

Conditions join with *and* and *or*, and a negation reverses one. Ask for the
larger orders that were still cheap to fulfill:

::: {.panel-tabset}

### run

```{r}
run('
  sales
    then keep where (([revenue] >= 150) and ([cost] < 100))
')
```

### R

```{r}
sales |> keep(revenue >= 150 & cost < 100)
```

### Python

```{python}
sales >> keep((col.revenue >= 150) & (col.cost < 100))
```

:::

Python needs the parentheses and R does not, because Python's `&` binds more
tightly than its `==`. That rule is Python's rather than the grammar's: the
preface holds the short list of rules Python brings of its own, and that
list is the whole of the subject.

## The words where the hosts disagree

Look at what the text form did to the R sentences above: `==` became `is`,
and `&` became `and`. That is not a style choice; it is a decision about
which words can be in a grammar at all.

A symbol that means different things in different tools is a word this
grammar refuses to own. `=` assigns in one language, compares in another, and
half the world's beginners have confused the two. So the text form spells
every contested symbol as the English word that means only one thing, and
each binding translates its host's habit into that word:

| you write in R | you write in Python | the grammar's word |
|---|---|---|
| `==` | `==` | `is` |
| `!=` | `!=` | `is not` |
| `&` | `&` | `and` |
| `\|` | `\|` | `or` |
| `!` | `~` | `not` |
| `%in%` | `.is_in([...])` | `in { }` |
| `is.na(x)` | `.is_missing()` | `[x] is missing` |

You keep writing your language's habits, and the grammar keeps one meaning
under them. When a pipeline moves to a database cell or a colleague's screen,
it is the unambiguous spelling that travels.

## One verb, both directions

There is no separate verb for dropping rows. Removing the rows where
something is true is keeping the rows where it is false, and one verb says
both:

::: {.panel-tabset}

### run

```{r}
run('
  sales
    then keep where (not ([product] is "Widget"))
')
```

### R

```{r}
sales |> keep(!(product == "Widget"))
```

### Python

```{python}
sales >> keep(~(col.product == "Widget"))
```

:::

This is the first appearance of a rule you will meet on every page: if the
grammar can already say something, it does not grow a second way to say it. A
`drop` verb would cost a word, and every reader who met `drop(x)` beside
`keep(!x)` would have to learn, and then remember, that they are the same
thing. The grammar spends its words on meanings, not on synonyms.

## What travels with it

- `where` is how the text form marks the condition, as in the gloss above.
  You never type it in R or Python; the condition sits inside the verb.
- The condition language travels. A `when` (choosing a value by a question)
  takes the conditions `keep` takes, unchanged. A `pick where` (choosing
  columns by a question) joins and negates its questions the same way, and
  asks them about a column's name and kind rather than its values.
- `and`, `or`, `not`, `is`, `in`, `missing` are the grammar's own words for
  the contested symbols, and the table above is their whole story.

## What it refuses

A condition that can never be true is a mistake, not a preference, and the
grammar says so before anything runs. Text is never greater than a number:

::: {.panel-tabset}

### R

```{r}
#| error: true
collect(sales |> keep(region > 100))
```

### Python

```{python}
#| classes: refusal
try:
    collect(sales >> keep(col.region > 100))
except GodError as refusal:
    print(refusal)
```

:::

Nothing was scanned and no rows came back empty. The pipeline was refused
whole, at the step that is wrong, with the fix in the message. A tool that
instead answered with zero rows would have told you something false about
your data, quietly, and you would have believed it.
