---
title: "Keeping rows"
---

{{< include ../_setup.qmd >}}

Which of the year's orders happened in the West? Most questions about a table
begin this way, by wanting fewer rows than you have. `keep` is the verb: it
takes a condition, asks it of every row, and keeps the rows where the answer
is yes.

::: {.panel-tabset}

### R

```{r}
sales |> keep(region == "West")
```

### Python

```{python}
sales >> keep(col.region == "West")
```

:::

*`sales then keep where [region] is "West"`*

Six of the fifteen rows survive. The other nine were not deleted from
anything; `sales` is untouched, and the pipeline handed back a new, smaller
table. Every verb in this grammar works that way, so no step can damage the
table a colleague is using behind you.

## Combining conditions

Conditions join with *and* and *or*, and a negation reverses one. Ask for the
larger orders that were still cheap to fulfill:

::: {.panel-tabset}

### R

```{r}
sales |> keep(revenue >= 150 & cost < 100)
```

### Python

```{python}
sales >> keep((col.revenue >= 150) & (col.cost < 100))
```

:::

Python needs the parentheses and R does not, because Python's `&` binds more
tightly than its `==`; that rule is Python's, not the grammar's, and the
preface's list of Python's own rules is the whole of that subject.

## The words where the hosts disagree

Look at what the gloss did to the R sentence above. `==` became `is`, and `&`
became `and`. That is not a stylistic flourish; it is a decision about which
words can be in a grammar at all.

A symbol that means different things in different tools is a word this
grammar refuses to own. `=` assigns in one language, compares in another, and
half the world's beginners have been burned by the difference. So the text
form spells every contested symbol as the English word that means only one
thing, and each binding translates its host's habit into that word:

| you write in R | you write in Python | the grammar's word |
|---|---|---|
| `==` | `==` | `is` |
| `!=` | `!=` | `is not` |
| `&` | `&` | `and` |
| `\|` | `\|` | `or` |
| `!` | `~` | `not` |
| `%in%` | `.is_in([...])` | `in { }` |
| `is.na(x)` | `.is_missing()` | `[x] is missing` |

You keep writing your language's habits, and the grammar keeps one meaning
under them. When a pipeline moves to a database cell or a colleague's screen,
it is the unambiguous spelling that travels.

## One verb, both directions

There is no separate verb for dropping rows. Removing the rows where
something is true is keeping the rows where it is false, and one verb says
both:

::: {.panel-tabset}

### R

```{r}
sales |> keep(!(product == "Widget"))
```

### Python

```{python}
sales >> keep(~(col.product == "Widget"))
```

:::

This is the first sighting of a rule you will meet on every page: if the
grammar can already say something, it does not grow a second way to say it. A
`drop` verb would cost a word, and every reader who met `drop(x)` beside
`keep(!x)` would have to learn, and then remember, that they are the same
thing. The grammar spends its words on meanings, not on synonyms.

## What travels with it

- `where` is how the text form marks the condition, as in the gloss above.
  You never type it in R or Python; the condition sits inside the verb.
- The condition language itself travels whole. Everything you can write in a
  `keep` you can later write inside `pick where` (choosing columns by a
  question) and inside `when` (choosing a value by a question), unchanged.
- `and`, `or`, `not`, `is`, `in`, `missing` are the grammar's own words for
  the contested symbols, and the table above is their whole story.

## What it refuses

A condition that can never be true is a mistake, not a preference, and the
grammar says so before anything runs. Text is never greater than a number:

::: {.panel-tabset}

### R

```{r}
#| error: true
collect(sales |> keep(region > 100))
```

### Python

```{python}
#| classes: refusal
try:
    collect(sales >> keep(col.region > 100))
except GodError as refusal:
    print(refusal)
```

:::

Nothing was scanned and no rows came back empty. The pipeline was refused
whole, at the step that is wrong, with the fix in the message. A tool that
instead answered with zero rows would have told you something false about
your data, quietly, and you would have believed it.
