---
title: "Missing values"
---

{{< include ../_setup.qmd >}}

What should a hole in the table mean for your answer? `drop_missing` drops
rows where a named column is missing. With no column named, it looks at
every column.

::: {.panel-tabset}

### R

```{r}
patchy <- data.frame(
  product = c("Widget", "Gadget", "Doohickey"),
  revenue = c(100, NA, 300)
)

patchy |> drop_missing(revenue)
```

### Python

```{python}
patchy = pd.DataFrame({
    "product": ["Widget", "Gadget", "Doohickey"],
    "revenue": [100.0, None, 300.0],
})

patchy >> drop_missing(col.revenue)
```

:::

*`patchy then drop_missing [revenue]`*

`fill_missing` replaces missing values with something.

::: {.panel-tabset}

### R

```{r}
patchy |> fill_missing(revenue = 0)
```

### Python

```{python}
patchy >> fill_missing(revenue = 0)
```


### run

```{r}
run('
  patchy
    then fill_missing [revenue] as 0
')
```

:::

The filler has to be the same kind of thing the column holds, or the column would
quietly stop being what it was.

::: {.panel-tabset}

### R

```{r}
#| error: true
collect(patchy |> fill_missing(revenue = "none"))
```

### Python

```{python}
#| classes: refusal
try:
    collect(patchy >> fill_missing(revenue = "none"))
except GodError as refusal:
    print(refusal)
```

:::

## Taking the first column that has a value

`fill_missing` puts one value in every gap. When the value you want is in another
column, `first_present` reads across the columns you name and takes the first one
that is there.

::: {.panel-tabset}

### R

```{r}
contacts <- data.frame(
  name     = c("ana", "bo", "cy"),
  mobile   = c("555-01", NA, NA),
  landline = c("555-99", "555-02", NA),
  email    = c("a@x", "b@x", "c@x"),
  stringsAsFactors = FALSE
)

contacts |> add(reach = first_present(mobile, landline, email))
```

### Python

```{python}
contacts = pd.DataFrame({
    "name":     ["ana", "bo", "cy"],
    "mobile":   ["555-01", None, None],
    "landline": ["555-99", "555-02", None],
    "email":    ["a@x", "b@x", "c@x"],
})

contacts >> add(reach = first_present(col.mobile, col.landline, col.email))
```

:::

Two things about it are worth knowing before you rely on it, and both are places
people are usually surprised.

The order is a priority rather than a set. It reads left to right and stops at the
first column that has a value, so ana is reached on the mobile number even though
a landline and an email are there too. Writing the same three columns in another
order gives a different answer, which is the point of writing them in an order at
all.

The only thing it skips is a missing value. A zero, an empty text and a `no`
are all values, and they come back.

::: {.panel-tabset}

### R

```{r}
readings <- data.frame(sensor = c(0, NA), backup = c(99, 99))
readings |> add(used = first_present(sensor, backup))
```

### Python

```{python}
readings = pd.DataFrame({"sensor": [0.0, None], "backup": [99.0, 99.0]})
readings >> add(used = first_present(col.sensor, col.backup))
```

:::

The first sensor read zero, which is a reading, so it is kept. Only the row where
the sensor recorded nothing at all falls through to the backup. If you want zero
treated as missing too, say so in a step of its own, because that is a decision
about your data rather than about the word.

Every column it looks in has to hold the same kind of thing, since one of them is
going to be the answer and a column holds one kind. If all of them are missing
for a row, the answer for that row is missing.

SQL and dplyr both call this `coalesce`. It is `first_present` here because
`missing` is already the grammar's word for the absent value, so `present` is its
opposite, and `first` says the part that trips people up.
