---
title: "Missing values"
---

{{< include ../_setup.qmd >}}

```{r}
#| include: false
patchy <- god_table("patchy")
contacts <- god_table("contacts")
sensors <- god_table("sensors")
```

```{python}
#| include: false
patchy = god_table("patchy")
contacts = god_table("contacts")
sensors = god_table("sensors")
```

What should a hole in the table mean for your answer? `drop_missing` drops
rows where a named column is missing. With no column named, it looks at
every column.

The table is `patchy`, three products with one revenue that never arrived:

```{r}
#| echo: false
patchy
```

::: {.panel-tabset}

### run

```{r}
run('
  patchy
    then drop_missing [revenue]
')
```

### R

```{r}
patchy |> drop_missing(revenue)
```

### Python

```{python}
patchy >> drop_missing(col.revenue)
```

:::

*`patchy then drop_missing [revenue]`*

`fill_missing` replaces missing values with something.

::: {.panel-tabset}

### run

```{r}
run('
  patchy
    then fill_missing [revenue] as 0
')
```

### R

```{r}
patchy |> fill_missing(revenue = 0)
```

### Python

```{python}
patchy >> fill_missing(revenue = 0)
```

:::

The filler has to be the same kind of thing the column holds, or the column would
quietly stop being what it was.

::: {.panel-tabset}

### R

```{r}
#| error: true
collect(patchy |> fill_missing(revenue = "none"))
```

### Python

```{python}
#| classes: refusal
try:
    collect(patchy >> fill_missing(revenue = "none"))
except GodError as refusal:
    print(refusal)
```

:::

## Taking the first column that has a value

`fill_missing` puts one value in every gap. When the value you want is in another
column, `first_present` reads across the columns you name and takes the first one
that is there.

`contacts` holds three ways of reaching someone and not everybody has all three:

```{r}
#| echo: false
contacts
```

::: {.panel-tabset}

### run

```{r}
run('
  contacts
    then add [reach] as first_present([mobile], [landline], [email])
')
```

### R

```{r}
contacts |> add(reach = first_present(mobile, landline, email))
```

### Python

```{python}
contacts >> add(reach = first_present(col.mobile, col.landline, col.email))
```

:::

*`contacts then add [reach] as first_present([mobile], [landline], [email])`*

Two things about it surprise people, and both matter before you rely on it.

The order is a priority rather than a set. It reads left to right and stops at the
first column that has a value, so ana is reached on the mobile number even though
a landline and an email are there too. Writing the same three columns in another
order gives a different answer, which is the point of writing them in an order at
all.

The only thing it skips is a missing value. A zero, an empty text and a `no`
are all values, and they come back.

```{r}
#| echo: false
sensors
```

::: {.panel-tabset}

### run

```{r}
run('
  sensors
    then add [used] as first_present([sensor], [backup])
')
```

### R

```{r}
sensors |> add(used = first_present(sensor, backup))
```

### Python

```{python}
sensors >> add(used = first_present(col.sensor, col.backup))
```

:::

*`sensors then add [used] as first_present([sensor], [backup])`*

The first sensor read zero, which is a reading, so it is kept. Only the row where
the sensor recorded nothing at all falls through to the backup. If you want zero
treated as missing too, say so in a step of its own, because that is a decision
about your data rather than about the word.

Every column it looks in has to hold the same kind of thing, since one of them is
going to be the answer and a column holds one kind. If all of them are missing
for a row, the answer for that row is missing.

SQL and dplyr both call this `coalesce`. It is `first_present` here because
`missing` is already the grammar's word for the absent value, so `present` is its
opposite, and `first` names the part people most often misread: the columns
are a priority order.

## Carrying the last value forward

A reading taken hourly, where the sensor only reports when the value changes,
leaves holes that are not really missing: the value is the last one you saw.
`latest` fills a hole with the most recent value above it.

::: {.panel-tabset}

### run

```{r}
readings <- data.frame(
  taken_at = 1:6,
  reading  = c(10, NA, NA, 40, NA, 60)
)

run('
  readings
    then sort [taken_at]
    then add [reading] as latest([reading])
', readings = readings)
```

### R

```{r}
readings |> sort(taken_at) |> add(reading = latest(reading))
```

### Python

```{python}
readings = pd.DataFrame({
    "taken_at": [1, 2, 3, 4, 5, 6],
    "reading":  [10, None, None, 40, None, 60],
})

readings >> sort(col.taken_at) >> add(reading = latest(col.reading))
```

:::

*`readings then sort [taken_at] then add [reading] as latest([reading])`*

It fills a run of holes, not just one. That is the difference between this
and reaching for `previous`, which looks back exactly one row and leaves the
second hole of a pair still open.

Like every value that walks the rows, it needs a `sort` first, because "the
last value" means nothing until something has said last in what order. And
`by` restarts it for each group, so one group never borrows another's value.

::: {.panel-tabset}

### run

```{r}
readings <- data.frame(
  sensor   = c("a", "a", "a", "b", "b", "b"),
  taken_at = c(1, 2, 3, 1, 2, 3),
  reading  = c(10, NA, NA, 50, NA, NA)
)

run('
  readings
    then sort [sensor], [taken_at]
    then add [reading] as latest([reading]) by [sensor]
', readings = readings)
```

### R

```{r}
readings |>
  sort(sensor, taken_at) |>
  add(reading = latest(reading), by = sensor)
```

### Python

```{python}
readings = pd.DataFrame({
    "sensor":   ["a", "a", "a", "b", "b", "b"],
    "taken_at": [1, 2, 3, 1, 2, 3],
    "reading":  [10, None, None, 50, None, None],
})

(readings
  >> sort(col.sensor, col.taken_at)
  >> add(reading = latest(col.reading), by = col.sensor))
```

:::

A row with nothing above it and nothing of its own stays missing, because
there is no earlier value to carry. `fill_missing` after it puts a value
there if you want one.

## Where the holes go when you sort

Holes arrive on their own once tables meet. `sales` sells a Sprocket and
`products` has never heard of one, so joining them leaves rows whose `maker`
nobody recorded:

::: {.panel-tabset}

### run

```{r}
run('
  sales
    then join products by [product]
    then summarize [total] as total([revenue]) by [maker]
    then sort [maker]
')
```

### R

```{r}
sales |>
  join(products, by = product) |>
  summarize(total = total(revenue), by = maker) |>
  sort(maker)
```

### Python

```{python}
(sales
  >> join(products, by = col.product)
  >> summarize(total = total(col.revenue), by = col.maker)
  >> sort(col.maker))
```

:::

*`sales then join products by [product] then summarize [total] as total([revenue]) by [maker] then sort [maker]`*

Sorting raises a question the sort itself does not answer: the row with no maker
has nothing to be ordered by, so where does it go? The grammar's answer is
**last**, and you get it without asking. Reading it aloud is the reason: the
values you have, then the ones you do not. It stays last when you reverse the
sort, too. `descending` reverses the values, and a row with no value has nothing
to reverse.

When you want the holes first, say so. `missing first` goes after the keys,
because it applies to the whole sort rather than to one of them:

::: {.panel-tabset}

### run

```{r}
run('
  sales
    then join products by [product]
    then summarize [total] as total([revenue]) by [maker]
    then sort [maker] missing first
')
```

### R

```{r}
sales |>
  join(products, by = product) |>
  summarize(total = total(revenue), by = maker) |>
  sort(maker, missing = "first")
```

### Python

```{python}
(sales
  >> join(products, by = col.product)
  >> summarize(total = total(col.revenue), by = col.maker)
  >> sort(col.maker, missing = "first"))
```

:::

*`sales then join products by [product] then summarize [total] as total([revenue]) by [maker] then sort [maker] missing first`*

This is a place where the tools underneath genuinely disagree, and the grammar
picks one answer for all of them. Left to themselves, DuckDB, dplyr and pandas
put the absent rows last; polars puts them first; Spark puts them first going up
and last coming down. The same question, five tools, three answers. So each one
is told explicitly where they go, and the sentence means the same thing wherever
it runs, which is the whole of what this grammar is for.
