---
title: "Missing values"
---

{{< include ../_setup.qmd >}}

```{r}
#| include: false
patchy <- god_table("patchy")
contacts <- god_table("contacts")
sensors <- god_table("sensors")
```

```{python}
#| include: false
patchy = god_table("patchy")
contacts = god_table("contacts")
sensors = god_table("sensors")
```

What should a hole in the table mean for your answer? `drop_missing` drops
rows where a named column is missing. With no column named, it looks at
every column.

The table is `patchy`, three products with one revenue that never arrived:

```{r}
#| echo: false
patchy
```

::: {.panel-tabset}

### run

```{r}
run('
  patchy
    then drop_missing [revenue]
')
```

### R

```{r}
patchy |> drop_missing(revenue)
```

### Python

```{python}
patchy >> drop_missing(col.revenue)
```

:::

*`patchy then drop_missing [revenue]`*

`fill_missing` replaces missing values with something.

::: {.panel-tabset}

### run

```{r}
run('
  patchy
    then fill_missing [revenue] as 0
')
```

### R

```{r}
patchy |> fill_missing(revenue = 0)
```

### Python

```{python}
patchy >> fill_missing(revenue = 0)
```

:::

The filler has to be the same kind of thing the column holds, or the column would
quietly stop being what it was.

::: {.panel-tabset}

### R

```{r}
#| error: true
collect(patchy |> fill_missing(revenue = "none"))
```

### Python

```{python}
#| classes: refusal
try:
    collect(patchy >> fill_missing(revenue = "none"))
except GodError as refusal:
    print(refusal)
```

:::

## Taking the first column that has a value

`fill_missing` puts one value in every gap. When the value you want is in another
column, `first_present` reads across the columns you name and takes the first one
that is there.

`contacts` holds three ways of reaching someone and not everybody has all three:

```{r}
#| echo: false
contacts
```

::: {.panel-tabset}

### run

```{r}
run('
  contacts
    then add [reach] as first_present([mobile], [landline], [email])
')
```

### R

```{r}
contacts |> add(reach = first_present(mobile, landline, email))
```

### Python

```{python}
contacts >> add(reach = first_present(col.mobile, col.landline, col.email))
```

:::

*`contacts then add [reach] as first_present([mobile], [landline], [email])`*

Two things about it are worth knowing before you rely on it, and both are places
people are usually surprised.

The order is a priority rather than a set. It reads left to right and stops at the
first column that has a value, so ana is reached on the mobile number even though
a landline and an email are there too. Writing the same three columns in another
order gives a different answer, which is the point of writing them in an order at
all.

The only thing it skips is a missing value. A zero, an empty text and a `no`
are all values, and they come back.

```{r}
#| echo: false
sensors
```

::: {.panel-tabset}

### run

```{r}
run('
  sensors
    then add [used] as first_present([sensor], [backup])
')
```

### R

```{r}
sensors |> add(used = first_present(sensor, backup))
```

### Python

```{python}
sensors >> add(used = first_present(col.sensor, col.backup))
```

:::

*`sensors then add [used] as first_present([sensor], [backup])`*

The first sensor read zero, which is a reading, so it is kept. Only the row where
the sensor recorded nothing at all falls through to the backup. If you want zero
treated as missing too, say so in a step of its own, because that is a decision
about your data rather than about the word.

Every column it looks in has to hold the same kind of thing, since one of them is
going to be the answer and a column holds one kind. If all of them are missing
for a row, the answer for that row is missing.

SQL and dplyr both call this `coalesce`. It is `first_present` here because
`missing` is already the grammar's word for the absent value, so `present` is its
opposite, and `first` names the part people most often misread: the columns
are a priority order.

## Carrying the last value forward

A reading taken hourly, where the sensor only reports when the value changes,
leaves holes that are not really missing: the value is the last one you saw.
`latest` fills a hole with the most recent value above it.

::: {.panel-tabset}

### run

```{r}
readings <- data.frame(
  taken_at = 1:6,
  reading  = c(10, NA, NA, 40, NA, 60)
)

run('
  readings
    then sort [taken_at]
    then add [reading] as latest([reading])
', readings = readings)
```

### R

```{r}
readings |> sort(taken_at) |> add(reading = latest(reading))
```

### Python

```{python}
readings = pd.DataFrame({
    "taken_at": [1, 2, 3, 4, 5, 6],
    "reading":  [10, None, None, 40, None, 60],
})

readings >> sort(col.taken_at) >> add(reading = latest(col.reading))
```

:::

*`readings then sort [taken_at] then add [reading] as latest([reading])`*

It fills a run of holes, not just one. That is the difference between this
and reaching for `previous`, which looks back exactly one row and leaves the
second hole of a pair still open.

Like every value that walks the rows, it needs a `sort` first, because "the
last value" means nothing until something has said last in what order. And
`by` restarts it for each group, so one group never borrows another's value.

::: {.panel-tabset}

### run

```{r}
readings <- data.frame(
  sensor   = c("a", "a", "a", "b", "b", "b"),
  taken_at = c(1, 2, 3, 1, 2, 3),
  reading  = c(10, NA, NA, 50, NA, NA)
)

run('
  readings
    then sort [sensor], [taken_at]
    then add [reading] as latest([reading]) by [sensor]
', readings = readings)
```

### R

```{r}
readings |>
  sort(sensor, taken_at) |>
  add(reading = latest(reading), by = sensor)
```

### Python

```{python}
readings = pd.DataFrame({
    "sensor":   ["a", "a", "a", "b", "b", "b"],
    "taken_at": [1, 2, 3, 1, 2, 3],
    "reading":  [10, None, None, 50, None, None],
})

(readings
  >> sort(col.sensor, col.taken_at)
  >> add(reading = latest(col.reading), by = col.sensor))
```

:::

A row with nothing above it and nothing of its own stays missing, because
there is no earlier value to carry. `fill_missing` after it puts a value
there if you want one.
