---
title: "Nothing runs until you ask"
---

{{< include ../_setup.qmd >}}

A verb does not return a table. It returns a pipeline, which is a plan for one.

::: {.panel-tabset}

### R

```{r}
answer <- sales |> keep(region == "West") |> take(2)
class(answer)
```

### Python

```{python}
answer = sales >> keep(col.region == "West") >> take(2)
type(answer).__name__
```

:::

Printing it runs it, which is why every example so far shows a table. When you
want the table itself, ask for it with `collect`.

::: {.panel-tabset}

### R

```{r}
nrow(collect(answer))
```

### Python

```{python}
len(collect(answer))
```

:::

In R this matters for more than speed. `sort` and `keep` are names other
packages use too, so a pipeline can land in some other package's `sort` by
mistake. A pipeline is not a table, so the stranger's `sort` has no meaning
for it, and the call stops rather than returning a plausible answer from the
wrong tool.

Forgetting `collect` is covered by the same protection. Ask a plan a table's
question, `nrow(answer)`, `names(answer)`, `answer$revenue`, and R's own
answer for each would be NULL, and NULL flows onward: `sum(answer$revenue)`
would be 0, with no complaint anywhere. So a pipeline refuses the question
and names the repair:

```{r}
#| error: true
answer$revenue
```

Python needs no guard built for this. A pipeline there has no length, no
brackets and no iteration, so the same slip stops itself, with the
language's own error.

You can also see what the verbs wrote, which is the text form, and it is the same
sentence in both languages.

::: {.panel-tabset}

### R

```{r}
cat(format(sales |> keep(region == "West") |> take(10)))
```

### Python

```{python}
print((sales >> keep(col.region == "West") >> take(10)).written())
```

:::

The two calls answer with the same bytes, and only their names differ:
`format` is the word R already owns for turning a thing into text, and
Python's `format` means something else, so its method says what it gives.
What comes back is the text form: the sentence with no pipe glyph and no
column syntax left, which is what
[the next chapter](what-it-wrote.qmd) stores, sends and runs.

Printing is also how a notebook shows a pipeline. In Jupyter, and on every
page of this book, a pipeline standing at the end of a cell renders as a
real table rather than as monospaced text. Each binding tells its notebook
how to display one, so the answer is the same everywhere and only the
rendering differs.

The whole rule is one sentence. Display runs a pipeline for free; anything
else asks with `collect`.
