---
title: "Nothing runs until you ask"
---

{{< include ../_setup.qmd >}}

A verb does not return a table. It returns a pipeline, which is a plan for one.

::: {.panel-tabset}

### R

```{r}
answer <- sales |> keep(region == "West") |> take(2)
class(answer)
```

### Python

```{python}
answer = sales >> keep(col.region == "West") >> take(2)
type(answer).__name__
```

:::

Printing it runs it, which is why the examples so far have shown tables
rather than plans. When you want the table itself, ask for it with `collect`.

::: {.panel-tabset}

### R

```{r}
nrow(collect(answer))
```

### Python

```{python}
len(collect(answer))
```

:::

The plan buys you speed. Steps cost nothing until you ask, and then the
whole sentence reaches the engine as one query. Six steps are not six
passes over the table, because the engine sees the whole question first.
[The machine underneath](../appendices/h-the-machine-underneath.qmd)
shows a pipeline arriving whole.

In R this matters for more than speed. `sort` and `keep` are names other
packages use too, so a pipeline can reach some other package's `sort` by
mistake. A pipeline is not a table, so that `sort` has no meaning for it,
and the call stops rather than returning a plausible answer from the wrong
tool.

Forgetting `collect` is covered by the same protection. Ask a plan a table's
question and R's own answers would be quietly wrong. `names(answer)` would
list the plan's own parts, and `nrow(answer)` and `answer$revenue` would be
NULL. NULL flows onward: `sum(answer$revenue)` would be 0, with no complaint
anywhere. So a pipeline refuses the question and names the repair:

```{r}
#| error: true
answer$revenue
```

Python needs no guard built for this. A pipeline there has no length, no
brackets and no iteration, so the same mistake stops itself, with the
language's own error.

You can also see what the verbs wrote, which is the text form, and it is the same
sentence in both languages.

::: {.panel-tabset}

### R

```{r}
cat(format(sales |> keep(region == "West") |> take(10)))
```

### Python

```{python}
print((sales >> keep(col.region == "West") >> take(10)).written())
```

:::

The two calls answer with the same bytes, and only their names differ.
`format` is the word R already owns for turning a thing into text. Python's
`format` means something else, so its method says what it gives.
What comes back is the text form: the sentence with no pipe glyph and
neither language's column spelling left, which is what
[the next chapter](what-it-wrote.qmd) stores, sends and runs.

Printing is also how a notebook shows a pipeline. In Jupyter, and on every
page of this book, a pipeline standing at the end of a cell renders as a
real table rather than as monospaced text. Each binding tells its notebook
how to display one, so the answer is the same everywhere and only the
rendering differs.

The whole rule is one sentence. Display runs a pipeline for free; anything
else asks with `collect`.
