---
title: "One way or another"
---

{{< include ../_setup.qmd >}}

```{r}
#| include: false
pupils <- god_table("pupils")
```

```{python}
#| include: false
pupils = god_table("pupils")
```

Is 95 an A or a B, and who decides in which order the rules apply? `when`
asks a question and gives the answer beside it, then asks the next one.

The table is `pupils`, three scores that fall in three different bands:

```{r}
#| echo: false
pupils
```

::: {.panel-tabset}

### run

```{r}
run('
  pupils
    then add [band] as when(([score] >= 90), "A", ([score] >= 70), "B", otherwise "C")
')
```

### R

```{r}
pupils |>
  add(band = when(score >= 90, "A", score >= 70, "B", otherwise = "C"))
```

### Python

```{python}
pupils >> add(band = when(col.score >= 90, "A", col.score >= 70, "B",
                          otherwise = "C"))
```

:::

*`pupils then add [band] as when(([score] >= 90), "A", ([score] >= 70), "B", otherwise "C")`*

The arguments come in pairs: a question, then what it gives. **The first question
that is true wins**, so the order is part of what the sentence means. Ask the
same two the other way round and the answers change, which is not a trap but the
thing you are choosing when you write them in an order:

::: {.panel-tabset}

### run

```{r}
run('
  pupils
    then add [band] as when(([score] >= 70), "B", ([score] >= 90), "A",
         otherwise "C")
')
```

### R

```{r}
pupils |>
  add(band = when(score >= 70, "B", score >= 90, "A", otherwise = "C"))
```

### Python

```{python}
pupils >> add(band = when(col.score >= 70, "B", col.score >= 90, "A",
                          otherwise = "C"))
```

:::

`ann` scored 95, met the first question that was asked, and got a B.

Leave out `otherwise` and a row that matched nothing is missing:

::: {.panel-tabset}

### run

```{r}
run('
  pupils
    then add [top] as when(([score] >= 90), "yes")
')
```

### R

```{r}
pupils |> add(top = when(score >= 90, "yes"))
```

### Python

```{python}
pupils >> add(top = when(col.score >= 90, "yes"))
```

:::

Every answer has to be the same kind of thing, because they all end up in one
column:

::: {.panel-tabset}

### R

```{r}
#| error: true
collect(pupils |> add(band = when(score >= 90, "A", otherwise = 0)))
```

### Python

```{python}
#| classes: refusal
try:
    collect(pupils >> add(band = when(col.score >= 90, "A", otherwise = 0)))
except GodError as refusal:
    print(refusal)
```

:::

Python has a conditional of its own, and it is worth saying why god does not use
it. Writing `"A" if col.score >= 90 else "B"` would decide the answer once, while
the pipeline was being built, and throw the question away. Nothing would report a
problem. R's `if` wants a single yes or no, so a column of answers is an error
there. So the word is `when` in both, and it is the same word in the text form.

## The lookup table

Half the time the question is the same question over and over: is the value
this, is it that. Recoding survey answers, expanding country codes, tidying a
categorical. `look_up` is that special case with the pairs side by side, each
written value next to what it becomes:

::: {.panel-tabset}

### run

```{r}
run('
  sales
    then add [short] as look_up([region], "West", "W", "East", "E", otherwise [region])
    then pick [region, short]
    then drop_duplicates
')
```

### R

```{r}
sales |>
  add(short = look_up(region, "West", "W", "East", "E", otherwise = region)) |>
  pick(region, short) |>
  drop_duplicates()
```

### Python

```{python}
(sales
 >> add(short = look_up(col.region, "West", "W", "East", "E",
                        otherwise = col.region))
 >> pick(col.region, col.short)
 >> drop_duplicates())
```

:::

`North` has no pair, and the `otherwise` said what happens to it: naming the
column keeps it as it was. Naming `missing` drops it instead, and a written
value is a default. Those three endings are the whole difference between the
two functions dplyr gives this idea and the two polars gives it, so god writes
the ending rather than naming the word twice.

The ending is required, and leaving it off is refused rather than guessed:

::: {.panel-tabset}

### R

```{r}
#| error: true
collect(sales |> add(short = look_up(region, "West", "W")))
```

### Python

```{python}
#| classes: refusal
try:
    collect(sales >> add(short = look_up(col.region, "West", "W")))
except GodError as refusal:
    print(refusal)
```

:::

Half of the neighbouring tools keep an unpaired value and half send it
missing, so whichever way god leaned, the sentence would surprise someone
arriving from the other half. Writing the ending is how you say you thought
about it, exactly as `unmatched` says it for a join's rows.

A pair may send a value missing, which is how "this value means absent" is
written: `look_up([code], "", missing, otherwise [code])` turns empty text
into a proper hole and leaves everything else alone. The pairs are written
values on both sides, and that is the boundary between the two words: the
moment a question is a comparison, or an answer is computed, the sentence
belongs to `when`.
