---
title: "One value per column"
---

{{< include ../_setup.qmd >}}

Why write the same calculation eight times? Repeating a thing for eight
columns is eight chances to write it slightly differently. `add where` takes
the same pattern the last two chapters used and one value to apply to each
column it matches. Inside that value, `value` stands for the column being
worked on.

::: {.panel-tabset}

### run

```{r}
run('
  survey
    then add where (name starts "q") as (value * 10)
')
```

### R

```{r}
survey |> add(where(startsWith(name, "q"), value * 10))
```

### Python

```{python}
survey >> add(where(name.starts("q"), value * 10))
```

:::

*`survey then add where (name starts "q") as (value * 10)`*

The matched columns keep their names. That is not a rule this had to invent:
`add` already covers making a column and replacing one, so replacing `q1_score`
with a new `q1_score` is the ordinary case.

`name` and `value` are a pair. One is what the column is called, the other is
what it holds, and they are the same two words the reshaping verbs use.

`summarize where` works the same way and collapses instead.

::: {.panel-tabset}

### run

```{r}
run('
  survey
    then summarize where (name ends "_score") as average(value) by [region]
')
```

### R

```{r}
survey |>
  summarize(where(endsWith(name, "_score"), average(value)), by = region)
```

### Python

```{python}
(survey
  >> summarize(where(name.ends("_score"), average(value)), by = col.region))
```

:::

dplyr calls this `across`. Here it needed no verb and one word, because the
selector was already there for `pick` and `where` already introduces a
condition.

Every rule that applies to a value written out by hand applies to this one,
because the grammar expands it before checking anything. Asking `summarize` for
something that does not collapse a group is refused, and the message names the
column it refused rather than the pattern.

::: {.panel-tabset}

### R

```{r}
#| error: true
collect(survey |> summarize(where(startsWith(name, "q"), value * 2)))
```

### Python

```{python}
#| classes: refusal
try:
    collect(survey >> summarize(where(name.starts("q"), value * 2)))
except GodError as refusal:
    print(refusal)
```

:::

Look at what it expanded to the first few times. `format` shows the sentence
you wrote; `show_as` shows the one the grammar settled on.

::: {.panel-tabset}

### R

```{r}
show_as(survey |> add(where(startsWith(name, "q"), value * 10)), "god")
```

### Python

```{python}
show_as(survey >> add(where(name.starts("q"), value * 10)), "god")
```

:::

The grammar's own printing expands the rule into the columns it matched, so
what actually ran is on the page rather than implied, which is honest to
whoever reads the pipeline after you.

## Asking one question of every column

The same argument works for a question. If eight columns hold scores, "who
answered above 3 anywhere?" is one question about eight columns, and writing
it eight times joined by `or` is eight chances to miss one. `keep` takes the
same selector, with `any` in front of it.

::: {.panel-tabset}

### run

```{r}
run('
  survey
    then keep where any name starts "q" as value > 4
    then pick [name, q1_score, q2_score]
')
```

### R

```{r}
survey |>
  keep(where_any(startsWith(name, "q"), value > 4)) |>
  pick(name, q1_score, q2_score)
```

### Python

```{python}
(survey
  >> keep(where_any(name.starts("q"), value > 4))
  >> pick(col.name, col.q1_score, col.q2_score))
```

:::

*`survey then keep where any name starts "q" as value > 4`*

`every` is the other half, and it asks the harder question: which respondents
scored above 3 on all eight?

::: {.panel-tabset}

### run

```{r}
run('
  survey
    then keep where every name starts "q" as value > 2
    then pick [name]
')
```

### R

```{r}
survey |> keep(where_every(startsWith(name, "q"), value > 2)) |> pick(name)
```

### Python

```{python}
survey >> keep(where_every(name.starts("q"), value > 2)) >> pick(col.name)
```

:::

The two are `or` and `and` written once instead of seven times, and the
expansion is on the page if you ask for it, exactly as it is for `add`.

A selector that matches no column is refused rather than answered. Both
answers would be defensible: nothing is true of no columns, and everything
is. Neither is guessable, so a mistyped selector would otherwise empty your
table or leave it whole without a word.

::: {.panel-tabset}

### R

```{r}
#| error: true
collect(survey |> keep(where_any(startsWith(name, "z"), value > 2)))
```

### Python

```{python}
#| classes: refusal
try:
    collect(survey >> keep(where_any(name.starts("z"), value > 2)))
except GodError as refusal:
    print(refusal)
```

:::

The words are `any` and `every` in the sentence, and `where_any` and
`where_every` in both languages. `any` is a function Python already has, and
the grammar has never taken a name a host was using.
