---
title: "One value, applied to every column that matches"
---

{{< include ../_setup.qmd >}}

Why write the same calculation eight times? Repeating a thing for eight
columns is eight chances to write it slightly differently. `add where` takes
the same pattern the last two chapters used and one value to apply to each
column it matches. Inside that value, `value` stands for the column being
worked on.

::: {.panel-tabset}

### R

```{r}
survey |> add(where(startsWith(name, "q"), value * 10))
```

### Python

```{python}
survey >> add(where(name.starts("q"), value * 10))
```

:::

*`survey then add where (name starts "q") as (value * 10)`*

The matched columns keep their names. That is not a rule this had to invent:
`add` already covers making a column and replacing one, so replacing `q1_score`
with a new `q1_score` is the ordinary case.

`name` and `value` are a pair. One is what the column is called, the other is
what it holds, and they are the same two words the reshaping verbs use.

`summarize where` works the same way and collapses instead.

::: {.panel-tabset}

### R

```{r}
survey |> summarize(where(endsWith(name, "_score"), average(value)), by = region)
```

### Python

```{python}
survey >> summarize(where(name.ends("_score"), average(value)), by = col.region)
```

:::

dplyr calls this `across`. Here it needed no verb and one word, because the
selector was already there for `pick` and `where` already introduces a
condition.

Every rule that applies to a value written out by hand applies to this one,
because the grammar expands it before checking anything. Asking `summarize` for
something that does not collapse a group is refused, and the message names the
column it refused rather than the pattern.

::: {.panel-tabset}

### R

```{r}
#| error: true
collect(survey |> summarize(where(startsWith(name, "q"), value * 2)))
```

### Python

```{python}
#| classes: refusal
try:
    collect(survey >> summarize(where(name.starts("q"), value * 2)))
except GodError as refusal:
    print(refusal)
```

:::

You can see what it expanded to, which is worth doing the first few times.
`format` shows the sentence you wrote; `show_as` shows the one the grammar
settled on.

::: {.panel-tabset}

### R

```{r}
show_as(survey |> add(where(startsWith(name, "q"), value * 10)), "god")
```

### Python

```{python}
show_as(survey >> add(where(name.starts("q"), value * 10)), "god")
```

:::

The grammar's own printing expands the rule into the columns it matched, so
what actually ran is on the page rather than implied, which is the honest
way to hand a sweep to the reader after you.
