---
title: "Answering one way or another"
---

{{< include ../_setup.qmd >}}

Is 95 an A or a B, and who decides in which order the rules apply? `when`
asks a question and gives the answer beside it, then asks the next one.

::: {.panel-tabset}

### R

```{r}
pupils <- data.frame(
  name = c("ann", "bob", "cat"),
  score = c(95, 75, 50),
  stringsAsFactors = FALSE
)

pupils |> add(band = when(score >= 90, "A", score >= 70, "B", otherwise = "C"))
```

### Python

```{python}
pupils = pd.DataFrame({
    "name": ["ann", "bob", "cat"],
    "score": [95, 75, 50],
})

pupils >> add(band = when(col.score >= 90, "A", col.score >= 70, "B", otherwise = "C"))
```

:::

*`pupils then add [band] as when(([score] >= 90), "A", ([score] >= 70), "B", otherwise "C")`*

The arguments come in pairs: a question, then what it gives. **The first question
that is true wins**, so the order is part of what the sentence means. Ask the
same two the other way round and the answers change, which is not a trap but the
thing you are choosing when you write them in an order:

::: {.panel-tabset}

### R

```{r}
pupils |> add(band = when(score >= 70, "B", score >= 90, "A", otherwise = "C"))
```

### Python

```{python}
pupils >> add(band = when(col.score >= 70, "B", col.score >= 90, "A", otherwise = "C"))
```


### run

```{r}
run('
  pupils
    then add [band] as when(([score] >= 70), "B", ([score] >= 90), "A", otherwise "C")
')
```

:::

`ann` scored 95, met the first question that was asked, and got a B.

Leave out `otherwise` and a row that matched nothing is missing:

::: {.panel-tabset}

### R

```{r}
pupils |> add(top = when(score >= 90, "yes"))
```

### Python

```{python}
pupils >> add(top = when(col.score >= 90, "yes"))
```


### run

```{r}
run('
  pupils
    then add [top] as when(([score] >= 90), "yes")
')
```

:::

Every answer has to be the same kind of thing, because they all end up in one
column:

::: {.panel-tabset}

### R

```{r}
#| error: true
collect(pupils |> add(band = when(score >= 90, "A", otherwise = 0)))
```

### Python

```{python}
#| classes: refusal
try:
    collect(pupils >> add(band = when(col.score >= 90, "A", otherwise = 0)))
except GodError as refusal:
    print(refusal)
```

:::

Python has a conditional of its own, and it is worth saying why god does not use
it. Writing `"A" if col.score >= 90 else "B"` would decide the answer once, while
the pipeline was being built, and throw the question away. Nothing would report a
problem. R has no such expression at all. So the word is `when` in both, and it
is the same word in the text form.
