---
title: "Reading practice"
---

{{< include ../_setup.qmd >}}

Can you read a pipeline you have never seen? That was the preface's day-one
promise, and this chapter is where it is tested rather than repeated. Eight
questions follow, each answered by one sentence built only from the words of
this part. Read each pipeline aloud before you look at its table, then check
the table against what you said. Nothing here is new; that is the point.

## Which orders were larger than usual?

::: {.panel-tabset}

### R

```{r}
sales |> keep(revenue >= 200)
```

### Python

```{python}
sales >> keep(col.revenue >= 200)
```


### run

```{r}
run('sales
  then keep where ([revenue] >= 200)')
```

:::

*`sales then keep where [revenue] >= 200`*

Four orders clear the bar, and every region appears at least once, so size
is not a story about one place.

## What did each region sell, product by product?

::: {.panel-tabset}

### R

```{r}
sales |> summarize(orders = row_count(), revenue = total(revenue), by = c(region, product))
```

### Python

```{python}
sales >> summarize(orders = row_count(), revenue = total(col.revenue), by = [col.region, col.product])
```


### run

```{r}
run('sales
  then summarize [orders] as row_count(), [revenue] as total([revenue]) by [region, product]')
```

:::

*`sales then summarize [orders] as row_count(), [revenue] as total([revenue]) by [region, product]`*

Every combination that occurred gets a row, and one combination is absent:
no North row mentions Sprocket, because the North never sold one. A grouped
summary can only report what happened, which is worth remembering when a
combination you expected is simply not there.

## What does each product sell for?

::: {.panel-tabset}

### R

```{r}
sales |>
  add(price = revenue / quantity) |>
  summarize(price = average(price), sold = total(quantity), by = product) |>
  sort(descending(price))
```

### Python

```{python}
(sales
  >> add(price = col.revenue / col.quantity)
  >> summarize(price = average(col.price), sold = total(col.quantity), by = col.product)
  >> sort(descending(col.price)))
```


### run

```{r}
run('sales
  then add [price] as ([revenue] / [quantity])
  then summarize [price] as average([price]), [sold] as total([quantity]) by [product]
  then sort [price] descending')
```

:::

*`sales then add [price] as [revenue] / [quantity] then summarize [price] as average([price]), [sold] as total([quantity]) by [product] then sort [price] descending`*

Three verbs you know, in an order you can hear: work out each order's unit
price, average it per product, read the dearest first. The averages come out
whole because each product held one price all year, which you can verify
from the table faster than you can doubt it.

## Which two orders cost the most to fulfill?

::: {.panel-tabset}

### R

```{r}
sales |> sort(descending(cost)) |> take(2)
```

### Python

```{python}
sales >> sort(descending(col.cost)) >> take(2)
```


### run

```{r}
run('sales
  then sort [cost] descending
  then take 2')
```

:::

*`sales then sort [cost] descending then take 2`*

## Who joined the survey before April?

::: {.panel-tabset}

### R

```{r}
survey |> keep(joined < "2026-04-01") |> pick(name, region, joined)
```

### Python

```{python}
survey >> keep(col.joined < "2026-04-01") >> pick(col.name, col.region, col.joined)
```


### run

```{r}
run('survey
  then keep where ([joined] < "2026-04-01")
  then pick [name, region, joined]')
```

:::

*`survey then keep where [joined] < "2026-04-01" then pick [name, region, joined]`*

The comparison is on text, and it works because these dates are written
largest unit first; [the chapter on dates](dates.qmd) gives you the real
date words for every shape that is not so cooperative.

## How many countries does each continent hold?

::: {.panel-tabset}

### R

```{r}
gapminder |> keep(year == 2007) |> summarize(countries = row_count(), by = continent)
```

### Python

```{python}
gapminder >> keep(col.year == 2007) >> summarize(countries = row_count(), by = col.continent)
```


### run

```{r}
run('gapminder
  then keep where ([year] is 2007)
  then summarize [countries] as row_count() by [continent]')
```

:::

*`gapminder then keep where [year] is 2007 then summarize [countries] as row_count() by [continent]`*

Same sentence shape as the sales questions, seventeen hundred rows instead
of fifteen, and nothing about the spelling had to know the difference.

## What was the cheapest order in each region?

::: {.panel-tabset}

### R

```{r}
sales |> sort(cost) |> take(1, by = region) |> pick(region, product, cost)
```

### Python

```{python}
sales >> sort(col.cost) >> take(1, by = col.region) >> pick(col.region, col.product, col.cost)
```


### run

```{r}
run('sales
  then sort [cost]
  then take 1 by [region]
  then pick [region, product, cost]')
```

:::

*`sales then sort [cost] then take 1 by [region] then pick [region, product, cost]`*

The sort has no `descending` this time, so first means smallest. One word's
absence flipped the question, which is what reading aloud is for.

## How much of its region is each order?

::: {.panel-tabset}

### R

```{r}
sales |>
  add(share = revenue / total(revenue), by = region) |>
  sort(region, descending(share)) |>
  pick(region, product, revenue, share)
```

### Python

```{python}
(sales
  >> add(share = col.revenue / total(col.revenue), by = col.region)
  >> sort(col.region, descending(col.share))
  >> pick(col.region, col.product, col.revenue, col.share))
```


### run

```{r}
run('sales
  then add [share] as ([revenue] / total([revenue])) by [region]
  then sort [region], [share] descending
  then pick [region, product, revenue, share]')
```

:::

*`sales then add [share] as [revenue] / total([revenue]) by [region] then sort [region], [share] descending then pick [region, product, revenue, share]`*

Each region's shares sum to one, and the biggest single share in the table
belongs to the West's largest order. This is the hardest sentence in the
chapter, and it is made of a verb from chapter 4 and a word from chapter 8.

## What you just did

Eight questions, no new words. Every sentence was the everyday core doing
its everyday work, and if you read them aloud, you have now heard the
grammar more times than you will need to keep it. Reading was recognition.
The next chapter asks for the harder thing, and it is the last thing this
part asks of you: writing.
