---
title: "Renaming, and dropping repeats"
---

{{< include ../_setup.qmd >}}

Why is this row in the table twice? And why is the column called `revenue`
when the report you owe someone says `earned`? Two small verbs do the
housekeeping the core six leave behind: `rename` changes what a column is
called, and `drop_duplicates` removes exact repeats.

::: {.panel-tabset}

### R

```{r}
sales |> rename(earned = revenue) |> pick(product, earned)
```

### Python

```{python}
sales >> rename(earned = col.revenue) >> pick(col.product, col.earned)
```

:::

*`sales then rename [earned] as [revenue] then pick [product, earned]`*

`rename` changes the name and nothing else: not the values, not the order,
not the kind. The new name goes first, the way assignment reads and the way
every naming verb in this grammar already works. If you are arriving from
pandas, read that direction twice: `rename(columns={"old": "new"})` is the
other way around, and both directions are legal spellings here, so only the
grammar's one-direction rule keeps a pipeline readable without a glossary.

## Dropping repeats

`drop_duplicates` drops the rows that are identical across every column. The
answer comes back in a settled order, because dropping repeats says nothing
about which order the rest should be in, and an answer that reorders itself
between runs is an answer you cannot test.

::: {.panel-tabset}

### R

```{r}
data.frame(a = c(1, 1, 2), b = c("x", "x", "y")) |> drop_duplicates()
```

### Python

```{python}
pd.DataFrame({"a": [1, 1, 2], "b": ["x", "x", "y"]}) >> drop_duplicates()
```

:::

## Why it takes no columns

Other tools let you write `distinct(g)` or `drop_duplicates(subset="g")`, and
the argument means two different things depending on the tool and the day:
sometimes "the distinct values of `g`", sometimes "whole rows, one per value
of `g`". The second reading has a hole in it that no tool can fill: *which*
whole row, out of the several that share a `g`? Any answer is an arbitrary
choice dressed as a function.

So here the verb takes no columns, and each of the two things you might have
meant has its own honest spelling. The distinct values of a column are a
`pick` followed by the drop:

::: {.panel-tabset}

### R

```{r}
sales |> pick(region, product) |> drop_duplicates()
```

### Python

```{python}
sales >> pick(col.region, col.product) >> drop_duplicates()
```

:::

*`sales then pick [region, product] then drop_duplicates`*

And one whole row per group is last chapter's idiom, `sort` then
`take 1 by`, which forces you to say *which* row, because "the latest" or
"the largest" was always the real question hiding inside "one per group".
Two meanings, two spellings, and neither has an arbitrary choice inside it.

## What travels with it

- `as` in the gloss is the same naming word it was in `add` and `summarize`,
  with the same direction: what it becomes, then what it was.
- `pick` then `drop_duplicates` is this grammar's spelling of every
  `distinct()` you have written elsewhere.
- `sort` then `take 1 by` is its spelling of every "first row per group",
  and the refusal that guards it was staged two chapters ago.

## What it refuses

Renaming a column onto a name the table already has would leave two columns
with one name, and every later step ambiguous:

::: {.panel-tabset}

### R

```{r}
#| error: true
sales |> rename(region = revenue) |> collect()
```

### Python

```{python}
try:
    collect(sales >> rename(region = col.revenue))
except GodError as refusal:
    print(refusal)
```

:::

The message carries the fix in both directions, because this mistake has two
common causes: you meant a different new name, or you had the direction
backward and were trying to rename `region` itself. Naming which side goes
first, in the message, at the moment of failure, is how a one-direction rule
earns the small strictness it asks of you.
