---
title: "The rows that are not there"
---

{{< include ../_setup.qmd >}}

```{r}
#| include: false
sittings <- god_table("sittings")
```

```{python}
#| include: false
sittings = god_table("sittings")
```

Which region sold no Sprockets? Ask the sales table what each region sold of
each product, and the answer comes back looking complete.

::: {.panel-tabset}

### run

```{r}
run('
  sales
    then summarize [sold] as total([revenue]) by [region, product]
')
```

### R

```{r}
sales |> summarize(sold = total(revenue), by = c(region, product))
```

### Python

```{python}
sales >> summarize(sold = total(col.revenue), by = [col.region, col.product])
```

:::

*`sales then summarize [sold] as total([revenue]) by [region, product]`*

Eleven rows, for three regions and four products. Three fours are twelve. The
missing row is North and Sprocket, and the reason it is missing is that the
North never sold one. That is the very thing the question was asking, and it
is the only answer that did not survive to be read.

This is the quiet failure of grouping. A group is made from the rows that are
there, so a group with no rows is not empty: it is absent. A total that
should read zero is not shown at all, and nothing tells you that a line is
missing. Sort the answer and the gap moves. Chart it and the bar is not
short, it is not drawn.

## Making them appear

`add_combinations` takes the values two columns already hold and makes every
combination of them a row.

::: {.panel-tabset}

### run

```{r}
run('
  sales
    then add_combinations [region, product]
    then summarize [sold] as total([revenue]) by [region, product]
')
```

### R

```{r}
sales |>
  add_combinations(region, product) |>
  summarize(sold = total(revenue), by = c(region, product))
```

### Python

```{python}
sales >> add_combinations(col.region, col.product) \
      >> summarize(sold = total(col.revenue), by = [col.region, col.product])
```

:::

Twelve rows now, and North/Sprocket is one of them. It reads as missing
rather than as zero, and that is right so far: the verb added a row and no
value, because nothing in the table says what the North sold. `fill_missing`
is where that gets said.

::: {.panel-tabset}

### run

```{r}
run('
  sales
    then add_combinations [region, product]
    then fill_missing [revenue] as 0
    then summarize [sold] as total([revenue]) by [region, product]
    then keep where [sold] < 100
')
```

### R

```{r}
sales |>
  add_combinations(region, product) |>
  fill_missing(revenue = 0) |>
  summarize(sold = total(revenue), by = c(region, product)) |>
  keep(sold < 100)
```

### Python

```{python}
sales >> add_combinations(col.region, col.product) \
      >> fill_missing(revenue = 0) \
      >> summarize(sold = total(col.revenue), by = [col.region, col.product]) \
      >> keep(col.sold < 100)
```

:::

Two steps rather than one argument, and deliberately. Whether an absent
combination means zero is a question about the data and not about the verb.
A region that sold no Sprockets sold zero of them. A sensor that was switched
off did not record a zero. The grammar cannot tell those apart, so it
declines to guess, and the second step is where you say which one you have.

## The values come from the table

Only the values already in a column are crossed. That is the boundary of what
this verb can do. A product nobody has ever sold anywhere is not in the
`product` column, so no row is made for it. A month with no orders at all
appears nowhere in the table, so no crossing can produce it.

The same rule settles what happens to a hole. A missing value is not a
category, so it makes no combinations, and nothing is lost by that: the rows
that were already there are handed on untouched. This verb never looks at
the holes.

## Inside a group

Crossing every value against every other is right for regions and products,
and wrong as soon as a third column says which combinations are real. Every
student took their own school's paper:

```{r}
#| echo: false
sittings
```

::: {.panel-tabset}

### run

```{r}
run('
  sittings
    then add_combinations [student, question] by [school]
')
```

### R

```{r}
sittings |> add_combinations(student, question, by = school)
```

### Python

```{python}
sittings >> add_combinations(col.student, col.question, by = col.school)
```

:::

*`sittings then add_combinations [student, question] by [school]`*

`by` names the columns to hold fixed, and the crossing happens inside each
group. Bob is given q2, because a student at bob's school took it; nobody is
given q3, because that is the west's paper. Cal gains nothing at all.

Without `by`, the same sentence would cross all three students against all
three questions and hand back nine rows with the school column empty in every
one that was made. That is a table saying something false in a shape that
looks tidy.

## What it refuses

Two columns or more, always. One column crossed with nothing is the values it
already holds, so the sentence would hand the table straight back, and a step
that cannot do anything is better refused than obeyed in silence.

::: {.panel-tabset}

### R

```{r}
#| error: true
collect(sales |> add_combinations(region))
```

### Python

```{python}
#| classes: refusal
try:
    collect(sales >> add_combinations(col.region))
except GodError as refusal:
    print(refusal)
```

:::

A column cannot be crossed and held fixed at once either, and the message
names both repairs. Crossing a column and holding it fixed mean different
tables, so the grammar leaves the choice to you.

::: {.panel-tabset}

### R

```{r}
#| error: true
collect(sales |> add_combinations(region, product, by = region))
```

### Python

```{python}
#| classes: refusal
try:
    collect(sales >> add_combinations(col.region, col.product, by = col.region))
except GodError as refusal:
    print(refusal)
```

:::

Another verb in the vocabulary takes a table rather than columns, and the two
words start alike. Writing a table's name here is answered by the verb that
wanted one:

```{r}
#| error: true
run('sales then add_combinations more')
```

That message belongs to the sentence rather than to either language. Written
in R or in Python, a bare name in this position is a *column*, so the two
bindings hand the grammar `add_combinations [more]`. The reply is the one for
a column that is not there. That is also the truth, and it leads to the same
repair.

`add_rows` stacks another table underneath this one; `add_combinations` works
on this table's own values and uses nothing outside it. Both make the table
taller, which is why they share a first word, and only one of them needs a
second table to do it.
