---
title: "Rows into columns"
---

{{< include ../_setup.qmd >}}

How do you get the wide table back, one row per student again? `widen` is the
inverse, and it reads the same two words the other way. `name` points at the
column holding column names in both verbs. Here it is being read rather than
made, and the verb is what says which.

The table is `marks`, the tall shape the last chapter made:

```{r}
#| echo: false
marks
```

::: {.panel-tabset}

### run

```{r}
run('
  marks
    then widen name [question], value [mark] by [student]
')
```

### R

```{r}
marks |> widen(name = question, value = mark, by = student)
```

### Python

```{python}
marks >> widen(name = col.question, value = col.mark, by = col.student)
```

:::

*`marks then widen name [question], value [mark] by [student]`*

`by` says which columns identify a row. Leaving it out means every column not
already named, which is usually right and is occasionally a surprise: one column
you had forgotten about makes every row distinct, and the table comes back as
tall as it went in. The grammar tells you what it assumed rather than deciding
quietly: a note arrives beside the answer, and [the laws](design-laws.qmd)
show one arriving.

Because the two verbs are inverses and share their defaults, the round trip is a
sentence with no arguments in it at all:

::: {.panel-tabset}

### run

```{r}
run('
  answers
    then lengthen [q1, q2, q3]
    then widen
')
```

### R

```{r}
answers |> lengthen(q1, q2, q3) |> widen()
```

### Python

```{python}
answers >> lengthen(col.q1, col.q2, col.q3) >> widen()
```

:::

## When two rows want the same cell

A cell holds one value. If two rows both claim it, there is no answer, and
choosing one of them silently is how a wrong number reaches a report.

This is the one refusal that arrives when the query runs rather than when the
pipeline is checked. Whether two rows collide is a fact about the data, and the
grammar reads column names rather than rows, so it cannot be known any earlier.

::: {.panel-tabset}

### R

```{r}
#| error: true
twice <- data.frame(
  student  = c("ann", "ann"),
  question = c("q1", "q1"),
  mark     = c(1, 9),
  stringsAsFactors = FALSE
)

collect(twice |> widen(name = question, value = mark, by = student))
```

### Python

```{python}
#| classes: refusal
try:
    twice = pd.DataFrame({
        "student":  ["ann", "ann"],
        "question": ["q1", "q1"],
        "mark":     [1, 9],
    })
    collect(twice
      >> widen(name = col.question, value = col.mark, by = col.student))
except GodError as refusal:
    # This refusal is raised by the engine as the query runs, and the
    # binding hands it over as a `GodError` like every other one. The words
    # are the grammar's, wrapped in the engine's report.
    print(refusal)
```

:::

Saying what should happen is a value rather than a new word. `value` takes an
expression, so an aggregation there is the answer:

::: {.panel-tabset}

### R

```{r}
twice |> widen(name = question, value = average(mark), by = student)
```

### Python

```{python}
(twice
  >> widen(name = col.question, value = average(col.mark),
           by = col.student))
```

:::

## What the columns became

`widen` is the one step whose new columns come from the data rather than from
the sentence, so what the drawing can say about it is worth knowing exactly.

::: {.panel-tabset}

### R

```{r}
marks |> widen(name = question, value = mark, by = student) |> show_steps()
```

### Python

```{python}
show_steps(marks >> widen(name = col.question, value = col.mark, by = col.student))
```

:::

`question` and `mark` leave, `student` stays, and the drawing says nothing at
all about what arrives. That silence is honest rather than a gap: the new column
names are values sitting in the `question` column, and nothing has read a row
yet. It is also why this is the last step here, and what the next section is
about.

## Saying what it makes

The columns `widen` produces come from the data, which nothing can know before
the query runs. Every other step in the grammar is checked against columns that
are known in advance, and this one cannot be.

So a `widen` that says nothing about what it makes is allowed to be the answer,
and is not allowed to be the middle of one:

::: {.panel-tabset}

### R

```{r}
#| error: true
marks |>
  widen(name = question, value = mark, by = student) |>
  take(1) |>
  collect()
```

### Python

```{python}
#| classes: refusal
try:
    collect(marks
      >> widen(name = col.question, value = col.mark, by = col.student)
      >> take(1))
except GodError as refusal:
    print(refusal)
```

:::

Write the columns down and everything after the `widen` is checked as usual:

::: {.panel-tabset}

### run

```{r}
run('
  marks
    then widen name [question], value [mark] by [student] giving [q1, q2]
    then add [gain] as ([q2] - [q1])
')
```

### R

```{r}
marks |>
  widen(name = question, value = mark, by = student, giving = c(q1, q2)) |>
  add(gain = q2 - q1)
```

### Python

```{python}
(marks
 >> widen(name = col.question, value = col.mark, by = col.student,
          giving = [col.q1, col.q2])
 >> add(gain = col.q2 - col.q1))
```

:::

Writing them down buys two more things. A value in the data that you did not list
stops the query instead of disappearing from the answer. And an empty cell can be
filled, which needs `giving` for the same reason: saying what an empty cell holds
means knowing which cells there are.

::: {.panel-tabset}

### R

```{r}
sparse <- data.frame(
  student  = c("ann", "ann", "bob"),
  question = c("q1", "q2", "q1"),
  mark     = c(1, 2, 4),
  stringsAsFactors = FALSE
)

sparse |> widen(name = question, value = mark, by = student,
                missing = 0, giving = c(q1, q2))
```

### Python

```{python}
sparse = pd.DataFrame({
    "student":  ["ann", "ann", "bob"],
    "question": ["q1", "q2", "q1"],
    "mark":     [1, 2, 4],
})

sparse >> widen(name = col.question, value = col.mark, by = col.student,
                missing = 0, giving = [col.q1, col.q2])
```

:::

Say what it makes with `giving`, say what fills the holes with `missing`,
and a sparse table widens with no surprise in it.
