---
title: "Choosing by the shape of a name"
---

{{< include ../_setup.qmd >}}

How do you pick eight score columns without writing eight names? Naming
columns one by one is tedious in a wide table, and it goes stale the moment
a ninth score arrives. `pick where` takes a question about a column's
**name** instead. The `survey` table is the case: six respondents, eight
score columns, and four columns of everything else.

::: {.panel-tabset}

### R

```{r}
survey |> take(2)

survey |> pick(where(startsWith(name, "q")))
```

### Python

```{python}
survey >> take(2)

survey >> pick(where(name.starts("q")))
```


### run

```{r}
run('survey
  then pick where (name starts "q")')
```

:::

*`survey then pick where (name starts "q")`*

`name` is the word for whichever column is being considered. There are three
tests you can ask about it, `starts`, `ends` and `contains`, and they join with
`and`, `or` and `not` like any other condition.

::: {.panel-tabset}

### R

```{r}
survey |> pick(where(endsWith(name, "_score") | name == "respondent"))
```

### Python

```{python}
survey >> pick(where(name.ends("_score") | (name == "respondent")))
```


### run

```{r}
run('survey
  then pick where ((name ends "_score") or (name is "respondent"))')
```

:::

`contains` is the third test, for a mark that sits in the middle of a name:

::: {.panel-tabset}

### R

```{r}
survey |> pick(where(grepl("q", name)))
```

### Python

```{python}
survey >> pick(where(name.contains("q")))
```


### run

```{r}
run('survey
  then pick where (name contains "q")')
```

:::

If no column's name matches, that is refused rather than handing back a table
with no columns at all.

::: {.panel-tabset}

### R

```{r}
#| error: true
collect(survey |> pick(where(startsWith(name, "zzz"))))
```

### Python

```{python}
#| classes: refusal
try:
    collect(survey >> pick(where(name.starts("zzz"))))
except GodError as refusal:
    print(refusal)
```

:::

An empty choice is refused rather than returned, because a table with no
columns answers nothing. The message lists what the table does have, which
is usually the spelling you meant.
