---
title: "Choosing by the shape of a name"
---

{{< include ../_setup.qmd >}}

When a table has thirty columns and you want the eight that begin with `q`,
naming them is tedious and goes stale the moment a ninth arrives. `pick where`
takes a question about a column's **name** instead.

::: {.panel-tabset}

### R

```{r}
survey <- data.frame(
  respondent = 1:3,
  q1_score   = c(4, 5, 3),
  q2_score   = c(2, 5, 5),
  region     = c("West", "East", "West"),
  stringsAsFactors = FALSE
)

survey |> pick(where(startsWith(name, "q")))
```

### Python

```{python}
survey = pd.DataFrame({
    "respondent": [1, 2, 3],
    "q1_score":   [4, 5, 3],
    "q2_score":   [2, 5, 5],
    "region":     ["West", "East", "West"],
})

survey >> pick(where(name.starts("q")))
```

:::

`name` is the word for whichever column is being considered. There are three
tests you can ask about it, `starts`, `ends` and `contains`, and they join with
`and`, `or` and `not` like any other condition.

::: {.panel-tabset}

### R

```{r}
survey |> pick(where(endsWith(name, "_score") | name == "respondent"))
```

### Python

```{python}
survey >> pick(where(name.ends("_score") | (name == "respondent")))
```

:::

`contains` is the third test, for a mark that sits in the middle of a name:

::: {.panel-tabset}

### R

```{r}
survey |> pick(where(grepl("q", name)))
```

### Python

```{python}
survey >> pick(where(name.contains("q")))
```

:::

If no column's name matches, that is refused rather than handing back a table
with no columns at all.

::: {.panel-tabset}

### R

```{r}
#| error: true
collect(survey |> pick(where(startsWith(name, "zzz"))))
```

### Python

```{python}
try:
    collect(survey >> pick(where(name.starts("zzz"))))
except GodError as refusal:
    print(refusal)
```

:::
