---
title: "The shape of a name"
---

{{< include ../_setup.qmd >}}

How do you pick eight score columns without writing eight names? Naming
columns one by one is tedious in a wide table, and the written list goes
stale the moment a ninth score arrives. `pick where` takes a question about
a column's **name** instead. The `survey` table is the example: six
respondents, eight score columns, and four columns of everything else.

::: {.panel-tabset}

### run

```{r}
run('
  survey
    then pick where (name starts "q")
')
```

### R

```{r}
survey |> take(2)

survey |> pick(where(startsWith(name, "q")))
```

### Python

```{python}
survey >> take(2)

survey >> pick(where(name.starts("q")))
```

:::

*`survey then pick where (name starts "q")`*

`name` is the word for whichever column is being considered. There are three
tests of its shape, `starts`, `ends` and `contains`, and they join with
`and`, `or` and `not` like any other condition. `is` still means an exact
match, so one named column can sit in the same question:

::: {.panel-tabset}

### run

```{r}
run('
  survey
    then pick where ((name ends "_score") or (name is "respondent"))
')
```

### R

```{r}
survey |> pick(where(endsWith(name, "_score") | name == "respondent"))
```

### Python

```{python}
survey >> pick(where(name.ends("_score") | (name == "respondent")))
```

:::

`contains` is the third test: it matches the quoted text anywhere in the
name, not only at the start or the end. In `survey` it finds the same
columns as `starts "q"`, because no other name has a `q` in it:

::: {.panel-tabset}

### run

```{r}
run('
  survey
    then pick where (name contains "q")
')
```

### R

```{r}
survey |> pick(where(grepl("q", name)))
```

### Python

```{python}
survey >> pick(where(name.contains("q")))
```

:::

If no column's name matches, the `pick` is refused before anything runs:

::: {.panel-tabset}

### R

```{r}
#| error: true
collect(survey |> pick(where(startsWith(name, "zzz"))))
```

### Python

```{python}
#| classes: refusal
try:
    collect(survey >> pick(where(name.starts("zzz"))))
except GodError as refusal:
    print(refusal)
```

:::

An empty choice is refused rather than returned, because a table with no
columns answers nothing. The message lists what the table does have, and
the spelling you meant is usually in that list.

The shape of a name is only the first question this part asks.
[What a column holds](what-a-column-holds.qmd) asks the next.
