---
title: "Somewhere else"
---

{{< include ../_setup.qmd >}}

Everything so far has read a table sitting in front of you. A warehouse names
its tables in parts instead, as a catalog, then a schema, then the table, and a
pipeline can say so. The dots are part of the name:

::: {.panel-tabset}

### R

```{r}
show_as("shop.orders then sort [revenue] descending then take 5",
        "spark",
        `shop.orders` = data.frame(product = "Widget", revenue = 10))
```

### Python

```{python}
show_as("shop.orders then sort [revenue] descending then take 5",
        "spark", **{"shop.orders": pd.DataFrame({"product": ["Widget"],
                                                 "revenue": [10]})})
```

:::

Each part is quoted on its own, because each part is a name in its own right.
Quoting the whole of `shop.orders` would ask for one table called `shop.orders`,
which is not a table anybody has.

A dot means this and nothing else. It joins the parts of a table's name, in the
four places a table is named: at the head of a pipeline, and in `join`,
`add_rows` and `matching`. Anywhere else it is refused, the way it always was.

## Where the pipeline runs

The sentence does not change. What changes is which engine answers it, and each
language says so in its own way, because the two reach a cluster differently.

In R, hand `use_engine` a connection. Anything `DBI` speaks will do, which
includes `sparklyr` and an `odbc` connection to a warehouse. Tables the
connection already holds are found by name, and nothing is copied up to it. The
connection below is a local database standing in for a remote one, so that this
page runs what it describes:

```{r}
warehouse <- DBI::dbConnect(duckdb::duckdb())
DBI::dbExecute(warehouse, "CREATE SCHEMA shop")
DBI::dbExecute(warehouse,
  "CREATE TABLE shop.orders (region VARCHAR, revenue DOUBLE)")
DBI::dbExecute(warehouse, "INSERT INTO shop.orders VALUES
                             ('West', 100), ('West', 250), ('East', 400)")

use_engine(warehouse)
run("shop.orders then summarize [total] as total([revenue]) by [region]")
```

The table was never in this session. It was described by asking the engine for
its columns and no rows, and the totalling happened where the data already was.

```{r}
use_engine()   # back to the engine on this machine
```

In Python there is nothing to say, because a Spark frame carries its session
with it. Give the verbs Spark tables and the pipeline runs on Spark; give them
pandas tables and it runs here, so `orders >> summarize(total = total(col.revenue),
by = col.region)` picks its engine from `orders`.

The answer comes back in the same currency as the question. Ask with pandas
frames and you get a pandas frame. Ask with Spark frames and you get a Spark
frame, still on the cluster, because bringing the answer of a pipeline over a
warehouse table down to one machine is rarely what you meant.

A pipeline that mixes the two is refused. One query cannot read a table on a
cluster and a table in your session at the same time, and saying so is better
than quietly moving one of them.

## When the table is nowhere

The lookups above are spoken about as if they always succeed. When neither
finds the table, the sentence is refused before any engine is asked, and the
message says how to hand the table over:

::: {.panel-tabset}

### R

```{r}
#| error: true
run('orders then take 1')
```

### Python

```{python}
#| classes: refusal
try:
    run('orders then take 1')
except GodError as refusal:
    print(refusal)
```

:::

Name a table only where a lookup can find it: the scope you are calling
from, or the engine a connection points at.
