---
title: "Sorting and taking"
---

{{< include ../_setup.qmd >}}

Which three orders were the largest of the year? Two verbs answer it
together, and they nearly always travel as a pair: `sort` puts the rows in an
order, and `take` keeps the first few of them.

::: {.panel-tabset}

### R

```{r}
sales |> sort(descending(revenue)) |> take(3)
```

### Python

```{python}
sales >> sort(descending(col.revenue)) >> take(3)
```

:::

*`sales then sort [revenue] descending then take 3`*

`take` on its own is honest but rarely what you meant: it keeps the first
rows *in whatever order the table happens to be in*, which is fine for
peeking at a table and wrong for answering a question. The sort in front is
what turns "some three rows" into "the three largest", and that distinction
returns with force at the end of this chapter.

## Why there is no ascending

`descending` reverses one key, and there is no matching word `ascending`,
because ascending is what happens when you do not ask for anything. A word
that means "do the default" is a second way to say the same thing, and every
second way costs every reader a small doubt: if both spellings exist, do
they differ? In this grammar, if two spellings existed they *would* differ,
so the redundant one is simply absent. This is the smallest example of the
one-way rule you will meet, and the cheapest place to get used to it.

## Several keys

Sorting by several keys reads left to right: the first key decides, the
second breaks its ties, and so on down.

::: {.panel-tabset}

### R

```{r}
sales |> sort(product, descending(revenue))
```

### Python

```{python}
sales >> sort(col.product, descending(col.revenue))
```

:::

`descending` wraps the key it reverses, not the whole sort, so mixed
directions are ordinary: products forward, revenue backward, exactly as
written.

## The latest one of each

Here is the pair doing its best work. Which was the most recent order in
each region? In most tools this question has a name you have to learn
(`slice_max`, `nlargest`, a window function with a filter); here it is the
two verbs you already have, plus `by`:

::: {.panel-tabset}

### R

```{r}
sales |> sort(descending(date)) |> take(1, by = region) |> pick(region, date, product)
```

### Python

```{python}
sales >> sort(descending(col.date)) >> take(1, by = col.region) >> pick(col.region, col.date, col.product)
```

:::

*`sales then sort [date] descending then take 1 by [region] then pick [region, date, product]`*

Sort newest first, take the first row of each region. The dates in this book
are written `2026-10-02` style, and text in that shape sorts correctly as
text, which is why no date machinery was needed; Part III introduces the real
date words for the shapes that do not sort on their own.

## What travels with it

- `descending` belongs to the key it wraps, and only `sort` accepts it.
- `by` after `take` means what `by` means everywhere: do this per group. The
  chapter on the small words returns to it.
- `take` composes with everything that settles an order, and the windowed
  words in Part V (`rank`, `row_number`) lean on `sort` in exactly the way
  `take ... by` does below.

## What it refuses

The first row *of each group* means nothing until something has said first by
what. `take ... by` without a sort in front is refused, not guessed at:

::: {.panel-tabset}

### R

```{r}
#| error: true
sales |> take(1, by = region) |> collect()
```

### Python

```{python}
#| classes: refusal
try:
    collect(sales >> take(1, by = col.region))
except GodError as refusal:
    print(refusal)
```

:::

A bare `take` is allowed and a grouped one is not, and the asymmetry is the
honest one. Peeking at some rows of a table is a real thing to want; "the
first of each group, by no particular order" is not a question, it is a
coin toss dressed as one, and a tool that answered it would be choosing your
answer for you.
