---
title: "Sorting and taking"
---

{{< include ../_setup.qmd >}}

Which three orders were the largest of the year? Two verbs answer it
together, and they nearly always travel as a pair: `sort` puts the rows in an
order, and `take` keeps the first few of them.

::: {.panel-tabset}

### run

```{r}
run('
  sales
    then sort [revenue] descending
    then take 3
')
```

### R

```{r}
sales |> sort(descending(revenue)) |> take(3)
```

### Python

```{python}
sales >> sort(descending(col.revenue)) >> take(3)
```

:::

*`sales then sort [revenue] descending then take 3`*

`take` on its own is honest but rarely what you meant. It keeps the first rows
*in whatever order the table happens to be in*, which is fine for peeking at a
table and wrong for answering a question. The sort in front is
what turns "some three rows" into "the three largest", and that distinction
returns with force at the end of this chapter.

## Why there is no ascending

`descending` reverses one key, and there is no matching word `ascending`,
because ascending is what happens when you do not ask for anything. A word that means "do the default" is a second way to say the same thing, and
every second way costs every reader a small doubt. If both spellings exist, do
they differ? In this grammar, if two spellings existed they *would* differ, so the
redundant one is absent. This is the smallest example of the one-way rule you
will meet, and the cheapest place to get used to it.

## Several keys

Sorting by several keys reads left to right: the first key decides, the
second breaks its ties, and so on down.

::: {.panel-tabset}

### run

```{r}
run('
  sales
    then sort [product], [revenue] descending
')
```

### R

```{r}
sales |> sort(product, descending(revenue))
```

### Python

```{python}
sales >> sort(col.product, descending(col.revenue))
```

:::

`descending` wraps the key it reverses, not the whole sort, so mixed
directions are ordinary: products forward, revenue backward, exactly as
written.

## The latest one of each

Here is the pair doing its best work. Which was the most recent order in
each region? In most tools this question has a name you have to learn: `slice_max`,
`nlargest`, a window function with a filter. Here it is the two verbs you
already have, plus `by`:

::: {.panel-tabset}

### run

```{r}
run('
  sales
    then sort [date] descending
    then take 1 by [region]
    then pick [region, date, product]
')
```

### R

```{r}
sales |>
  sort(descending(date)) |>
  take(1, by = region) |>
  pick(region, date, product)
```

### Python

```{python}
(sales
  >> sort(descending(col.date))
  >> take(1, by = col.region)
  >> pick(col.region, col.date, col.product))
```

:::

*`sales then sort [date] descending then take 1 by [region] then pick [region, date, product]`*

Sort newest first, take the first row of each region. The dates in this book
are written `2026-10-02` style, and text in that shape sorts correctly as
text, which is why no date machinery was needed. [The chapter on
dates](dates.qmd) introduces the real date words for the shapes that do not
sort on their own.

## The other end

`take` gives the rows a pipeline reaches first. `take_last` gives the ones at
the far end, and the pair reads as what they do.

::: {.panel-tabset}

### run

```{r}
run('
  sales
    then sort [revenue]
    then take_last 3
    then pick [product, revenue]
')
```

### R

```{r}
sales |> sort(revenue) |> take_last(3) |> pick(product, revenue)
```

### Python

```{python}
sales >> sort(col.revenue) >> take_last(3) >> pick(col.product, col.revenue)
```

:::

*`sales then sort [revenue] then take_last 3 then pick [product, revenue]`*

The answer comes back in the order the `sort` asked for, smallest first, and
the three rows are the three largest. Sorting the other way and taking the first three gets the same rows in the
opposite order. That is a different table, and usually not the one that was
wanted.

`by` works here as it does on `take`:

::: {.panel-tabset}

### run

```{r}
run('
  sales
    then sort [revenue]
    then take_last 1 by [region]
    then pick [region, product, revenue]
')
```

### R

```{r}
sales |> sort(revenue) |> take_last(1, by = region) |> pick(region, product, revenue)
```

### Python

```{python}
sales >> sort(col.revenue) >> take_last(1, by = col.region) >> pick(col.region, col.product, col.revenue)
```

:::

*`sales then sort [revenue] then take_last 1 by [region] then pick [region, product, revenue]`*

## When the cut falls inside a tie

`take 3` gives three rows. If the third and fourth are level on the sort key,
one of them is in and one is out, and nothing in the table says which. The
engine picked. That is fine when you wanted three rows and wrong when you
wanted the top three *scores*.

`with ties` says the second thing: keep every row level with the last one
taken.

::: {.panel-tabset}

### run

```{r}
scores <- data.frame(
  who    = c("ana", "ben", "cal", "dee"),
  points = c(10, 9, 9, 8)
)

run('
  scores
    then sort [points] descending
    then take 2 with ties
', scores = scores)
```

### R

```{r}
scores |> sort(descending(points)) |> take(2, ties = TRUE)
```

### Python

```{python}
scores = pd.DataFrame({
    "who":    ["ana", "ben", "cal", "dee"],
    "points": [10, 9, 9, 8],
})

scores >> sort(descending(col.points)) >> take(2, ties=True)
```

:::

*`scores then sort [points] descending then take 2 with ties`*

Three rows came back from `take 2`, and that is the point. Once you ask for ties you have stopped asking for a number of rows and started
asking for a cut. The count is no longer something you can read off the
sentence. That
is why it has to be asked for and is never the default.

It needs a `sort` in front of it, and more sharply than the others do: level
with the last row *on what*? With nothing sorted there is no answer, so the
sentence is refused rather than guessed.

::: {.panel-tabset}

### R

```{r}
#| error: true
collect(scores |> take(2, ties = TRUE))
```

### Python

```{python}
#| classes: refusal
try:
    collect(scores >> take(2, ties=True))
except GodError as refusal:
    print(refusal)
```

:::

`take_last` takes it too, and means the same thing at the other end.

If you are arriving from dplyr, this is the one place its default and this one
differ. `slice_max(points, n = 2)` keeps ties unless you tell it not to, and
`take 2` drops them unless you tell it to keep them. Neither is wrong.
Writing it down is what stops the two from quietly disagreeing.

## What travels with it

- `descending` belongs to the key it wraps, and only `sort` accepts it.
- `by` after `take` means what `by` means everywhere: do this per group. The
  chapter on [the small words](words-that-repeat.qmd) returns to it.
- `take` composes with everything that settles an order, and the windowed
  words in [giving each row a place](each-row-a-place.qmd) (`rank`,
  `row_number`) lean on `sort` in exactly the way `take ... by` does below.

## What it refuses

The first row *of each group* means nothing until something has said first by
what. `take ... by` without a sort in front is refused, not guessed at:

::: {.panel-tabset}

### R

```{r}
#| error: true
sales |> take(1, by = region) |> collect()
```

### Python

```{python}
#| classes: refusal
try:
    collect(sales >> take(1, by = col.region))
except GodError as refusal:
    print(refusal)
```

:::

A bare `take` is allowed and a grouped one is not, and the asymmetry is the
honest one. Peeking at some rows of a table is a real thing to want. But
"the first of each group, by no particular order" is not a question: it is
a coin toss dressed as one. A tool that answered it would be choosing your
answer for you.

`take_last` asks for a sort even ungrouped, and the reason is the same one
read from the other side. The rows a pipeline reaches first are at least the rows it reached first. The
rows at the far end are a claim about an end, and a table does not have one
until something says which way it runs.

::: {.panel-tabset}

### R

```{r}
#| error: true
sales |> take_last(3) |> collect()
```

### Python

```{python}
#| classes: refusal
try:
    collect(sales >> take_last(3))
except GodError as refusal:
    print(refusal)
```

:::

The message names `take` as the word that needs no sort. Wanting a few rows to
look at is the common reason to arrive here, and that is the verb for it.
