---
title: "Summarizing by group"
---

{{< include ../_setup.qmd >}}

How much did each product bring in over the year? The question is not about
any row; it is about the rows that belong together. `summarize` collapses
rows into answers: each new column is an aggregation, and `by` names the
columns that say which rows go together.

::: {.panel-tabset}

### run

```{r}
run('
  sales
    then summarize [revenue] as total([revenue]),
         [orders] as row_count() by [product]
')
```

### R

```{r}
sales |>
  summarize(revenue = total(revenue), orders = row_count(), by = product)
```

### Python

```{python}
(sales
  >> summarize(revenue = total(col.revenue), orders = row_count(),
               by = col.product))
```

:::

*`sales then summarize [revenue] as total([revenue]), [orders] as row_count() by [product]`*

Fifteen rows in, four out, one per product. This is the verb that turns a
record of what happened into a statement about it, and it is the single most
common reason anyone opens a table.

## Several groups, or none

Grouping by several columns is a list, and every combination that appears in
the data gets a row:

::: {.panel-tabset}

### run

```{r}
run('
  sales
    then summarize [revenue] as total([revenue]) by [region, product]
')
```

### R

```{r}
sales |> summarize(revenue = total(revenue), by = c(region, product))
```

### Python

```{python}
(sales
  >> summarize(revenue = total(col.revenue),
               by = [col.region, col.product]))
```

:::

Without `by`, the whole table is one group, and the answer is one row:

::: {.panel-tabset}

### run

```{r}
run('
  sales
    then summarize [biggest] as largest([revenue]),
         [average] as average([cost])
')
```

### R

```{r}
sales |> summarize(biggest = largest(revenue), average = average(cost))
```

### Python

```{python}
(sales
  >> summarize(biggest = largest(col.revenue), average = average(col.cost)))
```

:::

## Grouping is an argument, not a state

In this grammar, grouping is a word inside the step it affects. It is not a
mode you switch the table into. Nothing needs ungrouping afterwards, no verb
behaves differently because of what an earlier verb declared, and a pipeline
read a year from now says everything it does, where it does it. dplyr readers
will recognize the disease this avoids: a `group_by()` three steps up
silently changing what `summarise()` and `mutate()` mean, and the ritual
`ungroup()` that follows every recipe like an apology. Here there is nothing
to forget, because there is nothing being remembered.

## One spelling where you have learned five

This one operation is where the world's tools disagree the loudest. Group
and total, five ways:

```
dplyr       df |> group_by(g) |> summarise(total = sum(x))
pandas      df.groupby('g')['x'].sum().reset_index()
polars      df.group_by('g').agg(pl.col('x').sum().alias('total'))
data.table  df[, .(total = sum(x)), by = g]
SQL         SELECT g, SUM(x) AS total FROM df GROUP BY g
```

Every one of these describes the same act, the one Codd's algebra called
aggregation over a partition [@codd1970] and the R world later named
split-apply-combine [@wickham2011]: split the rows by a key, apply a
collapse, combine the answers. The ideas have been stable for fifty years.
The spelling is what churns, and the spelling is what you forget. god's bet
is that `summarize ... by ...`, read aloud, is the spelling you will still
have when the others have faded, because it is the one that says what the
operation does in the words you would use to describe it.

## The ten ways to collapse a group

`total` and `average` are the two you will use most. There are ten
altogether, and the list is closed on purpose: a small shelf you can hold
whole beats a long one you have to search.

`total`, `average` and `median` combine numbers into one, and
`standard_deviation` says how spread out they are: the sample deviation,
which is what R, pandas, polars and both SQL engines all mean by their own
bare word, so the answer does not change with the engine. `smallest` and
`largest` take the ends. `first` and `last` take a value by position, which
is why they want the rows in a known order. `row_count` counts the rows and
takes no column at all. `unique_count` counts how many different values a
column holds.

::: {.panel-tabset}

### run

```{r}
run('
  sales
    then summarize [cheapest] as smallest([cost]),
         [dearest] as largest([cost]), [middle] as median([revenue]),
         [spread] as standard_deviation([revenue]),
         [kinds] as unique_count([product]) by [region]
')
```

### R

```{r}
sales |> summarize(
  cheapest = smallest(cost),
  dearest  = largest(cost),
  middle   = median(revenue),
  spread   = standard_deviation(revenue),
  kinds    = unique_count(product),
  by = region
)
```

### Python

```{python}
sales >> summarize(
  cheapest = smallest(col.cost),
  dearest  = largest(col.cost),
  middle   = median(col.revenue),
  spread   = standard_deviation(col.revenue),
  kinds    = unique_count(col.product),
  by = col.region
)
```

:::

`first` and `last` answer by position, so a sort in front of them decides
which row they mean:

::: {.panel-tabset}

### run

```{r}
run('
  sales
    then sort [revenue] descending
    then summarize [best] as first([product]), [worst] as last([product])
         by [region]
')
```

### R

```{r}
sales |> sort(descending(revenue)) |> summarize(
  best  = first(product),
  worst = last(product),
  by = region
)
```

### Python

```{python}
sales >> sort(descending(col.revenue)) >> summarize(
  best  = first(col.product),
  worst = last(col.product),
  by = col.region
)
```

:::

## The same sentence at scale

Everything above ran on fifteen rows you can check by hand, which is how a
verb should be met. Here is the same verb meeting a table nobody checks by
hand: `gapminder`, the cast's large table, holding a half century of
countries and years.

::: {.panel-tabset}

### run

```{r}
run('
  gapminder
    then keep where ([year] is 2007)
    then summarize [life] as average([life]), [countries] as row_count()
         by [continent]
    then sort [life] descending
')
```

### R

```{r}
gapminder |>
  keep(year == 2007) |>
  summarize(life = average(life), countries = row_count(),
            by = continent) |>
  sort(descending(life))
```

### Python

```{python}
(gapminder
  >> keep(col.year == 2007)
  >> summarize(life = average(col.life), countries = row_count(),
               by = col.continent)
  >> sort(descending(col.life)))
```

:::

*`gapminder then keep where [year] is 2007 then summarize [life] as average([life]), [countries] as row_count() by [continent] then sort [life] descending`*

Gapminder's 1,704 rows went in; five came out, one per continent,
with the count beside each average so the answer carries its own weight of
evidence. The sentence is the same length it was on fifteen rows, and that is
the point of the demonstration: nothing about the grammar knows or cares how
big the table is. [Somewhere else](somewhere-else.qmd)
shows the same indifference reaching a warehouse.

## Where the other columns went

`summarize` keeps the columns you grouped by and the ones you named, and takes
away everything else. That is its whole behavior, and it is the thing readers
are most often surprised by. You can look at it before you run anything.

::: {.panel-tabset}

### R

```{r}
sales |>
  summarize(total = total(revenue), by = region) |>
  show_steps()
```

### Python

```{python}
show_steps(sales >> summarize(total = total(col.revenue), by = col.region))
```

:::

One line makes `total`, and the same line takes `date`, `product`, `quantity`,
`revenue` and `cost` away. When a later step asks for a column that is not
there, this is usually where it went. [What it does to the table](what-it-does.qmd) reads
the drawing in full.

## What travels with it

- `by` is the same word, with the same meaning, that grouped `take` in the
  last chapter and will group windows in [giving each row a
  place](each-row-a-place.qmd): these rows belong together.
- `as` names each answer, name first, exactly as it does in `add`.
- The ten aggregations reappear inside `add ... by`, where they broadcast a
  group's answer back onto its rows instead of collapsing them; the chapter
  on the small words shows the share-of-group idiom this buys.

## What it refuses

Every value in a `summarize` must span its group. A bare column does not,
and the grammar will not guess which of a group's values you meant:

::: {.panel-tabset}

### R

```{r}
#| error: true
sales |> summarize(revenue = revenue, by = region) |> collect()
```

### Python

```{python}
#| classes: refusal
try:
    collect(sales >> summarize(revenue = col.revenue, by = col.region))
except GodError as refusal:
    print(refusal)
```

:::

SQL answers this same mistake with an error about grouping clauses that has
confused beginners for decades; some tools answer it by silently taking the
first value, which is worse. This message names the actual problem, one row
per group against many values per group, and lists the words that resolve
it. Refusing with a direction is the grammar's answer to a whole class of
half-remembered rules.
