---
title: "Summarizing by group"
---

{{< include ../_setup.qmd >}}

`summarize` collapses rows. Each new column is an aggregation, and `by` names
the columns that say which rows go together.

::: {.panel-tabset}

### R

```{r}
sales |> summarize(revenue = total(revenue), orders = row_count(), by = product)
```

### Python

```{python}
sales >> summarize(revenue = total(col.revenue), orders = row_count(), by = col.product)
```

:::

Grouping by several columns is a list.

::: {.panel-tabset}

### R

```{r}
sales |> summarize(revenue = total(revenue), by = c(region, product))
```

### Python

```{python}
sales >> summarize(revenue = total(col.revenue), by = [col.region, col.product])
```

:::

Without `by`, the whole table is one group.

::: {.panel-tabset}

### R

```{r}
sales |> summarize(biggest = largest(revenue), average = average(cost))
```

### Python

```{python}
sales >> summarize(biggest = largest(col.revenue), average = average(col.cost))
```

:::

Grouping is an argument rather than a step, so there is nothing to ungroup
afterwards and no state to carry between verbs.

## The nine ways to collapse a group

`total` and `average` are the two you will reach for most. There are nine
altogether, and the list is closed.

`total`, `average` and `median` are the three that combine numbers into one.
`smallest` and `largest` take the ends. `first` and `last` take a value by
position, which is why they need the rows to be in a known order. `row_count`
counts the rows and takes no column at all. `unique_count` counts how many
different values a column holds.

::: {.panel-tabset}

### R

```{r}
sales |> summarize(
  cheapest = smallest(cost),
  dearest  = largest(cost),
  middle   = median(revenue),
  kinds    = unique_count(product),
  by = region
)
```

### Python

```{python}
sales >> summarize(
  cheapest = smallest(col.cost),
  dearest  = largest(col.cost),
  middle   = median(col.revenue),
  kinds    = unique_count(col.product),
  by = col.region
)
```

:::

`first` and `last` answer by position, so a sort in front of them is what decides
which row they mean. Without one they still answer, and the answer is whatever
order the rows arrived in.

::: {.panel-tabset}

### R

```{r}
sales |> sort(descending(revenue)) |> summarize(
  best  = first(product),
  worst = last(product),
  by = region
)
```

### Python

```{python}
sales >> sort(descending(col.revenue)) >> summarize(
  best  = first(col.product),
  worst = last(col.product),
  by = col.region
)
```

:::
