---
title: "The total so far"
---

{{< include ../_setup.qmd >}}

How much had accumulated by each day? `running_total` adds a column up as it
goes down the rows, so each row carries the total of everything to that
point.

::: {.panel-tabset}

### run

```{r}
run('
  diary
    then sort [on_]
    then add [so_far] as running_total([x])
')
```

### R

```{r}
diary |> sort(on_) |> add(so_far = running_total(x))
```

### Python

```{python}
diary >> sort(col.on_) >> add(so_far = running_total(col.x))
```

:::

*`diary then sort [on_] then add [so_far] as running_total([x])`*

`by` restarts it for each group, and `previous` and `following` look back and
on by one row:

::: {.panel-tabset}

### run

```{r}
run('
  diary
    then sort [on_]
    then add [so_far] as running_total([x]), [before] as previous([x]),
         [after] as following([x]) by [g]
')
```

### R

```{r}
diary |>
  sort(on_) |>
  add(so_far = running_total(x), before = previous(x),
      after = following(x), by = g)
```

### Python

```{python}
(diary
 >> sort(col.on_)
 >> add(so_far = running_total(col.x), before = previous(col.x),
        after = following(col.x), by = col.g))
```

:::

The first row of a group has nothing before it and the last has nothing after
it, so those are missing.

One row is the usual distance and not the only one. Say how far as a second
argument, which is how you compare a month against the same month a year
earlier:

::: {.panel-tabset}

### run

```{r}
run('
  diary
    then sort [on_]
    then add [two_back] as previous([x], 2), [two_on] as following([x], 2)
    then pick [on_, x, two_back, two_on]
')
```

### R

```{r}
diary |>
  sort(on_) |>
  add(two_back = previous(x, 2), two_on = following(x, 2)) |>
  pick(on_, x, two_back, two_on)
```

### Python

```{python}
(diary
 >> sort(col.on_)
 >> add(two_back = previous(col.x, 2), two_on = following(col.x, 2))
 >> pick(col.on_, col.x, col.two_back, col.two_on))
```

:::

How far is written out rather than worked out. A distance that changed from
row to row would be a different question, and no engine underneath answers it
here, so a column in that place is refused. Nought is refused too, because
that is the column itself, and a negative is refused with the name of the word
that goes the other way.

::: {.panel-tabset}

### R

```{r}
#| error: true
collect(diary |> sort(on_) |> add(back = previous(x, -2)))
```

### Python

```{python}
#| classes: refusal
try:
    collect(diary >> sort(col.on_) >> add(back = previous(col.x, -2)))
except GodError as refusal:
    print(refusal)
```

:::

All three need a `sort` in front of them, and say so if they do not get one. A
total *so far* means nothing until something has said so far in what order:

::: {.panel-tabset}

### R

```{r}
#| error: true
collect(diary |> add(so_far = running_total(x)))
```

### Python

```{python}
#| classes: refusal
try:
    collect(diary >> add(so_far = running_total(col.x)))
except GodError as refusal:
    print(refusal)
```

:::

`rank` is the exception, because its argument *is* the order: `rank(x)` says what
it goes by, so it needs no sort in front of it. The three words this chapter
added never get that shortcut: a total so far, the value before, the value
after, and all three begin with `sort` saying in what order the rows are
read.
