---
title: "The total so far"
---

{{< include ../_setup.qmd >}}

How much had accumulated by each day? `running_total` adds a column up as it
goes down the rows, so each row carries the total of everything to that
point.

::: {.panel-tabset}

### run

```{r}
run('
  diary
    then sort [on_]
    then add [so_far] as running_total([x])
')
```

### R

```{r}
diary |> sort(on_) |> add(so_far = running_total(x))
```

### Python

```{python}
diary >> sort(col.on_) >> add(so_far = running_total(col.x))
```

:::

*`diary then sort [on_] then add [so_far] as running_total([x])`*

`by` restarts it for each group, and `previous` and `following` look back and
on by one row:

::: {.panel-tabset}

### run

```{r}
run('
  diary
    then sort [on_]
    then add [so_far] as running_total([x]), [before] as previous([x]),
         [after] as following([x]) by [g]
')
```

### R

```{r}
diary |>
  sort(on_) |>
  add(so_far = running_total(x), before = previous(x),
      after = following(x), by = g)
```

### Python

```{python}
(diary
 >> sort(col.on_)
 >> add(so_far = running_total(col.x), before = previous(col.x),
        after = following(col.x), by = col.g))
```

:::

The first row of a group has nothing before it and the last has nothing after
it, so those are missing.

One row is the usual distance and not the only one. Say how far as a second
argument, which is how you compare a month against the same month a year
earlier:

::: {.panel-tabset}

### run

```{r}
run('
  diary
    then sort [on_]
    then add [two_back] as previous([x], 2), [two_on] as following([x], 2)
    then pick [on_, x, two_back, two_on]
')
```

### R

```{r}
diary |>
  sort(on_) |>
  add(two_back = previous(x, 2), two_on = following(x, 2)) |>
  pick(on_, x, two_back, two_on)
```

### Python

```{python}
(diary
 >> sort(col.on_)
 >> add(two_back = previous(col.x, 2), two_on = following(col.x, 2))
 >> pick(col.on_, col.x, col.two_back, col.two_on))
```

:::

How far is written out rather than worked out. A distance that changed from
row to row would be a different question, and no engine underneath answers it
here, so a column in that place is refused. Nought is refused too, because
that is the column itself, and a negative is refused with the name of the word
that goes the other way.

::: {.panel-tabset}

### R

```{r}
#| error: true
collect(diary |> sort(on_) |> add(back = previous(x, -2)))
```

### Python

```{python}
#| classes: refusal
try:
    collect(diary >> sort(col.on_) >> add(back = previous(col.x, -2)))
except GodError as refusal:
    print(refusal)
```

:::

`rolling` asks an aggregate of the last few rows and answers on every row, so
a moving average is one word wrapping another. The aggregate goes inside, one
of `total`, `average`, `median`, `smallest`, `largest` and
`standard_deviation`, and the number says how many rows the window holds, the
row itself included:

::: {.panel-tabset}

### run

```{r}
run('
  diary
    then sort [on_]
    then add [pair] as rolling(average([x]), 2)
    then pick [on_, x, pair]
')
```

### R

```{r}
diary |>
  sort(on_) |>
  add(pair = rolling(average(x), 2)) |>
  pick(on_, x, pair)
```

### Python

```{python}
(diary
 >> sort(col.on_)
 >> add(pair = rolling(average(col.x), 2))
 >> pick(col.on_, col.x, col.pair))
```

:::

The first row is missing, and that is the answer rather than a gap: one row is
not the two-row average, it is a different question wearing its name. A window
that is not yet full answers `missing`, and so does a full window with a
missing value inside it, which is how every library with a rolling word
answers too. `by` restarts the window inside each group, exactly as it does
for the totals above, and the aggregate inside reads a plain column: a
computed value is one `add` away, in a step of its own.

Every word here needs a `sort` in front of it, and says so if it does not get
one. A total *so far* means nothing until something has said so far in what
order:

::: {.panel-tabset}

### R

```{r}
#| error: true
collect(diary |> add(so_far = running_total(x)))
```

### Python

```{python}
#| classes: refusal
try:
    collect(diary >> add(so_far = running_total(col.x)))
except GodError as refusal:
    print(refusal)
```

:::

`rank` is the exception, because its argument *is* the order: `rank(x)` says what
it goes by, so it needs no sort in front of it. The words this chapter added
never get that shortcut. A total so far, the value before, the value after,
an aggregate of the last few rows: every one begins with `sort` saying in
what order the rows are read.
