---
title: "The total so far"
---

{{< include ../_setup.qmd >}}

How much had accumulated by each day? `running_total` adds a column up as it
goes down the rows, so each row carries the total of everything to that
point.

::: {.panel-tabset}

### R

```{r}
diary |> sort(on_) |> add(so_far = running_total(x))
```

### Python

```{python}
diary >> sort(col.on_) >> add(so_far = running_total(col.x))
```

:::

*`diary then sort [on_] then add [so_far] as running_total([x])`*

`by` restarts it for each group, and `previous` and `following` look one row back
and one row on:

::: {.panel-tabset}

### R

```{r}
diary |>
  sort(on_) |>
  add(so_far = running_total(x), before = previous(x), after = following(x), by = g)
```

### Python

```{python}
(diary
 >> sort(col.on_)
 >> add(so_far = running_total(col.x), before = previous(col.x),
        after = following(col.x), by = col.g))
```

:::

The first row of a group has nothing before it and the last has nothing after
it, so those are missing.

All three need a `sort` in front of them, and say so if they do not get one. A
total *so far* means nothing until something has said so far in what order:

::: {.panel-tabset}

### R

```{r}
#| error: true
collect(diary |> add(so_far = running_total(x)))
```

### Python

```{python}
try:
    collect(diary >> add(so_far = running_total(col.x)))
except GodError as refusal:
    print(refusal)
```

:::

`rank` is the exception, because its argument *is* the order: `rank(x)` says what
it goes by, so it needs no sort in front of it.
