---
title: "Tidying text"
---

{{< include ../_setup.qmd >}}

What do you do with `"  ann marie  "`? Real tables arrive untidy, and these
are the words for that.

::: {.panel-tabset}

### R

```{r}
messy <- data.frame(
  raw = c("  ann marie  ", "  bob  "),
  n = c(7L, 99L),
  stringsAsFactors = FALSE
)

messy |>
  add(name = trim(raw)) |>
  add(first = split_text(name, " ", 1),
      size = characters(name),
      fixed = replace_text(name, "a", "A"))
```

### Python

```{python}
messy = pd.DataFrame({
    "raw": ["  ann marie  ", "  bob  "],
    "n": [7, 99],
})

(messy
 >> add(name = trim(col.raw))
 >> add(first = split_text(col.name, " ", 1),
        size = characters(col.name),
        fixed = replace_text(col.name, "a", "A")))
```

:::

*`messy then add [name] as trim([raw]) then add [first] as split_text([name], " ", 1), [size] as characters([name]), [fixed] as replace_text([name], "a", "A")`*

`split_text` says which piece it wants, counting from 1, because every value in
the grammar is one value. Where there is no such piece the answer is empty text
rather than missing, so `bob` gives back `bob`.

`replace_text` looks for the text itself, not for a pattern, so nothing in it is
special. `characters` counts characters; it is not called `length`, because R's
`length` counts the elements of a vector and a word that reads as one thing and
does another is the trap this vocabulary is built to avoid.

## Case, both ways

`lower` and `upper` fold a value's case. They are ordinary functions on a value,
so they go anywhere a value goes.

::: {.panel-tabset}

### R

```{r}
messy |> add(shout = upper(trim(raw)), quiet = lower(trim(raw))) |> pick(shout, quiet)
```

### Python

```{python}
messy >> add(shout = upper(trim(col.raw)), quiet = lower(trim(col.raw))) >> pick(col.shout, col.quiet)
```

:::

The same two words also work on a column's *name* rather than on its contents,
which is how you match names whose capitalization you cannot rely on. That is
[choosing by what a column holds](what-a-column-holds.qmd), and it needs no
second spelling of either word.

## A number is not text

The repairs read text, and a number is not text, however much it looks like
one in a printout. Ask anyway and the refusal names the conversion, which is
[the next chapter](converting-kinds.qmd)'s subject:

::: {.panel-tabset}

### R

```{r}
#| error: true
collect(messy |> add(t = trim(n)))
```

### Python

```{python}
#| classes: refusal
try:
    collect(messy >> add(t = trim(col.n)))
except GodError as refusal:
    print(refusal)
```

:::

One kind goes in and the declared kind comes out, with nothing silently
coerced in between.
