---
title: "Tidying text"
---

{{< include ../_setup.qmd >}}

What do you do with `"  ann marie  "`? Real tables arrive untidy, and these
are the words for that.

The table is `messy`, two names with the whitespace still on them:

```{r}
#| echo: false
messy
```

::: {.panel-tabset}

### run

```{r}
run('
  messy
    then add [name] as trim([raw])
    then add [first] as split_text([name], " ", 1),
             [size] as characters([name]),
             [fixed] as replace_text([name], "a", "A")
')
```

### R

```{r}
messy |>
  add(name = trim(raw)) |>
  add(first = split_text(name, " ", 1),
      size = characters(name),
      fixed = replace_text(name, "a", "A"))
```

### Python

```{python}
(messy
 >> add(name = trim(col.raw))
 >> add(first = split_text(col.name, " ", 1),
        size = characters(col.name),
        fixed = replace_text(col.name, "a", "A")))
```

:::

*`messy then add [name] as trim([raw]) then add [first] as split_text([name], " ", 1), [size] as characters([name]), [fixed] as replace_text([name], "a", "A")`*

`split_text` says which piece it wants, counting from 1, because every value in
the grammar is one value. A value with no separator is all one piece, so asking
`bob` for piece 1 gives back `bob`. Asking it for piece 2 finds no such piece,
and the answer is empty text rather than missing.

`replace_text` looks for the text itself, not for a pattern, so no character
in it has a special meaning. `characters` counts characters; it is not called
`length`, because R's `length` counts the elements of a vector. A word that
reads as one thing and does another is the trap this vocabulary is built to
avoid.

## Putting text back together

`join_text` is `split_text` read the other way. It joins two or more values into
one, and a separator is written where it goes, as a value, rather than being set
somewhere else.

::: {.panel-tabset}

### run

```{r}
run('
  messy
    then add [name] as trim([raw])
    then add [greeting] as join_text("hello ", [name])
    then pick [greeting, n]
')
```

### R

```{r}
messy |>
  add(name = trim(raw)) |>
  add(greeting = join_text("hello ", name)) |>
  pick(greeting, n)
```

### Python

```{python}
(messy
  >> add(name = trim(col.raw))
  >> add(greeting = join_text("hello ", col.name))
  >> pick(col.greeting, col.n))
```

:::

*`messy then add [name] as trim([raw]) then add [greeting] as join_text("hello ", [name]) then pick [greeting, n]`*

Numbers are refused rather than converted. `join_text([raw], [n])` will not
run, and the message names `to_text`, because how a number should look is a
decision you make rather than one the grammar makes for you.

::: {.panel-tabset}

### R

```{r}
#| error: true
messy |> add(both = join_text(raw, n))
```

### Python

```{python}
#| classes: refusal
try:
    collect(messy >> add(both = join_text(col.raw, col.n)))
except GodError as refusal:
    print(refusal)
```

:::

A missing value anywhere makes the whole answer missing. That is the rule that
addition already follows. It is the one engines disagree about most quietly:
several of them drop the absent part and hand back the rest. So a label built
from a name nobody recorded comes back looking finished. If you would rather
fill the hole than lose the row, say what to fill it with:
`join_text([first], " ", first_present([last], ""))`.

## A whole group, run together

`join_text` reaches across the columns of one row. `join_rows` reaches down the
rows of a group, and hands back one piece of text for all of them. The names
carry the direction, because that is the real difference between them.

It is an aggregate, so it belongs in `summarize` beside `total` and `row_count`:

::: {.panel-tabset}

### run

```{r}
run('
  sales
    then sort [product]
    then summarize [products] as join_rows([product], ", ") by [region]
')
```

### R

```{r}
sales |>
  sort(product) |>
  summarize(products = join_rows(product, ", "), by = region)
```

### Python

```{python}
(sales
  >> sort(col.product)
  >> summarize(products = join_rows(col.product, ", "), by = col.region))
```

:::

*`sales then sort [product] then summarize [products] as join_rows([product], ", ") by [region]`*

The `sort` is what makes the answer the same every time. `join_rows` joins the
rows in the order they stand, and without a sort that order is whatever the
engine had them in. `first` and `last` make the same promise.

A missing value is skipped here, where `join_text` would have made the whole
answer missing. The two words answer differently on purpose. An aggregate
reduces the rows it is handed, and `total` already ignores the absent ones: it
does not come back missing because a single row was empty. A scalar
builds one value out of parts you named, and quietly dropping one would hand
back a label that looks finished.

The separator is written out rather than read from a column. The rows are being
collapsed into one answer, so a separator that varied by row would have no row
left to come from.

## Case, both ways

`lower` and `upper` fold a value's case. They are ordinary functions on a value,
so they go anywhere a value goes.

::: {.panel-tabset}

### run

```{r}
run('
  messy
    then add [shout] as upper(trim([raw])), [quiet] as lower(trim([raw]))
    then pick [shout, quiet]
')
```

### R

```{r}
messy |>
  add(shout = upper(trim(raw)), quiet = lower(trim(raw))) |>
  pick(shout, quiet)
```

### Python

```{python}
(messy
  >> add(shout = upper(trim(col.raw)), quiet = lower(trim(col.raw)))
  >> pick(col.shout, col.quiet))
```

:::

The same two words also work on a column's *name* rather than on its contents,
which is how you match names whose capitalization you cannot rely on. That is
shown in [What a column holds](what-a-column-holds.qmd), and it needs no
second spelling of either word.

## A number is not text

The repairs read text, and a number is not text, however much it looks like
one in a printout. Ask anyway and the refusal names the conversion, which is
[the next chapter](converting-kinds.qmd)'s subject:

::: {.panel-tabset}

### R

```{r}
#| error: true
collect(messy |> add(t = trim(n)))
```

### Python

```{python}
#| classes: refusal
try:
    collect(messy >> add(t = trim(col.n)))
except GodError as refusal:
    print(refusal)
```

:::

One kind goes in and the declared kind comes out, with nothing silently
coerced in between.
