---
title: "What it does to the table"
---

{{< include ../_setup.qmd >}}

A pipeline is a plan, and a plan can be looked at. The previous chapters showed
two ways to look at one: the sentence it wrote, and the query it becomes.
`show_steps` shows a third thing, which is what the table itself is doing.

::: {.column-body-outset}
::: {.panel-tabset}

### R

```{r}
sales |>
  add(margin = revenue - cost) |>
  keep(margin > 50) |>
  take(3) |>
  show_steps()
```

### Python

```{python}
show_steps(
  sales
  >> add(margin = col.revenue - col.cost)
  >> keep(col.margin > 50)
  >> take(3)
)
```

:::
:::

Nothing ran. The grammar reads a whole sentence against the table's columns
before any of it executes, which is the rule the rest of this part is built on,
and the drawing is that reading rather than a second one.

## What the marks mean

Read a drawing downward. The first band is the table you start from, and every
band after it is one step, showing the table as it stands once that step has
run.

Beside each step is what the table holds at that point, one name per column. A
column the step **makes** carries a `+`. A column the step **takes away**
carries a `-`, drawn on a line under the names that stayed, against the step
that removed it, not later. Everything else was already there and still is.

**The width is the point.** A step's columns are drawn as one block, so the
block is as wide as the table is. A `summarize` that takes five columns away
draws a third of the width of the one above it. A `join` draws one more
column than the table had. You can see what a step did to the table before
reading what it was.

A wide table wraps rather than running off the page, and the columns that left
wrap the same way. Nothing is counted or hidden, so a table with forty columns
is drawn with forty columns.

The type is written beside a column where you have not been told it yet. On
the table you start from it stands beside every column; on a column a step
has just made, it arrives with the `+`. Repeating it everywhere would double
the width of the drawing to say nothing new.

## Where a column went

The question this answers most often is where a column has gone. `summarize`
is where they usually go.

::: {.column-body-outset}
::: {.panel-tabset}

### R

```{r}
sales |>
  summarize(gross = total(revenue), by = region) |>
  show_steps()
```

### Python

```{python}
show_steps(sales >> summarize(gross = total(col.revenue), by = col.region))
```

:::
:::

One step makes `gross`, and the same step takes away everything that was not
grouped by. Here you can see it happen instead of finding out later.

The note underneath says what became of the rows. It appears only where you
could not have guessed: that `keep` returns fewer rows surprises nobody, so
no line says it, but `summarize` collapsing a table to one row per group gets
one.

## A second table

A table that arrives partway down gets a line of its own, indented under the
step that reads it, with its columns beside its name.

::: {.column-body-outset}
::: {.panel-tabset}

### R

```{r}
sales |> join(products, by = product) |> show_steps()
```

### Python

```{python}
show_steps(sales >> join(products, by = col.product))
```

:::
:::

Four things show here. The column both tables matched on carries an `=`, and
it carries it **twice**, once in each table's row: it is on both sides and it
arrives once, not as two columns with invented names. The column that crossed
over carries a `+`, the same mark any new column gets. One line says every
row of this table is kept: a `sales` row with no match in `products`
survives, with `maker` missing. And the last line says the rows may multiply,
because they may: whether `products` holds a product name more than once is a
fact about the data, and the grammar has read the column names without
reading a single row.

## What a second table sends

Not every second table sends columns. `matching` reads one and brings nothing
back.

::: {.column-body-outset}
::: {.panel-tabset}

### R

```{r}
sales |> keep(matching(products, by = product)) |> show_steps()
```

### Python

```{python}
show_steps(sales >> keep(matching(products, by = col.product)))
```

:::
:::

Compare the last line with the join above. A join *may multiply* the rows; a
`matching` never can, because it only decides which rows to keep. The two
sentences that ask for them read almost the same and mean different things,
and confusing them is the most common mistake anyone makes with two tables.
The drawing says which one you wrote.

`add_rows` is the third case, and it sends rows rather than columns: the two
tables already hold the same columns, so nothing arrives and the table only
gets longer.

## A sentence that will not run

A sentence the grammar refuses is still drawn, as far as it checked.

::: {.column-body-outset}
::: {.panel-tabset}

### R

```{r}
sales |>
  summarize(gross = total(revenue), by = region) |>
  keep(product == "Widget") |>
  show_steps()
```

### Python

```{python}
show_steps(
  sales
  >> summarize(gross = total(col.revenue), by = col.region)
  >> keep(col.product == "Widget")
)
```

:::
:::

The refusal says `product` is not a column, and it is right: one step
earlier, `summarize` took it away. The message alone tells you the column is
missing. The drawing tells you where it went, which is the part you need in
order to fix it.

This is the one place `show_steps` differs from every other way of looking at a
pipeline. It draws rather than raising, so a sentence you are still writing can
be looked at while it is wrong.

## The drawing's two forms

The drawing has two forms and you do not have to choose between them. At a
console it prints as text, because that is what the rest of a session looks
like. In a notebook, or on a page like this one, it draws itself, because a
page can hold a picture.

Ask for one outright when you want to keep it, with `format` in R and with
`.text` and `.svg` in Python.

::: {.panel-tabset}

### R

```{r}
cat(format(show_steps(sales |> take(2)), "text"))
```

### Python

```{python}
print(show_steps(sales >> take(2)).text)
```

:::

The picture is one file with its own colors inside it, so saving it and sending
it to somebody gives them the drawing you saw. `format(x, "svg")` in R, and
`.svg` in Python, hand you that file as text.

That is the whole drawing: the marks, the notes, and the two forms it takes.
The refusal's own words are [the next chapter](every-refusal.qmd).
