---
title: "What it does to the table"
---

{{< include ../_setup.qmd >}}

A pipeline is a plan, and a plan can be looked at. The previous chapters showed
two ways to look at one: the sentence it wrote, and the query it becomes.
`show_steps` shows a third thing, which is what the table itself is doing.

::: {.column-body-outset}
::: {.panel-tabset}

### R

```{r}
sales |>
  add(margin = revenue - cost) |>
  keep(margin > 50) |>
  take(3) |>
  show_steps()
```

### Python

```{python}
show_steps(
  sales
  >> add(margin = col.revenue - col.cost)
  >> keep(col.margin > 50)
  >> take(3)
)
```

:::
:::

Nothing ran. The grammar reads a whole sentence against the table's columns
before any of it executes, which is the rule the rest of this part is built on,
and the drawing is that reading rather than a second one.

## What the marks mean

Read a drawing downward. The first band is the table you start from, and every
band after it is one step, showing the table as it stands once that step has
run.

Beside each step is what the table holds at that point, one mark per column. A
column the step **makes** carries a `+`. A column the step **takes away**
carries a `-`, and it appears against the step that removed it, not later.
Everything else was already there and still is.

**The width is the point.** A step's columns are drawn as one block, so the
block is as wide as the table is: a `summarize` that takes five columns away
draws a third of the width of the one above it, and a `join` draws more. You can
see what a step did to the table before reading what it was.

A wide table wraps rather than running off the page, and the columns that left
wrap the same way. Nothing is counted or hidden, so a table with forty columns
is drawn with forty columns.

The type is written beside a column in two places only: on the table you start
from, and on a column a step has just made. Those are the two places you have
not been told yet. Repeating it everywhere would double the width of the drawing
to say nothing new.

## Where a column went

The question this answers most often is where a column has gone. `summarize`
is where they usually go.

::: {.column-body-outset}
::: {.panel-tabset}

### R

```{r}
sales |>
  summarize(gross = total(revenue), by = region) |>
  show_steps()
```

### Python

```{python}
show_steps(sales >> summarize(gross = total(col.revenue), by = col.region))
```

:::
:::

One line makes `gross`, and the same line takes away everything that was not
grouped by. That is `summarize`'s whole behavior, and it is the thing readers
are most often surprised by. Here you can see it happen instead of finding out
later.

The note underneath says what became of the rows. It appears only where you
could not have guessed: that `keep` returns fewer rows is not news, so no line
says it, but `summarize` collapsing a table to one row per group is worth
saying once.

## A second table

A table that arrives partway down gets a line of its own, indented under the
step that reads it.

::: {.column-body-outset}
::: {.panel-tabset}

### R

```{r}
sales |> join(products, by = product) |> show_steps()
```

### Python

```{python}
show_steps(sales >> join(products, by = col.product))
```

:::
:::

Three things are worth reading here. The column both tables matched on carries
an `=`, and it carries it **twice**, once in each table's row: it is on both
sides and it arrives once, not as two columns with invented names. The column
that crossed over carries a `+`, the same mark any new column gets. And the
last line says the rows may multiply, because they may: whether `products`
holds a product name more than once is a fact about the data, and the grammar
has read the column names without reading a single row.

## What a second table sends

Not every second table sends columns. `matching` reads one and brings nothing
back.

::: {.column-body-outset}
::: {.panel-tabset}

### R

```{r}
sales |> keep(matching(products, by = product)) |> show_steps()
```

### Python

```{python}
show_steps(sales >> keep(matching(products, by = col.product)))
```

:::
:::

Compare the last line with the join above. A join *may multiply* the rows; a
`matching` never can, because it only decides which rows to keep. Those two
sentences read almost the same and mean different things, and confusing them is
the most common mistake anyone makes with two tables. The drawing says which
one you wrote.

`add_rows` is the third case, and it sends rows rather than columns: the two
tables already hold the same columns, so nothing arrives and the table only
gets longer.

## A sentence that will not run

A sentence the grammar refuses is still drawn, as far as it checked.

::: {.column-body-outset}
::: {.panel-tabset}

### R

```{r}
sales |>
  summarize(gross = total(revenue), by = region) |>
  keep(product == "Widget") |>
  show_steps()
```

### Python

```{python}
show_steps(
  sales
  >> summarize(gross = total(col.revenue), by = col.region)
  >> keep(col.product == "Widget")
)
```

:::
:::

The refusal says `product` is not a column, and it is right: two lines above,
`summarize` took it away. The message alone tells you the column is missing.
The drawing tells you where it went, which is the part you need in order to fix
it.

This is the one place `show_steps` differs from every other way of looking at a
pipeline. It draws rather than raising, so a sentence you are still writing can
be looked at while it is wrong.

## At a console and on a page

The drawing has two forms and you do not choose between them. At a console it
prints as text, because that is what the rest of a session looks like. In a
notebook, or on a page like this one, it draws itself, because a page can hold
a picture.

Ask for one outright when you want to keep it, with `format` in R and the
matching name in Python.

::: {.panel-tabset}

### R

```{r}
cat(format(show_steps(sales |> take(2)), "text"))
```

### Python

```{python}
print(show_steps(sales >> take(2)).text)
```

:::

The picture is one file with its own colors inside it, so saving it and sending
it to somebody gives them the drawing you saw. `format(x, "svg")` in R, and
`.svg` in Python, hand you that file as text.
