---
title: "What a column holds"
---

{{< include ../_setup.qmd >}}

```{r}
#| include: false
mixed <- god_table("mixed")
notes <- god_table("notes")
```

```{python}
#| include: false
mixed = god_table("mixed")
notes = god_table("notes")
```

What kind of thing does a column hold, and can you choose by that alone? It
is the other question you can ask about a column without looking at a single
row. `kind` is one of `"number"`, `"text"`, `"truth"` or `"date"`. A
timestamp column answers as a date, because a timestamp is a date carrying a
time.

::: {.panel-tabset}

### run

```{r}
run('
  survey
    then pick where (kind is "number")
')
```

### R

```{r}
survey |> pick(where(kind == "number"))
```

### Python

```{python}
survey >> pick(where(kind == "number"))
```

:::

*`survey then pick where (kind is "number")`*

Both questions live in the same `where`, so they join.

::: {.panel-tabset}

### run

```{r}
run('
  survey
    then pick where ((kind is "number") and (name starts "q"))
')
```

### R

```{r}
survey |> pick(where(kind == "number" & startsWith(name, "q")))
```

### Python

```{python}
survey >> pick(where((kind == "number") & name.starts("q")))
```

:::

Put it with `summarize` and you have the sentence people reach for most: average
every number, whatever the numbers happen to be called.

::: {.panel-tabset}

### run

```{r}
run('
  survey
    then summarize where (kind is "number") as average(value)
')
```

### R

```{r}
survey |> summarize(where(kind == "number", average(value)))
```

### Python

```{python}
survey >> summarize(where(kind == "number", average(value)))
```

:::

That one line does not name a column at all, so it keeps working when the
table gains one. Inside the calculation, `value` stands for each column the
condition matched, and [the next chapter](one-value-per-column.qmd) is built
on that word.

## Case, and the two words that handle it

The name tests match exactly, so `name starts "q"` does not find a column called
`Q1_score`. Mixed-case names are ordinary enough that you meet this early.

There is no flag for it, because a name is text and text has a case. Fold the
case and ask the question.

`mixed` names one column `Q1_score` and the next `q2_score`:

```{r}
#| echo: false
mixed
```

::: {.panel-tabset}

### run

```{r}
run('
  mixed
    then pick where lower(name) starts "q"
')
```

### R

```{r}
mixed |> pick(where(startsWith(tolower(name), "q")))
```

### Python

```{python}
mixed >> pick(where(lower(name).starts("q")))
```

:::

*`mixed then pick where lower(name) starts "q"`*

`lower` and `upper` are ordinary words, so they work on a value in exactly the
same way.

::: {.panel-tabset}

### run

```{r}
run('
  mixed
    then keep where (lower([Region]) is "west")
')
```

### R

```{r}
mixed |> keep(tolower(Region) == "west")
```

### Python

```{python}
mixed >> keep(lower(col.Region) == "west")
```

:::

Folding the case of every name automatically would have been the other option,
and it is wrong. Two columns can differ only by case, and a grammar that
pretends otherwise cannot treat them as the two columns they are. Asking for
it is one word.

## Why you have to write `name`

The same three words also test what is *inside* a column, and there the subject
is the column itself.

::: {.panel-tabset}

### run

```{r}
run('
  survey
    then keep where ([region] starts "W")
')
```

### R

```{r}
survey |> keep(startsWith(region, "W"))
```

### Python

```{python}
survey >> keep(col.region.starts("W"))
```

:::

So `starts` means one thing, "this text begins with that text", and the subject
says what text is being asked about. Writing it is what keeps the two apart.

The alternative would have been `pick(starting("q"))`, which is shorter and
leaves nothing in the sentence to say whether `starting` is looking at a name or
at a value. You would learn it from the position, which is the kind of rule this
grammar removes rather than adds.

The tests are the same words in both languages. What differs is how a column's
name is reached, which is the same difference as everywhere else. R writes a
bare name and Python writes `col.name`, so R spells this with base R's
`startsWith` and Python with a method.

## A percent sign is a percent sign

The three tests read text literally, with no pattern language hiding inside.
R borrows `grepl`'s shape to ask the question, and the reading is still the
grammar's: the marks that pattern languages treat as instructions, `%` and
`.` among them, here only ever mean themselves.

```{r}
#| echo: false
notes
```

::: {.panel-tabset}

### run

```{r}
run('
  notes
    then keep where [note] contains "100%"
')
```

### R

```{r}
notes |> keep(grepl("100%", note))
```

### Python

```{python}
notes >> keep(col.note.contains("100%"))
```

:::

*`notes then keep where [note] contains "100%"`*

When the question really is a pattern, it belongs to the host language, a
boundary [the coverage appendix](../appendices/e-coverage.qmd) states with
the rest of them.

## One word, one seat

`kind` asks about a column, so the `where` that chooses columns is the one
seat it has. Written into an ordinary row question, it is refused rather
than guessed at:

::: {.panel-tabset}

### R

```{r}
#| error: true
collect(sales |> keep(kind == "number"))
```

### Python

```{python}
#| classes: refusal
try:
    collect(sales >> keep(kind == "number"))
except GodError as refusal:
    print(refusal)
```

:::

The two messages differ, because a bare `kind` is not the same thing in the
two languages. In Python it is the grammar's own word, so the message hands
the word back to its seat, with the working spelling written out. In R a
bare name inside `keep` is a column, so its message is the one a missing
column gets: there is no column called `kind`.
