---
title: "What it wrote, in a language you know"
---

{{< include ../_setup.qmd >}}

A small vocabulary covers most of what people do and never all of it, so the
question is not whether you reach its edge but what happens there. Ask any
pipeline what it would be in dplyr, or ask for the query. There are seven
answers: `sql`, `spark`, `dplyr`, `pandas`, `polars`, `pyspark`, and `god`
itself.

::: {.panel-tabset}

### R

```{r}
show_as(sales |> keep(region == "West") |> take(10), "dplyr")
```

### Python

```{python}
show_as(sales >> keep(col.region == "West") >> take(10), "dplyr")
```

:::

::: {.panel-tabset}

### R

```{r}
show_as(sales |> summarize(revenue = total(revenue), by = product), "sql")
```

### Python

```{python}
show_as(sales >> summarize(revenue = total(col.revenue), by = col.product), "sql")
```

:::

## The same sentence, on a cluster

`"sql"` is the query for the engine on your machine. `"spark"` is the same
pipeline for Spark, and the point is what does not change: the sentence you
wrote.

::: {.panel-tabset}

### R

```{r}
show_as(sales |> summarize(revenue = total(revenue), by = product), "spark")
```

### Python

```{python}
show_as(sales >> summarize(revenue = total(col.revenue), by = col.product), "spark")
```

:::

The two queries differ in five places, and only two of those are cosmetic. A
column is `` `revenue` `` on Spark and `"revenue"` here, and that one is worth
knowing about: a double-quoted name on Spark is not a column at all, it is the
text `'revenue'`, so a query written for the wrong engine would run happily and
put the column's own name in every row. Dropping a column is `EXCEPT` rather than
`EXCLUDE`, stopping a query is `raise_error` rather than `error`, and a backslash
inside a text value has to be written twice.

None of that is yours to remember. It is the difference between writing a
pipeline and writing a query.

## In a dataframe library instead

A query is one answer. The other is the dataframe library you already use, which
is what you want when the table is in front of you and there is no engine to hand
anything to. The same pipeline, three ways:

::: {.panel-tabset}

### R

```{r}
show_as(sales |>
          add(margin = revenue - cost) |>
          keep(margin > 50) |>
          sort(descending(margin)), "polars")
```

### Python

```{python}
show_as(sales
  >> add(margin = col.revenue - col.cost)
  >> keep(col.margin > 50)
  >> sort(descending(col.margin)), "polars")
```

:::

::: {.panel-tabset}

### R

```{r}
show_as(sales |>
          add(margin = revenue - cost) |>
          keep(margin > 50) |>
          sort(descending(margin)), "pandas")
```

### Python

```{python}
show_as(sales
  >> add(margin = col.revenue - col.cost)
  >> keep(col.margin > 50)
  >> sort(descending(col.margin)), "pandas")
```

:::

::: {.panel-tabset}

### R

```{r}
show_as(sales |>
          add(margin = revenue - cost) |>
          keep(margin > 50) |>
          sort(descending(margin)), "pyspark")
```

### Python

```{python}
show_as(sales
  >> add(margin = col.revenue - col.cost)
  >> keep(col.margin > 50)
  >> sort(descending(col.margin)), "pyspark")
```

:::

Each verb became one method, and the order of the steps did not move. What
changes is how a column is named. polars writes `pl.col`, PySpark writes `F.col`,
and pandas has no name of its own, so it borrows the frame it was handed and
writes `d["margin"]` inside a small function.

That last one is why the pandas rendering is the longest of the three, and it is
the honest answer rather than a failure to find a shorter one. pandas has no
single way to chain steps, so the idiom that does chain is `.assign` and
`.loc[lambda d: ...]`, and a column inside either of those has to name the frame
again. The grammar says `[margin]` once.

Where an engine cannot say what a sentence means, you get a refusal rather than
a query that says something close. Spark has to be told which columns a `widen`
makes, because its pivot cannot work them out from the data:

::: {.panel-tabset}

### R

```{r}
#| error: true
show_as(marks |> widen(name = question, value = mark, by = student), "spark")
```

### Python

```{python}
try:
    show_as(marks >> widen(name = col.question, value = col.mark, by = col.student),
            "spark")
except GodError as refusal:
    print(refusal)
```

:::

## When the query is all you want

`show_as` starts from a pipeline and a table you are holding. `god_sql` is the
shortcut for the other situation: you have the sentence and you know the
columns, and the query itself is the thing you are after, for a scheduler, a
dashboard, or a warehouse console that will run it far from here.

::: {.panel-tabset}

### R

```{r}
cat(god_sql(
  'sales then keep where [region] is "West" then take 3',
  "region:text,product:text,revenue:number"
))
```

### Python

```{python}
print(god_sql(
    'sales then keep where [region] is "West" then take 3',
    "region:text,product:text,revenue:number"
))
```

:::

No table is attached and none is needed. The column list is the contract, and
the same checks run against it before anything is written out.

## When the table is what you want

`god_sql` writes the query and stops. `run` writes it, executes it, and gives
back the table, starting from a sentence instead of from a pipeline you built
with verbs.

::: {.panel-tabset}

### R

```{r}
run('sales then keep where [region] is "West" then take 3')
```

### Python

```{python}
run('sales then keep where [region] is "West" then take 3')
```

:::

Read what sits inside the parentheses in each tab. It is the same string, to the
character. Everywhere else in this book the two languages differ in the pipe
glyph and in how a column is named, and a sentence in the text form has neither
of those, so nothing is left to respell.

The table at the head of the sentence is looked up where you are calling from,
which is how `sales` was found here. Pass it by name when it is somewhere else,
as `run(pipeline, sales = last_quarter)`.

## Who writes it, and who runs it

A sentence is data. It fits in a database cell, a configuration file, a
spreadsheet column, or a message to a colleague. So the person who writes a
question and the person who runs it do not have to be the same person.

Someone whose work is reading tables, and who will never write R or Python, can
still write this one and say whether it asks the right thing:

```
sales then keep where [region] is "West" then summarize [total] as total([revenue]) by [product]
```

An analyst calls `run` on it later, or a scheduled job does. The sentence
crosses that handoff intact, because it is not code in either language, and it
is still readable by the person who wrote it.

One boundary is worth stating plainly. `run` is an R function and a Python
function, so calling it needs one of those two languages installed. What travels
without them is the sentence, not the call around it.
