---
title: "Choosing by what a column holds"
---

{{< include ../_setup.qmd >}}

What kind of thing does a column hold, and can you choose by that alone? It
is the other question you can ask about a column without looking at a single
row. `kind` is one of `"number"`, `"text"`, `"truth"` or `"date"`.

::: {.panel-tabset}

### R

```{r}
survey |> pick(where(kind == "number"))
```

### Python

```{python}
survey >> pick(where(kind == "number"))
```

:::

*`survey then pick where (kind is "number")`*

Both questions live in the same `where`, so they join.

::: {.panel-tabset}

### R

```{r}
survey |> pick(where(kind == "number" & startsWith(name, "q")))
```

### Python

```{python}
survey >> pick(where((kind == "number") & name.starts("q")))
```

:::

Put it with `summarize` and you have the sentence people reach for most: average
every number, whatever the numbers happen to be called.

::: {.panel-tabset}

### R

```{r}
survey |> summarize(where(kind == "number", average(value)))
```

### Python

```{python}
survey >> summarize(where(kind == "number", average(value)))
```

:::

That one line does not name a column at all, so it keeps working when the table
gains one.

## Case, and the two words that handle it

The name tests match exactly, so `name starts "q"` does not find a column called
`Q1_score`. Mixed-case names are ordinary enough that this comes up quickly.

There is no flag for it, because a name is text and text has a case. Fold the
case and ask the question.

::: {.panel-tabset}

### R

```{r}
mixed <- data.frame(Q1_score = 4, q2_score = 5, Region = "West",
                    stringsAsFactors = FALSE)

mixed |> pick(where(startsWith(tolower(name), "q")))
```

### Python

```{python}
mixed = pd.DataFrame({"Q1_score": [4], "q2_score": [5], "Region": ["West"]})

mixed >> pick(where(lower(name).starts("q")))
```

:::

`lower` and `upper` are ordinary words, so they work on a value in exactly the
same way.

::: {.panel-tabset}

### R

```{r}
mixed |> keep(tolower(Region) == "west")
```

### Python

```{python}
mixed >> keep(lower(col.Region) == "west")
```

:::

Folding the case of every name automatically would have been the other option,
and it is wrong: two columns can differ only by case, and a grammar that pretends
otherwise cannot treat them as the two columns they are. Asking for it is one
word.

## Why you have to write `name`

The same three words also test what is **inside** a column, and there the subject
is the column itself.

::: {.panel-tabset}

### R

```{r}
survey |> keep(startsWith(region, "W"))
```

### Python

```{python}
survey >> keep(col.region.starts("W"))
```

:::

So `starts` means one thing, "this text begins with that text", and the subject
says what text is being asked about. Writing it is what keeps the two apart.

The alternative would have been `pick(starting("q"))`, which is shorter and
leaves nothing in the sentence to say whether `starting` is looking at a name or
at a value. You would learn it from the position, which is the kind of rule this
grammar removes rather than adds.

The tests are the same words in both languages. What differs is how a column's
name is reached, which is the same difference as everywhere else: R writes a bare
name and Python writes `col.name`, so R spells this with base R's `startsWith`
and Python with a method.

## A percent sign is a percent sign

The three tests read text literally, with no pattern language hiding inside.
R borrows `grepl`'s shape to ask the question, and the reading is still the
grammar's: the marks that regular expressions treat as instructions, `%` and
`.` among them, here only ever mean themselves.

::: {.panel-tabset}

### R

```{r}
notes <- data.frame(note = c("grew 100% this year", "flat since March"),
                    stringsAsFactors = FALSE)
notes |> keep(grepl("100%", note))
```

### Python

```{python}
notes = pd.DataFrame({"note": ["grew 100% this year", "flat since March"]})
notes >> keep(col.note.contains("100%"))
```

:::

When the question really is a pattern, it belongs to the host language, a
boundary the coverage appendix states with the rest of them.
