---
title: "What it wrote, in a language you know"
---

{{< include ../_setup.qmd >}}

A small vocabulary covers most of what people do and never all of it, so the
question is not whether you reach its edge but what happens there. Ask any
pipeline what it would be in dplyr, or ask for the query. There are seven
answers: `sql`, `spark`, `dplyr`, `pandas`, `polars`, `pyspark`, and `god`
itself.

::: {.panel-tabset}

### R

```{r}
show_as(sales |> keep(region == "West") |> take(10), "dplyr")
```

### Python

```{python}
show_as(sales >> keep(col.region == "West") >> take(10), "dplyr")
```

:::

`show_as` also takes the sentence itself, with any tables passed by name, as
`show_as('sales then take 3', "dplyr", sales = sales)`; the phrasebook
appendix is generated that way. And the two languages hand the result back
differently, on purpose. R prints the code and returns it invisibly, the way
R's printers behave. Python returns it as a string, for `print` or for
storing, because printing and returning both made a notebook say everything
twice.

::: {.panel-tabset}

### R

```{r}
show_as(sales |> summarize(revenue = total(revenue), by = product), "sql")
```

### Python

```{python}
show_as(sales >> summarize(revenue = total(col.revenue), by = col.product), "sql")
```

:::

## The same sentence, on a cluster

`"sql"` is the query for the engine on your machine. `"spark"` is the same
pipeline for Spark, and the point is what does not change: the sentence you
wrote.

::: {.panel-tabset}

### R

```{r}
show_as(sales |> summarize(revenue = total(revenue), by = product), "spark")
```

### Python

```{python}
show_as(sales >> summarize(revenue = total(col.revenue), by = col.product), "spark")
```

:::

The two dialects differ in six ways, and the queries above show the first. A
column is `` `revenue` `` on Spark and `"revenue"` here, and that one is worth
knowing about. A double-quoted name on Spark is not a column at all: it is the
text `'revenue'`. A query written for the wrong engine would therefore run
happily and put the column's own name in every row. Two more are spelling:
dropping a column is `EXCEPT` rather than `EXCLUDE`, and stopping a query is
`raise_error` rather than `error`. A backslash inside a text value has to be
written twice on Spark. The word `weekday` is asked for through a different
function on each engine, so that Monday can stay 1 on both. And a `widen`
that names no columns is a sentence Spark cannot be given at all, which is
the refusal staged later on this page.

None of that is yours to remember. It is the difference between writing a
pipeline and writing a query.

## In a dataframe library instead

A query is one answer. The other is the dataframe library you already use, which
is what you want when the table is in front of you and there is no engine to hand
anything to. The same pipeline, three ways:

::: {.panel-tabset}

### R

```{r}
show_as(sales |>
          add(margin = revenue - cost) |>
          keep(margin > 50) |>
          sort(descending(margin)), "polars")
```

### Python

```{python}
show_as(sales
  >> add(margin = col.revenue - col.cost)
  >> keep(col.margin > 50)
  >> sort(descending(col.margin)), "polars")
```

:::

::: {.panel-tabset}

### R

```{r}
show_as(sales |>
          add(margin = revenue - cost) |>
          keep(margin > 50) |>
          sort(descending(margin)), "pandas")
```

### Python

```{python}
show_as(sales
  >> add(margin = col.revenue - col.cost)
  >> keep(col.margin > 50)
  >> sort(descending(col.margin)), "pandas")
```

:::

::: {.panel-tabset}

### R

```{r}
show_as(sales |>
          add(margin = revenue - cost) |>
          keep(margin > 50) |>
          sort(descending(margin)), "pyspark")
```

### Python

```{python}
show_as(sales
  >> add(margin = col.revenue - col.cost)
  >> keep(col.margin > 50)
  >> sort(descending(col.margin)), "pyspark")
```

:::

Each verb became one method, and the order of the steps did not move. What
changes is how a column is named. polars writes `pl.col`, PySpark writes `F.col`,
and pandas has no name of its own, so it borrows the frame it was handed and
writes `d["margin"]` inside a small function.

That last one is why the pandas rendering is the longest of the three, and it is
the honest answer rather than a failure to find a shorter one. pandas has no
single way to chain steps, so the idiom that does chain is `.assign` and
`.loc[lambda d: ...]`, and a column inside either of those has to name the frame
again. The grammar says `[margin]` once.

Where an engine cannot say what a sentence means, you get a refusal rather than
a query that says something close. Spark has to be told which columns a `widen`
makes, because its pivot cannot work them out from the data:

::: {.panel-tabset}

### R

```{r}
#| error: true
show_as(marks |> widen(name = question, value = mark, by = student), "spark")
```

### Python

```{python}
#| classes: refusal
try:
    show_as(marks >> widen(name = col.question, value = col.mark, by = col.student),
            "spark")
except GodError as refusal:
    print(refusal)
```

:::

## When the query is all you want

`show_as` starts from a pipeline and a table you are holding. `god_sql` is the
shortcut for the other situation: you have the sentence and you know the
columns, and the query itself is the thing you are after, for a scheduler, a
dashboard, or a warehouse console that will run it far from here.

::: {.panel-tabset}

### R

```{r}
cat(god_sql(
  'sales then keep where [region] is "West" then take 3',
  "region:text,product:text,revenue:number"
))
```

### Python

```{python}
print(god_sql(
    'sales then keep where [region] is "West" then take 3',
    "region:text,product:text,revenue:number"
))
```

:::

No table is attached and none is needed. The column list is the contract, and
the same checks run against it before anything is written out.

## The same engine at the command line

Everything on this page is one binary, `god-cli`, which the transcripts in
this book shorten to `god`. It reads a sentence and writes text, so it needs
no R and no Python: give it the columns, and name the target with `--as`.

```
$ god --columns 'region:text,revenue:number' --as dplyr 'sales then keep where [region] is "West" then take 3'
sales |>
  filter((region == "West")) |>
  head(3)
```

Wrap the pipeline in single quotes, since a text value inside it uses double
ones. The pipeline can also arrive on standard input, as
`god [options] < pipeline.god`, which is the shape a scheduled job wants.

Two smaller questions have flags of their own. `--needs` prints the tables a
pipeline reads and stops, which is how a launcher knows what to describe
before describing anything:

```
$ god --needs 'sales then join products by [product] then take 3'
sales
products
```

And `--vocabulary` prints every word the grammar has, one per line, tagged
with its kind; [Appendix B](../appendices/b-every-word.qmd) is generated
from it.

The exit code says what happened, so a script can ask. `0` is an answer, on
standard output. `1` is a command line the tool could not read, a flag given
twice included, with the usage on standard error. `2` is a refusal:

```
$ god --columns 'region:text' --as excel 'sales then take 1'
illegal: there is no backend called `excel`. There is: sql, spark, dplyr, pandas, polars, pyspark, god
```

An assumption is the third voice, and it costs nothing. Where the grammar
decided something you did not say, the note lands on standard error while
the answer still arrives with exit `0`, so nothing is decided silently.

## When the table is what you want

`god_sql` writes the query and stops. `run` writes it, executes it, and gives
back the table, starting from a sentence instead of from a pipeline you built
with verbs.

::: {.panel-tabset}

### R

```{r}
run('sales then keep where [region] is "West" then take 3')
```

### Python

```{python}
run('sales then keep where [region] is "West" then take 3')
```

:::

Read what sits inside the parentheses in each tab. It is the same string, to the
character. Everywhere else in this book the two languages differ in the pipe
glyph and in how a column is named, and a sentence in the text form has neither
of those, so nothing is left to respell.

The single quotes around the sentence are the transcripts' convention, kept
for the same reason. A text value inside a sentence is always written with
double quotes, so wrapping the whole of it in single ones means neither
language ever needs an escape, and the string on this page pastes into a
shell, an R script or a Python one unchanged.

Swap the two out of habit and the grammar catches the half that reaches it.
A text value in single quotes is refused with the spelling named, so the
mistake costs a message rather than a wrong answer:

::: {.panel-tabset}

### R

```{r}
#| error: true
run("sales then keep where [region] is 'West' then take 3")
```

### Python

```{python}
#| classes: refusal
try:
    run("sales then keep where [region] is 'West' then take 3")
except GodError as refusal:
    print(refusal)
```

:::

The other half belongs to the hosts: a single quote inside a single-quoted
string is a syntax error R and Python report themselves, before god is ever
called. And when a text value genuinely needs an apostrophe, each host has
its own door. Python's triple quotes, `run('''…''')`, and R's raw string,
`run(r"(…)")`, both hold either kind of quote unescaped.

The table at the head of the sentence is looked up where you are calling from,
which is how `sales` was found here. Pass it by name when it is somewhere else,
as `run(pipeline, sales = last_quarter)`.

## Who writes it, and who runs it

A sentence is data. It fits in a database cell, a configuration file, a
spreadsheet column, or a message to a colleague. So the person who writes a
question and the person who runs it do not have to be the same person.

Someone whose work is reading tables, and who will never write R or Python, can
still write this one and say whether it asks the right thing:

```
sales
  then keep where [region] is "West"          # this quarter's question
  then summarize [total] as total([revenue]) by [product]
```

The `#` starts a comment that runs to the end of its line, here exactly as it
does in both languages, so a stored sentence can carry its reasoning along.
An analyst calls `run` on it later, or a scheduled job does. The sentence
crosses that handoff intact, because it is not code in either language, and it
is still readable by the person who wrote it.

One boundary is worth stating plainly. `run` is an R function and a Python
function, so calling it needs one of those two languages installed. What travels
without them is the sentence, not the call around it.
