---
title: "Where this ends, and gog begins"
---

{{< include ../_setup.qmd >}}

```{r}
#| include: false
suppressMessages(library(gog))
# Gapminder's country panel (CC BY 4.0), vendored beside the book: this render
# used to fetch it from the sibling book's site, which made one URL the single
# point of failure for the whole build.
nations <- read.csv(find_up("book/data/gm_all.csv"))
```

```{python}
#| include: false
import gog
_data = next(p / "book" / "data" / "gm_all.csv"
             for p in (pathlib.Path.cwd(), *pathlib.Path.cwd().parents)
             if (p / "book" / "data" / "gm_all.csv").exists())
nations = pd.read_csv(_data)
```

god manipulates tables. It does not draw them, and it is not going to, because
drawing is a different job with a different vocabulary. A grammar that tried to
do both would stop being small enough to hold in your head, which is the one
thing this project will not trade.

That other grammar exists. **gog** is this project's sibling: a grammar of
graphics, by the same author, built on the same discipline and holding the same
promise about day one and day two. Most real questions use both, because a
question usually starts as a table that is the wrong shape and ends as a picture.

This chapter is one question answered end to end. Nothing else is loaded.

## The table

Gapminder's country panel: 142 countries, twelve observations each from 1952 to
2007, 1,704 rows in all. It is a real table rather than a demonstration, which
is the point of using it here. The numbers are Gapminder's, CC BY 4.0, and the
file ships beside the book, so rendering this page needs no network.

::: {.panel-tabset}

### R

```{r}
nations |> take(5)
```

### Python

```{python}
nations >> take(5)
```

:::

Five rows of 1,704. Every table on this page is shown that way, because a
chapter that prints 1,704 rows is not showing you anything.

## The question

> Which countries gained the most life expectancy, and what did the world look
> like while it happened?

The first half is the table's job. Each country has twelve rows and the answer
needs two of them, the earliest and the latest, which is what `first` and `last`
are for. They answer by position, so a `sort` in front of them is what decides
which position that is.

::: {.panel-tabset}

### R

```{r}
gains <- nations |>
  sort(year) |>
  summarize(started = first(life), ended = last(life), by = c(country, continent)) |>
  add(gained = ended - started) |>
  sort(descending(gained))

gains |> take(5)
```

### Python

```{python}
gains = (nations
  >> sort(col.year)
  >> summarize(started = first(col.life), ended = last(col.life),
               by = [col.country, col.continent])
  >> add(gained = col.ended - col.started)
  >> sort(descending(col.gained)))

gains >> take(5)
```

:::

Oman gained more than thirty years of life expectancy in half a century. That is
one sentence of god: sort, summarize by two columns, subtract, sort again.

The other end of the same table is a different question and the same pipeline
with one word changed.

::: {.panel-tabset}

### R

```{r}
gains |> sort(gained) |> take(5)
```

### Python

```{python}
gains >> sort(col.gained) >> take(5)
```

:::

## When you need `collect`

Nothing above called `collect`, and the tables still appeared. That is worth
explaining, because it is the question everyone asks once.

A pipeline is a plan, and it does not run when you write it. Printing one runs
it. That is why every example so far has shown a table: this book prints each
pipeline, so each one is asked for its answer.

`collect` is how you ask for the answer when something *other than printing*
needs it. Handing a table to another function is exactly that case, and gog is
about to be handed one.

::: {.panel-tabset}

### R

```{r}
top <- gains |> take(12) |> collect()
class(top)
```

### Python

```{python}
top = collect(gains >> take(12))
type(top).__name__
```

:::

So the rule is short. If you want to *look* at it, print it. If you want to
*use* it, collect it.

## The picture

Now gog. The table goes in, a mark and some channels say what to draw, and the
sentence reads the way a god pipeline does.

::: {.panel-tabset}

### R

```{r}
data(top) + bar + x(gained) + y(country) + color(continent)
```

### Python

```{python}
gog.data(top) + gog.bar + gog.x(gog.col.gained) + gog.y(gog.col.country) + gog.color(gog.col.continent)
```

:::

## What the whole world was doing

The bar chart answers the question. This next one is why the pairing is worth
having: the same panel, unsummarized, with one more channel.

`play` turns a column into time. Every country is a point, income runs along the
bottom on a log scale, life expectancy up the side, and the year becomes the
animation rather than another axis.

::: {.panel-tabset}

### R

```{r}
data(nations) + point + x(gdp, scale = "log") + y(life) +
  color(continent) + size(population) + play(year)
```

### Python

```{python}
(gog.data(nations) + gog.point + gog.x(gog.col.gdp, scale = "log") + gog.y(gog.col.life)
   + gog.color(gog.col.continent) + gog.size(gog.col.population) + gog.play(gog.col.year))
```

:::

That is the Rosling chart, and neither package needed a special case to draw it.
`play` is a channel like any other, so the year binds the way a color does.

## What it replaced

Written in the tools this book keeps comparing itself against, those two steps
need a reshaping function from one package and a plotting system from another.
Each brings its own vocabulary, its own idea of what a column reference is,
and its own rules about which arguments are quoted. The animation usually
needs a third.

The saving is not the line count. It is that `summarize` and `play` come from
the same design, so the second half of the analysis does not ask you to change
how you are thinking. `by` means in gog what it means here. So does `color`.

## The one thing to know about using both

In R, load them and go. R looks a column name up in your data before it looks
anywhere else, and the two packages take care not to collide.

In Python, one name is claimed by both. `col` is how each of them names a column,
so `from god import *` followed by `from gog import *` would leave you with one
of them and no warning. Import gog by name instead, which is what the examples
above do:

```python
from god import *
import gog
```

Then `col.life` is god's and `gog.col.life` is gog's, and nothing is ambiguous.
[Appendix A](../appendices/a-two-languages.qmd) says the same thing about every
other name god takes over.

## Where to go next

The graphics grammar has its own manual, at
[the gog book](https://psychometrician.github.io/gog-book/), written the way this
one is: every plot on the page was drawn by the engine while the page was built.

Read it in the order you would use them. You start in god, with a table that is
the wrong shape. You end in gog, looking at the answer.

The table on this page is the Gapminder panel, by way of the `gapminder` R
package. It is public domain, and credited anyway: the package is Jennifer
Bryan's, and the figures are the Gapminder Foundation's.
