---
title: "With gog"
---

{{< include ../_setup.qmd >}}

```{r}
#| include: false
suppressMessages(library(gog))
# The gapminder panel is read on stage in "The table" below, from the copy
# vendored beside the book: this render used to fetch it from the sibling
# book's site, which made one URL the single point of failure for the build.
#
# The three interactive plots leave six lines in the render log, one per tab:
# "WARNING ... Unable to parse table from raw html block: skipping." Both
# languages emit it rather than Python alone, counted in the rendered page.
# The `<table` pandoc finds is a template literal inside gog's own selection
# script, so there is no table to parse; the parse is skipped and the block
# passes through whole, which is why every plot still draws and why this
# page's own tables keep their styling. Chunk-level `html-table-processing:
# none` does not reach reticulate's output on Quarto 1.8.25, measured rather
# than assumed. The cure is splitting one string literal in gog's runtime,
# which is the sibling project's change to make. The workflow's log gate
# matches `WARN:` with its colon, so these lines do not block a publish.
```

```{python}
#| include: false
import gog
```

god manipulates tables. It does not draw them, and it is not going to, because
drawing is a different job with a different vocabulary. A grammar that tried to
do both would stop being small enough to hold in your head, which is the one
thing this project will not trade.

That other grammar exists. **gog** is this project's sibling: a grammar of
graphics, by the same author, built on the same discipline and holding the same
promise about day one and day two. Most real questions use both, because a
question usually starts as a table that is the wrong shape and ends as a picture.

This chapter is one question answered end to end. Nothing else is loaded.

## The table

Gapminder's country panel: 142 countries, twelve observations each from 1952 to
2007, 1,704 rows in all. It is a real table rather than a demonstration, which
is the point of using it here. The numbers are Gapminder's, CC BY 4.0, and the
file ships beside the book, so rendering this page needs no network.

It is also the one table this book reads from a file on the page, because a
file is how your own table will arrive. One line in either language, and the
panel the preface's cast declared is an ordinary data frame:

::: {.panel-tabset}

### R

```{r}
gapminder <- read.csv("../data/gapminder.csv")
```

### Python

```{python}
gapminder = pd.read_csv("../data/gapminder.csv")
```

:::

From here it pipes like every table before it.

::: {.panel-tabset}

### run

```{r}
run('
  gapminder
    then take 5
')
```

### R

```{r}
gapminder |> take(5)
```

### Python

```{python}
gapminder >> take(5)
```

:::

Five rows of 1,704. Every table on this page is shown that way, because a
chapter that prints 1,704 rows is not showing you anything.

## The question

> Which countries gained the most life expectancy, and what did the world look
> like while it happened?

The first half is the table's job. Each country has twelve rows and the answer
needs two of them, the earliest and the latest, which is what `first` and `last`
are for. They answer by position, so a `sort` in front of them is what decides
which position that is.

::: {.panel-tabset}

### R

```{r}
gains <- gapminder |>
  sort(year) |>
  summarize(started = first(life), ended = last(life),
            by = c(country, continent)) |>
  add(gained = ended - started) |>
  sort(descending(gained))

gains |> take(5)
```

### Python

```{python}
gains = (gapminder
  >> sort(col.year)
  >> summarize(started = first(col.life), ended = last(col.life),
               by = [col.country, col.continent])
  >> add(gained = col.ended - col.started)
  >> sort(descending(col.gained)))

gains >> take(5)
```

:::

Oman gained more than thirty years of life expectancy in half a century. That is
one sentence of god: sort, summarize by two columns, subtract, sort again.

The other end of the same table is a different question and the same pipeline
with one word changed.

::: {.panel-tabset}

### R

```{r}
gains |> sort(gained) |> take(5)
```

### Python

```{python}
gains >> sort(col.gained) >> take(5)
```

:::

## When you need `collect`

Nothing above called `collect`, and the tables still appeared. That is worth
explaining, because it is the question everyone asks once.

A pipeline is a plan, and it does not run when you write it. Printing one runs
it. That is why every example so far has shown a table: this book prints each
pipeline, so each one is asked for its answer.

`collect` is how you ask for the answer when something *other than printing*
needs it. Handing a table to another function is exactly that case, and gog is
about to be handed one.

::: {.panel-tabset}

### R

```{r}
top <- gains |> take(12) |> collect()
class(top)
```

### Python

```{python}
top = collect(gains >> take(12))
type(top).__name__
```

:::

So the rule is short. If you want to *look* at it, print it. If you want to
*use* it, collect it.

## The picture

Now gog. The table goes in, a mark and some channels say what to draw, and the
sentence reads the way a god pipeline does. The table going in is `top`, the
first twelve rows of the ranking you have already read. On the Python side
every gog name arrives through the module, as `gog.point` and `gog.col`; the
section near the end of this chapter says why, and why R needs no prefix at
all.

::: {.panel-tabset}

### R

```{r}
data(top) + bar + x(gained) + y(country) + color(continent)
```

### Python

```{python}
(gog.data(top) + gog.bar + gog.x(gog.col.gained) + gog.y(gog.col.country)
   + gog.color(gog.col.continent))
```

:::

## What the gain looked like while it happened

The ranking says who gained. The shape of the gain is a second god sentence,
one `summarize` with two columns after `by`:

::: {.panel-tabset}

### R

```{r}
trends <- gapminder |>
  summarize(life = average(life), by = c(continent, year)) |>
  sort(year) |>
  collect()

trends |> take(5)
```

### Python

```{python}
trends = collect(gapminder
  >> summarize(life = average(col.life), by = [col.continent, col.year])
  >> sort(col.year))

trends >> take(5)
```

:::

Sixty rows came back, one per continent per year, and the first five are the
five continents in 1952. A line per continent is then one gog sentence:

::: {.panel-tabset}

### R

```{r}
data(trends) + line + x(year) + y(life) + color(continent)
```

### Python

```{python}
(gog.data(trends) + gog.line + gog.x(gog.col.year) + gog.y(gog.col.life)
   + gog.color(gog.col.continent))
```

:::

Every continent climbs across the half century, and the distance between the
top line and the bottom one is the story the rest of this page moves through.

## What the whole world was doing

The bar chart answers the question. This next one is why the pairing is worth
having: the same panel, unsummarized, with one more channel.

`play` turns a column into time. Every country is a point, GDP per person runs
along the bottom on a log scale, life expectancy up the side, and the year
becomes the animation rather than another axis.

::: {.panel-tabset}

### R

```{r}
data(gapminder) + point + x(gdp, scale = "log") + y(life) +
  color(continent) + size(population) + play(year)
```

### Python

```{python}
(gog.data(gapminder) + gog.point + gog.x(gog.col.gdp, scale = "log")
   + gog.y(gog.col.life) + gog.color(gog.col.continent)
   + gog.size(gog.col.population) + gog.play(gog.col.year))
```

:::

That is the Rosling chart, and neither package needed a special case to draw it.
`play` is a channel like any other, so the year binds the way a color does.

## One panel per continent

`facet` splits a plot into panels, one per value of a column. It joins with
`|` rather than `+`, because it acts on the whole plot rather than adding a
channel to it. The moving chart above, five times:

::: {.panel-tabset}

### R

```{r}
data(gapminder) + point + x(gdp, scale = "log") + y(life) +
  color(continent) + play(year) | facet(continent)
```

### Python

```{python}
(gog.data(gapminder) + gog.point + gog.x(gog.col.gdp, scale = "log")
   + gog.y(gog.col.life) + gog.color(gog.col.continent)
   + gog.play(gog.col.year)) | gog.facet(gog.col.continent)
```

:::

Five panels, one clock. The year advances everywhere at once, so a continent
that stalls is read against the ones that do not.

## A third position

`z` is the third position, and writing it is what turns a plot into a cube.
The slice of one year is a god step:

::: {.panel-tabset}

### R

```{r}
latest <- gapminder |> keep(year == 2007) |> collect()

latest |> take(5)
```

### Python

```{python}
latest = collect(gapminder >> keep(col.year == 2007))

latest >> take(5)
```

:::

One row per country, 142 in all, and the cube binds three of the columns.
`space` names the viewing angle, `turn` and `tilt`, for when the default
angle hides what you came to see:

::: {.panel-tabset}

### R

```{r}
data(latest) + point + x(gdp, scale = "log") + y(life) + z(population) +
  color(continent) + space(turn = -50, tilt = 15)
```

### Python

```{python}
(gog.data(latest) + gog.point + gog.x(gog.col.gdp, scale = "log")
   + gog.y(gog.col.life) + gog.z(gog.col.population)
   + gog.color(gog.col.continent) + gog.space(turn = -50, tilt = 15))
```

:::

And `play` is still a channel, so the cube moves the way the flat chart did,
with nothing taught about the combination:

::: {.panel-tabset}

### R

```{r}
data(gapminder) + point + x(gdp, scale = "log") + y(life) + z(population) +
  color(continent) + play(year)
```

### Python

```{python}
(gog.data(gapminder) + gog.point + gog.x(gog.col.gdp, scale = "log")
   + gog.y(gog.col.life) + gog.z(gog.col.population)
   + gog.color(gog.col.continent) + gog.play(gog.col.year))
```

:::

## Handing the reader the choice

`brush` lights a region and pushes the rest back, and it removes nothing.
Bounded in the sentence, it prints with the selection already made; written
bare, as `+ brush`, it waits for a reader to drag:

::: {.panel-tabset}

### R

```{r}
data(latest) + point + x(gdp, scale = "log") + y(life) +
  color(continent) + brush(gdp, at = c(20000, 60000))
```

### Python

```{python}
(gog.data(latest) + gog.point + gog.x(gog.col.gdp, scale = "log")
   + gog.y(gog.col.life) + gog.color(gog.col.continent)
   + gog.brush(gog.col.gdp, at = [20000, 60000]))
```

:::

The countries above twenty thousand dollars sit almost entirely above
seventy-five years, and everything else stays in view behind them. That is
the working difference between the two grammars' questions: god's `keep`
removes rows before the answer, gog's `brush` highlights them inside it, and
the comparison above needs the context a `keep` would have thrown away.

## What it replaced

Written in the tools this book keeps comparing itself against, the page above
needs a reshaping function from one package and a plotting system from
another. Each brings its own vocabulary, its own idea of what a column
reference is, and its own rules about which arguments are quoted. The
animation usually needs a third package, and the cube and the brush a fourth.

The saving is not the line count. It is that `summarize` and `play` come from
the same design, so the second half of the analysis does not ask you to change
how you are thinking. `by` means in gog what it means here. So does `color`.

## The one thing to know about using both

In R, load them and go. R looks a column name up in your data before it looks
anywhere else, and the two packages take care not to collide.

Python is a different shape. **The rule there is short:** bring god in whole, and
reach gog through its name, which is what the examples above do.

```python
from god import *
import gog
```

Then `col.life` is god's and `gog.col.life` is gog's, and nothing is ambiguous.

Two things make that the right way round, rather than a preference for whichever
book you happen to be reading. Two names are claimed by both: `col`, which is how
each of them names a column, and `median`, a word both vocabularies wanted.
Bringing in both whole would leave you holding gog's pair and no warning.

The second reason would still apply if god did not exist. A grammar of graphics
wants words the language has already spent: `sum`, `max` and several more are
transforms in gog and builtins in Python, so taking gog whole takes those from
you as well. god's vocabulary overlaps none of Python's, and that is what makes
it the one that is safe to bring in whole.

So every gog name on this page carries the prefix, not only the two that collide.
One rule is easier to keep than a list of which names happen to be free this year.

It reverses cleanly when a file is mostly pictures. Take gog whole, reach god
through its name, and take back the builtins you use with `from builtins import
sum`. What does not reverse is taking both whole, which is the one arrangement
here that fails without saying so.

[Appendix A](../appendices/a-two-languages.qmd) says the same thing about every
other name god takes over.

## The question that needs two steps

Halfway between the table and the plot, a tempting shortcut appears: keep
the countries above the average, in one step. `keep` decides one row at a
time, so it cannot ask a question about a whole group, and the refusal says
which two steps to write instead:

::: {.panel-tabset}

### R

```{r}
#| error: true
collect(gapminder |> keep(average(life) > 50))
```

### Python

```{python}
#| classes: refusal
try:
    collect(gapminder >> keep(average(col.life) > 50))
except GodError as refusal:
    print(refusal)
```

:::

The two-step spelling it offers is the shape this chapter's own pipeline
already has.

## Where to go next

The graphics grammar has its own manual, at
[the gog book](https://psychometrician.github.io/gog-book/), written the way this
one is: every plot on the page was drawn by the engine while the page was built.

Read it in the order you would use them. You start in god, with a table that is
the wrong shape. You end in gog, looking at the answer.

The table on this page is the Gapminder panel, in the excerpt that Jennifer
Bryan's `gapminder` R package made standard. The figures are the Gapminder
Foundation's, CC BY 4.0, and the credit belongs in both places.
