---
title: "The words that repeat"
---

{{< include ../_setup.qmd >}}

Why did the eighth verb take minutes, when the first one took a chapter? You
have met eight verbs now, and a handful of small words that traveled between
them: `by`, `where`, `as`, `all_but`, `descending`. This chapter is about
those small words, and it is the shortest useful thing in the book, because
the answer to the question above is that the small words never changed.

Each of them means one thing. Not one thing per verb, one thing. When you
meet a verb you have never seen, the small words in it already make sense,
and that is most of what makes this grammar quick to get back after a year
away.

## `by` says which rows go together

You saw `by` on `summarize`, where it made the groups. It is the same word on
`add`, where it says which rows a value is worked out within, and on `take`,
where it says which rows the first few are taken from.

::: {.panel-tabset}

### run

```{r}
run('
  sales
    then summarize [total] as total([revenue]) by [product]
')
```

### R

```{r}
sales |> summarize(total = total(revenue), by = product)
```

### Python

```{python}
sales >> summarize(total = total(col.revenue), by = col.product)
```

:::

*`sales then summarize [total] as total([revenue]) by [product]`*

On `add`, the same word turns a collapse into a broadcast. The aggregation is
still worked out per group, and instead of one row per group, every row gets
its group's answer, which is exactly what a share needs:

::: {.panel-tabset}

### run

```{r}
run('
  sales
    then add [share] as ([revenue] / total([revenue])) by [product]
    then pick [product, revenue, share]
')
```

### R

```{r}
sales |>
  add(share = revenue / total(revenue), by = product) |>
  pick(product, revenue, share)
```

### Python

```{python}
(sales
  >> add(share = col.revenue / total(col.revenue), by = col.product)
  >> pick(col.product, col.revenue, col.share))
```

:::

Read a product's rows and their shares sum to one: each row's revenue over
its own group's total. In dplyr this idiom needs a grouped mutate and an
ungroup; in SQL it needs a window function; here it is the `by` you already
know, sitting on a different verb. And on `take`, the word means the same
thing a third time:

::: {.panel-tabset}

### run

```{r}
run('
  sales
    then sort [revenue] descending
    then take 1 by [product]
')
```

### R

```{r}
sales |> sort(descending(revenue)) |> take(1, by = product)
```

### Python

```{python}
sales >> sort(descending(col.revenue)) >> take(1, by = col.product)
```

:::

Three verbs, one word, one meaning. Later you will meet `by` on `join` and on
`widen`, and it will mean the same thing there: the columns that say which rows
correspond.

## `as` gives something a name

`add [name] as [value]`. The name goes first, the way assignment reads, and the
same shape names the result of `rename` and the filler in `fill_missing`.

::: {.panel-tabset}

### run

```{r}
run('
  sales
    then add [margin] as ([revenue] - [cost])
    then rename [earned] as [revenue]
    then pick [product, earned, margin]
')
```

### R

```{r}
sales |>
  add(margin = revenue - cost) |>
  rename(earned = revenue) |>
  pick(product, earned, margin)
```

### Python

```{python}
(sales
  >> add(margin = col.revenue - col.cost)
  >> rename(earned = col.revenue)
  >> pick(col.product, col.earned, col.margin))
```

:::

## `all_but` turns a list inside out

Wrap the names you do *not* want. It is the same word wherever a set of columns
is chosen, which so far is `pick` and, in the next part, `lengthen`.

::: {.panel-tabset}

### run

```{r}
run('
  sales
    then pick all_but [cost, region]
')
```

### R

```{r}
sales |> pick(all_but(cost, region))
```

### Python

```{python}
sales >> pick(all_but(col.cost, col.region))
```

:::

## `descending` reverses an ordering

It is a modifier on a column, in a position where an ordering is being given. So
it reads the same on `sort` and inside `rank`, and beyond the column it marks,
it has no options of its own on either.

::: {.panel-tabset}

### run

```{r}
run('
  sales
    then add [place] as rank([revenue] descending)
    then sort [place]
    then pick [product, revenue, place]
')
```

### R

```{r}
sales |>
  add(place = rank(descending(revenue))) |>
  sort(place) |>
  pick(product, revenue, place)
```

### Python

```{python}
(sales
  >> add(place = rank(descending(col.revenue)))
  >> sort(col.place)
  >> pick(col.product, col.revenue, col.place))
```

:::

## `where` asks a question

On `keep` the question is about a row. Later you will see it on `pick`, where the
question is about a column, and the difference is written down rather than
guessed: inside `pick` you say `name` or `kind` to name what you are asking
about.

::: {.panel-tabset}

### run

```{r}
run('
  sales
    then keep where ([revenue] > 150)
')
```

### R

```{r}
sales |> keep(revenue > 150)
```

### Python

```{python}
sales >> keep(col.revenue > 150)
```

:::

## Where `by` does not go

A word that travels still has a boundary, and the boundary is stated rather
than guessed. `by` means per group, and `sort` already orders every row, so
the word has nothing to add there and is refused by name. The sentence below
is the text form, so the call is the same string in both tabs:

::: {.panel-tabset}

### R

```{r}
#| error: true
run('sales then sort by [region]')
```

### Python

```{python}
#| classes: refusal
try:
    run('sales then sort by [region]')
except GodError as refusal:
    print(refusal)
```

:::

The message says what `sort` does take, which is the whole of it.

## Why this matters more than it looks

A vocabulary is easy to learn and easy to lose. What survives a year away is
not the word list, it is the rules for putting words together, and those only
survive if there are few of them and they have no exceptions.

So the test this grammar sets itself is not whether it has a word for
everything. It is whether a word you learned in one place still means the
same thing in the next place you meet it. Every part after this one is that
test being run. And you now hold the working form of the preface's motto: if
you can say something two ways in this grammar, one of them is a bug, and
the author wants to hear about it. That sentence is not rhetoric. It is the
standard each of the last seven chapters answered to, from the missing
`ascending` to the missing `drop`, and it is falsifiable by any reader with
an afternoon.

The eight verbs and five small words you now have are the everyday core of
the grammar, and the preface's promise has reached its first checkpoint:
day one was reading, and you have been reading since the first pipeline.
Day two is writing, and the practice pair ahead closes with it. Those two
chapters are where this part's claims stop being the book's and start being
yours.
