---
title: "Writing practice"
---

{{< include ../_setup.qmd >}}

Reading back a sentence and producing one are different skills, so which one
do you actually have? Recognition came free with the last eight chapters.
Recall is the day-two promise, it is the one that pays for itself every
working day, and the only way to get it is the way this chapter is built:
you write, look, and refine.

The method is always the same. Start from the smallest legal sentence, which
is the table's name and one verb. Look at what came back. Add one word or one
step, and look again. A pipeline is a plan, so every intermediate look is
free, and nobody who works this way needs to hold a whole sentence in their
head before typing.

## First question: what is slow-moving stock?

Say the products that sold fewer than ten units all year. Start smaller than
the question: totals per product.

::: {.panel-tabset}

### R

```{r}
sales |> summarize(sold = total(quantity), by = product)
```

### Python

```{python}
sales >> summarize(sold = total(col.quantity), by = col.product)
```

:::

*`sales then summarize [sold] as total([quantity]) by [product]`*

That is most of the answer already. One `keep` finishes it:

::: {.panel-tabset}

### R

```{r}
sales |> summarize(sold = total(quantity), by = product) |> keep(sold < 10)
```

### Python

```{python}
sales >> summarize(sold = total(col.quantity), by = col.product) >> keep(col.sold < 10)
```

:::

*`sales then summarize [sold] as total([quantity]) by [product] then keep where [sold] < 10`*

Note where the `keep` went: after the summarize, asking about the new `sold`
column. A `keep` before it would have asked a different question, about
single orders rather than about products.

## Second question: where does the margin come from?

Say each region's total margin, largest first. Build it in the order you
would explain it: make the margin, collapse it, order the answer.

::: {.panel-tabset}

### R

```{r}
sales |>
  add(margin = revenue - cost) |>
  summarize(margin = total(margin), by = region) |>
  sort(descending(margin))
```

### Python

```{python}
(sales
  >> add(margin = col.revenue - col.cost)
  >> summarize(margin = total(col.margin), by = col.region)
  >> sort(descending(col.margin)))
```

:::

*`sales then add [margin] as [revenue] - [cost] then summarize [margin] as total([margin]) by [region] then sort [margin] descending`*

## Third question: who is the strongest respondent?

From the survey, say each respondent's best single score and read the top
three people. You have not seen this question, and you own every word it
needs.

::: {.panel-tabset}

### R

```{r}
survey |>
  add(best = largest(q1_score), by = respondent) |>
  pick(name, region, best) |>
  sort(descending(best), name) |>
  take(3)
```

### Python

```{python}
(survey
  >> add(best = largest(col.q1_score), by = col.respondent)
  >> pick(col.name, col.region, col.best)
  >> sort(descending(col.best), col.name)
  >> take(3))
```

:::

*`survey then add [best] as largest([q1_score]) by [respondent] then pick [name, region, best] then sort [best] descending, [name] then take 3`*

The second sort key is doing quiet, honest work: two respondents share a
score at the cut, and without a tiebreak "top three" would be leaving the
choice to chance. Naming the tiebreak is naming the choice.

One honest limitation is in view: `best` here reads one score column, and
the survey has eight. Asking for the largest across all eight, without
naming them one by one, is a real question and this part cannot say it yet.
[Choosing columns without naming them](../parts/choosing-columns.qmd) can,
in one sentence, and knowing exactly where your current words stop is part
of owning them.

## When the engine says no

Writing means meeting refusals, so meet three on purpose. Each is the
grammar keeping a promise from an earlier chapter, and each message carries
its own repair.

A column that a step removed is gone for every later step, because each step
sees only the table the last one made:

::: {.panel-tabset}

### R

```{r}
#| error: true
sales |>
  summarize(t = total(revenue), by = region) |>
  keep(cost > 100) |>
  collect()
```

### Python

```{python}
#| classes: refusal
try:
    collect(sales
      >> summarize(t = total(col.revenue), by = col.region)
      >> keep(col.cost > 100))
except GodError as refusal:
    print(refusal)
```

:::

The message lists what the table holds *at that step*: two columns, not six.
Read the list and the mistake explains itself.

A grouped `take` still needs its order, in any pipeline, however long:

::: {.panel-tabset}

### R

```{r}
#| error: true
sales |> keep(region != "East") |> take(2, by = product) |> collect()
```

### Python

```{python}
#| classes: refusal
try:
    collect(sales >> keep(col.region != "East") >> take(2, by = col.product))
except GodError as refusal:
    print(refusal)
```

:::

And a typo is a typo wherever it happens, with the nearest real name
offered back:

::: {.panel-tabset}

### R

```{r}
#| error: true
survey |> keep(regoin == "West") |> collect()
```

### Python

```{python}
#| classes: refusal
try:
    collect(survey >> keep(col.regoin == "West"))
except GodError as refusal:
    print(refusal)
```

:::

None of these cost you anything. Nothing ran, nothing half-completed, and in
each case the message is shorter than the documentation you did not have to
open. Refusals are how this grammar teaches at the moment of need, which is
why this book keeps putting them on stage.

## The checkpoint

This part made three claims about you, and they are now checkable. Read any
pipeline in this part aloud, cold. Given a question about one table, name
the verbs that answer it before typing. And predict, before running, whether
a sentence will answer or refuse. If any of the three fails, the chapter
that repairs it is at most eight back, and they are short.

From here the book forks, and both branches are honest. Read on in order,
and [the next part](../parts/shape.qmd) changes the shape of a table. Or jump to the cookbook near the
end of the book, which answers questions in exactly the vocabulary you now
hold, and come back to the middle parts as your questions demand them. A
grammar is not a syllabus; from this page on, you are allowed to need
things in your own order.
