---
# A preface is not chapter zero, and numbering it said otherwise: its sections
# read as 0.1 through 0.9 while carrying no argument a reader would ever cite.
# The heading below carries `.unnumbered` for the same reason, which is also
# what keeps the first real chapter from being chapter two.
number-sections: false

# Re-stamped on every render, so the landing page reports how current this copy
# is. Its fixed counterpart, the first-published date, is `book: date:` in
# `_quarto.yml`. This one belongs here because it must reach the landing page
# alone; the comment beside its partner says what the other placements do.
#
# **Quarto's own keywords do not work here and print themselves.** `today`,
# `last-modified` and `now` each reach the page as the literal word, so the date
# is computed in R instead. Measured on this book rather than assumed: with
# `date-modified: last-modified` the row read "Updated last-modified".
#
# R formats the date as well as computing it, because `date-format` reaches
# `date:` and not a value substituted into the front matter. Left as an ISO
# string the two rows disagree: "August 7, 2026" above "2026-08-08". The `sub()`
# drops the leading zero `%d` puts on a single-digit day, which is the one place
# the two would still differ.
date-modified: "`r sub(' 0', ' ', format(Sys.Date(), '%B %d, %Y'))`"
---

# Preface {.unnumbered}

<!-- The include below has two jobs beyond its tables. Its chunks are what
make the date above a date: Quarto evaluates inline R in the front matter only
when the document holds at least one executable chunk, and without one the
expression reaches the page as its own source text and the "Updated" row never
appears. And it has to sit below the heading, not above it: Quarto reads the
file for its first `#` line to name the chapter before knitr has run, so a
comment at column 0 inside a fence is a heading as far as that scan is
concerned. Put the preamble first and the sidebar takes its first line as the
title, the preface takes a number, and every chapter shifts by one. This page
used to carry a chunk that computed nothing instead, because the preface
showed no live example; the first pipeline below runs now, in both languages,
so the full preamble stopped being an extravagance and became the
requirement. -->

{{< include _setup.qmd >}}

::: {.god-motto}
Say it once. Run it anywhere.
:::

god is one data-manipulation engine written in [Rust](https://www.rust-lang.org/),
callable from [R](https://www.r-project.org/) and
[Python](https://www.python.org/). You write your pipeline as a sentence, and
the same sentence works in both. Its vocabulary is plain everyday English, and its shape comes from
[dplyr](https://dplyr.tidyverse.org/) and [tidyr](https://tidyr.tidyverse.org/),
which are the tools that got this right.

Behind both is one promise:

> Anyone who works with data should be able to read a pipeline aloud on day one,
> write one themselves on day two, and still have it after a year away.

That last clause is the hard one, and it is the reason this project exists.

A pipeline in god is a sequence of plain English verbs. You keep some rows, add a
column, group and summarize, sort the result, and take the first ten. Each step
takes a table and returns a table, so you can read a pipeline from top to bottom
and stop anywhere.

Most of what anyone does to a table is ordinary work of exactly that kind,
plus a join and a reshape. god is for that work, every day, in whichever of
the two languages you happen to be in. It is not for the exotic case, and the
chapter on what it refuses says so.

## About the name

god is **G**rammar **O**f **D**ata. Its sibling is [gog](https://psychometrician.github.io/gog-book/),
a grammar of graphics, and the two are named the same way for the same reason.

It is written lowercase, always, the way dplyr and pandas and npm are. Nobody
writes DPLYR. No claim of any other kind is made by the letters, and none is
intended. They are three initials that happened to spell a word, kept because
the sibling naming is what makes the pair legible.

You will see it lowercase everywhere in this book, including at the start of a
sentence.

You do not have to read this preface. Nothing in it is needed to write a
pipeline. [Install god](#installing-it), then go to
[your first pipeline](chapters/first-pipeline.qmd), which begins the six
everyday verbs. Come back here when you want to know why the grammar refuses
what it refuses. If you would rather see all of it at once first,
[Appendix B](appendices/b-every-word.qmd) is every word there is, on one page.

## One sentence, three ways

The main way to write a pipeline in god is to say it. `run` takes the sentence,
gives back the table, and is spelled exactly the same in R and in Python.

```{r}
run('sales
       then keep where [region] is "West"
       then add [margin] as [revenue] - [cost]
       then summarize [margin] as total([margin]),
            [orders] as row_count() by [product]
       then sort [margin] descending
       then take 10')
```

Read it aloud and it is a sentence. Keep the rows where the region is West. Add
a margin column. For each product, total the margin and count the orders. Sort
by margin, largest first. Take the top ten.

If you already write R, the same sentence has a pipe spelling.

```{r}
sales |>
  keep(region == "West") |>
  add(margin = revenue - cost) |>
  summarize(margin = total(margin), orders = row_count(), by = product) |>
  sort(descending(margin)) |>
  take(10)
```

And if you already write Python, so does it there.

```{python}
(sales
  >> keep(col.region == "West")
  >> add(margin = col.revenue - col.cost)
  >> summarize(margin = total(col.margin), orders = row_count(),
               by = col.product)
  >> sort(descending(col.margin))
  >> take(10))
```

None of the three is an illustration. All three ran when this page was built,
and the table under each is the answer that came back.

**A matter of taste.** Which one you write is exactly that, and there is
nothing you can reach with a pipe that you cannot say in a sentence. Both pipe
spellings work by turning your code into the sentence above and handing that to
the engine. Anything you can pipe, you can say.

**Start with `run`.** If this kind of work is new to you, it is the shortest
thing to learn, it is identical in both languages, and it is the one spelling
that still means something where there is no R or Python at all. The pipe
flavors are here for readers who already think in dplyr or pandas, and who
would rather not change hands to use this.

## Two differences

The two pipe versions differ from each other in two things.

- **The pipe.** R uses its native `|>`. Python uses `>>`.
- **The column.** R writes a bare name, `revenue`. Python writes `col.revenue`.

No word in the vocabulary is spelled differently in the two languages. Every
verb, every marker, every argument name is the same in both, and that is meant
literally rather than approximately. One word broke it once, and it was renamed
rather than excused.

What is left is not vocabulary but syntax, and all of it is Python's own. This
book states it here rather than letting you find it in an error message.
Comparisons joined by `&` need parentheses, because `&` binds more tightly than
`==`. Negation is `~`, because Python cannot redefine `not`. Asking whether a
value is one of several is `col.region.is_in([...])`, and asking whether one is
missing is `col.cost.is_missing()`, because Python's `in` cannot be reached by an
object at all.

## A sentence with no host

A sentence is data. That is the one reason to prefer it that is not about
taste. Some places
have no language to bind into: a database cell, a configuration file, a
pipeline that was stored or generated rather than typed. All of them can hold
one of these, and it is the same grammar sitting there.

```
sales
  then keep where [region] is "West"
  then add [margin] as [revenue] - [cost]
  then summarize [margin] as total([margin]),
       [orders] as row_count() by [product]
  then sort [margin] descending
  then take 10
```

Both pipe versions above produce exactly this, which is why a pipeline can move
between a script, a notebook and a warehouse without being rewritten.

Where there is no R or Python, the sentence is still the sentence, but
something has to run it for you. `god-cli` reads one and writes the query for
your engine, which you then run where your data already is. [What it
wrote](chapters/what-it-wrote.qmd) shows both routes.

Notice that split early. Whoever writes a question and whoever runs it do not
have to be the same person, and only one of them needs a language to run it
in.

## The rules this book is held to

The preface makes large claims, so the book binds itself to five rules you can
check, and every one of them is checked by a script each time the book is
built.

- **Both tabs are one sentence.** Every example is written twice, R above
  Python, and both run when the book is built. Two tabs disagreeing on a page
  is a defect the build refuses to ship.
- **Questions first.** A teaching chapter opens with a question about data,
  never with a feature. The verb arrives as an answer.
- **A small cast.** The same few tables carry the whole book, and three
  quarters of its pipelines start from one of them.
- **Read it aloud.** The first pipeline of every teaching chapter is followed
  by the same sentence in the grammar's own words, in italics. The gloss is
  not decoration: it is a pipeline too, and the build parses every one.
- **Refusals on stage.** What the grammar refuses is shown refusing, live, in
  both languages, so an error message is something you have already read
  before one finds you.

## The cast

Every example in this book runs on the same few tables, declared here once.

- `sales`: fifteen rows of orders. Three regions, four products, a year of
  dates, a quantity and the money. Small enough to check a total by eye,
  large enough that no group is a single row.
- `products`: the lookup table beside it. One product nobody sold, and one
  sale of a product nobody lists; the mismatch is kept on purpose, for the
  chapters on joining.
- `survey`: the wide one. Six respondents, eight score columns, and four
  columns of everything else, which is what choosing columns by the shape of
  a name is for.
- `answers` and `marks`: the same facts twice, once wide and once long, for
  the chapters on reshaping.
- `messy` and `diary`: text that needs repair, and dates that need reading.
- `gapminder`: 1,704 rows of countries and years [@bryan2025gapminder], the
  one table you cannot check by eye, for the moments a verb has to prove it
  survives scale.

The cast is closed, because a new table costs you attention. When a chapter
needs a special shape, such as two rows that collide or a table with one
column, it builds the smallest table that shows it. The next example returns to
the cast.

The three tables that carry most pages are typed out below in full, so your
session can hold the same rows this book computes from. The rest of the cast
is smaller, and each of those tables is written out in the chapter that
introduces it.

::: {.panel-tabset}

### R

```{r}
sales <- data.frame(
  date     = c("2025-11-03", "2025-11-17", "2025-12-05", "2025-12-22",
               "2026-01-09", "2026-01-26", "2026-02-14", "2026-03-03",
               "2026-04-11", "2026-05-06", "2026-06-19", "2026-07-08",
               "2026-08-24", "2026-09-15", "2026-10-02"),
  region   = c("West", "East", "West", "North", "East", "West", "North",
               "East", "West", "North", "East", "West", "North", "East",
               "West"),
  product  = c("Widget", "Gadget", "Doohickey", "Widget", "Widget",
               "Gadget", "Doohickey", "Doohickey", "Widget", "Gadget",
               "Sprocket", "Sprocket", "Widget", "Gadget", "Widget"),
  quantity = c(4L, 2L, 3L, 6L, 8L, 5L, 2L, 5L, 2L, 4L, 10L, 6L, 3L, 1L, 5L),
  revenue  = c(100, 120, 120, 150, 200, 300, 80, 200, 50, 240, 150, 90, 75,
               60, 125),
  cost     = c(40, 80, 75, 60, 80, 200, 50, 125, 20, 160, 120, 72, 30, 40,
               50),
  stringsAsFactors = FALSE
)

products <- data.frame(
  product = c("Widget", "Gadget", "Doohickey", "Gizmo"),
  maker   = c("Acme", "Globex", "Initech", "Initrode"),
  stringsAsFactors = FALSE
)

survey <- data.frame(
  respondent = 1:6,
  name       = c("ana", "ben", "cal", "dee", "eli", "fay"),
  region     = c("West", "East", "North", "West", "East", "North"),
  joined     = c("2026-01-12", "2026-02-03", "2026-02-27", "2026-03-15",
                 "2026-04-01", "2026-04-22"),
  q1_score   = c(4L, 5L, 3L, 4L, 2L, 5L),
  q2_score   = c(2L, 5L, 5L, 3L, 4L, 4L),
  q3_score   = c(5L, 4L, 2L, 5L, 3L, 4L),
  q4_score   = c(3L, 3L, 4L, 2L, 5L, 5L),
  q5_score   = c(4L, 2L, 5L, 4L, 4L, 3L),
  q6_score   = c(5L, 5L, 3L, 3L, 2L, 4L),
  q7_score   = c(2L, 4L, 4L, 5L, 3L, 5L),
  q8_score   = c(3L, 5L, 2L, 4L, 5L, 4L)
)
```

### Python

```{python}
sales = pd.DataFrame({
    "date":     ["2025-11-03", "2025-11-17", "2025-12-05", "2025-12-22",
                 "2026-01-09", "2026-01-26", "2026-02-14", "2026-03-03",
                 "2026-04-11", "2026-05-06", "2026-06-19", "2026-07-08",
                 "2026-08-24", "2026-09-15", "2026-10-02"],
    "region":   ["West", "East", "West", "North", "East", "West", "North",
                 "East", "West", "North", "East", "West", "North", "East",
                 "West"],
    "product":  ["Widget", "Gadget", "Doohickey", "Widget", "Widget",
                 "Gadget", "Doohickey", "Doohickey", "Widget", "Gadget",
                 "Sprocket", "Sprocket", "Widget", "Gadget", "Widget"],
    "quantity": [4, 2, 3, 6, 8, 5, 2, 5, 2, 4, 10, 6, 3, 1, 5],
    "revenue":  [100, 120, 120, 150, 200, 300, 80, 200, 50, 240, 150, 90,
                 75, 60, 125],
    "cost":     [40, 80, 75, 60, 80, 200, 50, 125, 20, 160, 120, 72, 30,
                 40, 50],
})

products = pd.DataFrame({
    "product": ["Widget", "Gadget", "Doohickey", "Gizmo"],
    "maker":   ["Acme", "Globex", "Initech", "Initrode"],
})

survey = pd.DataFrame({
    "respondent": [1, 2, 3, 4, 5, 6],
    "name":       ["ana", "ben", "cal", "dee", "eli", "fay"],
    "region":     ["West", "East", "North", "West", "East", "North"],
    "joined":     ["2026-01-12", "2026-02-03", "2026-02-27", "2026-03-15",
                   "2026-04-01", "2026-04-22"],
    "q1_score":   [4, 5, 3, 4, 2, 5],
    "q2_score":   [2, 5, 5, 3, 4, 4],
    "q3_score":   [5, 4, 2, 5, 3, 4],
    "q4_score":   [3, 3, 4, 2, 5, 5],
    "q5_score":   [4, 2, 5, 4, 4, 3],
    "q6_score":   [5, 5, 3, 3, 2, 4],
    "q7_score":   [2, 4, 4, 5, 3, 5],
    "q8_score":   [3, 5, 2, 4, 5, 4],
})
```

:::

Typing them in is one way. The other is one call. Every table in the cast is
published beside this book, and `god_table` reads one by name in either
language, so the rows in your session are the rows every page here computed
from. It looks for a `data/<name>.csv` in your working directory or any
directory above it before it reaches for the published copy, so a folder of
your own takes precedence and nothing needs a connection once you have one.

::: {.panel-tabset}

### R

```{r}
sales <- god_table("sales")
nrow(sales)
```

### Python

```{python}
sales = god_table("sales")
len(sales)
```

:::

Those are the same fifteen rows typed out above. This page runs both calls and
the tables agree, which is the only reason to believe the sentence before them.

## Why there is another one of these

You have probably met at least one tool for this already. Here is one operation,
grouping and totaling, written five ways:

```
dplyr       df |> group_by(g) |> summarise(total = sum(x))
pandas      df.groupby('g')['x'].sum().reset_index()
polars      df.group_by('g').agg(pl.col('x').sum().alias('total'))
data.table  df[, .(total = sum(x)), by = g]
SQL         SELECT g, SUM(x) AS total FROM df GROUP BY g
```

Every one of those is a fine tool and none of them is the problem. The problem is
that there are five, and that each has collected rules that have to be memorized
rather than worked out. pandas has three ways to filter a row, and an index that
changes what square brackets mean. SQL's clauses are not executed in the order
you write them. dplyr has two ways to group.

None of that is difficult in the sense that the *ideas* are difficult. The ideas
are small: about fifteen operations, and you already know most of them.

The ideas are also old, which matters before you learn a sixth spelling of
them. Choosing rows, choosing columns, joining, grouping and totaling were set
out as an algebra by E. F. Codd in 1970 [@codd1970], and the ambition to write
that algebra as English sentences is nearly as old: SEQUEL, the query language
that became SQL, stated it in its title in 1974 [@chamberlin1974]. Little has
been added since except window functions and the two directions of reshaping,
and the pattern behind `summarize ... by` had a name, split-apply-combine,
before this project existed [@wickham2011]. What changes every few years is
the syntax, not the operations. pandas [@mckinney2010] and dplyr
[@wickham2023dplyr; @wickham2019tidyverse] are the two homes this vocabulary
answers to, and the shape its tables arrive in has a name and a paper of its
own [@wickham2014tidy].

## Five engines, one vocabulary

There is evidence that a grammar outlives the engine underneath it, and it comes
from the tool this one borrows from. dplyr's own documentation names five other
backends it runs on: `arrow`, `dbplyr`, `dtplyr`, `duckplyr` and `sparklyr`. The
same verbs reach Arrow, a SQL database, data.table, DuckDB and Spark.

Read that list again, because it is the argument of this book made by somebody
else. Five engines, one vocabulary, and the sentence you write does not change
when the thing underneath it does. Nobody had to invent new words to reach
DuckDB, and nobody will have to invent new words to reach whatever replaces it.

What that list does not cross is a language. Every one of those backends is
reached from R, so a person who opens a Python notebook starts again from the
beginning. That gap is the whole of what this project is aimed at, and it is the
reason a sixth spelling is worth learning.

## Where this one came from

The rest of this preface is written in one voice about a system. This section is
not, because the honest answer to "why build another one of these" is personal.

dplyr and tidyr are what I use for my own work, most days. Nothing here is a
complaint about them and nothing here is meant to replace them. If R is the only
place you work, they are better than this is and they will stay better. They have
years of care in them and a large community around them.

The difficulty starts when I leave R. Some work has to happen in Python, and some
has to happen in Databricks.

I have tried to learn pandas more than once, and PySpark more than once. Each
time I learned it, and each time, about a week later, most of the syntax was gone
again. Not the ideas. The ideas are the same in every tool.
What I lost was which method, in which order, taking which argument.

That is twenty years of pattern, not one bad month. pandas, data.table, polars,
and every other package I have worked in have gone the same way. dplyr and tidyr
are the exception, and they are the only exception.

That is not a claim about pandas, which is a serious tool that a great deal of
the world runs on. It is a claim about what I can remember. A tool you have to
relearn every time you pick it up is a tool you put off using, and the work goes
back to the language you already know, which is not always the right place for
it.

### The data.table case

You could reasonably stop here and say this is just someone who likes R. The
piece of evidence that says otherwise is data.table, which is in R.

data.table is fast, concise, and admired by people who work with large tables
every day. I tried to learn it and lost it the same way, on the same schedule as
pandas.

That is the useful part. The tools I could not retain do not sort by language.
They sort by shape: `df[i, j, by]` and `df.groupby('g')['x'].sum()` are
punctuation and brackets, and the two tools I did retain read from left to right
as a sentence.

So this is a preference, and it is worth naming as one rather than presenting it
as a finding. It is a preference about shape, and data.table is how I know it is
not about the language.

### The polars question

polars deserves its own paragraph, because it is what a reader should ask about
first. It has an R port and a Python port, so one API in two languages already
exists. It is fast, lazy, and carefully designed. It did not solve this for me
either.

What did not survive the week was the words. polars is written as method chains
over column objects, `pl.col('x').sum().alias('total')`, and its vocabulary comes
from the dataframe tradition rather than from English. That is a defensible
design and it suits people who work in it every day. It was not the thing I could
still read six months later.

dbplyr is the other answer worth naming, and it comes closest. It translates
dplyr into SQL, so one grammar already reaches a database. It is R only. It does
not help in a Python notebook and it does not help in a Databricks Python job.

[Ibis](https://ibis-project.org/) is the same idea approached from the other
side, and it is the nearest thing to this in Python. One API reaches more than
twenty engines, it compiles a query rather than running one itself, and it is a
serious piece of work by people who have thought about this longer than I have.
It is Python only. Its sentences are method chains over an expression object,
which is the shape I have already said I cannot hold.

So the two best answers to this each solve one half of the problem. One grammar
over many engines has been built twice, once in each language, and neither one
crosses to the other. Standing in both languages at once is the part nobody had
done, and it is a smaller ambition than either of them: not twenty engines, but
two languages saying the same sentence.

### What was actually wanted

All three attempts have one thing in common, and it is the connector. pandas and
polars chain with `.`, polars in R chains with `$`, and data.table nests inside
brackets.

A dot is how English ends a sentence. Using it to join the steps of one works
against what the mark already means, and what I had left after a week was a row
of dots with no memory of which order they went in.

`|>` does not read that way. It points at the next step, and what it says is
*then*.

So the ask was narrow. dplyr and tidyr's sentence, in Python and in a SQL job,
with as little changed as the languages allow.

That turned out to be two things: `>>` where R writes `|>`, and `col.revenue`
where R writes `revenue`. Every verb, every keyword and every argument position
is the same in both. `>>` was picked to point the same way `|>` points, which is
the whole of its job here. Python cannot have `|>` itself, and the next chapter
says why.

So this is not a better dplyr, and it is not trying to be. It is dplyr's reading
experience in the places dplyr cannot go, and the price of carrying it there is a
much smaller vocabulary than dplyr has. Whether that trade is worth making is
yours to judge, and the chapter after this one is where you can see the size of
it.

### The bar this sets

Reading it on day one and writing it on day two is the promise made above. There
is a second half, and it is the harder one. You should still have it after a year
away. Not remember that it exists. Be able to write a pipeline in it,
after twelve months of not touching it, without opening this book.

That is the standard a physical skill meets and a syntax usually does not. It is
the reason the vocabulary is closed, the reason a word means one thing in every
sentence, and the reason several useful things are deliberately missing.

It also sets what counts as success, and that is not what a feature list would
suggest. No vocabulary this small covers everything anyone does with data, and
this one is not trying to. Covering enough of it, in a form that stays
learned, is what this one is for. What happens when you reach the edge is the
next thing this preface talks about.

## *A* grammar, not *the* grammar

The cover says *a*, and that is a claim rather than modesty. There are many
grammars of data that could be built, and several good ones exist. This is one
of them, with stated limits: a closed vocabulary, a set of laws it answers to,
and a list of things it refuses and says why. Those limits are what make the
claim testable. A tool with no stated boundary cannot be measured against
anything, and nothing can be said to cover more of the job than it does.

So this book will tell you what god cannot do about as often as it tells you what
it can. That is not an apology. A grammar that refuses nothing has no shape.

The **hex sticker** at the top of this page says the same thing without a
sentence. R packages have carried these for years, hexagonal so that a wall of
them tiles, and handing one over is how the community introduces a package. If
you have come from Python, that is all it is: a logo with a shape convention.

What sits inside this one is a face. Its eyes are `|>` and `>>`, the pipe in R
and the pipe in Python. Its nose is `then`. Its mouth is the package name.

Those marks are chosen and not decorative, and the test they had to pass is that
you actually write them. A face has two eyes and this grammar has two bindings:
one language on the left, one on the right, saying the same thing. Both eyes
point right, so the face is looking at the next step rather than at you.

The nose is the third spelling. `|>` in R and `>>` in Python both mean *and then
do this to it*, and the plain text form spells that same thing `then`. So the
three marks on the face are not three ideas. They are one idea in each of the
three places you can write god, which is the claim this whole book is making,
sitting on the cover without a sentence.

They are also the same size as each other in practice. Across the sentences this
book is tested against, `|>` appears 172 times, `>>` appears 173, and `then`
appears 172. That is not a coincidence worth remarking on so much as a
consequence: a sentence has as many `then`s as it has pipes, because they are
the same word.

The nose was `(` and `)` until this was noticed, and parentheses do pass the
test below. Every verb in both languages is a call, so you type them constantly.
What they do not do is say the same thing as the eyes. They hold what happens
*inside* a step, where the eyes join one step to the *next*, and a face whose
three marks say one thing is worth more than a face whose marks are merely all
frequent.

The column brackets are not on the cover, and a measurement is the reason.
Across the sentences this book is tested against, `[` and `]` appear *not once*
in R. The few times they appear in Python, they are Python's own list syntax
rather than anything god asks for. They are the plain text form's way of naming a
column, which is a real job, and it is not a job most readers will ever see. A
cover should carry a mark its reader types.

A full stop was considered for the nose and rejected for the opposite reason.
You would type it, and this book argues at some length that joining steps with
a dot is the thing god was built to avoid. A cover should not say the opposite
of its book.

## What this grammar borrows, and from where

Its vocabulary and its working shape come from **dplyr** and **tidyr**, which
solved this for R and are the reason the author still writes R. Its verbs, its
grouping and its two reshaping directions are all recognizable from there, and
where god differs it is usually because it had to say the same thing in two
languages at once.

One further idea comes from **Hangeul**, the Korean alphabet designed in 1443,
by way of this project's sibling. It is one idea rather than a theme: parts that
combine the same way every time are easy to read and easy to write. That is the
whole of the borrowing, and nothing else about the alphabet is claimed here.

Applied, it means a small set of verbs, functions and delimiters where a rule you
learn in one place holds in every other. The promise at the top of this preface
is what that regularity is *for*. And there is a working test that keeps it
honest rather than aspirational: if you can say something two ways, one of them
is a bug. Report it.

## Naming a column whatever you like

You can call a column `sort`, or `total`, or even `then`, and nothing breaks.
Each spelling has its own unambiguous answer: R looks names up in your data
first, Python writes `col.sort`, and the text form puts it in brackets.

In the text form, four delimiters each do exactly one job.

| | Means |
|---|---|
| `[ … ]` | a column, or a list of columns |
| `" … "` | a text value |
| `( … )` | a group; a name in front of one applies that name to it |
| `{ … }` | a set of values to match |

Inside `[ ]` it is a column, always. Outside, it is grammar.

```
orders then sort [sort] descending
```

No backticks, no escape word, and no list of names you are not allowed to use.
This grammar does not restrict what your columns are called; it marks where a
name is being used.

The two quote marks divide the same way: double quotes belong to the grammar,
for a text value, and single ones to the host, around a sentence. The chapter
on what it wrote gathers the whole quoting story in one place.

## Words instead of symbols, where the languages disagree

A symbol that means different things in different tools is not in this grammar. A
plain word that means one thing everywhere is.

| Instead of | Write |
|---|---|
| `==` or `=` | `is` |
| `!=` or `<>` | `is not` |
| `TRUE` / `True` / `true` | `yes` |
| `FALSE` / `False` / `false` | `no` |
| `NA` / `None` / `NULL` | `missing` |
| `%in%` / `.isin()` / `IN` | `in { … }` |
| `&` and `\|` | `and` and `or` |
| `!` or `~` or `NOT` | `not` |

`<`, `<=`, `>` and `>=` stay as they are, because those already agree everywhere.

## It will tell you what it wrote

No small vocabulary covers everything, and this one does not try. When you reach
its edge, you can ask what your pipeline would be in a tool you already use:

```
$ god --columns 'region:text,revenue:number' --as dplyr 'sales then keep where [region] is "West" then take 10'
sales |>
  filter((region == "West")) |>
  head(10)
```

The pipeline is wrapped in single quotes because a text value inside it is
written with double quotes, which keeps the shell out of it. The `--columns`
list is the table the pipeline is checked against: nothing is written out
until the whole sentence has passed the same checks a running one would.

Learning this grammar is meant to make the tools you already have easier, not to
replace them.

## When something is wrong

The whole pipeline is read and checked before any of it runs, so a mistake at
step two is reported as a mistake at step two, with the word underlined:

```
illegal: there is no column called `reveune`. Did you mean `revenue`? The table has: region, product, revenue, cost, ordered_on
  |
2 |   then keep where [reveune] > 100
  |                    ^^^^^^^
```

A refusal always names what would have happened and what to write instead. And
nothing is ever quietly ignored: a pipeline that runs, returns a number, and
silently dropped one of your lines is the way a tool loses your trust.

## What god does not do

god is not an engine. The work of joining, grouping, sorting and handling missing
values is done underneath by software built for it. god owns the words, the
checking and the error messages.

god also does not own your data. A table arrives from your language and leaves as
one. There is no god table type to convert to, and no session to open.

Anything that does not return a table belongs to your language rather than to
god. Fitting a model, running a test, drawing a plot, writing a file: those are
things R and Python already do well, and a grammar that tried to absorb them
would stop being small enough to read.

### How fast is it?

The question has two answers, and keeping them apart is the honest way to ask it.

What god costs is compiling your pipeline: reading it, checking it against the
columns, and writing the query. That is a fixed cost per pipeline rather than per
row, so it does not grow with your data. A hundred rows and a hundred million
rows compile in the same time.

Everything after that belongs to the engine you pointed god at. It hands DuckDB
or Spark a query and does nothing else. So a comparison of god against polars
is really DuckDB against polars, and reporting it as god's number would take
credit, or blame, for somebody else's engine.

The claim this book makes is therefore narrow: the grammar costs almost
nothing to put in front of an engine, and your query reaches that engine
unchanged. [Where the speed lives](appendices/g-where-the-speed-lives.qmd)
holds a measurement made to exactly that framing, in two halves that are
never mixed. The grammar's own cost is timed live while these pages are
built. The engine timings beside it are a dated record, so each engine is
credited with its own result. For today's number on your own engine, ask
`show_as` for the exact query and measure it where your data lives.

Speed was never the goal here. The goal is that you can still write a pipeline a
year after the last one.

## Installing it

In R, from R-universe, as a binary with nothing else to set up:

```r
install.packages("god",
  repos = c("https://psychometrician.r-universe.dev",
            "https://cloud.r-project.org"))
```

Then `library(god)` attaches the verbs, and every example in this book runs
as printed.

In Python, from PyPI, one wheel per platform with the engine inside:

```bash
pip install grammar-of-data
```

The install name and the import name differ, because `god` was taken on PyPI
in 2016 by an unrelated package: you install `grammar-of-data` and write
`import god`. The examples in this book use `from god import *`; `import god`
works too, with a prefix on every name.

Every example shown in both languages ran in both when this book was built,
exactly as printed, and a page whose two tabs disagree is a build the book
refuses to ship.

## Who this book is for

You have worked with data in a table before. You know what a row is, what a
column is, and what it means to group by something and count. That is the whole
assumption.

If you are near the start, read the book in order. [The six
verbs](parts/six-verbs.qmd) open it, one short chapter at a time, and when
you finish them you can answer real questions about a table. Nothing before
them is required, including this preface.

If you have used several of these tools already, you are the other reader this
book is written for, and you will get more out of it from the back. Appendix
C writes the same pipelines out side by side in every dialect god prints,
which is the fastest way to see what is being proposed. Appendix B is the
entire vocabulary on one page. [How god works](parts/how-it-works.qmd) is
where the claims in this preface are demonstrated rather than asserted,
including the one about a pipeline running unchanged on a cluster.

You do not need to know dplyr, and you do not need to know pandas. If you know
one of them, some chapters will read as a vocabulary list, and that is fine. If
you know neither, nothing here assumes you do.

## Where this stops, and what picks it up

god manipulates tables. It does not draw them, and it never will, because
drawing is a different grammar with a different vocabulary.

That grammar exists, and it is this project's sibling.
[gog](https://psychometrician.github.io/gog-book/) is a grammar of graphics
built on the same discipline, by the same author, with the same promise about
day one and day two. The two are designed to sit next to each other. A question
usually starts as a table you have to reshape and ends as a picture, so you
start in god and finish in gog.

[The closing chapter](chapters/with-gog.qmd) shows one question answered end
to end that way. Read it even if you never draw a plot, because it is the
clearest statement of where this grammar's job ends.

## How to read the examples

Every example that manipulates data is written twice, once in R and once in
Python, and the two sit in tabs above the same explanation. Click the one you
write.

Many tabsets carry a third tab, `run`. It holds the same sentence a third
way: the grammar's own text form, inside the one call that is spelled
identically in both languages. The sentence in it is the engine's canonical
printing of the pipeline beside it, and the chunk runs like any other, so
all three tabs answer with the same table.

The tabs are not illustrations of each other. Every tab is executed when the
book is built, so a tab showing a table is a tab whose code ran, and the
tables come back from the same engine. Suppose a sentence stopped meaning the
same thing in the two languages, or the text form drifted from either. The
tabs would stop agreeing on this page, which is the easiest place to notice
it.

The appendix on [the two languages](appendices/a-two-languages.qmd) is the
exception, and deliberately so: it is about the two places the languages
differ, and there the two are not the same sentence.

## The state of this book

The vocabulary is closed, and every word in it is demonstrated here by a
chunk that runs. That claim is checked rather than remembered: when the book
builds, a script asks the engine for its word list and fails the build if
any word lacks a running example. A second check hands every pipeline the
book shows outside a chunk to the engine to parse.

Prose can name a verb that was never built and still render cleanly, which
is why those checks exist. What no script can catch is handled the way the
coverage appendix promises: dated, and corrected in public when a reader
finds the mistake.
